### 2020-02-04 manuscript.Rmd
init

---
title: 'Title'
author: 'Florian Muthreich'
date: '2020/02/04'
output: pdf_document
---

### 2020-02-04 readme.Rmd
init

---
title: 'Read Me'
author: 'Florian Muthreich'
date: '2020/02/04'
output: pdf_document
---

### 2020-02-04 manuscript.Rmd
calculated workload for FTIR lab

@@
 ---
+
+Notes:
+- Time to measure 50 pollen grains on single taxa slides: 60 min. 
+- includes Mounting, selecting pollen, measuring time FTIR
+
+Fossil pollen will require more searching in the selection phase.
+Have to measure 75 grains if we target 50 good spectra.
+Estimate time for mounting, selecting, measuring 75 grains of one taxa: 90 - 120 min
+
+10 days in the lab; 8+ hrs a day: 80 hrs
+Target: 5 taxa per sample; 5 x 90: 7.5 hrs `r 5*90/60`
+
+9-12 13-24
+10 days in the lab; 14 hrs a day: 140
+Target: 5 taxa per sample; 140/7.5 = 18; 14/10 = 14
+
+```{r}
+total_samples = function(hrs_per_sample, no_taxa, total_hrs = 140){
+  tmp = total_hrs / (no_taxa * hrs_per_sample)
+  return(signif(tmp,3))
+}
+M = outer(c(1,1.5,2),c(3,4,5),total_samples, total_hrs = 140)
+row.names(M) = c(1,1.5,2)
+colnames(M) = c(3,4,5)
+M
+```
+
+Anywhere from 18 to 30 samples is reasonable.
+At 3 cores 5 taxa:
+Minimum 4 samples per core 
+Maximum 9 samples per core
+```{r}
+signif(M/3, 3)
+```
+At 2 cores 5 taxa:
+Minimum 7 samples per core
+Maximum 14 samples per core
+```{r}
+signif(M/2, 3)
+```

### 2020-02-05 manuscript.Rmd
added some print strings

@@
 ```{r}
+print('No. of samples at Three Cores')
 signif(M/3, 3)
@@
 ```{r}
+print('No. of samples at Two Cores')
 signif(M/2, 3)

### 2020-02-06 manuscript.Rmd
added research Qs

@@
 ---
-title: 'Title'
+title: 'Viability of FTIR on fossil pollen'
 author: 'Florian Muthreich'
@@
 
+extracted fossil pollen from sediment.
+ca 5 common taxa: Alnus, Picea, Pinus,
+
+Research Questions:
+1. chemical variation between taxa in the same cores.
+2. chemical variation between taxa between cores 
+3. chemical variation of pollen with depths/age
+
 Notes:
-- Time to measure 50 pollen grains on single taxa slides: 60 min. 
+- Time to measure 50 pollen grains on single taxa slides: min 60 min. 
 - includes Mounting, selecting pollen, measuring time FTIR
@@
 Fossil pollen will require more searching in the selection phase.
-Have to measure 75 grains if we target 50 good spectra.
+Have to measure ~75 grains if we target 50 good spectra.
 Estimate time for mounting, selecting, measuring 75 grains of one taxa: 90 - 120 min
@@
 colnames(M) = c(3,4,5)
+print('Number of samples at One Core')
 M

### 2020-02-11 manuscript.Rmd
added references.bib; wrote notes on possible cores

@@
 date: '2020/02/04'
-output: pdf_document
+output:
+  pdf_document: default
+bibliography: references.bib
 ---
@@
 3. chemical variation of pollen with depths/age
+## Cores
+Dalman Lado:
 
-Notes:
+Totenmoos [@heissUltenValleySouth2005]:
+- bog-sediment core
+- min 1000 grains counted
+- Picea, Pinus very common, switch at 10000 BP from Pinus to Picea
+- Alnus, Betula throughout
+- Abies, Fagus after 9000 BP
+- trace Quercus
+
+Younger Dryas [@brauerHighResolutionSediment1999]:
+- 12 samples laminated sediment
+- from  11400 to 13120
+- min 500 grains
+- pollen conc highest in 11000-11600 and 12700-12900
+- Pinus betula most common (ca 150 grains counted)
+- salix (ca 25 counted)
+- quercus throughout in low conc (<1%)
+- graminoids/cyperaceae/artemisia common
+
+Tinner alpine:
+
+Tinner meso-mediterranean:
+Quercus sections Quercus, Cerris and Ilex
+
+Tinner thermo-mediterranean:
+Quercus sections Quercus, Cerris and Ilex
+
+## Time to measure notes:
 - Time to measure 50 pollen grains on single taxa slides: min 60 min. 

### 2020-02-11 manuscript.Rmd
minor changes after meeting with Al

@@
 ## Cores
-Dalman Lado:
+Dalmuttlado:
+- pine
+- holocene
 
 Totenmoos [@heissUltenValleySouth2005]:
-- bog-sediment core
+- bog-sediment core; past 15000 yrs
 - min 1000 grains counted
@@
 - graminoids/cyperaceae/artemisia common
+
+Tiefer See:
+- 1870-2010
+- laminated sediments
+- Pinus pollen
 
@@
 Maximum 9 samples per core
+
 ```{r}
@@
 Maximum 14 samples per core
+
 ```{r}

### 2020-02-18 readme.Rmd
lab notes

@@
 ---
+
+#Lab-Notes
+Samples 2020-02-13
+Vial  Sample         Pinus  Betula  Salix  Spores
+5*    A4-DL 138-140  +++    +              +++
+4     D4-DL 161-163  ++     +              +++
+4*    A5-UR  64- 66  ++     +       1      +++
+
+1. add 1 cube to vial (~1cm^-3^)
+2. wash with water -> dissolve sediment vortex
+3. centrifuge 3min 3000rpm
+4. decant; save decant
+5. add 3 ml 1.43 SPT solution
+6. vortex
+7. centrifuge 3min 3000rpm
+8. decant supernatant into new vial
+9. repeat 5-7; decant into same vial
+10. wash with water; save decant for checking
+11. wash with water; save decant x2
+12. wash residue with water; save in vial
+
+2nd Run
+Vial  Sample
+2*    A4-DL 188-190
+3*    A5-UR  72- 74
+5     D4-DL 182-184
+6     A4-DL 130-132
+7*    A4-DL 163-165
+
+3rd Run
+Vial  Sample
+4     A4-DL 175-177
+5     D4-DL 149-151
+6     A4-DL 145-147
+7*    D4-DL 138-140

### 2020-02-21 manuscript.Rmd
started intro and methods

@@
 ---
-
+# Notes
 extracted fossil pollen from sediment.
@@
 
-Research Questions:
-1. chemical variation between taxa in the same cores.
-2. chemical variation between taxa between cores 
-3. chemical variation of pollen with depths/age
 ## Cores
@@
 ```
+
+# Introduction
+Research Questions:
+1. chemical variation between taxa in the same cores.
+2. chemical variation between taxa between cores 
+3. chemical variation of pollen with depths/age
+
+# Methods and Materials
+
+Pollen extracted from sediment samples 
+[@kampsAgentSeparationDissolved1985] SPT patent
+[@nakagawaDensemediaSeparationMore1998] 
+[@regnellPreparativeCentrifugationNew1996] density of pollen 1.4

### 2020-03-03 manuscript.Rmd
added references for Tinner samples

@@
 Tinner meso-mediterranean:
+[@bisculmHoloceneVegetationFire2012; @vanniereClimateHumandrivenFire2008]
 Quercus sections Quercus, Cerris and Ilex
 
-Tinner thermo-mediterranean:
+Tinner sub-mediterranean:
+[@zoller19601960]
 Quercus sections Quercus, Cerris and Ilex

### 2020-04-29 readme.Rmd
ran Tiefer See samples
short notes from meeting with boris

@@
 7*    D4-DL 138-140
+
+Samples 2020-02-19
+Vial  Samples
+1     TSK15-K6 42.0-42.8
+2*    TSK15-K6 17.5-18.2
+3*    TSK15-K6 38.7-39.2
+4*    TSK15-K6  9.8-10.5
+4     TSK15-K6 26.0-26.8
+5     TSK15-K6  0.0- 1.5
+5*    TSK15-K6 14.0-14.5
+6     TSK15-K6 34.0-35.0
+7     TSK15-K6 29.5-31.0
+8     TSK15-K6 21.5-22.2
+
+Meeting with Boris
+higher microscope
+tradeoff more time less image
+BSi region showing up on the spectra
+36x 1 px = 1 µm
+testing of objective

### 2020-04-29 readme.Rmd
added plan for potential lab work in may

@@
 testing of objective
+
+#Plan for Ås Lab work
+
+dependant on results from Boris tests
+
+Status after last update that binding medium is working:
+- much better picture than on ZnSe slides
+- only measured empty sample
+- normal objectiv works, no switching objectives needed
+- if comparison to modern material, needs to be measured on new setup as well, cant use old data
+Problems:
+- Pollen density on slides
+- unknownd layer over pollen; BSi? (maybe more washing, switching to store in alcohol instead of water?)
+
+Possible set ups:
+1. Measure one core. Multiple samples (5+), multiple species (3+) -> 15 samples
+2. measure multiple cores (5+), reduced number of samples (<3) and red species (1; only Pinus) -> 15 samples
+
+RQ for both setups:
+Description of method for FTIR analysis of single grain fossil pollen
+
+specific to 1: 
+RQ1 Taxonomic differences between fossil pollen
+RQ2 change in pollen chem with age (finer scale 100s of years)
+
+specific to 2:
+RQ1 chem differences of same species between cores (assuming P. sylvestris everwhere)
+RQ2 change in pollen with age (broader scale 1000s of years)

### 2020-05-15 readme.Rmd
Lab work notes last day in Ås

@@
 #Lab-Notes
-Samples 2020-02-13
+##Samples 2020-02-13
 Vial  Sample         Pinus  Betula  Salix  Spores
@@
 
-Samples 2020-02-19
+##Samples 2020-02-19
 Vial  Samples
@@
 8     TSK15-K6 21.5-22.2
+
+##Samples 2020-05-04
+Urio Zuazzrocchi
+Vial  Samples
+1     UQC-20
+2     UQC-70
+2*    UQC-150
+3*    UQC-268
+4     UQC-365
+5*    UQC-468
+6     UQC-588
+7     UQC-667
+8     UQC-767
+8*    UQC-882
+
+Notes: after first waterwash vial 5-8 spilled in the centrifuge
+
+##Samples 2020-05-05
+Sa Curcurica
+Vial  Samples
+1     SCUR-12
+2     SCUR-120
+2*    SCUR-232
+3*    SCUR-328
+4     SCUR-430
+5*    SCUR-526
+6     SCUR-664
+7     SCUR-728
+8     SCUR-824
+8*    SCUR-944
+
+##Samples 2020-05-06
+Lago del Sangiatto
+Vial  Samples
+1     SNG-32
+2     SNG-72
+2*    SNG-128
+3*    SNG-176
+4     SNG-224
+5*    SNG-264
+6     SNG-308
+7     SNG-328
+8     SNG-360
+8*    SNG-390
+
+##Samples 2020-05-07
+Dalmutladdo
+Vial  Samples
+1     DL-20
+2     DL-40
+2*    DL-60
+3*    DL-90
+4     DL-100
+5*    DL-110
+6     DL-120
+7     DL-130
+8     DL-140
+8*    DL-150
 
@@
 RQ2 change in pollen with age (broader scale 1000s of years)
+
+#Ås Labwork
+
+2020-05-11
+Bergen PS7_02
+all sideways pollen
+polar pollen left 2x2
+
+INN_PS3_01
+sideways pollen: all
+polar pollen: 1x1 and 1x2
+
+INN_PS19_01
+sideways_pollen: 1x1, 1x3
+polar pollen: 1x1,1x2
+
+DAL_20_01
+all Pinus
+Alnus (6)
+BetCor (8)
+
+DAL_20_02
+all Pinus
+Alnus (1)
+BetCor (7)
+
+BGO_PS08
+sideways: 25 grains
+polar: 25 grains
+
+DAL_90_01
+all Pinus (8)
+blanks (6)
+
+DAL_90_02
+all Pinus (9)
+blanks (6)
+
+DAL_90_03
+all Pinus (19)
+blanks (6)
+
+DAL_130_01
+all Pinus (12)
+blanks (6)
+
+DAL_130_02
+all Pinus (27)
+blanks (6)
+
+DAL_60_01
+Pinus 1x1 to 6x4 (47)
+blanks (6)
+
+DAL_110_01
+Pinus (11)
+blanks (6)
+
+DAL_110_02
+Pinus (21)
+blanks (6)
+
+DAL_150_03-09
+Pinus (7)
+
+Cor_ave 14B1 skip 1x2 quadrant
+
+DAL_08_01
+Pinus (7)
+blanks (4)
+
+DAL_08_02
+Pinus (2)
+blanks (4)
+
+DAL_08_03
+Pinus (2)
+
+DAL_40_01
+Pinus 1x1 to 2x6 (37)
+blanks (5)
+
+DAL_08_04-15
+Pinus (12)
+
+DAL_140_01_17
+Pinus (17)

### 2020-05-18 manuscript.Rmd
analysis notes after talking to Boris
wrote short report and overview tables for meeting with Boris/Al/Valeria

@@
 # Methods and Materials
+## Labwork Report
+
+did µFTIR on fresh and fossil pollen using Hyperion-FPA
+
+This is very sample dependent.
+organic rich cores are problematic (requires a better extraction protocol), pollen density is another.
+Dalmutladdo was a very good core for this: low in organic and relatively rich in pollen.
+BSi increased with depth, highest in oldest samples (extracted blanks).
+Still, some depth were more time intensive (130/150), low amount of pollen .
+I managed about 2-3 core samples per day, the fresh samples can be done in a fraction of the time, all done within 1 half day.
+
+9 depth from Dalmutladdo core. targeted and extracted Pinus pollen (see table)
+
+possibilities for Alnus and Betula, even though scattering visible in spectra
+
+```{r}
+dalmutladdo = data.frame(ID = c("DAL_08","DAL_20","DAL_40","DAL_60","DAL_90", "DAL_110","DAL_130","DAL_140","DAL_150"),
+                         pinus_pollen = c(23,49,37,47,36,32,39,17,7),
+                         alnus_pollen = c(0,7,0,0,0,0,0,0,0),
+                         betula_pollen = c(0,15,0,0,0,0,0,0,0))
+dalmutladdo
+```
+
+```{r}
+fresh_pollen = data.frame(ID = c("BGO_PS7", "BGO_PS8", "INN_PS13", "INN_PS19A","NMBU_Pin_syl_14E1", "NMBU_Pin_syl_14H1", "NMBU_Aln_glu_14E1", "NMBU_Aln_glu_14J1", "NMBU_Aln_inc_14B1", "NMBU_Aln_inc_14G1", "NMBU_Bet_pub_14A1", "NMBU_Bet_pub_14B1", "NMBU_Cor_ave_14B1", "NMBU_Cor_ave_14C1"),
+                          extracted = c(109,55,51,24,0,0,0,0,0,0,0,0,0,0))
+fresh_pollen
+```
+
+Research questions we could answer with this set:
+
+- Differences between fresh and fossil/near fossil grains of Pinus
+- chemical changes in Pinus pollen with age/depth
+(- chemical differences between Pinus/Alnus/Betula)
+
+extensions to this set:
+- add more samples from Dalmuttladdo highest and lowest depth (maybe labwork)
+- add cores (e.g. MFM, Tiefer See or Italian cores), Festi core was quite organic rich
+-
+
+# Introduction
+
+# Methods and Materials 
+PCA on grains for exploration
+
+regression of spectra on depth
+
+repeat on different grain averages (1, 3, 5, 10)
+
+average all spectra and stack plot
 

### 2020-05-18 manuscript.Rmd
excluded pdfs from tracking; removed time to measure segment from manus

@@
 
-## Time to measure notes:
-- Time to measure 50 pollen grains on single taxa slides: min 60 min. 
-- includes Mounting, selecting pollen, measuring time FTIR
-
-Fossil pollen will require more searching in the selection phase.
-Have to measure ~75 grains if we target 50 good spectra.
-Estimate time for mounting, selecting, measuring 75 grains of one taxa: 90 - 120 min
-
-10 days in the lab; 8+ hrs a day: 80 hrs
-Target: 5 taxa per sample; 5 x 90: 7.5 hrs `r 5*90/60`
-
-9-12 13-24
-10 days in the lab; 14 hrs a day: 140
-Target: 5 taxa per sample; 140/7.5 = 18; 14/10 = 14
-
-```{r}
-total_samples = function(hrs_per_sample, no_taxa, total_hrs = 140){
-  tmp = total_hrs / (no_taxa * hrs_per_sample)
-  return(signif(tmp,3))
-}
-M = outer(c(1,1.5,2),c(3,4,5),total_samples, total_hrs = 140)
-row.names(M) = c(1,1.5,2)
-colnames(M) = c(3,4,5)
-print('Number of samples at One Core')
-M
-```
-
-Anywhere from 18 to 30 samples is reasonable.
-At 3 cores 5 taxa:
-Minimum 4 samples per core 
-Maximum 9 samples per core
-
-```{r}
-print('No. of samples at Three Cores')
-signif(M/3, 3)
-```
-At 2 cores 5 taxa:
-Minimum 7 samples per core
-Maximum 14 samples per core
-
-```{r}
-print('No. of samples at Two Cores')
-signif(M/2, 3)
-```
-
 # Introduction
@@
 - add cores (e.g. MFM, Tiefer See or Italian cores), Festi core was quite organic rich
--
+- add more grains from already measured samples (to hit min 25 or 50)
 

### 2020-05-22 manuscript.Rmd
updated core segment in manus; sent to Al

@@
 
-Totenmoos [@heissUltenValleySouth2005]:
-- bog-sediment core; past 15000 yrs
-- min 1000 grains counted
-- Picea, Pinus very common, switch at 10000 BP from Pinus to Picea
-- Alnus, Betula throughout
-- Abies, Fagus after 9000 BP
-- trace Quercus
-
 Younger Dryas [@brauerHighResolutionSediment1999]:
@@
 
-Tinner alpine:
-
-Tinner meso-mediterranean:
-[@bisculmHoloceneVegetationFire2012; @vanniereClimateHumandrivenFire2008]
-Quercus sections Quercus, Cerris and Ilex
-
-Tinner sub-mediterranean:
-[@zoller19601960]
-Quercus sections Quercus, Cerris and Ilex
+Tinner/Cesar cores:
+Urio Quattrocchi (supra mediterranean) [@bisculmHoloceneVegetationFire2012]
+Sa Curcurica (thermo mediterranean) [@beffaVegetationFireHistory2016]
+Lago del Sangiatto (southern alps)
 
@@
 # Methods and Materials
+
+Pollen extraction:
+
+1. add 1 cube to vial (~1cm^-3^)
+2. wash with water -> dissolve sediment vortex
+3. centrifuge 3min 3000rpm
+4. decant; save decant
+5. add 3 ml 1.43 SPT solution
+6. vortex
+7. centrifuge 3min 3000rpm
+8. decant supernatant into new vial
+9. repeat 5-7; decant into same vial
+10. wash with water; save decant for checking
+11. wash with water; save decant x2
+12. wash residue with water; save in vial
+
+Preprocessing:
+- SG 2nd degree, 2 poly, windowsize 5 or 6
+- EMSC linear (probably needs to involve background spectra to filter them out)
+
 PCA on grains for exploration
 
-regression of spectra on depth
+PLS regression of spectra on depth
 
@@
 
-average all spectra and stack plot
+average all spectra and stack plots of different depth
 
@@
 [@regnellPreparativeCentrifugationNew1996] density of pollen 1.4
+
+# References

### 2020-10-06 manuscript.Rmd
added cross-ref with bookdown

@@
 output:
-  pdf_document: default
+  bookdown::pdf_document2:
+    extra_dependencies: ["flafter"]
 bibliography: references.bib
 ---
+
 # Notes
@@
 
-9 depth from Dalmutladdo core. targeted and extracted Pinus pollen (see table)
+9 depth from Dalmutladdo core. targeted and extracted Pinus pollen (Table \@ref(tab:pollencounts), Table \@ref(tab:fresh))
 
@@
 
-```{r}
+```{r pollencounts, echo=FALSE}
 dalmutladdo = data.frame(ID = c("DAL_08","DAL_20","DAL_40","DAL_60","DAL_90", "DAL_110","DAL_130","DAL_140","DAL_150"),
@@
                          betula_pollen = c(0,15,0,0,0,0,0,0,0))
-dalmutladdo
+knitr::kable(dalmutladdo, caption = "Pollen counts")
 ```
 
-```{r}
+```{r fresh, echo=FALSE}
 fresh_pollen = data.frame(ID = c("BGO_PS7", "BGO_PS8", "INN_PS13", "INN_PS19A","NMBU_Pin_syl_14E1", "NMBU_Pin_syl_14H1", "NMBU_Aln_glu_14E1", "NMBU_Aln_glu_14J1", "NMBU_Aln_inc_14B1", "NMBU_Aln_inc_14G1", "NMBU_Bet_pub_14A1", "NMBU_Bet_pub_14B1", "NMBU_Cor_ave_14B1", "NMBU_Cor_ave_14C1"),
                           extracted = c(109,55,51,24,0,0,0,0,0,0,0,0,0,0))
-fresh_pollen
+knitr::kable(fresh_pollen, caption = "Fresh pollen counts")
 ```

### 2020-10-07 manuscript.Rmd
added table formatting; added Biology Letters CSL

@@
   bookdown::pdf_document2:
+    toc: false
     extra_dependencies: ["flafter"]
 bibliography: references.bib
+csl: biology-letters.csl
 ---
@@
                          betula_pollen = c(0,15,0,0,0,0,0,0,0))
-knitr::kable(dalmutladdo, caption = "Pollen counts")
+knitr::kable(dalmutladdo, format="latex", booktabs=TRUE,
+             caption = "Pollen counts")
 ```
@@
                           extracted = c(109,55,51,24,0,0,0,0,0,0,0,0,0,0))
-knitr::kable(fresh_pollen, caption = "Fresh pollen counts")
+knitr::kable(fresh_pollen, format="latex", booktabs=TRUE,
+             caption = "Fresh pollen counts")
 ```
+
+9 fossil samples from 1 core
+
+2 trees of Pinus at each location
 

### 2020-10-07 manuscript.Rmd
wrote on introduction

@@
 
-# Introduction
 Research Questions:
@@
 
-# Methods and Materials
 ## Labwork Report
@@
 
+# TODO
+
+- extract fresh Pinus from images
+- measure samples from 2 more cores (9 each)
+- mark Alnus and BetCor on images
+- extract Alnus BetCor from Dal samples
+- extract fresh Alnus BetCor
+
 # Introduction
+
+Main Goal: Explore the chemical differences between fresh and untreated fossil pollen.
+
+The study of plant remains is an integral part of Palaeoecology and Palaeobotany.
+Spores and pollengrains, besides macrofossils, are utilized to study past environmental and climatic change, migration of plant communities as well as the development and evolution of plants. 
+EXAMPLES
+
+Pollen has shown to be resistant to diagenetic processes and found in most palaeoecological records.
+
+FTIR used on fossil papers in the past use acetolysed pollen or other chemically extracted pollen. 
+There has not been a comparison between fresh and untreated fossil pollen.
+In order to compare which compounds are missing and still present in fossil pollen.
+
+RQs
+i. Difference between fresh and fossil grains.
+ii. effect of orientation and depth/age on chemistry
+iii. difference to other species
+    -> needs extracting of spectra from files (marking of grains on images)
+    -> add another core
 

### 2020-10-07 manuscript.Rmd
more on introduction

@@
 
-The study of plant remains is an integral part of Palaeoecology and Palaeobotany.
-Spores and pollengrains, besides macrofossils, are utilized to study past environmental and climatic change, migration of plant communities as well as the development and evolution of plants. 
+The study of plant micro and macrofossil (spores, seed and pollen),recovered from lake sediments and peat bogs, is an integral part of Palaeoecology and Palaeobotany.
+Pollengrains are an important tool to reconstruct past vegetational, environmental and climatic changes. 
 EXAMPLES
 
-Pollen has shown to be resistant to diagenetic processes and found in most palaeoecological records.
+Palaeo relies on taxonomic identification of pollen based on morphology.
+Not always possible.
+Recently IR spectro has emerged as a promising candidate to improve taxonomic resolution and automation of pollen counting.
+
+Pollen is preserved in lake sediments and has shown to be resistant to diagenetic processes and found in most palaeoecological records.
 

### 2020-10-08 manuscript.Rmd
wrote more on introduction; started writing read data file

@@
 The study of plant micro and macrofossil (spores, seed and pollen),recovered from lake sediments and peat bogs, is an integral part of Palaeoecology and Palaeobotany.
-Pollengrains are an important tool to reconstruct past vegetational, environmental and climatic changes. 
+Pollengrains are resistant to the diagenetic processes that occur in the formation of sediments, bogs and other deposits, which makes pollen records 
+an important tool to reconstruct past vegetational, environmental and climatic changes. 
 EXAMPLES
 
+The main component, which makes pollen grains so resistant is sporopollenin, a highly resistant biopolymer, making up the majority of the outer pollen grain wall (exine).
+
+In recent years, the study of pollen grain chemistry has lead to new information on the chemical structure of sporopollenin as well as taxonomic and environmental information stored in the chemistry of pollen.
+
+pollengrains are higly resistant to the processes that occur in the formation of sediments, bogs and other deposits (diagenesis).
+Sporopollenin is a biopolymer, which is the main building block for pollen grains. 
 Palaeo relies on taxonomic identification of pollen based on morphology.
 Not always possible.
-Recently IR spectro has emerged as a promising candidate to improve taxonomic resolution and automation of pollen counting.
-
-Pollen is preserved in lake sediments and has shown to be resistant to diagenetic processes and found in most palaeoecological records.
+Recently IR spectro has emerged as a promising candidate to improve taxonomic resolution and automation ofpollen counting.
+EXAMPLES
 
@@
 In order to compare which compounds are missing and still present in fossil pollen.
+chemical degradation of pollen during diagenesis or depth.
 

### 2020-10-08 R/read_data.R

@@ -1,0 +1,7 @@
+library(here)
+
+raw = read.csv(here("data", "input", "spectra", "DAL_spectra_raw.csv"), check.names = F, colClasses = "character")
+
+spectra_raw = as.matrix((raw[-c(1,2),c(1:765)]))
+spectra_raw = apply(spectra_raw, 2, as.numeric)
+spectra_meta = raw[,c(766:773)]

### 2020-10-08 manuscript.Rmd
wrote on introdution

@@
 
-The main component, which makes pollen grains so resistant is sporopollenin, a highly resistant biopolymer, making up the majority of the outer pollen grain wall (exine).
+The high preservation of pollen grains in the geological record is because of the resistant biopolymer sporopollenin, which is the main building block of the outer pollen grain wall (exine).
+The exact structure of sporopollenin is unknown, but recent studies give a more detailed picture of this complex substance [@liMolecularStructurePlant2019].
 
-In recent years, the study of pollen grain chemistry has lead to new information on the chemical structure of sporopollenin as well as taxonomic and environmental information stored in the chemistry of pollen.
+In recent years, the study of pollen grain chemistry has lead to new information on the chemical structure of sporopollenin.
+Taxonomic differences have been identified in the chemistry of pollen, as well as environmental information stored in the chemistry of pollen (UV Proxy).
 

### 2020-10-12 manuscript.Rmd
changes from Fri

--- 2020-10-07 manuscript.Rmd
@@
 The study of plant micro and macrofossil (spores, seed and pollen),recovered from lake sediments and peat bogs, is an integral part of Palaeoecology and Palaeobotany.
-Pollengrains are an important tool to reconstruct past vegetational, environmental and climatic changes. 
+Pollengrains are resistant to the diagenetic processes that occur in the formation of sediments, bogs and other deposits, which makes pollen records 
+an important tool to reconstruct past floristic, vegetational and environmental changes. 
 EXAMPLES
 
-Palaeo relies on taxonomic identification of pollen based on morphology.
-Not always possible.
-Recently IR spectro has emerged as a promising candidate to improve taxonomic resolution and automation of pollen counting.
+The high preservation of pollen grains in the geological record is because of the resistant biopolymer sporopollenin, which is the main building block of the outer pollen grain wall (exine).
 
-Pollen is preserved in lake sediments and has shown to be resistant to diagenetic processes and found in most palaeoecological records.
+More and more details have been revealed about the exact structure of sporopollenin, e.g.
+importance of aromatic compounds, cross-linked aliphatic components [@liMolecularStructurePlant2019], similarity to lignins.
+
+In recent years, the study of pollen grain chemistry has lead to new information important for biological and palaeoecological research.
+Taxonomic differences have been identified in the chemistry of pollen, which highlight the potential of pollen identification with chemical methods in palaeoecological research.
+Chemical fingerprinting of pollen has the potential of higher taxonomic resolution, automation of pollen identification and offers a more objective method than traditional morphological methods.
+as well as an environmental signal stored in the chemistry of pollen (UV Proxy), which could be used as a UV-proxy.
+
+acetolysis alters sporopollenin [@dominguezIsolationIntactPollen1998; @schulzeosthoffPhenolsIntegratedCompounds1987] affecting solubility
 

### 2020-10-12 R/read_data.R
wrote data reading file; assigns unique ID; read DAL samples

@@ -1,7 +1,23 @@
 library(here)
+library(stringi)
 
 raw = read.csv(here("data", "input", "spectra", "DAL_spectra_raw.csv"), check.names = F, colClasses = "character")
 
 spectra_raw = as.matrix((raw[-c(1,2),c(1:765)]))
 spectra_raw = apply(spectra_raw, 2, as.numeric)
-spectra_meta = raw[,c(766:773)]
+spectra_meta = raw[-c(1,2),c(766:773)]
+
+ids = stri_rand_strings(nrow(spectra_meta),5)
+# Check IDs for duplicates
+while (sum(duplicated(ids))!=0) {
+  ids = stri_rand_strings(nrow(spectra_meta),5)
+}
+
+spectra_meta$ids = ids
+rownames(spectra_raw) = ids
+
+DAL = spectra_meta
+DAL$raw = spectra_raw
+DAL = DAL[,c("ids","type","species","depth","raw")]
+
+saveRDS(DAL, file = here("data","output","DAL_spectra.rds"))

### 2020-12-03 manuscript.Rmd
read_data file; loading core data chunk; added authors and affiliations

@@
 ---
+
+```{r message=FALSE, warning=FALSE, include=FALSE}
+library(here)
+source(here("R","authors_and_affiliations.R"))
+```
+
+---
+author: `r Authors`
+---
+`r Affiliations`
+
+as
+
+```{r}
+if (file.exists(here("data","output","DAL_spectra.rds"))) {
+  DAL = readRDS(here("data","output","DAL_spectra.rds"))
+} else {
+  source(here("R","read_data.R"))
+  read_data("DAL")
+  DAL = readRDS(here("data","output","DAL_spectra.rds"))
+}
+
+if (file.exists(here("data","output","MFM_spectra.rds"))) {
+  MFM = readRDS(here("data","output","MFM_spectra.rds"))
+} else {
+  source(here("R","read_data.R"))
+  read_data("MFM")
+  MFM = readRDS(here("data","output","MFM_spectra.rds"))
+}
+
+if (file.exists(here("data","output","TSK_spectra.rds"))) {
+  TSK = readRDS(here("data","output","TSK_spectra.rds"))
+} else {
+  source(here("R","read_data.R"))
+  read_data("TSK")
+  TSK = readRDS(here("data","output","TSK_spectra.rds"))
+}
+```
 

### 2020-12-03 R/authors_and_affiliations.R

@@ -1,0 +1,73 @@
+library(tidyr)
+library(dplyr)
+
+dat <- data.frame(FirstName = c("Florian", 
+                                "Valeria", 
+                                "Boris", 
+                                "Alistair W R"),
+                  LastName = c("Muthreich", 
+                               "Tafintseva", 
+                               "Zimmermann", 
+                               "Seddon"),
+                  Affiliation1 = c("University of Bergen, Department of Biological Sciences, Norway", 
+                                   "University of Life Sciences, Ås, Norway",
+                                   "University of Life Sciences, Ås, Norway",
+                                   "University of Bergen, Department of Biological Sciences, Norway"),
+                  Affiliation2 = c(NA, 
+                                   NA, 
+                                   NA, 
+                                   "Bjerknes Centre for Climate Research, Bergen"))
+dat <- dat %>% 
+  # gather all affiliations in one column
+  gather(key = Number, value = Affiliation, Affiliation1, Affiliation2) %>%
+  # remove rows with no Affiliations
+  dplyr::filter(!is.na(Affiliation))
+
+# Function to get affiliations ranked from 1 to n (this function was found on Stack Overflow)
+rankID <- function(x){
+  su=sort(unique(x))
+  for (i in 1:length(su)) x[x==su[i]] = i
+  return(x)
+}
+
+
+NameAffList <- dat %>% 
+  dplyr::arrange(LastName, Affiliation) %>% 
+  rowwise() %>% 
+  # extract the first letter of each first name and put a dot after each letter
+  mutate(
+    Initials = paste(stringi::stri_extract_all(regex = "\\b([[:alpha:]])", str = FirstName, simplify = TRUE), collapse = ". "),
+    Initials = paste0(Initials, ".")) %>%
+  ungroup() %>% 
+  # add a column from 1 to n
+  mutate(ID = 1:n()) %>%
+  group_by(Affiliation) %>% 
+  # replace ID with min number (same affiliations become the same number)
+  mutate(ID = min(ID)) %>% 
+  ungroup() %>% 
+  # use function above to assign new ID from 1 to n
+  mutate(ID = rankID(ID)) %>%
+  #Paste Last and Initials
+  mutate(name = paste0(LastName, ", ", Initials)) %>%
[…]

### 2020-12-03 R/read_data.R

@@ -9,33 +9,30 @@ repl = function(data, c){
   return(tmp)
 }
 
-# assign ids
-# assign metadata
-# save as rds
-
-raw = read.csv(here("data", "input", "spectra", "DAL_spectra_raw.csv"), check.names = F, colClasses = "character")
-
-spectra_raw = as.matrix((raw[-c(1,2),c(1:765)]))
-spectra_raw = apply(spectra_raw, 2, as.numeric)
-spectra_meta = raw[-c(1,2),c(766:ncol(raw))]
-sc = strtrim(spectra_meta$`Sample Name`[1], 3)
-ids = paste0(sc,stri_rand_strings(nrow(spectra_meta),5))
-
-# Check IDs for duplicates
-while (sum(duplicated(ids))!=0) {
+read_data = function(sc){
+  raw = read.csv(here("data", "input", "spectra", paste0(sc,"_spectra_raw.csv")), 
+                 check.names = F, 
+                 colClasses = "character")
+  spectra_raw = as.matrix((raw[-c(1,2),c(1:765)]))
+  spectra_raw = apply(spectra_raw, 2, as.numeric)
+  spectra_meta = raw[-c(1,2),c(766:ncol(raw))]
+  sc = strtrim(spectra_meta$`Sample Name`[1], 3)
   ids = paste0(sc,"_",stri_rand_strings(nrow(spectra_meta),5))
+  
+  # Check IDs for duplicates
+  while (sum(duplicated(ids))!=0) {
+    ids = paste0(sc,"_",stri_rand_strings(nrow(spectra_meta),5))
+  }
+  
+  spectra_meta$ids = ids
+  spectra_meta$species = repl(spectra_meta, c("Blank", "Pinus", "Abies", "BetCor", "Alnus"))
+  spectra_meta$treatment = repl(spectra_meta, c("acet"))
+  spectra_meta$orientation = repl(spectra_meta,c("_side"))
+  
+  rownames(spectra_raw) = ids
+  
+  data = spectra_meta
+  data$raw = spectra_raw
+  
+  saveRDS(data, file = here("data","output",paste0(sc,"_spectra.rds")))
 }
-
-spectra_meta$ids = ids
-spectra_meta$species = repl(DAL, c("Blank", "Pinus", "Abies", "BetCor", "Alnus"))
-spectra_meta$treatment = repl(DAL, c("acet"))
-spectra_meta$orientation = repl(DAL,c("_side"))
-
[…]

### 2020-12-03 manuscript.Rmd
dedicated packages file; function to extract sample depth from filename

@@
 library(here)
+source(here("R","packages.R"))
 source(here("R","authors_and_affiliations.R"))
@@
 
-as
+Abstract
 
@@
 }
+```
+
+```{r}
+count_grains = function(core){
+  counts = core %>%
+    group_by(type, depth, treatment) %>%
+    summarise(count=n())
+  return(counts)
+}
+count_grains(TSK)
 ```

### 2020-12-03 R/authors_and_affiliations.R

@@ -1,6 +1,3 @@
-library(tidyr)
-library(dplyr)
-
 dat <- data.frame(FirstName = c("Florian", 
                                 "Valeria", 
                                 "Boris", 

### 2020-12-03 R/packages.R

@@ -1,0 +1,8 @@
+library(tidyr)
+library(stringi)
+library(dplyr)
+library(ggplot2)
+library(EMSC)
+library(pls)
+library(kableExtra)
+library(reshape2)

### 2020-12-03 R/read_data.R

@@ -1,6 +1,3 @@
-library(here)
-library(stringi)
-
 repl = function(data, c){
   tmp = rep("", nrow(data))
   for (i in c) {
@@ -28,6 +25,13 @@ read_data = function(sc){
   spectra_meta$species = repl(spectra_meta, c("Blank", "Pinus", "Abies", "BetCor", "Alnus"))
   spectra_meta$treatment = repl(spectra_meta, c("acet"))
   spectra_meta$orientation = repl(spectra_meta,c("_side"))
+  dpth = str_split_fixed(spectra_meta$`Sample Name`, "_", 4)
+  if (sc == "MFM") {
+    dpth = paste0(dpth[,2],"_",dpth[,3])
+  } else {
+    dpth = dpth[,2]
+  }
+  spectra_meta$depth = dpth
   
   rownames(spectra_raw) = ids
   

### 2020-12-07 manuscript.Rmd
pollen count tables; fixed authors; fixed utf formatting in references; added information from biology letters guidelines to manuscript.

@@
 title: 'Viability of FTIR on fossil pollen'
-author: 'Florian Muthreich'
 date: '2020/02/04'
@@
 
-Abstract
+2500 words not including in tables title page references
+2 figs and 2 tables
+figures separate legends in text
+separate cover letter
 
+# Chemistry of fossil Pinus pollen and the effect of acetolysis
+
+## Abstract
+200 words no abreviations/refs
+
+Keywords: FTIR, fossil pollen, Holocene, density separation, acetolysis, pollen chemistry
+
+<!-- READ DATA -->
 ```{r}
@@
 
-```{r}
+```{r DALcounts, echo=FALSE, results='asis'}
 count_grains = function(core){
   counts = core %>%
+    subset(type == "pollen" & species == "Pinus") %>% 
     group_by(type, depth, treatment) %>%
@@
 }
-count_grains(TSK)
+kable(count_grains(DAL), booktabs = T, format = "latex",
+      caption = "Pinus pollen grain count in Dalmutladdo core.")
+#      digits = 1, format.args = list(scientific = F),
+#      col.names = c("")) %>% 
+#  add_header_above()
+```
+
+```{r TSKcounts, echo=FALSE, results='asis'}
+kable(count_grains(TSK), booktabs = T, format = "latex",
+      caption = "Pinus pollen grain count in Tiefer See core.")
+```
+
+```{r MFMcounts, echo=FALSE, results='asis'}
+kable(count_grains(MFM), booktabs = T, format = "latex",
+      caption = "Pinus pollen grain count in Meerfelder Maar core.")
 ```

### 2020-12-07 R/read_data.R

@@ -1,5 +1,5 @@
-repl = function(data, c){
-  tmp = rep("", nrow(data))
+repl = function(data, c, default=""){
+  tmp = rep(default, nrow(data))
   for (i in c) {
     tmp[grep(i,data$Filename)] = i
   }
@@ -23,7 +23,7 @@ read_data = function(sc){
   
   spectra_meta$ids = ids
   spectra_meta$species = repl(spectra_meta, c("Blank", "Pinus", "Abies", "BetCor", "Alnus"))
-  spectra_meta$treatment = repl(spectra_meta, c("acet"))
+  spectra_meta$treatment = repl(spectra_meta, c("acet"),c("SPT"))
   spectra_meta$orientation = repl(spectra_meta,c("_side"))
   dpth = str_split_fixed(spectra_meta$`Sample Name`, "_", 4)
   if (sc == "MFM") {

### 2020-12-07 manuscript.Rmd
moved notes section; changed title; adjusted RQs.

@@
 ---
-title: 'Viability of FTIR on fossil pollen'
+title: 'Chemistry of fossil Pinus pollen and the effect of acetolysis'
 date: '2020/02/04'
@@
 
-# Chemistry of fossil Pinus pollen and the effect of acetolysis
-
-## Abstract
+# Abstract
 200 words no abreviations/refs
@@
 
+# Introduction
+
+Main Goal: Explore the chemical differences between fresh and untreated fossil pollen.
+
+The study of plant micro and macrofossil (spores, seed and pollen), recovered from lake sediments and peat bogs, is an integral part of Palaeoecology and Palaeobotany.
+Pollengrains are resistant to the diagenetic processes that occur in the formation of sediments, bogs and other deposits, which makes pollen records an important tool to reconstruct past floristic, vegetational and environmental changes.
+EXAMPLES
+
+The high preservation of pollen grains in the geological record is because of the resistant biopolymer sporopollenin, which is the main building block of the outer pollen grain wall (exine).
+
+More and more details have been revealed about the exact structure of sporopollenin, e.g.
+importance of aromatic compounds, cross-linked aliphatic components [@liMolecularStructurePlant2019], similarity to lignins.
+
+In recent years, the study of pollen grain chemistry has lead to new information important for biological and palaeoecological research.
+Taxonomic differences have been identified in the chemistry of pollen, which highlight the potential of pollen identification with chemical methods in palaeoecological research.
+Chemical fingerprinting of pollen has the potential of higher taxonomic resolution, automation of pollen identification and offers a more objective method than traditional morphological methods.
+as well as an environmental signal stored in the chemistry of pollen (UV Proxy), which could be used as a UV-proxy.
+
+acetolysis alters sporopollenin [@dominguezIsolationIntactPollen1998; @schulzeosthoffPhenolsIntegratedCompounds1987] affecting solubility
+
+FTIR used on fossil papers in the past use acetolysed pollen or other chemically extracted pollen. 
+There has not been a comparison between fresh and untreated fossil pollen.
+In order to compare which compounds are missing and still present in fossil pollen.
+chemical degradation of pollen during diagenesis or depth.
+
+RQs
+i. Difference between fresh and fossil grains.
+i.i effect of orientation on chemistry
+i.ii effect of depth/age on chemistry
+ii Differences between cores
+
+# Methods and Materials
+
+Pollen extraction:
+
+1. add 1 cube to vial (~1cm^-3^)
+2. wash with water -> dissolve sediment vortex
+3. centrifuge 3min 3000rpm
+4. decant; save decant
+5. add 3 ml 1.43 SPT solution
+6. vortex
+7. centrifuge 3min 3000rpm
+8. decant supernatant into new vial
+9. repeat 5-7; decant into same vial
+10. wash with water; save decant for checking
+11. wash with water; save decant x2
+12. wash residue with water; save in vial
+
+Preprocessing:
+- SG 2nd degree, 2 poly, windowsize 5 or 6
+- EMSC linear (probably needs to involve background spectra to filter them out)
+
+PCA on grains for exploration
+
+PLS regression of spectra on depth
+
+repeat on different grain averages (1, 3, 5, 10)
+
+average all spectra and stack plots of different depth
+
+Pollen extracted from sediment samples 
+[@kampsAgentSeparationDissolved1985] SPT patent
+[@nakagawaDensemediaSeparationMore1998] 
+[@regnellPreparativeCentrifugationNew1996] density of pollen 1.4
+
 # Notes
@@
 
-Tiefer See:
+Tiefer See [@theuerkaufEffectsChangesLand2015]:
 - 1870-2010
@@
 - Pinus pollen
-
-Tinner/Cesar cores:
-Urio Quattrocchi (supra mediterranean) [@bisculmHoloceneVegetationFire2012]
-Sa Curcurica (thermo mediterranean) [@beffaVegetationFireHistory2016]
-Lago del Sangiatto (southern alps)
 
@@
 
-# Introduction
-
-Main Goal: Explore the chemical differences between fresh and untreated fossil pollen.
-
-The study of plant micro and macrofossil (spores, seed and pollen),recovered from lake sediments and peat bogs, is an integral part of Palaeoecology and Palaeobotany.
-Pollengrains are resistant to the diagenetic processes that occur in the formation of sediments, bogs and other deposits, which makes pollen records 
-an important tool to reconstruct past floristic, vegetational and environmental changes. 
-EXAMPLES
-
-The high preservation of pollen grains in the geological record is because of the resistant biopolymer sporopollenin, which is the main building block of the outer pollen grain wall (exine).
-
-More and more details have been revealed about the exact structure of sporopollenin, e.g.
-importance of aromatic compounds, cross-linked aliphatic components [@liMolecularStructurePlant2019], similarity to lignins.
-
-In recent years, the study of pollen grain chemistry has lead to new information important for biological and palaeoecological research.
-Taxonomic differences have been identified in the chemistry of pollen, which highlight the potential of pollen identification with chemical methods in palaeoecological research.
-Chemical fingerprinting of pollen has the potential of higher taxonomic resolution, automation of pollen identification and offers a more objective method than traditional morphological methods.
-as well as an environmental signal stored in the chemistry of pollen (UV Proxy), which could be used as a UV-proxy.
-
-acetolysis alters sporopollenin [@dominguezIsolationIntactPollen1998; @schulzeosthoffPhenolsIntegratedCompounds1987] affecting solubility
-
-FTIR used on fossil papers in the past use acetolysed pollen or other chemically extracted pollen. 
-There has not been a comparison between fresh and untreated fossil pollen.
-In order to compare which compounds are missing and still present in fossil pollen.
-chemical degradation of pollen during diagenesis or depth.
-
-RQs
-i. Difference between fresh and fossil grains.
-ii. effect of orientation and depth/age on chemistry
-iii. difference to other species
-    -> needs extracting of spectra from files (marking of grains on images)
-    -> add another core
-
-# Methods and Materials
-
-Pollen extraction:
-
-1. add 1 cube to vial (~1cm^-3^)
-2. wash with water -> dissolve sediment vortex
-3. centrifuge 3min 3000rpm
-4. decant; save decant
-5. add 3 ml 1.43 SPT solution
-6. vortex
-7. centrifuge 3min 3000rpm
-8. decant supernatant into new vial
-9. repeat 5-7; decant into same vial
-10. wash with water; save decant for checking
-11. wash with water; save decant x2
-12. wash residue with water; save in vial
-
-Preprocessing:
-- SG 2nd degree, 2 poly, windowsize 5 or 6
-- EMSC linear (probably needs to involve background spectra to filter them out)
-
-PCA on grains for exploration
-
-PLS regression of spectra on depth
-
-repeat on different grain averages (1, 3, 5, 10)
-
-average all spectra and stack plots of different depth
-
-Pollen extracted from sediment samples 
-[@kampsAgentSeparationDissolved1985] SPT patent
-[@nakagawaDensemediaSeparationMore1998] 
-[@regnellPreparativeCentrifugationNew1996] density of pollen 1.4
-
 # References

### 2020-12-07 R/helper_functions.R
wrote preproc and plotting function

@@ -1,0 +1,19 @@
+pre_processing = function(spectra, ws){
+  foo_sg = SavitzkyGolay(spectra, poly = 2, width = 5, deriv = 2)
+  foo_emsc = EMSC(foo_sg, degree = 1)$corrected
+  rownames(foo_emsc) = rownames(spectra)
+  colnames(foo_emsc) = colnames(spectra)
+  return(foo_emsc)
+}
+
+plot_spectra = function(data, spec_id, sel, id_vars = c("ids","type","species","treatment","orientation","depth")){
+  pdata = subset(data, eval(sel))
+  gdata = melt(cbind(pdata[,id_vars],
+                       as.data.frame(pdata[,c(spec_id)])),
+                 id.vars=id_vars)
+  gdata$variable = as.numeric(as.character(gdata$variable))
+  ggplot(gdata, aes(variable, value, color = depth, group = ids)) +
+    geom_line() + scale_x_reverse() +
+    theme_bw()
+  #return(gdata)
+}

### 2020-12-07 R/helper_functions.R
preproc function cuts spectra and carries row/colnames

@@ -1,8 +1,15 @@
-pre_processing = function(spectra, ws){
-  foo_sg = SavitzkyGolay(spectra, poly = 2, width = 5, deriv = 2)
+pre_processing = function(spectra, ws, cut=T){
+  wn = as.numeric(colnames(spectra))
+  foo_sg = SavitzkyGolay(spectra, poly = 2, width = ws, deriv = 2)
+  if (cut) {
+    foo_sg = foo_sg[,wn<1900 & wn>700]
+    colnames(foo_sg) = wn[wn<1900 & wn>700]
+  } else {
+    colnames(foo_sg) = wn
+  }
   foo_emsc = EMSC(foo_sg, degree = 1)$corrected
   rownames(foo_emsc) = rownames(spectra)
-  colnames(foo_emsc) = colnames(spectra)
+  colnames(foo_emsc) = colnames(foo_sg)
   return(foo_emsc)
 }
 

### 2020-12-10 manuscript.Rmd
wrote preproc and plotting function
wrote preprocess file and hook into manuscript

@@
   TSK = readRDS(here("data","output","TSK_spectra.rds"))
+}
+```
+
+<!-- PRE-PROCESS DATA -->
+```{r}
+pre_process = TRUE
+source(here("R","helper_functions.R"))
+if (!hasName(DAL, "sg2")) {
+  source(here("R","pre_process.R"))
+} else {
+  TSK = readRDS(here("data","output","TSK_spectra.rds"))
+  MFM = readRDS(here("data","output","MFM_spectra.rds"))
+  DAL = readRDS(here("data","output","DAL_spectra.rds"))
 }
@@
 
-# TODO
-
-- extract fresh Pinus from images
-- measure samples from 2 more cores (9 each)
-- mark Alnus and BetCor on images
-- extract Alnus BetCor from Dal samples
-- extract fresh Alnus BetCor
-
 # References

### 2020-12-10 R/pre_process.R

@@ -1,0 +1,22 @@
+pre_processing = function(spectra, ws, cut=T){
+  wn = as.numeric(colnames(spectra))
+  foo_sg = SavitzkyGolay(spectra, poly = 2, width = ws, deriv = 2)
+  if (cut) {
+    foo_sg = foo_sg[,wn<1900 & wn>700]
+    colnames(foo_sg) = wn[wn<1900 & wn>700]
+  } else {
+    colnames(foo_sg) = wn
+  }
+  foo_emsc = EMSC(foo_sg, degree = 1)$corrected
+  rownames(foo_emsc) = rownames(spectra)
+  colnames(foo_emsc) = colnames(foo_sg)
+  return(foo_emsc)
+}
+
+DAL$sg2 = pre_processing(DAL$raw, 6)
+TSK$sg2 = pre_processing(TSK$raw, 6)
+MFM$sg2 = pre_processing(MFM$raw, 6)
+
+saveRDS(DAL, here("data","output","DAL_spectra.rds"))
+saveRDS(TSK, here("data","output","TSK_spectra.rds"))
+saveRDS(MFM, here("data","output","DAL_spectra.rds"))

### 2020-12-10 R/pre_process.R
removed superfluous code from precessing chunk; changed to ws 5 from 6

@@ -13,9 +13,9 @@ pre_processing = function(spectra, ws, cut=T){
   return(foo_emsc)
 }
 
-DAL$sg2 = pre_processing(DAL$raw, 6)
-TSK$sg2 = pre_processing(TSK$raw, 6)
-MFM$sg2 = pre_processing(MFM$raw, 6)
+DAL$sg2 = pre_processing(DAL$raw, 5)
+TSK$sg2 = pre_processing(TSK$raw, 5)
+MFM$sg2 = pre_processing(MFM$raw, 5)
 
 saveRDS(DAL, here("data","output","DAL_spectra.rds"))
 saveRDS(TSK, here("data","output","TSK_spectra.rds"))

### 2020-12-10 manuscript.Rmd
removed superfluous code from precessing chunk; changed to ws 5 from 6
prepping PCA

--- 2020-12-07 manuscript.Rmd
@@
 }
+```
+
+<!-- PRE-PROCESS DATA -->
+```{r}
+if (!hasName(DAL, "sg2")) {
+  source(here("R","pre_process.R"))
+}
+```
+
+<!-- PCA -->
+```{r}
+DAL$Label = "DAL"
+TSK$Label = "TSK"
+MFM$Label = "MFM"
+
+pca_data = rbind(subset(DAL, species == "Pinus"), 
+        subset(TSK, species == "Pinus"), 
+        subset(MFM, species == "Pinus"))
+
 ```

### 2021-02-08 manuscript.Rmd
added missing package to packages.R; fixed typo in pre_process; wrote pca and pc plotting code

@@
         subset(MFM, species == "Pinus"))
-
+pca_data[pca_data$treatment == "","treatment"] = "SPT"
+pca_res = princomp(pca_data$sg2, scores = T)
+scores =  as.data.frame(pca_res$scores[,1:10]) %>%
+  mutate(ids = rownames(.), 
+         core = pca_data$Label, 
+         treatment = pca_data$treatment
+         ) #%>%
+  #melt(.,id.vars = "ids")
+ggplot(scores, aes(Comp.1, Comp.2, color = core)) + geom_point()
+# pca_data[scores[scores[,2] < -.2,"ids"]==pca_data$ids,1:10]
 ```

### 2021-02-08 R/packages.R

@@ -1,5 +1,6 @@
 library(tidyr)
 library(stringi)
+library(stringr)
 library(dplyr)
 library(ggplot2)
 library(EMSC)

### 2021-02-08 R/pre_process.R

@@ -19,4 +19,4 @@ MFM$sg2 = pre_processing(MFM$raw, 5)
 
 saveRDS(DAL, here("data","output","DAL_spectra.rds"))
 saveRDS(TSK, here("data","output","TSK_spectra.rds"))
-saveRDS(MFM, here("data","output","DAL_spectra.rds"))
+saveRDS(MFM, here("data","output","MFM_spectra.rds"))

### 2021-02-08 R/pca.R
checking outliers

@@ -1,0 +1,44 @@
+DAL$Label = "DAL"
+TSK$Label = "TSK"
+MFM$Label = "MFM"
+pca_wrap = function(pca_data,sel){
+  pca_data[pca_data$treatment == "","treatment"] = "SPT"
+  if (!sel=="None") {
+    pca_data = subset(pca_data, Label == sel)
+  }
+  pca_res = prcomp(pca_data$sg2)
+  scores =  as.data.frame(pca_res$x[,1:10]) %>%
+    mutate(ids = rownames(.),
+           core = pca_data$Label,
+           treatment = pca_data$treatment,
+           depth = pca_data$depth
+           )
+  return(scores)
+}
+pca_data = rbind(subset(DAL, species == "Pinus"), 
+                 subset(TSK, species == "Pinus"), 
+                 subset(MFM, species == "Pinus"))
+ #%>%
+#melt(.,id.vars = "ids")
+scores = pca_wrap(pca_data,"DAL")
+ggplot(scores, aes(PC1, PC2, color = depth)) + geom_point(aes(shape = treatment), size = 4)
+pca_data[pca_data$ids %in% scores[scores[,2] < -.05 | scores[,1] > .035,"ids"],1:10]
+plot_spectra(DAL,"raw",sel = expression(ids == "DAL_rx1GZ")) 
+plot_spectra(DAL,"raw",sel = expression(ids == "DAL_2cxp7"))
+plot_spectra(DAL,"raw",sel = expression(ids == "DAL_VzFi0"))
+plot_spectra(DAL,"raw",sel = expression(ids == "DAL_ANdP0"))
+
+scores = pca_wrap(pca_data,"MFM")
+ggplot(scores, aes(PC1, PC2, color = depth)) + geom_point(aes(shape = treatment), size = 2)
+pca_data[pca_data$ids %in% scores[scores[,1] > 0.01 & scores[,2] < 0.0,"ids"],1:10]
+plot_spectra(MFM,"raw",sel = expression(ids == "MFM_av67v")) 
+plot_spectra(MFM,"raw",sel = expression(ids == "MFM_OftM2"))
+plot_spectra(MFM,"raw",sel = expression(ids == "MFM_xogJr"))
+plot_spectra(MFM,"raw",sel = expression(ids == "MFM_hoG4y"))
+plot_spectra(MFM,"raw",sel = expression(ids == "MFM_5UNIJ"))
+
[…]

### 2021-03-08 manuscript.Rmd
worked on intro; added renv to project

@@
 
-In recent years, the study of pollen grain chemistry has lead to new information important for biological and palaeoecological research.
-Taxonomic differences have been identified in the chemistry of pollen, which highlight the potential of pollen identification with chemical methods in palaeoecological research.
-Chemical fingerprinting of pollen has the potential of higher taxonomic resolution, automation of pollen identification and offers a more objective method than traditional morphological methods.
-as well as an environmental signal stored in the chemistry of pollen (UV Proxy), which could be used as a UV-proxy.
+In recent years, the study of pollen grain chemistry has shown promising results for biological and palaeoecological research.
+Fourier Transform Infrares (FTIR) spectroscopy has become a popular method for analysing pollen chemsitry
+A number of FTIR studies have identified pollen chemical differences on genus level, which may extent to species level differences. 
+This is a promising finding for the application of pollen identification with chemical methods in palaeoecological research, because it has the potential to increase taxomonic resolution.
+Further, chemical methods are suitable for automation and offer a more objective method than traditional morphological methods.
+In addition to taxonomic differences in pollen chemistry, there is evidence of environmental signals that are recorded in the chemistry of pollen.
+Several studies have shown links between solar and UV radiation and pollen chemistry as well as responses in pollen chemistry to temperature changes.
 
-acetolysis alters sporopollenin [@dominguezIsolationIntactPollen1998; @schulzeosthoffPhenolsIntegratedCompounds1987] affecting solubility
+So far, most FTIR studies have used modern material or material extracted from herbarium specimens.
+In some cases the pollen samples are processed with acetolysis following standard palynological methods (Erdman) to simulate fossilization.
+The assumption is that acetolysis does not alter the chemistry of sporopollenin and is not affecting any chemical signal.
+To date, there has not been a comparison between acetolysed and untreated fossil pollen to examine any effect of acetolysis on sporopollenin or chemical degradation of pollen during diagenesis.
 
-FTIR used on fossil papers in the past use acetolysed pollen or other chemically extracted pollen. 
-There has not been a comparison between fresh and untreated fossil pollen.
-In order to compare which compounds are missing and still present in fossil pollen.
-chemical degradation of pollen during diagenesis or depth.
+In this study we investigate the chemistry of fossil pollen to explore three potential sources of chemical alteration: i. the effects of diagenesis, ii. acetolysis and iii. age on pollen chemistry.
+We extracted pollen from three sediment cores using density separation to obtain chemically untreated fossil spectra.
+The pollen was collected from three different cores in three locations (Germany and Norway), which allows us examine potential differences in diagenesis on pollen chemistry at different locations.
+One replicate of the pollen samples in each core were acetolysed to analyse any changes caused by the strong acid treatment and compared to the untreated replicate.
+Each core was sub sampled at different depths to allows us to investigate chemical changes in the pollen with age.
 
@@
 
+acetolysis alters sporopollenin [@dominguezIsolationIntactPollen1998; @schulzeosthoffPhenolsIntegratedCompounds1987] affecting solubility
+
 # Methods and Materials
+
+Three cores from different environments and age ranges
 

### 2021-03-08 manuscript.Rmd
updated lockfile; had to install citr

@@
 
+\newpage
+
 # Introduction
+<!--Main Goal: Explore the chemical differences between fresh and untreated fossil pollen.-->
 
-Main Goal: Explore the chemical differences between fresh and untreated fossil pollen.
-
-The study of plant micro and macrofossil (spores, seed and pollen), recovered from lake sediments and peat bogs, is an integral part of Palaeoecology and Palaeobotany.
-Pollengrains are resistant to the diagenetic processes that occur in the formation of sediments, bogs and other deposits, which makes pollen records an important tool to reconstruct past floristic, vegetational and environmental changes.
+The study of plant micro and macrofossil (spores, seed and pollen) is an integral part of Palaeoecology and Palaeobotany.
+Pollengrains are extracted from sediments, bogs and other deposits and counted.
+These pollen records are an important tool to reconstruct past floristic, vegetational and environmental changes. 
 EXAMPLES
@@
 The high preservation of pollen grains in the geological record is because of the resistant biopolymer sporopollenin, which is the main building block of the outer pollen grain wall (exine).
-
-More and more details have been revealed about the exact structure of sporopollenin, e.g.
-importance of aromatic compounds, cross-linked aliphatic components [@liMolecularStructurePlant2019], similarity to lignins.
+The structure of sporopollenin has been described as similar to lignins, e.g.
+importance of aromatic compounds, cross-linked aliphatic components [@liMolecularStructurePlant2019], but the exact structure remains unknown.
 
 In recent years, the study of pollen grain chemistry has shown promising results for biological and palaeoecological research.
-Fourier Transform Infrares (FTIR) spectroscopy has become a popular method for analysing pollen chemsitry
+Fourier Transform Infrares (FTIR) spectroscopy has become a popular method for analysing pollen chemistry.
 A number of FTIR studies have identified pollen chemical differences on genus level, which may extent to species level differences. 
@@
 So far, most FTIR studies have used modern material or material extracted from herbarium specimens.
-In some cases the pollen samples are processed with acetolysis following standard palynological methods (Erdman) to simulate fossilization.
+In some cases the pollen samples are processed with acetolysis following standard palynological methods [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989] to simulate fossilization.
 The assumption is that acetolysis does not alter the chemistry of sporopollenin and is not affecting any chemical signal.
@@
 # Methods and Materials
+
+The pollen was extracted from three cores: Dalmuttlado in Norwa
 

### 2021-03-08 manuscript.Rmd
added spectra read files to trackign

@@
 
+\newpage
+
 # Abstract
@@
 
+\newpage
 <!-- READ DATA -->

### 2021-03-09 manuscript.Rmd
added refs to intro

@@
 The study of plant micro and macrofossil (spores, seed and pollen) is an integral part of Palaeoecology and Palaeobotany.
-Pollengrains are extracted from sediments, bogs and other deposits and counted.
-These pollen records are an important tool to reconstruct past floristic, vegetational and environmental changes. 
-EXAMPLES
+Pollen records are an important tool to reconstruct past floristic, vegetational and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @nogues-bravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013].
 
@@
 Fourier Transform Infrares (FTIR) spectroscopy has become a popular method for analysing pollen chemistry.
-A number of FTIR studies have identified pollen chemical differences on genus level, which may extent to species level differences. 
-This is a promising finding for the application of pollen identification with chemical methods in palaeoecological research, because it has the potential to increase taxomonic resolution.
-Further, chemical methods are suitable for automation and offer a more objective method than traditional morphological methods.
-In addition to taxonomic differences in pollen chemistry, there is evidence of environmental signals that are recorded in the chemistry of pollen.
+A number of FTIR studies have identified pollen chemical differences on genus level, which may extent to species level differences [@zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019; @muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020].
+This is a promising finding for the application of pollen identification with chemical methods in palaeoecological research, because it has the potential to increase taxomonic resolution @dellannaPollenDiscriminationClassification2009; @zimmermannAnalysisAllergenicPollen2016.
+Further, chemical methods are suitable for automation and offer a more objective method than traditional morphological methods [ @mondolApplicationHighThroughputScreening2019; @zimmermannAnalysisAllergenicPollen2016; @guedesPollenRamanSpectra2014].
+In addition to taxonomic differences in pollen chemistry, there is evidence of environmental signals that are recorded in the chemistry of pollen [@bagciogluMultiscaleVibrationalSpectroscopic2015; @kendelChemicalAnalysisPollen2020; @diehnCombiningChemicalInformation2020].
 Several studies have shown links between solar and UV radiation and pollen chemistry as well as responses in pollen chemistry to temperature changes.
 
-So far, most FTIR studies have used modern material or material extracted from herbarium specimens.
-In some cases the pollen samples are processed with acetolysis following standard palynological methods [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989] to simulate fossilization.
+So far, most FTIR studies have used modern material or material extracted from herbarium specimens .
+In some cases the pollen samples are processed with acetolysis following standard palynological methods [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989] to simulate fossilization [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018].
 The assumption is that acetolysis does not alter the chemistry of sporopollenin and is not affecting any chemical signal.

### 2021-03-09 manuscript.Rmd
added ages to TSK and MFM; provided by Brauer and Franscesco

@@
 ```{r}
+source(here("R","ages.R"))
 if (file.exists(here("data","output","DAL_spectra.rds"))) {
@@
   MFM = readRDS(here("data","output","MFM_spectra.rds"))
+  MFM[MFM$depth=="A4UR_72-74", 'depth'] = "A5UR_72-74" #fixing typo in sample name
+  MFM$age = fn_age(MFM$depth)
 } else {
@@
   TSK = readRDS(here("data","output","TSK_spectra.rds"))
+  TSK$age = fn_age(TSK$depth)
 } else {

### 2021-03-09 R/ages.R

@@ -1,0 +1,7 @@
+fn_age = function(foo){
+  ages = data.frame(name = c(unique(MFM$depth),unique(TSK$depth)),
+           age = c(11400, 12080, 13120, 12340, 12850,
+                   2011, 1998, 1966, 1944, 1933))
+  tmp = sapply(foo, function(x){ ages[ages$name==x, 'age'] })
+  return(tmp)
+}

### 2021-03-09 readme.Rmd
moved notes to readme; added section on renv; included important renv file for tracking

@@
 ---
+
+# Reproducability
+The file 'packages.R' is a list of essential packages that are needed to run the analysis and figures in this manuscript.
+This manuscript is also enabled for the package 'renv' which enables a sandboxed, version controlled library of all packages used during writing and analysis.
+The 'renv.lock' file can be used to initiate the library after 'renv' is installed by calling 'renv::restore()'.
+
+# Detailed Methods and Materials
+
+## SPT extraction
+
+Pollen extraction:
+
+1. add 1 cube to vial (~1cm^-3^)
+2. wash with water -> dissolve sediment vortex
+3. centrifuge 3min 3000rpm
+4. decant; save decant
+5. add 3 ml 1.43 SPT solution
+6. vortex
+7. centrifuge 3min 3000rpm
+8. decant supernatant into new vial
+9. repeat 5-7; decant into same vial
+10. wash with water; save decant for checking
+11. wash with water; save decant x2
+12. wash residue with water; save in vial
 
@@
 Pinus (17)
+
+# Notes
+extracted fossil pollen from sediment.
+ca 5 common taxa: Alnus, Picea, Pinus,
+
+## Cores
+Dalmuttlado:
+- pine
+- holocene
+
+Younger Dryas [@brauerHighResolutionSediment1999]:
+- 12 samples laminated sediment
+- from  11400 to 13120
+- min 500 grains
+- pollen conc highest in 11000-11600 and 12700-12900
+- Pinus betula most common (ca 150 grains counted)
+- salix (ca 25 counted)
+- quercus throughout in low conc (<1%)
+- graminoids/cyperaceae/artemisia common
+
+Tiefer See [@theuerkaufEffectsChangesLand2015]:
+- 1870-2010
+- laminated sediments
+- Pinus pollen
+
+Research Questions:
+1. chemical variation between taxa in the same cores.
+2. chemical variation between taxa between cores 
+3. chemical variation of pollen with depths/age
+
+## Labwork Report
+
+did µFTIR on fresh and fossil pollen using Hyperion-FPA
+
+This is very sample dependent.
+organic rich cores are problematic (requires a better extraction protocol), pollen density is another.
+Dalmutladdo was a very good core for this: low in organic and relatively rich in pollen.
+BSi increased with depth, highest in oldest samples (extracted blanks).
+Still, some depth were more time intensive (130/150), low amount of pollen .
+I managed about 2-3 core samples per day, the fresh samples can be done in a fraction of the time, all done within 1 half day.
+
+9 depth from Dalmutladdo core. targeted and extracted Pinus pollen (Table \@ref(tab:pollencounts), Table \@ref(tab:fresh))
+
+possibilities for Alnus and Betula, even though scattering visible in spectra
+
+```{r pollencounts, echo=FALSE}
+dalmutladdo = data.frame(ID = c("DAL_08","DAL_20","DAL_40","DAL_60","DAL_90", "DAL_110","DAL_130","DAL_140","DAL_150"),
+                         pinus_pollen = c(23,49,37,47,36,32,39,17,7),
+                         alnus_pollen = c(0,7,0,0,0,0,0,0,0),
+                         betula_pollen = c(0,15,0,0,0,0,0,0,0))
+knitr::kable(dalmutladdo, format="latex", booktabs=TRUE,
+             caption = "Pollen counts")
+```
+
+```{r fresh, echo=FALSE}
+fresh_pollen = data.frame(ID = c("BGO_PS7", "BGO_PS8", "INN_PS13", "INN_PS19A","NMBU_Pin_syl_14E1", "NMBU_Pin_syl_14H1", "NMBU_Aln_glu_14E1", "NMBU_Aln_glu_14J1", "NMBU_Aln_inc_14B1", "NMBU_Aln_inc_14G1", "NMBU_Bet_pub_14A1", "NMBU_Bet_pub_14B1", "NMBU_Cor_ave_14B1", "NMBU_Cor_ave_14C1"),
+                          extracted = c(109,55,51,24,0,0,0,0,0,0,0,0,0,0))
+knitr::kable(fresh_pollen, format="latex", booktabs=TRUE,
+             caption = "Fresh pollen counts")
+```
+
+9 fossil samples from 1 core
+
+2 trees of Pinus at each location
+
+Research questions we could answer with this set:
+
+- Differences between fresh and fossil/near fossil grains of Pinus
+- chemical changes in Pinus pollen with age/depth
+(- chemical differences between Pinus/Alnus/Betula)
+
+extensions to this set:
+- add more samples from Dalmuttladdo highest and lowest depth (maybe labwork)
+- add cores (e.g. MFM, Tiefer See or Italian cores), Festi core was quite organic rich
+- add more grains from already measured samples (to hit min 25 or 50)

### 2021-03-09 manuscript.Rmd
wrote on methods; moved bits to readme

@@
 separate cover letter
+
+<!-- 
+A  200
+I  500
+MM 400
+R  500
+D  800
+-->
 
@@
 
-The pollen was extracted from three cores: Dalmuttlado in Norwa
-
-Three cores from different environments and age ranges
-
-Pollen extraction:
-
-1. add 1 cube to vial (~1cm^-3^)
-2. wash with water -> dissolve sediment vortex
-3. centrifuge 3min 3000rpm
-4. decant; save decant
-5. add 3 ml 1.43 SPT solution
-6. vortex
-7. centrifuge 3min 3000rpm
-8. decant supernatant into new vial
-9. repeat 5-7; decant into same vial
-10. wash with water; save decant for checking
-11. wash with water; save decant x2
-12. wash residue with water; save in vial
+The pollen was extracted from three cores: Dalmuttlado in Norway, Tiefer See [@dragerVarveMicrofaciesVarve2017] and Meerfelder Maar [@brauerHighResolutionSediment1999] in Germany.
+Tiefer See formed during the last glacial period, while the Meerfelder Maar record is from the largest maar of the Westeifel Volcaniv field and formed ~35000 BP.
+Tiefer See and Meerfelder Maar are varved cores and were the samples used in this study span two different time-periods.
+The sub samples from Tiefer see are from the last 80 yrs of the core (1933 - 2011), while the samples from Meerfelder Maar cover the timeperiod from the late Allerød to the Younger Dryas/Holocene transition (~12900 - ~11000 BP).
+The samples were extracted using density separation using dissolved Sodium Polytungstate (SPT) at 1.43 (g/ml)? as separation medium. <!--FM: check Unit-->
+During density separation a known volume of SPT (density 1.43) solution is added to the sample and homogenized/vortexed, followed by centrifugation.
+After centrifugation, all particulates/material heavier than 1.43 has deposited at the bottom of the vial and the lighter fraction can be decanted and collected.
+In our case the lighter fraction contained most of the pollen.
+The density of pollen is reported at <1.4 UNIT [@nakagawaDensemediaSeparationMore1998; @regnellPreparativeCentrifugationNew1996].
 
@@
 
-# Notes
-extracted fossil pollen from sediment.
-ca 5 common taxa: Alnus, Picea, Pinus,
-
-## Cores
-Dalmuttlado:
-- pine
-- holocene
-
-Younger Dryas [@brauerHighResolutionSediment1999]:
-- 12 samples laminated sediment
-- from  11400 to 13120
-- min 500 grains
-- pollen conc highest in 11000-11600 and 12700-12900
-- Pinus betula most common (ca 150 grains counted)
-- salix (ca 25 counted)
-- quercus throughout in low conc (<1%)
-- graminoids/cyperaceae/artemisia common
-
-Tiefer See [@theuerkaufEffectsChangesLand2015]:
-- 1870-2010
-- laminated sediments
-- Pinus pollen
-
-Research Questions:
-1. chemical variation between taxa in the same cores.
-2. chemical variation between taxa between cores 
-3. chemical variation of pollen with depths/age
-
-## Labwork Report
-
-did µFTIR on fresh and fossil pollen using Hyperion-FPA
-
-This is very sample dependent.
-organic rich cores are problematic (requires a better extraction protocol), pollen density is another.
-Dalmutladdo was a very good core for this: low in organic and relatively rich in pollen.
-BSi increased with depth, highest in oldest samples (extracted blanks).
-Still, some depth were more time intensive (130/150), low amount of pollen .
-I managed about 2-3 core samples per day, the fresh samples can be done in a fraction of the time, all done within 1 half day.
-
-9 depth from Dalmutladdo core. targeted and extracted Pinus pollen (Table \@ref(tab:pollencounts), Table \@ref(tab:fresh))
-
-possibilities for Alnus and Betula, even though scattering visible in spectra
-
-```{r pollencounts, echo=FALSE}
-dalmutladdo = data.frame(ID = c("DAL_08","DAL_20","DAL_40","DAL_60","DAL_90", "DAL_110","DAL_130","DAL_140","DAL_150"),
-                         pinus_pollen = c(23,49,37,47,36,32,39,17,7),
-                         alnus_pollen = c(0,7,0,0,0,0,0,0,0),
-                         betula_pollen = c(0,15,0,0,0,0,0,0,0))
-knitr::kable(dalmutladdo, format="latex", booktabs=TRUE,
-             caption = "Pollen counts")
-```
-
-```{r fresh, echo=FALSE}
-fresh_pollen = data.frame(ID = c("BGO_PS7", "BGO_PS8", "INN_PS13", "INN_PS19A","NMBU_Pin_syl_14E1", "NMBU_Pin_syl_14H1", "NMBU_Aln_glu_14E1", "NMBU_Aln_glu_14J1", "NMBU_Aln_inc_14B1", "NMBU_Aln_inc_14G1", "NMBU_Bet_pub_14A1", "NMBU_Bet_pub_14B1", "NMBU_Cor_ave_14B1", "NMBU_Cor_ave_14C1"),
-                          extracted = c(109,55,51,24,0,0,0,0,0,0,0,0,0,0))
-knitr::kable(fresh_pollen, format="latex", booktabs=TRUE,
-             caption = "Fresh pollen counts")
-```
-
-9 fossil samples from 1 core
-
-2 trees of Pinus at each location
-
-Research questions we could answer with this set:
-
-- Differences between fresh and fossil/near fossil grains of Pinus
-- chemical changes in Pinus pollen with age/depth
-(- chemical differences between Pinus/Alnus/Betula)
-
-extensions to this set:
-- add more samples from Dalmuttladdo highest and lowest depth (maybe labwork)
-- add cores (e.g. MFM, Tiefer See or Italian cores), Festi core was quite organic rich
-- add more grains from already measured samples (to hit min 25 or 50)
-
 # References

### 2021-03-09 manuscript.Rmd
testing EMSC interferent filter

@@
 During density separation a known volume of SPT (density 1.43) solution is added to the sample and homogenized/vortexed, followed by centrifugation.
-After centrifugation, all particulates/material heavier than 1.43 has deposited at the bottom of the vial and the lighter fraction can be decanted and collected.
+After centrifugation, particulates/material heavier than 1.43 has deposited at the bottom of the vial and the lighter fraction can be decanted and collected.
 In our case the lighter fraction contained most of the pollen.

### 2021-03-09 R/helper_functions.R

@@ -20,7 +20,7 @@ plot_spectra = function(data, spec_id, sel, id_vars = c("ids","type","species","
                  id.vars=id_vars)
   gdata$variable = as.numeric(as.character(gdata$variable))
   ggplot(gdata, aes(variable, value, color = depth, group = ids)) +
-    geom_line() + scale_x_reverse() +
-    theme_bw()
+    geom_line() + scale_x_reverse() + xlim(1800,900) +
+    theme_bw() + theme(legend.position = "bottom")
   #return(gdata)
 }

### 2021-03-09 R/testing_EMSC.R

@@ -1,0 +1,17 @@
+foo_spectra = subset(MFM, species == "Pinus" & depth == "A4DL_130-132")
+foo_blank = subset(MFM, species == "Blank" & depth == "A4DL_130-132")
+sel = c(as.numeric(colnames(foo_spectra$raw)) < 1300)
+
+emsc_spectra = EMSC(foo_spectra$raw[,sel], interferent = colMeans(foo_blank$raw[,sel]), degree = 1)
+
+plot(as.numeric(colnames(emsc_spectra$X)),emsc_spectra$X[1,], type = "l" )
+plot(as.numeric(colnames(emsc_spectra$X)),emsc_spectra$corrected[1,], type = "l" )
+plot(as.numeric(colnames(emsc_spectra$X)),colMeans(foo_blank$raw[,sel]), type = "l" )
+
+emsc_spectra = EMSC(foo_spectra$raw[,sel], interferent = colMeans(MFM$raw[MFM$species == "Blank",sel]), degree = 1)
+
+plot(as.numeric(colnames(emsc_spectra$X)),emsc_spectra$X[1,], type = "l" )
+plot(as.numeric(colnames(emsc_spectra$X)),emsc_spectra$corrected[1,], type = "l" )
+plot(as.numeric(colnames(emsc_spectra$X)),colMeans(foo_blank$raw[,sel]), type = "l" )
+
+plot_spectra(MFM, "raw", expression(species=="Blank" & depth == unique(MFM$depth)[1]))

### 2021-03-09 R/testing_EMSC.R
wrote some todo plans for analysis

@@ -15,3 +15,16 @@ plot(as.numeric(colnames(emsc_spectra$X)),emsc_spectra$corrected[1,], type = "l"
 plot(as.numeric(colnames(emsc_spectra$X)),colMeans(foo_blank$raw[,sel]), type = "l" )
 
 plot_spectra(MFM, "raw", expression(species=="Blank" & depth == unique(MFM$depth)[1]))
+
+#' check difference in blanks between SPT and acet
+#' 
+#' Pre-process:
+#'
+#' 1. linear EMSC of raw spectra with meaned Blanks
+#' 2. SG
+#' 
+#' Analysis:
+#' PLSR depth regression or class
+#' PCA on all cores 
+#' PCA acet vs SPT
+#' 

### 2021-03-10 R/testing_EMSC.R
some more testing EMSC

@@ -14,7 +14,7 @@ plot(as.numeric(colnames(emsc_spectra$X)),emsc_spectra$X[1,], type = "l" )
 plot(as.numeric(colnames(emsc_spectra$X)),emsc_spectra$corrected[1,], type = "l" )
 plot(as.numeric(colnames(emsc_spectra$X)),colMeans(foo_blank$raw[,sel]), type = "l" )
 
-plot_spectra(MFM, "raw", expression(species=="Blank" & depth == unique(MFM$depth)[1]))
+plot_spectra(MFM, "raw", expression(species=="Blank" & depth == unique(MFM$depth)[1] & treatment == "SPT"))
 
 #' check difference in blanks between SPT and acet
 #' 

### 2021-03-10 manuscript.Rmd
hardcoded depth names into ages.R

@@
 }
+
+DAL$Label = "DAL"
+TSK$Label = "TSK"
+MFM$Label = "MFM"
 ```

### 2021-03-10 R/ages.R

@@ -1,5 +1,6 @@
 fn_age = function(foo){
-  ages = data.frame(name = c(unique(MFM$depth),unique(TSK$depth)),
+  ages = data.frame(name = c("A4DL_130-132", "A4DL_188-190", "A5UR_72-74",   "D4DL_138-140", "D4DL_182-184", 
+                             "9-10", "17-18", "29-31", "38-39", "42-43"),
            age = c(11400, 12080, 13120, 12340, 12850,
                    2011, 1998, 1966, 1944, 1933))
   tmp = sapply(foo, function(x){ ages[ages$name==x, 'age'] })

### 2021-03-10 R/ages.R
carried over ages.R change from master

@@ -1,7 +1,8 @@
 fn_age = function(foo){
-  ages = data.frame(name = c(unique(MFM$depth),unique(TSK$depth)),
-           age = c(11400, 12080, 13120, 12340, 12850,
-                   2011, 1998, 1966, 1944, 1933))
+  ages = data.frame(name = c("A4DL_130-132", "A4DL_188-190", "A5UR_72-74",   "D4DL_138-140", "D4DL_182-184", 
+                             "9-10", "17-18", "29-31", "38-39", "42-43"),
+                    age = c(11400, 12080, 13120, 12340, 12850,
+                            2011, 1998, 1966, 1944, 1933))
   tmp = sapply(foo, function(x){ ages[ages$name==x, 'age'] })
   return(tmp)
 }

### 2021-03-10 R/pca.R

@@ -1,6 +1,3 @@
-DAL$Label = "DAL"
-TSK$Label = "TSK"
-MFM$Label = "MFM"
 pca_wrap = function(pca_data,sel){
   pca_data[pca_data$treatment == "","treatment"] = "SPT"
   if (is.expression(sel)) {

### 2021-03-18 manuscript.Rmd
changed count grain function to check pollen species counts; experimenting on spls with caret pkg

@@
 MFM$Label = "MFM"
-
-pca_data = rbind(subset(DAL, species == "Pinus"), 
-        subset(TSK, species == "Pinus"), 
-        subset(MFM, species == "Pinus"))
-pca_data[pca_data$treatment == "","treatment"] = "SPT"
-pca_res = princomp(pca_data$sg2, scores = T)
-scores =  as.data.frame(pca_res$scores[,1:10]) %>%
-  mutate(ids = rownames(.), 
-         core = pca_data$Label, 
-         treatment = pca_data$treatment
-         ) #%>%
-  #melt(.,id.vars = "ids")
-ggplot(scores, aes(Comp.1, Comp.2, color = core)) + geom_point()
-# pca_data[scores[scores[,2] < -.2,"ids"]==pca_data$ids,1:10]
 ```
@@
     group_by(type, depth, treatment) %>%
+    summarise(count=n())
+  return(counts)
+}
+count_grains = function(core){
+  counts = core %>%
+    subset(type == "pollen") %>% 
+    group_by(treatment, depth, species) %>%
     summarise(count=n())

### 2021-03-18 R/pca.R

@@ -8,7 +8,8 @@ pca_wrap = function(pca_data,sel){
     mutate(ids = rownames(.),
            core = pca_data$Label,
            treatment = pca_data$treatment,
-           depth = pca_data$depth
+           depth = pca_data$depth,
+           age = pca_data$age
            )
   return(scores)
 }
@@ -76,3 +77,59 @@ plsr_data = subset(pca_data_outliers_removed_2, Label == "DAL")
 plsr_data$depth = as.numeric(plsr_data$depth)
 pls_res = plsr(depth ~ sg2, data = plsr_data, ncomp = 4, validation = "LOO")
 plot(pls_res)
+
+plsr_data = subset(pca_data_outliers_removed_2, Label == "TSK" & treatment == "SPT")
+#plsr_data$depth = as.numeric(plsr_data$depth)
+pls_res = plsr(age ~ sg2, data = plsr_data, ncomp = 4, validation = "LOO")
+plot(pls_res)
+
+plsr_data = subset(pca_data_outliers_removed_2, treatment == "SPT")
+M = model.matrix(~ Label -1, plsr_data)
+pls_res = plsr(M ~ sg2, data = plsr_data, ncomp = 4, validation = "LOO")
+biplot(pls_res, which = c("y"))
+
+#SPLS
+library(spls)
+
+sparse.cv = cv.spls(plsr_data$sg2, M, K = c(1:10), eta = seq(.8,.99,.01))
+sparse.spsl = spls(plsr_data$sg2, M, sparse.cv$K.opt, sparse.cv$eta.opt, scale.x = F)
+
+#caret:::splsda
+
+splsFit <- caret:::splsda(plsr_data$sg2, M, 
+                          K = 5, eta = .9)
+preds = caret:::predict.splsda(splsFit, plsr_data$sg2)
+levels(preds) = c("DAL","MFM","TSK")
+confusionMatrix(preds,
+                as.factor(plsr_data$Label))
+
+## Not run: 
+data(mdrr)
+set.seed(1)
+inTrain <- sample(seq(along = mdrrClass), 450)
+
+nzv <- nearZeroVar(mdrrDescr)
+filteredDescr <- mdrrDescr[, -nzv]
+
+training <- filteredDescr[inTrain,]
+test <- filteredDescr[-inTrain,]
+trainMDRR <- mdrrClass[inTrain]
+testMDRR <- mdrrClass[-inTrain]
+
+preProcValues <- preProcess(training)
+
+trainDescr <- predict(preProcValues, training)
[…]

### 2021-03-24 manuscript.Rmd
testing spls; started writing bootstrap cv file

@@
 RQs
-i. Difference between fresh and fossil grains.
-i.i effect of orientation on chemistry
-i.ii effect of depth/age on chemistry
-ii Differences between cores
+i. Differences between cores
+ii.i Difference between acetolysed and fossil
+ii.ii Difference between fresh and fossil grains.
+iii effect of depth/age on chemistry
+
+iv effect of orientation on chemistry
 

### 2021-03-24 R/bootstrap.R

@@ -1,0 +1,24 @@
+
+
+shuffle_train = function(n, p) runif(n) < p
+
+fit_pls
+
+bootstrap = function(data, i) {
+  res = list()
+  
+  #start loop
+  for (j in 1:i) {
+    paste0("fitting model "j"/"i)
+    #shuffle train/test split
+    data$train = shuffle_train(nrow(data), .7)
+    #fit model
+    train_ = fit_pls
+    #predict test set
+    test_ = caret::predict()
+    #save to list
+    res[[paste0("iter_",j)]] = list("train" = train_, "test" = test_)
+  }
+
+  return(res)
+}

### 2021-03-24 R/pca.R

@@ -92,7 +92,7 @@ biplot(pls_res, which = c("y"))
 library(spls)
 
 sparse.cv = cv.spls(plsr_data$sg2, M, K = c(1:10), eta = seq(.8,.99,.01))
-sparse.spsl = spls(plsr_data$sg2, M, sparse.cv$K.opt, sparse.cv$eta.opt, scale.x = F)
+sparse.spsl = spls:::spls(plsr_data$sg2, M, sparse.cv$K.opt, sparse.cv$eta.opt, scale.x = F)
 
 #caret:::splsda
 

### 2021-03-24 R/spls.R

@@ -1,0 +1,71 @@
+pca_data = rbind(subset(DAL, species == "Pinus"), 
+                 subset(TSK, species == "Pinus"), 
+                 subset(MFM, species == "Pinus"))
+
+outliers =  c("DAL_rx1GZ", "DAL_2cxp7", "DAL_VzFi0", "DAL_ANdP0", "MFM_av67v", "MFM_OftM2", "MFM_xogJr", "MFM_hoG4y", "MFM_5UNIJ", "TSK_vvgOZ", "DAL_xgYjM", "TSK_m4hN6", "TSK_e04Go")
+
+pca_data_outliers_removed_2 = pca_data[!pca_data$ids %in% outliers,]
+
+pca_wrap = function(pca_data,sel){
+  pca_data[pca_data$treatment == "","treatment"] = "SPT"
+  if (is.expression(sel)) {
+    pca_data = subset(pca_data, eval(sel))
+  }
+  pca_res = prcomp(pca_data$sg2)
+  scores =  as.data.frame(pca_res$x[,1:10]) %>%
+    mutate(ids = rownames(.),
+           core = pca_data$Label,
+           treatment = pca_data$treatment,
+           depth = pca_data$depth,
+           age = pca_data$age
+    )
+  loadings = pca_res$rotation
+  return(list("scores" = scores, "loadings" = loadings))
+}
+
+M_wave = function(M){
+  return(as.numeric(as.character(rownames(M))))
+}
+
+#PCA on SPT cores samples
+#pca_data = subset(pca_data_outliers_removed_2, treatment == "SPT")
+
+scores_cores = pca_wrap(pca_data_outliers_removed_2, expression(treatment == "SPT"))
+ggplot(scores_cores[["scores"]], aes(PC1, PC2, color = core)) + geom_point(aes(), size = 4)
+
+pca_data = pca_data_outliers_removed_2[pca_data_outliers_removed_2$depth %in% unique(pca_data_outliers_removed_2[pca_data_outliers_removed_2$treatment == "acet","depth"]),]
+
+scores_treatment = pca_wrap(pca_data, "")
+
+ggplot(scores_treatment, aes(PC1, PC2, color = core)) + geom_point(aes(shape = treatment), size = 4)
+
+plsr_data = subset(pca_data_outliers_removed_2, treatment == "SPT")
+M = model.matrix(~ Label -1, plsr_data)
+
+splsFit <- caret:::splsda(plsr_data$sg2, M, 
[…]

### 2021-03-30 manuscript.Rmd
added fresh spectra; wrote bootstrap draft function; other changes

@@
 
+if (file.exists(here("data","output","fresh_spectra.rds"))) {
+  FRE = readRDS(here("data","output","fresh_spectra.rds"))
+} else {
+  source(here("R","read_data.R"))
+  read_data_fresh()
+  FRE = readRDS(here("data","output","fresh_spectra.rds"))
+}
+
+FRE$Label = "FRE"
 DAL$Label = "DAL"
@@
 ```{r}
-if (!hasName(DAL, "sg2")) {
+if (!hasName(DAL, "emsc")) {
   source(here("R","pre_process.R"))

### 2021-03-30 R/bootstrap.R

@@ -9,7 +9,7 @@ bootstrap = function(data, i) {
   
   #start loop
   for (j in 1:i) {
-    paste0("fitting model "j"/"i)
+    paste0("fitting model ",j,"/",i)
     #shuffle train/test split
     data$train = shuffle_train(nrow(data), .7)
     #fit model

### 2021-03-30 R/helper_functions.R

@@ -24,3 +24,41 @@ plot_spectra = function(data, spec_id, sel, id_vars = c("ids","type","species","
     theme_bw() + theme(legend.position = "bottom")
   #return(gdata)
 }
+
+plot_mean_spectra = function(data, subset_sel, spec_id, id.vars = c("ids", "treatment", "Lable", "depth", "age", "type")){
+  # subset data
+  pdata = subset(data, eval(subset_sel))
+  
+  # melt data  
+  pdata = melt(cbind(pdata[,1:7],
+               as.data.frame(pdata[,c(spec_id)])),
+         id.vars=colnames(data)[1:7])
+  
+  # wavenumbers to numeric  
+  pdata$variable = as.numeric(as.character(pdata$variable))
+
+  # mean data over groups
+  pdata_mean = pdata %>%
+    group_by(type, variable) %>%
+    summarise(mean = mean(value))
+  
+  #offset
+  
+  g= ggplot(pdata, aes(variable, value, color = type))+
+    geom_line(aes(group = ids), alpha = .05) + #mean spectra
+    geom_line(data = pdata_mean, aes(variable, mean, color = type), inherit.aes = F, size = 1) + #all spectra
+    xlim(1800,900) +
+    theme_bw() + theme(legend.position = "bottom")  
+  
+  return(g)
+}  
+
+df = data.frame(x = rep(c("a","b"), c(5,10)), y = c(runif(5,3,7),runif(10,)))
+
+offset_spectra = function(data, variable, value, offset) {
+  for (i in 1:length(unique(df[, variable]))) {
+    data[data[, variable] == unique(data[, variable])[i],value] = data[data[, variable] == unique(data[, variable])[i],value] + offset[i]
+  }
+  return(data)
+}
+

### 2021-03-30 R/pre_process.R

@@ -1,22 +1,44 @@
-pre_processing = function(spectra, ws, cut=T){
+pre_processing = function(spectra, ws, cut=T, emsc = F){
   wn = as.numeric(colnames(spectra))
-  foo_sg = SavitzkyGolay(spectra, poly = 2, width = ws, deriv = 2)
-  if (cut) {
-    foo_sg = foo_sg[,wn<1900 & wn>700]
-    colnames(foo_sg) = wn[wn<1900 & wn>700]
+  
+  if (emsc){
+    foo_emsc = EMSC(spectra, degree = 1)$corrected
+    rownames(foo_emsc) = rownames(spectra)
+    if (cut) {
+      foo_emsc = foo_emsc[,wn<1900 & wn>700]
+      colnames(foo_emsc) = wn[wn<1900 & wn>700]
+    } else {
+      colnames(foo_emsc) = wn
+    }    
+    
   } else {
-    colnames(foo_sg) = wn
+    foo_sg = SavitzkyGolay(spectra, poly = 2, width = ws, deriv = 2)
+    if (cut) {
+      foo_sg = foo_sg[,wn<1900 & wn>700]
+      colnames(foo_sg) = wn[wn<1900 & wn>700]
+    } else {
+      colnames(foo_sg) = wn
+    }
+    foo_emsc = EMSC(foo_sg, degree = 1)$corrected
+    rownames(foo_emsc) = rownames(spectra)
+    colnames(foo_emsc) = colnames(foo_sg)
   }
-  foo_emsc = EMSC(foo_sg, degree = 1)$corrected
-  rownames(foo_emsc) = rownames(spectra)
-  colnames(foo_emsc) = colnames(foo_sg)
+  
+
   return(foo_emsc)
 }
 
 DAL$sg2 = pre_processing(DAL$raw, 5)
 TSK$sg2 = pre_processing(TSK$raw, 5)
 MFM$sg2 = pre_processing(MFM$raw, 5)
+FRE$sg2 = pre_processing(FRE$raw, 5)
+
+DAL$emsc = pre_processing(DAL$raw, 5, T, T)
+TSK$emsc = pre_processing(TSK$raw, 5, T, T)
+MFM$emsc = pre_processing(MFM$raw, 5, T, T)
+FRE$emsc = pre_processing(FRE$raw, 5, T, T)
 
 saveRDS(DAL, here("data","output","DAL_spectra.rds"))
[…]

### 2021-03-30 R/read_data.R

@@ -6,6 +6,32 @@ repl = function(data, c, default=""){
   return(tmp)
 }
 
+read_data_fresh = function() {
+  raw = read.csv(here("data", "input", "spectra", "fresh_spectra_raw.csv"), 
+                 check.names = F, 
+                 colClasses = "character")
+  spectra_raw = as.matrix((raw[,c(1:765)]))
+  spectra_raw = apply(spectra_raw, 2, as.numeric)
+  spectra_meta = raw[,c(766:ncol(raw))]
+  ids = paste0("FRE_",stri_rand_strings(nrow(spectra_meta),5))
+  
+  while (sum(duplicated(ids))!=0) {
+    ids = paste0(sc,"_",stri_rand_strings(nrow(spectra_meta),5))
+  }
+  spectra_meta$ids = ids
+  spectra_meta$treatment = "fresh"
+  spectra_meta$depth = 0
+  spectra_meta$age = 2019
+  spectra_meta$species = "Pinus"
+  
+  rownames(spectra_raw) = ids
+  
+  data = spectra_meta
+  data$raw = spectra_raw
+  
+  saveRDS(data, file = here("data","output",paste0("fresh_spectra.rds")))
+}
+
 read_data = function(sc){
   raw = read.csv(here("data", "input", "spectra", paste0(sc,"_spectra_raw.csv")), 
                  check.names = F, 

### 2021-03-30 manuscript.Rmd
added missing references to bib-file third time

@@
 We extracted pollen from three sediment cores using density separation to obtain chemically untreated fossil spectra.
-The pollen was collected from three different cores in three locations (Germany and Norway), which allows us examine potential differences in diagenesis on pollen chemistry at different locations.
+The pollen was collected from three different cores in three locations (Germany and Norway), which allows us to examine potential differences in diagenesis on pollen chemistry at different locations.
 One replicate of the pollen samples in each core were acetolysed to analyse any changes caused by the strong acid treatment and compared to the untreated replicate.
 Each core was sub sampled at different depths to allows us to investigate chemical changes in the pollen with age.
-
+[@dellannaPollenDiscriminationClassification2009; @nogues-bravoCrackingCodeBiodiversity2018]
 RQs

### 2021-03-30 manuscript.Rmd
added missing references to bib-file fourth time
added missing references to bib-file fifth time

@@
 The study of plant micro and macrofossil (spores, seed and pollen) is an integral part of Palaeoecology and Palaeobotany.
-Pollen records are an important tool to reconstruct past floristic, vegetational and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @nogues-bravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013].
+Pollen records are an important tool to reconstruct past floristic, vegetational and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @noguesbravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013].
 
@@
 Each core was sub sampled at different depths to allows us to investigate chemical changes in the pollen with age.
-[@dellannaPollenDiscriminationClassification2009; @nogues-bravoCrackingCodeBiodiversity2018]
+[@noguesbravoCrackingCodeBiodiversity2018]
 RQs

### 2021-04-07 manuscript.Rmd
mean spectra plotting function

@@
   MFM$age = fn_age(MFM$depth)
+  MFM[MFM$treatment == "","treatment"] = "SPT"
 } else {
@@
 Each core was sub sampled at different depths to allows us to investigate chemical changes in the pollen with age.
-[@noguesbravoCrackingCodeBiodiversity2018]
+
 RQs

### 2021-04-07 R/helper_functions.R

@@ -25,28 +25,29 @@ plot_spectra = function(data, spec_id, sel, id_vars = c("ids","type","species","
   #return(gdata)
 }
 
-plot_mean_spectra = function(data, subset_sel, spec_id, id.vars = c("ids", "treatment", "Lable", "depth", "age", "type")){
+plot_mean_spectra = function(data, subset_sel, spec_id, id.vars = c("ids", "treatment", "Label", "depth", "age", "type"), group = "type"){
   # subset data
   pdata = subset(data, eval(subset_sel))
   
   # melt data  
-  pdata = melt(cbind(pdata[,1:7],
+  pdata = melt(cbind(pdata[,id.vars],
                as.data.frame(pdata[,c(spec_id)])),
-         id.vars=colnames(data)[1:7])
+         id.vars=id.vars)
   
   # wavenumbers to numeric  
   pdata$variable = as.numeric(as.character(pdata$variable))
-
-  # mean data over groups
-  pdata_mean = pdata %>%
-    group_by(type, variable) %>%
-    summarise(mean = mean(value))
   
   #offset
+  pdata = offset_spectra(pdata, group, "value", seq(0, by=max(pdata$value)/2, length.out = length(unique(pdata[,group]))))
   
-  g= ggplot(pdata, aes(variable, value, color = type))+
-    geom_line(aes(group = ids), alpha = .05) + #mean spectra
-    geom_line(data = pdata_mean, aes(variable, mean, color = type), inherit.aes = F, size = 1) + #all spectra
+  # mean data over groups
+  pdata_mean = pdata %>%
+    group_by_at(c(group, "variable")) %>%
+    summarise(mean = mean(value))
+  
+  g = ggplot(pdata, aes_string("variable", "value", color = group))+
+    geom_line(aes(group = ids), alpha = .15) + #mean spectra
+    geom_line(data = pdata_mean, aes_string("variable", "mean", group = group), inherit.aes = F, size = 1) + #all spectra
     xlim(1800,900) +
     theme_bw() + theme(legend.position = "bottom")  
   
@@ -56,7 +57,7 @@ plot_mean_spectra = function(data, subset_sel, spec_id, id.vars = c("ids", "trea
 df = data.frame(x = rep(c("a","b"), c(5,10)), y = c(runif(5,3,7),runif(10,)))
 
 offset_spectra = function(data, variable, value, offset) {
-  for (i in 1:length(unique(df[, variable]))) {
[…]

### 2021-04-12 manuscript.Rmd
wrote plotting functions and plots of mean spectra; moved outlier removal into manuscript data loading code

@@
   DAL = readRDS(here("data","output","DAL_spectra.rds"))
+  DAL$age = 0
 } else {
@@
 MFM$Label = "MFM"
+
+outliers = c("DAL_rx1GZ", "DAL_2cxp7", "DAL_VzFi0", "DAL_ANdP0", "MFM_av67v", "MFM_OftM2", "MFM_xogJr", "MFM_hoG4y", "MFM_5UNIJ", "TSK_vvgOZ", "DAL_xgYjM", "TSK_m4hN6", "TSK_e04Go")
+
+DAL = DAL[!DAL$ids %in% outliers,]
+TSK = TSK[!TSK$ids %in% outliers,]
+MFM = MFM[!MFM$ids %in% outliers,]
 ```
@@
 
+# Results
+
+```{r echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="100%"}
+mean_plots = readRDS(here("data","output","mean_plots.rds"))
+
+mean_plots[[7]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
+mean_plots[[1]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL acet"}
+mean_plots[[2]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK SPT"}
+mean_plots[[3]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK acet"}
+mean_plots[[4]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM SPT"}
+mean_plots[[5]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM acet"}
+mean_plots[[6]]
+```
+
+\newpage
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
+blank_mean_plots = readRDS(here("data","output","blank_mean_plots.rds"))
+blank_mean_plots[[1]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL acet"}
+blank_mean_plots[[2]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK SPT"}
+blank_mean_plots[[3]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK acet"}
+blank_mean_plots[[4]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM SPT"}
+blank_mean_plots[[5]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM acet"}
+blank_mean_plots[[6]]
+```
+
 # References

### 2021-04-12 R/helper_functions.R

@@ -24,42 +24,3 @@ plot_spectra = function(data, spec_id, sel, id_vars = c("ids","type","species","
     theme_bw() + theme(legend.position = "bottom")
   #return(gdata)
 }
-
-plot_mean_spectra = function(data, subset_sel, spec_id, id.vars = c("ids", "treatment", "Label", "depth", "age", "type"), group = "type"){
-  # subset data
-  pdata = subset(data, eval(subset_sel))
-  
-  # melt data  
-  pdata = melt(cbind(pdata[,id.vars],
-               as.data.frame(pdata[,c(spec_id)])),
-         id.vars=id.vars)
-  
-  # wavenumbers to numeric  
-  pdata$variable = as.numeric(as.character(pdata$variable))
-  
-  #offset
-  pdata = offset_spectra(pdata, group, "value", seq(0, by=max(pdata$value)/2, length.out = length(unique(pdata[,group]))))
-  
-  # mean data over groups
-  pdata_mean = pdata %>%
-    group_by_at(c(group, "variable")) %>%
-    summarise(mean = mean(value))
-  
-  g = ggplot(pdata, aes_string("variable", "value", color = group))+
-    geom_line(aes(group = ids), alpha = .15) + #mean spectra
-    geom_line(data = pdata_mean, aes_string("variable", "mean", group = group), inherit.aes = F, size = 1) + #all spectra
-    xlim(1800,900) +
-    theme_bw() + theme(legend.position = "bottom")  
-  
-  return(g)
-}  
-
-df = data.frame(x = rep(c("a","b"), c(5,10)), y = c(runif(5,3,7),runif(10,)))
-
-offset_spectra = function(data, variable, value, offset) {
-  for (i in 1:length(unique(data[, variable]))) {
-    data[data[, variable] == unique(data[, variable])[i],value] = data[data[, variable] == unique(data[, variable])[i],value] + offset[i]
-  }
-  return(data)
-}
-

### 2021-04-12 R/plotting.R

@@ -1,0 +1,100 @@
+plot_mean_spectra = function(data, subset_sel, spec_id, id.vars = c("ids", "treatment", "Label", "depth", "age", "type"), group = "type"){
+  # subset data
+  pdata = subset(data, eval(subset_sel))
+  
+  # melt data  
+  pdata = melt(cbind(pdata[,id.vars],
+               as.data.frame(pdata[,c(spec_id)])),
+         id.vars=id.vars)
+  
+  # wavenumbers to numeric  
+  pdata$variable = as.numeric(as.character(pdata$variable))
+  
+  #offset
+  pdata = offset_spectra(pdata, group, "value", seq(0, by=max(pdata$value)/3, length.out = length(unique(pdata[,group]))))
+  
+  # mean data over groups
+  pdata_mean = pdata %>%
+    group_by_at(c(group, "variable")) %>%
+    summarise(mean = mean(value))
+  
+  g = ggplot(pdata, aes_string("variable", "value", color = group))+
+    geom_line(aes(group = ids), alpha = .15) + #mean spectra
+    geom_line(data = pdata_mean, aes_string("variable", "mean", group = group), inherit.aes = F, size = 1) + #all spectra
+    xlim(1800,900) +
+    theme_bw() + theme(legend.position = "bottom")  
+  
+  return(g)
+}  
+
+offset_spectra = function(data, variable, value, offset) {
+  for (i in 1:length(unique(data[, variable]))) {
+    data[data[, variable] == unique(data[, variable])[i],value] = data[data[, variable] == unique(data[, variable])[i],value] + offset[i]
+  }
+  return(data)
+}
+
+### creating Plots
+
+DAL_acet = DAL_ %>%
+  subset(species == "Pinus") %>%
+  subset(depth %in% unique(DAL_[DAL_$treatment == "acet", "depth"]))
+
+TSK_acet = TSK_ %>%
+  subset(species == "Pinus") %>%
+  subset(depth %in% unique(TSK_[TSK_$treatment == "acet", "depth"]))
+
+MFM_acet = MFM_ %>%
+  subset(species == "Pinus") %>%
+  subset(depth %in% unique(MFM_[MFM_$treatment == "acet", "depth"]))
+
+l = list()
+
[…]

### 2021-04-12 R/pca.R
added loadings extration from pca_wrap; wrote selecter function

@@ -4,15 +4,27 @@ pca_wrap = function(pca_data,sel){
     pca_data = subset(pca_data, eval(sel))
   }
   pca_res = prcomp(pca_data$sg2)
-  scores =  as.data.frame(pca_res$x[,1:10]) %>%
+  l = list
+  l[["scores"]] =  as.data.frame(pca_res$x[,1:10]) %>%
     mutate(ids = rownames(.),
            core = pca_data$Label,
            treatment = pca_data$treatment,
            depth = pca_data$depth,
            age = pca_data$age
            )
+  l[["loadings"]] = as.data.frame(pca_res$rotation)
   return(scores)
 }
+
+acet_sel = function(data){
+  data_ = filter(data, (treatment == "acet" & species == "Pinus"))
+  data__ = data %>%
+    filter(species == "Pinus") %>%
+    filter(depth %in% unique(data_$depth))
+  return(rbind(data_, data__))
+}
+
+
 pca_data = rbind(subset(DAL, species == "Pinus"), 
                  subset(TSK, species == "Pinus"), 
                  subset(MFM, species == "Pinus"))

### 2021-04-12 manuscript.Rmd
added loadings extration from pca_wrap; wrote selecter function
wrote selectors in pca-file2

@@
 if (file.exists(here("data","output","fresh_spectra.rds"))) {
-  FRE = readRDS(here("data","output","fresh_spectra.rds"))
+  FRE = readRDS(here("data","output","FRE_spectra.rds"))
 } else {
@@
   read_data_fresh()
-  FRE = readRDS(here("data","output","fresh_spectra.rds"))
+  FRE = readRDS(here("data","output","FRE_spectra.rds"))
 }
@@
 
-outliers = c("DAL_rx1GZ", "DAL_2cxp7", "DAL_VzFi0", "DAL_ANdP0", "MFM_av67v", "MFM_OftM2", "MFM_xogJr", "MFM_hoG4y", "MFM_5UNIJ", "TSK_vvgOZ", "DAL_xgYjM", "TSK_m4hN6", "TSK_e04Go")
+outliers = c("DAL_rx1GZ", "DAL_2cxp7", "DAL_VzFi0", "DAL_ANdP0", "MFM_av67v", "MFM_OftM2", "MFM_xogJr", "MFM_hoG4y", "MFM_5UNIJ", "TSK_vvgOZ", "DAL_xgYjM", "TSK_m4hN6", "TSK_e04Go") #THese MFM ids dont exist anymore
 
@@
 ```{r}
-DAL$Label = "DAL"
-TSK$Label = "TSK"
-MFM$Label = "MFM"
+
 ```

### 2021-04-12 R/pca.R

@@ -4,16 +4,16 @@ pca_wrap = function(pca_data,sel){
     pca_data = subset(pca_data, eval(sel))
   }
   pca_res = prcomp(pca_data$sg2)
-  l = list
-  l[["scores"]] =  as.data.frame(pca_res$x[,1:10]) %>%
+  l = list()
+  l[["scores"]] =  as.data.frame(pca_res$x[,1:5]) %>%
     mutate(ids = rownames(.),
            core = pca_data$Label,
            treatment = pca_data$treatment,
            depth = pca_data$depth,
            age = pca_data$age
            )
-  l[["loadings"]] = as.data.frame(pca_res$rotation)
-  return(scores)
+  l[["loadings"]] = as.data.frame(pca_res$rotation[,1:5])
+  return(l)
 }
 
 acet_sel = function(data){
@@ -23,6 +23,27 @@ acet_sel = function(data){
     filter(depth %in% unique(data_$depth))
   return(rbind(data_, data__))
 }
+
+# Cores selected for acet and SPT depths and fresh
+pca_data = rbind(FRE,
+                 acet_sel(DAL),
+                 acet_sel(TSK),
+                 acet_sel(MFM))
+
+l = pca_wrap(pca_data, "")
+ggplot(l[["scores"]], aes(PC1, PC2, color = core)) + geom_point(aes(shape = treatment), size = 4)
+
+# Cores SPT only 
+pca_data = rbind(FRE,
+                 filter(DAL, treatment == "SPT" & species == "Pinus"),
+                 filter(TSK, treatment == "SPT" & species == "Pinus"),
+                 filter(MFM, treatment == "SPT" & species == "Pinus"))
+
+# Cores acet only 
+pca_data = rbind(FRE,
+                 filter(DAL, treatment == "acet" & species == "Pinus"),
+                 filter(TSK, treatment == "acet" & species == "Pinus"),
+                 filter(MFM, treatment == "acet" & species == "Pinus"))
 
 
 pca_data = rbind(subset(DAL, species == "Pinus"), 

### 2021-04-12 R/read_data.R

@@ -23,6 +23,7 @@ read_data_fresh = function() {
   spectra_meta$depth = 0
   spectra_meta$age = 2019
   spectra_meta$species = "Pinus"
+  spectra_meta$orientation = repl(spectra_meta,c("_side"))
   
   rownames(spectra_raw) = ids
   

### 2021-04-15 manuscript.Rmd
added Als comments; wrote pca bootstrap to test orientation and depths

@@
 
-The study of plant micro and macrofossil (spores, seed and pollen) is an integral part of Palaeoecology and Palaeobotany.
-Pollen records are an important tool to reconstruct past floristic, vegetational and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @noguesbravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013].
+The study of plant micro and macrofossil (spores, seed and pollen) is an integral part of palaeoecology and palaeobotany.
+Pollen records are an important tool to reconstruct past floristic, vegetational and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @noguesbravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013] and are important tools to address ecological processes with timescales that operate on centennial/decadal timescales.
 
@@
 
-In recent years, the study of pollen grain chemistry has shown promising results for biological and palaeoecological research.
-Fourier Transform Infrares (FTIR) spectroscopy has become a popular method for analysing pollen chemistry.
+In recent years, the study of pollen-grain chemistry has shown promising results for biological and palaeoecological research.
+Fourier Transform Infrared (FTIR) spectroscopy has become a popular method  for characterising the chemical variation within pollen and spores.
 A number of FTIR studies have identified pollen chemical differences on genus level, which may extent to species level differences [@zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019; @muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020].

### 2021-04-15 R/pca.R

@@ -3,16 +3,19 @@ pca_wrap = function(pca_data,sel){
   if (is.expression(sel)) {
     pca_data = subset(pca_data, eval(sel))
   }
-  pca_res = prcomp(pca_data$sg2)
+  pca_res = prcomp(pca_data$sg2, center = T, scale. = F)
   l = list()
-  l[["scores"]] =  as.data.frame(pca_res$x[,1:5]) %>%
+  l[["pca"]] = pca_res
+  l[["scores"]] =  as.data.frame(pca_res$x[,1:10]) %>%
     mutate(ids = rownames(.),
            core = pca_data$Label,
            treatment = pca_data$treatment,
            depth = pca_data$depth,
-           age = pca_data$age
+           age = pca_data$age,
+           orientation = pca_data$orientation,
+           type = pca_data$type
            )
-  l[["loadings"]] = as.data.frame(pca_res$rotation[,1:5])
+  l[["loadings"]] = as.data.frame(pca_res$rotation[,1:10])
   return(l)
 }
 
@@ -28,24 +31,129 @@ acet_sel = function(data){
 pca_data = rbind(FRE,
                  acet_sel(DAL),
                  acet_sel(TSK),
-                 acet_sel(MFM))
+                 acet_sel(MFM)
+                 )
 
 l = pca_wrap(pca_data, "")
 ggplot(l[["scores"]], aes(PC1, PC2, color = core)) + geom_point(aes(shape = treatment), size = 4)
 
+ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC1)) + geom_line()
+ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC2)) + geom_line()
+
+
 # Cores SPT only 
-pca_data = rbind(FRE,
+pca_data = rbind(
                  filter(DAL, treatment == "SPT" & species == "Pinus"),
                  filter(TSK, treatment == "SPT" & species == "Pinus"),
-                 filter(MFM, treatment == "SPT" & species == "Pinus"))
+                 filter(MFM, treatment == "SPT" & species == "Pinus")
+                 )
+
+l = pca_wrap(pca_data, "")
+ggplot(l[["scores"]], aes(PC1, PC2, color = depth)) + geom_point(aes(shape = treatment), size = 4)
+
[…]

### 2021-05-18 R/pca.R
lda on treatment; before deletion of old pca code (clean up)

@@ -49,7 +49,7 @@ pca_data = rbind(
                  )
 
 l = pca_wrap(pca_data, "")
-ggplot(l[["scores"]], aes(PC1, PC2, color = depth)) + geom_point(aes(shape = treatment), size = 4)
+ggplot(l[["scores"]], aes(PC1, PC2, color = core)) + geom_point( size = 4)
 
 ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC1)) + geom_line()
 ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC2)) + geom_line()
@@ -66,6 +66,15 @@ ggplot(l[["scores"]], aes(PC1, PC2, color = core)) + geom_point(aes(shape = orie
 
 ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC1)) + geom_line()
 ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC2)) + geom_line()
+
+#DAL
+pca_data = rbind(
+  filter(DAL, species == "Pinus")
+)
+
+l = pca_wrap(pca_data, "")
+ggplot(l[["scores"]], aes(PC1, PC2, color = depth)) + geom_point(aes(shape = treatment
+                                                                     ), size = 4)
 
 
 
@@ -97,7 +106,18 @@ bootstrap_pca = function(data, target = "orientation", n = 50){
   return(mean(res))
 }
 
-pca_lda = data.frame(core = c("FRE","DAL","DAL","TSK","TSK","MFM","MFM"),
+pca_lda_treatment = data.frame(core = c("DAL","TSK","MFM"),
+                                 acc = c(bootstrap_pca(filter(DAL, species == "Pinus"), target = "treatment"),
+                                         bootstrap_pca(filter(TSK, species == "Pinus"), target = "treatment"),
+                                         bootstrap_pca(filter(MFM, species == "Pinus"), target = "treatment")
+                                           )*100,
+                                 n = c(nrow(filter(DAL, species == "Pinus")),
+                                       nrow(filter(TSK, species == "Pinus")),
+                                       nrow(filter(MFM, species == "Pinus"))
+                                       )
+)
+
+pca_lda_orientation = data.frame(core = c("FRE","DAL","DAL","TSK","TSK","MFM","MFM"),
                      treatment = c("fresh","SPT","acet","SPT","acet","SPT","acet"),
[…]

### 2021-05-18 R/pca.R
deleted old code

@@ -27,6 +27,34 @@ acet_sel = function(data){
   return(rbind(data_, data__))
 }
 
+bootstrap_pca = function(data, target = "orientation", n = 50){
+  res = rep(0,n)
+  for (i in 1:n) {
+    train = runif(length(data[,target])) <=.7
+    while (sum(train) == length(train)) {
+      train = runif(length(data[,target])) <=.7
+    }
+    
+    #PCA
+    l = pca_wrap(data[train, ],"")
+    #LDA
+    z = lda(as.formula(paste(target,"~ PC1 + PC2")) , l[["scores"]])
+    res[i] =sum(predict(z, as.data.frame(predict(l[["pca"]],data[!train,"sg2"]))[,1:2])$class == data[!train,target])/length(data[!train, target])
+    
+  }
+  
+  return(mean(res))
+}
+
+#show difference between acetolyzed and non
+#show depth differences
+#show difference to fresh
+#show difference in orientation fresh vs spt vs acet
+# pca on each core with lda
+
+#LDAs
+#Orientation
+
 # Cores selected for acet and SPT depths and fresh
 pca_data = rbind(FRE,
                  acet_sel(DAL),
@@ -35,11 +63,83 @@ pca_data = rbind(FRE,
                  )
 
 l = pca_wrap(pca_data, "")
-ggplot(l[["scores"]], aes(PC1, PC2, color = core)) + geom_point(aes(shape = treatment), size = 4)
+pca_plot = ggplot(l[["scores"]], aes(PC1, PC2, color = core)) + geom_point(aes(shape = treatment), size = 4)
 
 ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC1)) + geom_line()
 ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC2)) + geom_line()
 
+# LDA
+pca_lda_treatment = data.frame(core = c("DAL","TSK","MFM"),
+                               acc = c(bootstrap_pca(filter(DAL, species == "Pinus"), target = "treatment"),
+                                       bootstrap_pca(filter(TSK, species == "Pinus"), target = "treatment"),
+                                       bootstrap_pca(filter(MFM, species == "Pinus"), target = "treatment")
+                               )*100,
+                               n = c(nrow(filter(DAL, species == "Pinus")),
+                                     nrow(filter(TSK, species == "Pinus")),
+                                     nrow(filter(MFM, species == "Pinus"))
[…]

### 2021-05-18 manuscript.Rmd
wrote save portion for pca results; added loading those results into manuscript

@@
 <!-- READ DATA -->
-```{r}
+```{r message=FALSE, warning=FALSE, include=FALSE}
 source(here("R","ages.R"))
@@
 
+\newpage
+
 # Results
 
+```{r}
+readRDS(here("data","output","pca_plots.rds"))
+readRDS(here("data","output","lda_tables.rds"))
+```
+
+## Spectra mean plots
 ```{r echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="100%"}

### 2021-05-18 R/pca.R

@@ -137,6 +137,13 @@ pca_lda_depth = data.frame(core = c("DAL","DAL","DAL","TSK","TSK","TSK","MFM","M
                            )
 )
 
+figures = list("pca acet select" = pca_plot)
+tables = list("treatment" = pca_lda_treatment,
+              "orientation" = pca_lda_orientation,
+              "depth" = pca_lda_depth)
+
+saveRDS(figures, here("data","output","pca_plots.rds"))
+saveRDS(tables, here("data","output","lda_tables.rds"))
 #####################################END###############################################
 
 # more PCA code
@@ -149,10 +156,10 @@ pca_data = rbind(
                  )
 
 l = pca_wrap(pca_data, "")
-ggplot(l[["scores"]], aes(PC1, PC2, color = core)) + geom_point( size = 4)
+#ggplot(l[["scores"]], aes(PC1, PC2, color = core)) + geom_point( size = 4)
 
-ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC1)) + geom_line()
-ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC2)) + geom_line()
+#ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC1)) + geom_line()
+#ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC2)) + geom_line()
 
 # Cores acet only 
 pca_data = rbind(
@@ -162,10 +169,10 @@ pca_data = rbind(
                  )
 
 l = pca_wrap(pca_data, "")
-ggplot(l[["scores"]], aes(PC1, PC2, color = core)) + geom_point(aes(shape = orientation), size = 4)
+#ggplot(l[["scores"]], aes(PC1, PC2, color = core)) + geom_point(aes(shape = orientation), size = 4)
 
-ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC1)) + geom_line()
-ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC2)) + geom_line()
+#ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC1)) + geom_line()
+#ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC2)) + geom_line()
 
 #DAL
 pca_data = rbind(
[…]

### 2021-05-18 manuscript.Rmd
wrote some on results

@@
 
-```{r}
-readRDS(here("data","output","pca_plots.rds"))
-readRDS(here("data","output","lda_tables.rds"))
+i. Differences between cores
+ii.i Difference between acetolysed and fossil
+ii.ii Difference between fresh and fossil grains.
+iii effect of depth/age on chemistry
+
+iv effect of orientation on chemistry
+
+There are differences between fresh and fossil grains.
+Most notibly the lipid peak at 1745 is not visible in the fossil grains, neither acetolysed or raw fossil pollen.
+Changes in the carbohydrate region from 1000 to 1250.
+in SPT pollen there is a dominant peak there.
+The most prominent and characteristic sporopollenin peaks are clearly visible in the fossil pollen, 1600, 1510 and 1170, while the rest of the spectra have changed substantially.
+
+1170 is more pronounced in spt than acet
+
+```{r message=FALSE, warning=FALSE, include=FALSE}
+pca_acet_select = readRDS(here("data","output","pca_plots.rds"))
+lda_tables = readRDS(here("data","output","lda_tables.rds"))
+```
+
+```{r pca_acet_sel echo=FALSE, message=FALSE, warning=FALSE, fig.cap = "PCA all cores, only samples present in both acet and SPT"}
+pca_acet_select[1]
 ```
@@
 
+# Discussion
+
 # References

### 2021-05-19 manuscript.Rmd
wrote on results and discussion; rough notes of main findings

@@
 
+we see a peak in the SPT wide spectra at 1090 - 1100 that is not caused by pollen.
+wide peak indicates inorganic compound.
+This peak is not present in either fresh pollen or acetolysed pollen.
+1090 peak is present in all three cores, in addition Tiefer See shows a similar lower peak at 1620 - 1630.
+The most prominent and characteristic sporopollenin peaks are visible in the fossil pollen, 1600, 1510 and 1170 in all three cores.
+PCA ordination shows that the fossil pollen are clearly separated from the fresh pollen, but not clearly separable into cores (Fig pca_acet_sel)
+
 There are differences between fresh and fossil grains.
@@
 in SPT pollen there is a dominant peak there.
-The most prominent and characteristic sporopollenin peaks are clearly visible in the fossil pollen, 1600, 1510 and 1170, while the rest of the spectra have changed substantially.
+The most prominent and characteristic sporopollenin peaks are visible in the fossil pollen, 1600, 1510 and 1170, while the rest of the spectra have changed substantially.
 
-1170 is more pronounced in spt than acet
+1170 is more pronounced in spt than acet.
+the 1090 peak dissappear after acetolysis from all spectra as does the 1630 in TSK. 
+a peak at 1040 is now visible in acetolysed pollen that was covered by the wide peak in the SPT pollen.
+The region 1000 - 1200 is now dominated by two peaks 1040 and 1170 and its shoulders.
+In the PCA ordination the acetolysed pollen is separated from the SPT pollen, stronger separation for TSK than for DAL and MFM.
+A LDA on each core separately had `r signif(mean(lda_tables[[1]]$acc),1)` % accuracy to distinguish between sample treatment (Table LDA treatment).
+
+LDA to classify the pollen into depths showed higher accuracy for acetolysed pollen than SPT pollen.
+For the SPT pollen the accuracy increased when the depths were restricted to the same depths as the acetolysed samples.
+Generally lower number of depths to classify increased accuracy.
+sporopollenin peaks are similar between depths, changes in the other parts of the spectra are driving separation.
 
@@
 
+strange peak in SPT samples most likely biogenic silica.
+is removed with acetolysis (check references of this happening anywhere else)
+
+some effect of the core on chemistry, because TSK is more dissimilar than the other two cores and TSK is laminated sediment.
+BSi peak at 1630 was only present here.
+TSK samples are also youngest, most dissimilar between SPT and acetolysed pollen, maybe because these pollen had still remnants of non sporopollenin parts that were removed by acetolysis
+
+acetolysation had an effect on pollen chemistry.
+especially in the 1000-1200 region, which had the BSi peak but also the region from 1150 to 1250 changes, which is outside of the main effects of the BSi peak.
+These regions are mainly caused by carbohydrates, which could be cellulose from the grain wall.
+
+the pollen spectra are way different from fresh pollen.
+the sporopollenin peaks are all present, but since so much of the rest of the spectra changes, we cant clearly associate any relative changes in these peaks to changes in composition of the sporopollenin or simply changes that only affect other components of the pollen (e.g.
+carbohydrates).
+GCMS on these pollen can give more answers
+
+no clear patterns with depths chemically. the more depths are added, the more inaccurate the LDA became.
+for acetolysed pollen samples with the youngest and oldest pollen were used, while SPT had all samples.
+
+Paragraph on practicality
+This is an improvement on previous single grain measurements with hyperion, where each grain had to be targeted individually.
+instead a large area could be measured and spectra selected post-hoc.
+This introduced also the possibilities of automization, such as machinelearning, which have shown to be very adapt at extracting object from larger images.
+the spectra can be seen as images, suitable for image segmentation/object location deep neural networks such as RCNN/YOLO etc.
+Nevertheless this introduces another challenge: concentration of pollen.
+Some of our samples had very low concentration of pollen
+
 # References

### 2021-05-21 manuscript.Rmd
wrote on results and reorganized discussion

@@
 
+There are differences between fresh and fossil grains.
+The most prominent and characteristic sporopollenin peaks are visible in the fossil pollen, 1600, 1510 and 1170, while the rest of the spectra have changed substantially.
+Most notibly the lipid peak at 1745 is not visible in the fossil grains, neither acetolysed or raw fossil pollen.
+Changes in the carbohydrate region from 1000 to 1250.
+
 we see a peak in the SPT wide spectra at 1090 - 1100 that is not caused by pollen.
-wide peak indicates inorganic compound.
+
 This peak is not present in either fresh pollen or acetolysed pollen.
-1090 peak is present in all three cores, in addition Tiefer See shows a similar lower peak at 1620 - 1630.
+1090 peak is present in all three cores, in addition Tiefer See shows a similar lower peak at 1630.
 The most prominent and characteristic sporopollenin peaks are visible in the fossil pollen, 1600, 1510 and 1170 in all three cores.
 PCA ordination shows that the fossil pollen are clearly separated from the fresh pollen, but not clearly separable into cores (Fig pca_acet_sel)
-
-There are differences between fresh and fossil grains.
-Most notibly the lipid peak at 1745 is not visible in the fossil grains, neither acetolysed or raw fossil pollen.
-Changes in the carbohydrate region from 1000 to 1250.
-in SPT pollen there is a dominant peak there.
-The most prominent and characteristic sporopollenin peaks are visible in the fossil pollen, 1600, 1510 and 1170, while the rest of the spectra have changed substantially.
 
@@
 
-strange peak in SPT samples most likely biogenic silica.
+.Overall intro to discussion
+
+.difference to fresh pollen
+the pollen spectra are way different from fresh pollen.
+the sporopollenin peaks are all present, but since so much of the rest of the spectra changes, we cant clearly associate any relative changes in these peaks to changes in composition of the sporopollenin or simply changes that only affect other components of the pollen (e.g.
+carbohydrates).
+
+.on bsi peak in spt samples
+wide peak indicates inorganic compound.
+strange peak in SPT samples most likely biogenic silica [@alipourInfraredMicrospectroscopyBionanomaterials2016].
 is removed with acetolysis (check references of this happening anywhere else)
 
+.core effects
 some effect of the core on chemistry, because TSK is more dissimilar than the other two cores and TSK is laminated sediment.
@@
 
+.acetolysis effects
 acetolysation had an effect on pollen chemistry.
@@
 
-the pollen spectra are way different from fresh pollen.
-the sporopollenin peaks are all present, but since so much of the rest of the spectra changes, we cant clearly associate any relative changes in these peaks to changes in composition of the sporopollenin or simply changes that only affect other components of the pollen (e.g.
-carbohydrates).
 GCMS on these pollen can give more answers
 
+.depth patterns
 no clear patterns with depths chemically. the more depths are added, the more inaccurate the LDA became.
@@
 
+.application for palaeoecology
 Paragraph on practicality
@@
 
+.conclusions
+
 # References

### 2021-05-21 manuscript.Rmd
wrote on results and discussion; some core differences lda code in PCA.R

@@
 
+(NEEDS SHORT PARAGRAPH ON NEW MICROSCOPE SETUP. capturing of larger area instead of single grain targeting)
+
 In this study we investigate the chemistry of fossil pollen to explore three potential sources of chemical alteration: i. the effects of diagenesis, ii. acetolysis and iii. age on pollen chemistry.
@@
 
-iv effect of orientation on chemistry
-
 There are differences between fresh and fossil grains.
@@
 The most prominent and characteristic sporopollenin peaks are visible in the fossil pollen, 1600, 1510 and 1170 in all three cores.
-PCA ordination shows that the fossil pollen are clearly separated from the fresh pollen, but not clearly separable into cores (Fig pca_acet_sel)
+PCA ordination shows that the fossil pollen are clearly separated from the fresh pollen, but not clearly separable into three cores (Fig \@ref(fig:pca_acet_sel)).
+Tiefer See pollen is separated from the other two cores, which cluster closer together.
+This is the same for acetolysed and SPT pollen.
 
+## Differences between acetolysed and SPT pollen
 1170 is more pronounced in spt than acet.
@@
 In the PCA ordination the acetolysed pollen is separated from the SPT pollen, stronger separation for TSK than for DAL and MFM.
-A LDA on each core separately had `r signif(mean(lda_tables[[1]]$acc),1)` % accuracy to distinguish between sample treatment (Table LDA treatment).
+A LDA on each core separately had `r signif(mean(lda_tables[[1]]$acc),1)` % accuracy to distinguish between sample treatment (Table \@ref(tab:ldatreatment)).
 
-LDA to classify the pollen into depths showed higher accuracy for acetolysed pollen than SPT pollen.
+LDA to classify the pollen into depths showed higher accuracy for acetolysed pollen than SPT pollen (Table \@ref(tab:ldadepth)).
 For the SPT pollen the accuracy increased when the depths were restricted to the same depths as the acetolysed samples.
@@
 pca_acet_select[1]
+```
+
+```{r ldatreatment, echo=FALSE, results='asis'}
+
+kable(lda_tables[["treatment"]], booktabs = T, format = "latex",
+      caption = "Accuracy of bootstrapped PCA LDA to distinguish treatment (SPT vs acet)")
+```
+
+```{r ldadepth, echo=FALSE, results='asis'}
+
+kable(lda_tables[["depth"]], booktabs = T, format = "latex",
+      caption = "Accuracy of bootstrapped PCA LDA to distinguish depths for each core. treatments are separate.")
 ```
@@
 \newpage
-
+<!--
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
@@
 ```
+-->
 
@@
 .difference to fresh pollen
-the pollen spectra are way different from fresh pollen.
-the sporopollenin peaks are all present, but since so much of the rest of the spectra changes, we cant clearly associate any relative changes in these peaks to changes in composition of the sporopollenin or simply changes that only affect other components of the pollen (e.g.
+the fossil pollen spectra show changes from fresh pollen.
+lipid and protein peaks have dissappeared and there is a lot of differences in the carbohydrate region.
+this part of the spectra has changed substantially and most fresh components are most likely removed.
+Nevertheless some peaks in the carb region indicate some amount of cellulose or grain material still present.
+the characteristic sporopollenin peaks are all present, but since so much of the rest of the spectra changes, we cant clearly associate any relative changes in these peaks to changes in composition of the sporopollenin or simply changes that only affect other components of the pollen (e.g.
 carbohydrates).
@@
 wide peak indicates inorganic compound.
-strange peak in SPT samples most likely biogenic silica [@alipourInfraredMicrospectroscopyBionanomaterials2016].
-is removed with acetolysis (check references of this happening anywhere else)
+strange peak in SPT samples most likely biogenic silica, when compared to this biogenic silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016].
+we observed a crystalline layer over our pollen grains, when observed under the microscope. 
+this peak is removed with acetolysis (check references of this happening anywhere else)
 
 .core effects
-some effect of the core on chemistry, because TSK is more dissimilar than the other two cores and TSK is laminated sediment.
+some effect of the core on chemistry, because TSK is more dissimilar than the other two core, could be connected to the fact that TSK is laminated sediment.
 BSi peak at 1630 was only present here.
@@
 instead a large area could be measured and spectra selected post-hoc.
-This introduced also the possibilities of automization, such as machinelearning, which have shown to be very adapt at extracting object from larger images.
+This introduces also the possibilities of automization, such as machinelearning, which have shown to be very adapt at extracting object from larger images.
 the spectra can be seen as images, suitable for image segmentation/object location deep neural networks such as RCNN/YOLO etc.
 Nevertheless this introduces another challenge: concentration of pollen.
-Some of our samples had very low concentration of pollen
+Some of our samples had very low concentration of pollen and would need a concentration step in sample processing.
+Our results do show that acetolysis does not remove any of the sporopollenin peaks, but some of the other parts of the spectra, and could therefore be seen as a standardizing procedure.
+Further studies with GCMS methods on fossil pollen and several species have to be performed.
 

### 2021-05-21 R/pca.R

@@ -8,7 +8,7 @@ pca_wrap = function(pca_data,sel){
   l[["pca"]] = pca_res
   l[["scores"]] =  as.data.frame(pca_res$x[,1:10]) %>%
     mutate(ids = rownames(.),
-           core = pca_data$Label,
+           Label = pca_data$Label,
            treatment = pca_data$treatment,
            depth = pca_data$depth,
            age = pca_data$age,
@@ -145,6 +145,18 @@ tables = list("treatment" = pca_lda_treatment,
 saveRDS(figures, here("data","output","pca_plots.rds"))
 saveRDS(tables, here("data","output","lda_tables.rds"))
 #####################################END###############################################
+spt = rbind(
+  filter(DAL, treatment == "SPT" & species == "Pinus"),
+  filter(TSK, treatment == "SPT" & species == "Pinus"),
+  filter(MFM, treatment == "SPT" & species == "Pinus")
+)
+bootstrap_pca(spt, target = "Label")
+acet = rbind(
+    filter(DAL, treatment == "acet" & species == "Pinus"),
+  filter(TSK, treatment == "acet" & species == "Pinus"),
+  filter(MFM, treatment == "acet" & species == "Pinus")
+)
+bootstrap_pca(acet, target = "Label")
 
 # more PCA code
 

### 2021-05-21 manuscript.Rmd
adjusted some chunk option

@@
 
+```{r message=FALSE, warning=FALSE, include=FALSE}
+pca_acet_select = readRDS(here("data","output","pca_plots.rds"))
+lda_tables = readRDS(here("data","output","lda_tables.rds"))
+```
+
 # Results
@@
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-pca_acet_select = readRDS(here("data","output","pca_plots.rds"))
-lda_tables = readRDS(here("data","output","lda_tables.rds"))
-```
-
-```{r pca_acet_sel echo=FALSE, message=FALSE, warning=FALSE, fig.cap = "PCA all cores, only samples present in both acet and SPT"}
+```{r pca_acet_sel, echo=FALSE, message=FALSE, warning=FALSE, fig.cap = "PCA all cores, only samples present in both acet and SPT"}
 pca_acet_select[1]

### 2021-05-21 manuscript.Rmd
fixed error in whole core mean plot generation; commented out the other mean plots

@@
 ## Spectra mean plots
+
 ```{r echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="100%"}
@@
 
-```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
+```{r echo=FALSE, fig.cap="DAL whole core mean plots", message=FALSE, warning=FALSE, out.width="100%"}
+mean_plots = readRDS(here("data","output","mean_plots_whole.rds"))
+
 mean_plots[[1]]
@@
 
-```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL acet"}
+```{r echo=FALSE, fig.cap="MFM whole core mean plots", message=FALSE, warning=FALSE, out.width="100%"}
 mean_plots[[2]]
@@
 
+```{r echo=FALSE, fig.cap="TSK whole core mean plots", message=FALSE, warning=FALSE, out.width="100%"}
+mean_plots[[3]]
+```
+<!--
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
+#mean_plots[[1]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL acet"}
+#mean_plots[[2]]
+```
+
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK SPT"}
-mean_plots[[3]]
+#mean_plots[[3]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK acet"}
-mean_plots[[4]]
+#mean_plots[[4]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM SPT"}
-mean_plots[[5]]
+#mean_plots[[5]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM acet"}
-mean_plots[[6]]
+#mean_plots[[6]]
 ```
-
+-->
 \newpage
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
-blank_mean_plots = readRDS(here("data","output","blank_mean_plots.rds"))
-blank_mean_plots[[1]]
+#blank_mean_plots = readRDS(here("data","output","blank_mean_plots.rds"))
+#blank_mean_plots[[1]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL acet"}
-blank_mean_plots[[2]]
+#blank_mean_plots[[2]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK SPT"}
-blank_mean_plots[[3]]
+#blank_mean_plots[[3]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK acet"}
-blank_mean_plots[[4]]
+#blank_mean_plots[[4]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM SPT"}
-blank_mean_plots[[5]]
+#blank_mean_plots[[5]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM acet"}
-blank_mean_plots[[6]]
+#blank_mean_plots[[6]]
 ```

### 2021-05-21 R/plotting.R

@@ -34,6 +34,18 @@ offset_spectra = function(data, variable, value, offset) {
   return(data)
 }
 
+### Treatments mean plots over whole core
+
+l = list()
+
+l[["DAL"]] = plot_mean_spectra(DAL, expression(species == "Pinus"), spec_id = "emsc", group = "treatment")
+
+l[["MFM"]] = plot_mean_spectra(MFM, expression(species == "Pinus"), spec_id = "emsc", group = "treatment")
+
+l[["TSK"]] = plot_mean_spectra(TSK, expression(species == "Pinus"), spec_id = "emsc", group = "treatment")
+
+saveRDS(l, here("data","output","mean_plots_whole.rds"))
+
 ### creating Plots
 
 DAL_acet = DAL_ %>%

### 2021-05-25 manuscript.Rmd
rewriting of some results and discussion bits

@@
 
-There are differences between fresh and fossil grains.
-The most prominent and characteristic sporopollenin peaks are visible in the fossil pollen, 1600, 1510 and 1170, while the rest of the spectra have changed substantially.
-Most notibly the lipid peak at 1745 is not visible in the fossil grains, neither acetolysed or raw fossil pollen.
-Changes in the carbohydrate region from 1000 to 1250.
+There are differences between fresh and fossil grains (Fig. \@ref(fig:meanDAL), \@ref(fig:meanMFM) and \@ref(fig:meanTSK).
+The prominent and characteristic sporopollenin peaks (1600, 1510 and 1170 cm^-1^) are visible in the fossil pollen, while other parts of the spectra show substantial differences.
+Most notibly the lipid peak at 1745 cm^-1^, caused by triglycerides, is not visible in the fossil pollen grains, neither acetolysed or raw.
+There are also several differences in the carbohydrate region from 1000 to 1250 cm^-1^ between fresh and fossil pollen grains.
+A peak at 1030  and ~1200 cm^-1^ is visible in the acetolysed pollen but not in the fresh spectra, while we see a broad peak at 1090-1100 cm^-1^ in the SPT spectra of all three cores.
+This peak is not present in either fresh pollen or acetolysed pollen, additionally Tiefer See shows a similar peak at 1630.
 
-we see a peak in the SPT wide spectra at 1090 - 1100 that is not caused by pollen.
-
-This peak is not present in either fresh pollen or acetolysed pollen.
-1090 peak is present in all three cores, in addition Tiefer See shows a similar lower peak at 1630.
-The most prominent and characteristic sporopollenin peaks are visible in the fossil pollen, 1600, 1510 and 1170 in all three cores.
-PCA ordination shows that the fossil pollen are clearly separated from the fresh pollen, but not clearly separable into three cores (Fig \@ref(fig:pca_acet_sel)).
+PCA ordination shows that the fossil pollen are clearly separated from the fresh pollen, but not clearly separable into the three cores (Fig \@ref(fig:pcaacetsel)).
 Tiefer See pollen is separated from the other two cores, which cluster closer together.
 This is the same for acetolysed and SPT pollen. 
+A PCA-LDA model to differentiate fossil pollen into the three cores achieved ~60% accuracy to classify MFM and DAL pollen and 85% accuracy to identify TSK pollen (REF TO SM).
 
-## Differences between acetolysed and SPT pollen
-1170 is more pronounced in spt than acet.
-the 1090 peak dissappear after acetolysis from all spectra as does the 1630 in TSK. 
-a peak at 1040 is now visible in acetolysed pollen that was covered by the wide peak in the SPT pollen.
-The region 1000 - 1200 is now dominated by two peaks 1040 and 1170 and its shoulders.
+the peak at 1170 cm^-1^ is more pronounced in spt than acetolyzed pollen.
+
+the 1090 cm^-1^ peak dissappears after acetolysis from all spectra as does the 1630 cm^-1^ peak in TSK. 
+a peak at 1040 cm^-1^ is now visible in acetolysed pollen that is not visible in the SPT pollen.
+The region 1000 - 1200 cm^-1^ is now dominated by two peaks at 1040 and 1170 and its shoulders.
 In the PCA ordination the acetolysed pollen is separated from the SPT pollen, stronger separation for TSK than for DAL and MFM.
@@
 
-```{r pca_acet_sel, echo=FALSE, message=FALSE, warning=FALSE, fig.cap = "PCA all cores, only samples present in both acet and SPT"}
+```{r pcaacetsel, echo=FALSE, message=FALSE, warning=FALSE, fig.cap = "PCA all cores, only samples present in both acet and SPT"}
 pca_acet_select[1]
@@
 
-```{r echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="100%"}
-mean_plots = readRDS(here("data","output","mean_plots.rds"))
+```{r meanFRE, echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="100%"}
+mean_plotsSM = readRDS(here("data","output","mean_plots.rds"))
 
-mean_plots[[7]]
+mean_plotsSM[[7]]
 ```
 
-```{r echo=FALSE, fig.cap="DAL whole core mean plots", message=FALSE, warning=FALSE, out.width="100%"}
+```{r meanDAL, echo=FALSE, fig.cap="DAL whole core mean plots", message=FALSE, warning=FALSE, out.width="100%"}
 mean_plots = readRDS(here("data","output","mean_plots_whole.rds"))
@@
 
-```{r echo=FALSE, fig.cap="MFM whole core mean plots", message=FALSE, warning=FALSE, out.width="100%"}
+```{r meanMFM, echo=FALSE, fig.cap="MFM whole core mean plots", message=FALSE, warning=FALSE, out.width="100%"}
 mean_plots[[2]]
@@
 
-```{r echo=FALSE, fig.cap="TSK whole core mean plots", message=FALSE, warning=FALSE, out.width="100%"}
+```{r meanTSK, echo=FALSE, fig.cap="TSK whole core mean plots", message=FALSE, warning=FALSE, out.width="100%"}
 mean_plots[[3]]
 ```
-<!--
+
+# Discussion
+
+.Overall intro to discussion
+
+.difference to fresh pollen
+the fossil pollen spectra show changes from fresh pollen.
+lipid and protein peaks have dissappeared and there is a lot of differences in the carbohydrate region.
+this part of the spectra has changed substantially and most fresh components are most likely removed.
+Nevertheless some peaks in the carb region indicate some amount of cellulose or grain material still present.
+the characteristic sporopollenin peaks are all present, but since so much of the rest of the spectra changes, we cant clearly associate any relative changes in these peaks to changes in composition of the sporopollenin or simply changes that only affect other components of the pollen (e.g.
+carbohydrates).
+
+.on bsi peak in spt samples
+wide peak indicates inorganic compound.
+This peak in the SPT samples is most likely biogenic silica, when compared to this biogenic silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016].
+we observed a crystalline layer over our pollen grains, when observed under the microscope. 
+this peak is removed with acetolysis (check references of this happening anywhere else)
+
+.acetolysis effects
+acetolysation had an effect on pollen chemistry.
+especially in the 1000-1200 region, which had the BSi peak but also the region from 1150 to 1250 changes, which is outside of the main effects of the BSi peak.
+These regions are mainly caused by carbohydrates, which could be cellulose from the grain wall.
+
+GCMS on these pollen can give more answers
+
+.core effects
+some effect of the core on chemistry, because TSK is more dissimilar than the other two core, could be connected to the fact that TSK is laminated sediment.
+BSi peak at 1630 was only present here.
+
+We saw large overlap between MFM and DAL in the PCA ordination and lower accuracies in LDA differentiating between cores, both in SPT and acetolysed pollen.
+This seems to indicate that pine pollen from different cores are chemically quite similar, even cores that are far apart.
+TSK pollen on the other hand could be clearly separated from the other two cores, in both SPT and acetolysed pollen. 
+In fact acetolysed TSK pollen is even further apart from the rest of the acetolysed pollen than SPT pollen is.
+There seem to be chemical differences between TSK and the other two cores that are present in SPT and acetolysed pollen.
+One reason for the dissimilarity of SPT TSK pollen is the BSi peak at 1630 which is only present in TSK, but even acetolysed pollen from TSK is quite different from the other cores according to PCA and LDA analysis.
+The main difference of TSK pollen to the other cores is that TSK pollen is much younger than the other two cores.
+Our TSK samples are from the past 80 years, while MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene.
+The other difference is that TSK is a varved core, while MFM and DAL are not.
+Our findings suggest a chemical difference in fossil pollen grains, either caused by diagenesis (varved vs non-varved), age or combination of the two that is persistent even after acetolysis.
+
+.depth patterns
+no clear patterns with depths chemically. the more depths are added, the more inaccurate the LDA became.
+for acetolysed pollen samples with the youngest and oldest pollen were used, while SPT had all samples.
+
+.application for palaeoecology
+Paragraph on practicality
+This is an improvement on previous single grain measurements with hyperion, where each grain had to be targeted individually.
+instead a large area could be measured and spectra selected post-hoc.
+This introduces also the possibilities of automization, such as machinelearning, which have shown to be very adapt at extracting object from larger images.
+the spectra can be seen as images, suitable for image segmentation/object location deep neural networks such as RCNN/YOLO etc.
+Nevertheless this introduces another challenge: concentration of pollen.
+Some of our samples had very low concentration of pollen and would need a concentration step in sample processing.
+Our results do show that acetolysis does not remove any of the sporopollenin peaks, but some of the other parts of the spectra, and could therefore be seen as a standardizing procedure.
+Further studies with GCMS methods on fossil pollen and several species have to be performed.
+
+.conclusions
+
+# Mean spectra figures SM
+
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
-#mean_plots[[1]]
+mean_plotsSM[[1]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL acet"}
-#mean_plots[[2]]
+mean_plotsSM[[2]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK SPT"}
-#mean_plots[[3]]
+mean_plotsSM[[3]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK acet"}
-#mean_plots[[4]]
+mean_plotsSM[[4]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM SPT"}
-#mean_plots[[5]]
+mean_plotsSM[[5]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM acet"}
-#mean_plots[[6]]
+mean_plotsSM[[6]]
 ```
--->
+
 \newpage
@@
 
-# Discussion
-
-.Overall intro to discussion
-
-.difference to fresh pollen
-the fossil pollen spectra show changes from fresh pollen.
-lipid and protein peaks have dissappeared and there is a lot of differences in the carbohydrate region.
-this part of the spectra has changed substantially and most fresh components are most likely removed.
-Nevertheless some peaks in the carb region indicate some amount of cellulose or grain material still present.
-the characteristic sporopollenin peaks are all present, but since so much of the rest of the spectra changes, we cant clearly associate any relative changes in these peaks to changes in composition of the sporopollenin or simply changes that only affect other components of the pollen (e.g.
-carbohydrates).
-
-.on bsi peak in spt samples
-wide peak indicates inorganic compound.
-strange peak in SPT samples most likely biogenic silica, when compared to this biogenic silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016].
-we observed a crystalline layer over our pollen grains, when observed under the microscope. 
-this peak is removed with acetolysis (check references of this happening anywhere else)
-
-.core effects
-some effect of the core on chemistry, because TSK is more dissimilar than the other two core, could be connected to the fact that TSK is laminated sediment.
-BSi peak at 1630 was only present here.
-TSK samples are also youngest, most dissimilar between SPT and acetolysed pollen, maybe because these pollen had still remnants of non sporopollenin parts that were removed by acetolysis
-
-.acetolysis effects
-acetolysation had an effect on pollen chemistry.
-especially in the 1000-1200 region, which had the BSi peak but also the region from 1150 to 1250 changes, which is outside of the main effects of the BSi peak.
-These regions are mainly caused by carbohydrates, which could be cellulose from the grain wall.
-
-GCMS on these pollen can give more answers
-
-.depth patterns
-no clear patterns with depths chemically. the more depths are added, the more inaccurate the LDA became.
-for acetolysed pollen samples with the youngest and oldest pollen were used, while SPT had all samples.
-
-.application for palaeoecology
-Paragraph on practicality
-This is an improvement on previous single grain measurements with hyperion, where each grain had to be targeted individually.
-instead a large area could be measured and spectra selected post-hoc.
-This introduces also the possibilities of automization, such as machinelearning, which have shown to be very adapt at extracting object from larger images.
-the spectra can be seen as images, suitable for image segmentation/object location deep neural networks such as RCNN/YOLO etc.
-Nevertheless this introduces another challenge: concentration of pollen.
-Some of our samples had very low concentration of pollen and would need a concentration step in sample processing.
-Our results do show that acetolysis does not remove any of the sporopollenin peaks, but some of the other parts of the spectra, and could therefore be seen as a standardizing procedure.
-Further studies with GCMS methods on fossil pollen and several species have to be performed.
-
-.conclusions
-
 # References

### 2021-05-27 manuscript.Rmd
after meeting with Boris and AL

@@
 
-So far, most FTIR studies have used modern material or material extracted from herbarium specimens .
-In some cases the pollen samples are processed with acetolysis following standard palynological methods [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989] to simulate fossilization [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018].
-The assumption is that acetolysis does not alter the chemistry of sporopollenin and is not affecting any chemical signal.
-To date, there has not been a comparison between acetolysed and untreated fossil pollen to examine any effect of acetolysis on sporopollenin or chemical degradation of pollen during diagenesis.
+So far, most FTIR studies focused on identification of pollen have used modern material and bulk measurements, while some used FTIR microscopes.
+These FTIR microscopes required manual targeting of small areas of the sample slides to record spectra of single or groups of pollen grains.
+FTIR as a method has advanced on a technical level over the past 5 to 10 years.
+The development and capabilities of microscopes with connected FTIR spectrometers now allow easier and faster capture of spectra of pollen grains.
+With these newer machines a larger area is irradiated with FTIR and spectra are recorded at higher resolution over the entire irradiated area.
+This allows capture of an image and matrix of spectra over the same sample area, which enables e.g.
+the extraction of spectra from several single grains from the same area on the sample without the need to relocate the detector of the microscope or move the slide.
 
-(NEEDS SHORT PARAGRAPH ON NEW MICROSCOPE SETUP. capturing of larger area instead of single grain targeting)
+Palaeoecologists use a combination of several chemical extractions to isolate and concentrate pollen from the sediment matrix.
+Most common is acetolysis in addition to HF treatment [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
+These strong acid treatments aim to remove other organic particulates and silicates, while preserving morphological integrity of pollen grains [REF].
+A limited number of studies have used acetolysis and other chemical treatments in combination with FTIR on pollen to simulate fossilization [@julierChemotaxonomyToolInterpreting2016; @jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or study the effects of acid extractions [@jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
+@jardineImpactOxidationSpore2015 found limited to no effect of short acetolysis (<10 min) on *Lycopodium* spore exine chemistry, while labile outer compounds and the intine/protoplast where removed.
 
-In this study we investigate the chemistry of fossil pollen to explore three potential sources of chemical alteration: i. the effects of diagenesis, ii. acetolysis and iii. age on pollen chemistry.
-We extracted pollen from three sediment cores using density separation to obtain chemically untreated fossil spectra.
+An alternative to chemical extraction is density separation using e.g.
+sodium polytungstate (SPT), which is used to isolate pollen from minerogenic sediments [REFS] or as preparations for pollen isotope studies [REFS].
+Density separation may be a an alternative to acid extractions for pollen studies concerned with chemical alteration of pollen grain chemistry.
+SPT is an inorganic compound, which should not alter pollen chemistry in any way and is removed with several washes after extraction.
+Contamination of the sample is easily identifiable by characteristic wavebands visible in the spectra.
+
+In this study we explore the use of SPT density separation as an alternative to acetolysis by investigating the chemical alteration of SPT and acetolyis on modern pollen and apply both extraction methods to three fossil cores.
+With this, we aim to investigate changes in fossil pollen chemistry, due to: i.
+differences between modern and fossil material, ii.
+extraction method and iii.
+effect of diagenesis (age of fossil pollen).
+The effect of treatment with SPT on pollen chemistry is tested on modern material to establish if any chemical changes occur.
 The pollen was collected from three different cores in three locations (Germany and Norway), which allows us to examine potential differences in diagenesis on pollen chemistry at different locations.
-One replicate of the pollen samples in each core were acetolysed to analyse any changes caused by the strong acid treatment and compared to the untreated replicate.
 Each core was sub sampled at different depths to allows us to investigate chemical changes in the pollen with age.
+The pollen was extracted from the sediment cores using density separation, with one replicate being acetolysed afterwards.
 
@@
 \newpage
-<!--
+
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
-#blank_mean_plots = readRDS(here("data","output","blank_mean_plots.rds"))
-#blank_mean_plots[[1]]
+blank_mean_plots = readRDS(here("data","output","blank_mean_plots.rds"))
+blank_mean_plots[[1]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL acet"}
-#blank_mean_plots[[2]]
+blank_mean_plots[[2]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK SPT"}
-#blank_mean_plots[[3]]
+blank_mean_plots[[3]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK acet"}
-#blank_mean_plots[[4]]
+blank_mean_plots[[4]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM SPT"}
-#blank_mean_plots[[5]]
+blank_mean_plots[[5]]
 ```
@@
 ```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM acet"}
-#blank_mean_plots[[6]]
+blank_mean_plots[[6]]
 ```
--->
 

### 2021-05-27 R/pca.R

@@ -63,7 +63,7 @@ pca_data = rbind(FRE,
                  )
 
 l = pca_wrap(pca_data, "")
-pca_plot = ggplot(l[["scores"]], aes(PC1, PC2, color = core)) + geom_point(aes(shape = treatment), size = 4)
+pca_plot = ggplot(l[["scores"]], aes(PC1, PC2, color = core)) + geom_point(aes(shape = treatment), size = 2)
 
 ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC1)) + geom_line()
 ggplot(l[["loadings"]], aes(as.numeric(rownames(l[["loadings"]])),PC2)) + geom_line()

### 2021-05-27 R/Regression.R
trying to find the right regression method

@@ -1,0 +1,40 @@
+# Stepwise Regression
+
+library(pls)
+
+pca_data = rbind(
+                 acet_sel(DAL),
+                 acet_sel(TSK),
+                 acet_sel(MFM)
+)
+pca_data$core = model.matrix(~Label -1, data=pca_data)
+plsr = plsr(model.matrix(~treatment -1, data=acet_sel(DAL))~sg2,
+            6, data = acet_sel(DAL), 
+            validation = "LOO")
+
+DAL_ = acet_sel(DAL)
+colnames(DAL_$sg2)=colnames(DAL$sg2)
+plsr = plsr(model.matrix(~treatment -1, data=DAL_)~sg2,
+           6, data = DAL_,
+           validation = "LOO", center=F, scale=F)
+plot(rownames(plsr$loadings), plsr$loadings[,1], type = "l")
+plot(plsr$scores[,1], plsr$scores[,2])
+
+
+colnames(pca_data$sg2)=colnames(DAL$sg2)
+plsr = plsr(model.matrix(~Label -1, data=pca_data)~sg2,
+            6, data = pca_data,
+            validation = "LOO", center=F, scale=T)
+plot(rownames(plsr$loadings), plsr$loadings[,1], type = "l")
+plot(plsr$scores[,1], plsr$scores[,2])
+
+reg_data = cbind(pca_data$Label, as.data.frame(pca_data$sg2))
+colnames(reg_data)[1] = "core"
+reg_data$core = model.matrix(~core -1, data=reg_data)
+intercept_only = lm(core ~ 1, data=reg_data)
+all = lm(core ~ ., data = reg_data)
+both <- step(intercept_only, direction='backward', scope=formula(all))
+
+data(yarn)
+## Default methods:
+yarn.pcr <- pcr(density ~ NIR, 6, data = yarn, validation = "CV")

### 2021-05-28 manuscript.Rmd
rewrote introduction

@@
 This is a promising finding for the application of pollen identification with chemical methods in palaeoecological research, because it has the potential to increase taxomonic resolution @dellannaPollenDiscriminationClassification2009; @zimmermannAnalysisAllergenicPollen2016.
-Further, chemical methods are suitable for automation and offer a more objective method than traditional morphological methods [ @mondolApplicationHighThroughputScreening2019; @zimmermannAnalysisAllergenicPollen2016; @guedesPollenRamanSpectra2014].
+<!--MOVE TO DISCUSSION: Further, chemical methods are suitable for automation and offer a more objective method than traditional morphological methods [ @mondolApplicationHighThroughputScreening2019; @zimmermannAnalysisAllergenicPollen2016; @guedesPollenRamanSpectra2014].
 In addition to taxonomic differences in pollen chemistry, there is evidence of environmental signals that are recorded in the chemistry of pollen [@bagciogluMultiscaleVibrationalSpectroscopic2015; @kendelChemicalAnalysisPollen2020; @diehnCombiningChemicalInformation2020].
-Several studies have shown links between solar and UV radiation and pollen chemistry as well as responses in pollen chemistry to temperature changes.
+Several studies have shown links between solar and UV radiation and pollen chemistry as well as responses in pollen chemistry to temperature changes. -->
 
-So far, most FTIR studies focused on identification of pollen have used modern material and bulk measurements, while some used FTIR microscopes.
-These FTIR microscopes required manual targeting of small areas of the sample slides to record spectra of single or groups of pollen grains.
+So far, most FTIR studies focused on identification of pollen have used modern material and bulk measurements, while some have used FTIR microscopes.
+These FTIR microscopes required manual targeting of small areas of the sample slides to record spectra of single grains or groups of pollen grains.
 FTIR as a method has advanced on a technical level over the past 5 to 10 years.
 The development and capabilities of microscopes with connected FTIR spectrometers now allow easier and faster capture of spectra of pollen grains.
-With these newer machines a larger area is irradiated with FTIR and spectra are recorded at higher resolution over the entire irradiated area.
+With these newer machines a larger area is irradiated and spectra are recorded at higher resolution over the entire irradiated area.
 This allows capture of an image and matrix of spectra over the same sample area, which enables e.g.
@@
 Palaeoecologists use a combination of several chemical extractions to isolate and concentrate pollen from the sediment matrix.
-Most common is acetolysis in addition to HF treatment [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
-These strong acid treatments aim to remove other organic particulates and silicates, while preserving morphological integrity of pollen grains [REF].
-A limited number of studies have used acetolysis and other chemical treatments in combination with FTIR on pollen to simulate fossilization [@julierChemotaxonomyToolInterpreting2016; @jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or study the effects of acid extractions [@jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
-@jardineImpactOxidationSpore2015 found limited to no effect of short acetolysis (<10 min) on *Lycopodium* spore exine chemistry, while labile outer compounds and the intine/protoplast where removed.
+Most common is acetolysis in addition to HF treatment and HCl [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
+These strong acid treatments aim to remove other organic particulates, calcium carbonates and silicates, while preserving morphological integrity of pollen grains [REF].
+A limited number of studies have used acetolysis and other chemical treatments in combination with FTIR on pollen to simulate fossilization [@julierChemotaxonomyToolInterpreting2016; @jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or study the effects of acid extractions [@dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000;@schulzeosthoffPhenolsIntegratedCompounds1987;@jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
+These results suggest that acetolysis may alter the chemical composition of sporopollenin.
 
@@
 sodium polytungstate (SPT), which is used to isolate pollen from minerogenic sediments [REFS] or as preparations for pollen isotope studies [REFS].
-Density separation may be a an alternative to acid extractions for pollen studies concerned with chemical alteration of pollen grain chemistry.
-SPT is an inorganic compound, which should not alter pollen chemistry in any way and is removed with several washes after extraction.
-Contamination of the sample is easily identifiable by characteristic wavebands visible in the spectra.
+Density separation may be a an alternative to acid extractions for pollen studies concerned with chemical alteration of sporomorph chemistry.
+SPT is used to adjust the density of sample suspension and facilitace separation of less dense particulates (e.g. pollen and other microfossils) over mineral particles.
+SPT is removed from the sample with several washes after extraction and any contamination of the sample is easily identifiable by characteristic wavebands in the spectra.
 
@@
 iv effect of orientation on chemistry
-
-acetolysis alters sporopollenin [@dominguezIsolationIntactPollen1998; @schulzeosthoffPhenolsIntegratedCompounds1987] affecting solubility
 

### 2021-05-28 manuscript.Rmd
expanded on methods

@@
 The sub samples from Tiefer see are from the last 80 yrs of the core (1933 - 2011), while the samples from Meerfelder Maar cover the timeperiod from the late Allerød to the Younger Dryas/Holocene transition (~12900 - ~11000 BP).
-The samples were extracted using density separation using dissolved Sodium Polytungstate (SPT) at 1.43 (g/ml)? as separation medium. <!--FM: check Unit-->
-During density separation a known volume of SPT (density 1.43) solution is added to the sample and homogenized/vortexed, followed by centrifugation.
-After centrifugation, particulates/material heavier than 1.43 has deposited at the bottom of the vial and the lighter fraction can be decanted and collected.
+The samples were extracted using density separation using dissolved Sodium Polytungstate (SPT) [@kampsAgentSeparationDissolved1985] at 1.43 g cm^-3^ as separation medium.
+During density separation a known volume of SPT (density 1.43 g cm^-3^) solution is added to the sample and homogenized/vortexed, followed by centrifugation.
+After centrifugation, particulates/material heavier than 1.43 g cm^-3^ has deposited at the bottom of the vial and the lighter fraction can be decanted and collected.
 In our case the lighter fraction contained most of the pollen.
-The density of pollen is reported at <1.4 UNIT [@nakagawaDensemediaSeparationMore1998; @regnellPreparativeCentrifugationNew1996].
+There is no agreement on the density of the pollen exine and reported values vary between  <1.4 g cm^-3^ [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996], 1.88 cm^-3^ [@nakagawaDensemediaSeparationMore1998] and 2.2 cm^-3^  [@juvigneDensiteExinesQuelques1973].
+
+## FTIR measurements
+
+The spectra were measured with a Vertex 70 FTIR spectrometer connected to a Hyperion 3000 IR microscope (Bruker Optik, Ettlingen, Germany) equipped with a focal plane array (FPA) nitrogen cooled mercury cadmium telluride (MCT) detector and a 15x magnification objective.
+The FPA-MCT detector allows the capture of FTIR images, where each pixel is a full spectra, instead point measurements.
+For this study images of 128x128 pixels were taken and each spectra was recorded with 64 scans.
+The accompanied visible images were obtained by a charge coupled device (CCD) camera coupled to the microscope.
+Single grain pollen grain spectra were extracted from the FTIR images, by averaging all pixel spectra in the image area of the pollen grain. 
+Data acquisition and instrument control was performed using the OPUS 7.5 software (Bruker Optik GmbH, Germany).
+
+The extracted pollen was applied to ZnSe slides and a background spectra was measured before each sample on an empty area of the slide.
+The pollen treated with SPT and acetolysis were applied onto the slide suspended in water and dried out before measurement under a fume hood.
+The extracted sediment samples contained other small material than pollen (e.g. diatom fragments, organic material fragments, etc).
+We recorded several "blank" spectra of dried residue that formed over the dried pollen residue (see also SM).
+
+## Spectra preprocessing and analysis
 
@@
 average all spectra and stack plots of different depth
-
-Pollen extracted from sediment samples 
-[@kampsAgentSeparationDissolved1985] SPT patent
-[@nakagawaDensemediaSeparationMore1998] 
-[@regnellPreparativeCentrifugationNew1996] density of pollen 1.4
 

### 2021-05-31 R/Regression.R
core for regression coeff using plsr

@@ -2,6 +2,50 @@
 
 library(pls)
 
+FRE$treat = "fresh"
+DAL$treat = "fossil"
+MFM$treat = "fossil"
+TSK$treat = "fossil"
+
+# PLS fresh vs fossil
+
+pls_data = filter(rbind(FRE,DAL,MFM,TSK), species == "Pinus")
+pls_data$y = model.matrix(~treatment -1, data = pls_data)
+colnames(pls_data$sg2) = colnames(DAL$sg2)
+
+results = plsr(y ~sg2, 6, data = pls_data, validation = "LOO")
+plot(results, plottype = "coefficients")
+
+# PLS SPT vs acet
+## all cores
+pls_data = filter(
+  rbind(
+    acet_sel(DAL),
+    acet_sel(MFM),
+    acet_sel(TSK)
+    ), 
+  species == "Pinus")
+pls_data$y = model.matrix(~treatment -1, data = pls_data)
+colnames(pls_data$sg2) = colnames(DAL$sg2)
+
+results = plsr(y ~sg2, 6, data = pls_data, validation = "LOO")
+plot(results, plottype = "coefficients")
+
+## DAL
+pls_data = filter(
+  rbind(
+    acet_sel(DAL)
+  ), 
+  species == "Pinus")
+pls_data$y = model.matrix(~treatment -1, data = pls_data)
+colnames(pls_data$sg2) = colnames(DAL$sg2)
+
+results = plsr(y ~sg2, 6, data = pls_data, validation = "LOO")
+plot(results, plottype = "coefficients")
+
+
+
+## 
 pca_data = rbind(
                  acet_sel(DAL),
                  acet_sel(TSK),

### 2021-05-31 R/Regression.R
wrote coef plot function

@@ -1,5 +1,22 @@
 # Stepwise Regression
 
+plot_coef = function(obj, comp = 1){
+  tmp = as.data.frame(obj[,,comp])
+  n = ncol(tmp)
+  tmp$wavenumbers = as.numeric(rownames(obj[,,comp]))
+  for (i in 1:n) {
+    tmp[,i] = tmp[,i] + i * max(abs(obj[,,comp]))
+  }
+  tmp2 = melt(tmp, id.vars = c("wavenumbers"))
+  colnames(tmp2) = c("wavenumbers", "treatment", "coef")
+  g = ggplot(tmp2, aes(x = wavenumbers, y = coef, color = treatment)) + 
+    geom_line() + xlim(1900,900) + theme_bw() +
+    theme(axis.ticks.y = element_blank(),
+          axis.text.y = element_blank())
+    
+  return(g)
+}
+
 library(pls)
 
 FRE$treat = "fresh"
@@ -43,42 +60,3 @@ colnames(pls_data$sg2) = colnames(DAL$sg2)
 results = plsr(y ~sg2, 6, data = pls_data, validation = "LOO")
 plot(results, plottype = "coefficients")
 
-
-
-## 
-pca_data = rbind(
-                 acet_sel(DAL),
-                 acet_sel(TSK),
-                 acet_sel(MFM)
-)
-pca_data$core = model.matrix(~Label -1, data=pca_data)
-plsr = plsr(model.matrix(~treatment -1, data=acet_sel(DAL))~sg2,
-            6, data = acet_sel(DAL), 
-            validation = "LOO")
-
-DAL_ = acet_sel(DAL)
-colnames(DAL_$sg2)=colnames(DAL$sg2)
-plsr = plsr(model.matrix(~treatment -1, data=DAL_)~sg2,
-           6, data = DAL_,
-           validation = "LOO", center=F, scale=F)
-plot(rownames(plsr$loadings), plsr$loadings[,1], type = "l")
-plot(plsr$scores[,1], plsr$scores[,2])
-
-
-colnames(pca_data$sg2)=colnames(DAL$sg2)
-plsr = plsr(model.matrix(~Label -1, data=pca_data)~sg2,
-            6, data = pca_data,
-            validation = "LOO", center=F, scale=T)
-plot(rownames(plsr$loadings), plsr$loadings[,1], type = "l")
-plot(plsr$scores[,1], plsr$scores[,2])
-
[…]

### 2021-06-01 R/Regression.R
troubleshooting coef plot function,

@@ -27,11 +27,11 @@ TSK$treat = "fossil"
 # PLS fresh vs fossil
 
 pls_data = filter(rbind(FRE,DAL,MFM,TSK), species == "Pinus")
-pls_data$y = model.matrix(~treatment -1, data = pls_data)
+pls_data$y = model.matrix(~treat -1, data = pls_data)
 colnames(pls_data$sg2) = colnames(DAL$sg2)
 
 results = plsr(y ~sg2, 6, data = pls_data, validation = "LOO")
-plot(results, plottype = "coefficients")
+plot_coef(results$coefficients)
 
 # PLS SPT vs acet
 ## all cores
@@ -46,7 +46,7 @@ pls_data$y = model.matrix(~treatment -1, data = pls_data)
 colnames(pls_data$sg2) = colnames(DAL$sg2)
 
 results = plsr(y ~sg2, 6, data = pls_data, validation = "LOO")
-plot(results, plottype = "coefficients")
+plot_coef(results$coefficients)
 
 ## DAL
 pls_data = filter(

### 2021-06-01 R/Regression.R
added a plsr to file for testing

@@ -48,6 +48,19 @@ colnames(pls_data$sg2) = colnames(DAL$sg2)
 results = plsr(y ~sg2, 6, data = pls_data, validation = "LOO")
 plot_coef(results$coefficients)
 
+## SPT core only lda on cores
+pls_data = filter(
+  rbind(
+    acet_sel(DAL),
+    acet_sel(MFM),
+    acet_sel(TSK)
+  ), 
+  species == "Pinus" & treatment == "SPT")
+pls_data$y = model.matrix(~Label -1, data = pls_data)
+colnames(pls_data$sg2) = colnames(DAL$sg2)
+
+results = plsr(y ~sg2, 6, data = pls_data, validation = "LOO")
+
 ## DAL
 pls_data = filter(
   rbind(

### 2021-06-01 manuscript.Rmd
rewrote first results paragraph to include plsr coefficients

@@
 
-PCA on grains for exploration
+PLSR second deriv on treatment (fresh vs fossil)
 
@@
 
-There are differences between fresh and fossil grains (Fig. \@ref(fig:meanDAL), \@ref(fig:meanMFM) and \@ref(fig:meanTSK).
+There are differences between modern and fossil grains (Fig. \@ref(fig:meanDAL), \@ref(fig:meanMFM) and \@ref(fig:meanTSK).
 The prominent and characteristic sporopollenin peaks (1600, 1510 and 1170 cm^-1^) are visible in the fossil pollen, while other parts of the spectra show substantial differences.
-Most notibly the lipid peak at 1745 cm^-1^, caused by triglycerides, is not visible in the fossil pollen grains, neither acetolysed or raw.
-There are also several differences in the carbohydrate region from 1000 to 1250 cm^-1^ between fresh and fossil pollen grains.
-A peak at 1030  and ~1200 cm^-1^ is visible in the acetolysed pollen but not in the fresh spectra, while we see a broad peak at 1090-1100 cm^-1^ in the SPT spectra of all three cores.
-This peak is not present in either fresh pollen or acetolysed pollen, additionally Tiefer See shows a similar peak at 1630.
+Most notibly the lipid peak at 1745 cm^-1^, caused by triglycerides, is not visible in the fossil pollen grains, neither in the SPT treated or acetolysed pollen.
+There are also several differences in the "fingerprint" region (1500 to 1800 cm^-1^) and carbohydrate region from 1000 to 1250 cm^-1^ between modern and fossil pollen grains.
+A peak at 1030 and ~1200 cm^-1^ is visible in the acetolysed pollen but not in the modern spectra, while we see a broad peak at 1090-1100 cm^-1^ in the SPT spectra of all three cores.
+This peak is not present in either modern pollen or acetolysed pollen, additionally Tiefer See shows a similar peak at 1630.
 
-PCA ordination shows that the fossil pollen are clearly separated from the fresh pollen, but not clearly separable into the three cores (Fig \@ref(fig:pcaacetsel)).
+After visual inspection of the spectra, the chemical differences between modern and fossil polle were explored with a PLSR.
+The ordination showed that the fossil pollen are clearly separated from the modern pollen (Fig \@ref(fig:plsrord)).
+The regression coefficients reflect our observations from the mean spectra and show high importance of the 1745 lipid peak, the 1500 - 1800 region and the 1170 peak.
+
+<!-- 
 Tiefer See pollen is separated from the other two cores, which cluster closer together.
@@
 A PCA-LDA model to differentiate fossil pollen into the three cores achieved ~60% accuracy to classify MFM and DAL pollen and 85% accuracy to identify TSK pollen (REF TO SM).
+
+-->
 

### 2021-06-01 manuscript.Rmd
wrote score plot function

@@
 
-After visual inspection of the spectra, the chemical differences between modern and fossil polle were explored with a PLSR.
+After visual inspection of the spectra, the chemical differences between modern and fossil pollen were explored with a PLSR model.
 The ordination showed that the fossil pollen are clearly separated from the modern pollen (Fig \@ref(fig:plsrord)).
-The regression coefficients reflect our observations from the mean spectra and show high importance of the 1745 lipid peak, the 1500 - 1800 region and the 1170 peak.
+The regression coefficients reflect our observations from the visual inspection and show high importance of the 1745 lipid peak, the 1500 - 1800 region and the 1170 peak (Fig \@ref(fig:plsrcoef)).
 

### 2021-06-01 R/Regression.R

@@ -17,6 +17,17 @@ plot_coef = function(obj, comp = 1){
   return(g)
 }
 
+plot_scor = function(obj, meta){
+  meta = meta[,c("Label","treatment")]
+  tmp = as.data.frame(obj[])
+  n = ncol(tmp)
+  colnames(tmp) = c("PC1","PC2","PC3","PC4","PC5","PC6")
+  tmp = cbind(tmp, meta)
+  g = ggplot(tmp, aes_string(x = "PC1", y =  "PC2", color = "Label")) +
+    geom_point(aes_string(shape = "treatment"), size = 1) + scale_shape_manual(values = c(17, 19, 15)) + theme_bw()
+  return(g)
+}
+
 library(pls)
 
 FRE$treat = "fresh"
@@ -32,6 +43,7 @@ colnames(pls_data$sg2) = colnames(DAL$sg2)
 
 results = plsr(y ~sg2, 6, data = pls_data, validation = "LOO")
 plot_coef(results$coefficients)
+plot_scor(results$scores, pls_data)
 
 # PLS SPT vs acet
 ## all cores

### 2021-06-02 manuscript.Rmd
figures export to png

@@
 There are also several differences in the "fingerprint" region (1500 to 1800 cm^-1^) and carbohydrate region from 1000 to 1250 cm^-1^ between modern and fossil pollen grains.
-A peak at 1030 and ~1200 cm^-1^ is visible in the acetolysed pollen but not in the modern spectra, while we see a broad peak at 1090-1100 cm^-1^ in the SPT spectra of all three cores.
-This peak is not present in either modern pollen or acetolysed pollen, additionally Tiefer See shows a similar peak at 1630.
 
@@
 
-the peak at 1170 cm^-1^ is more pronounced in spt than acetolyzed pollen.
+A peak at 1030 and ~1200 cm^-1^ is visible in the acetolysed pollen but not in the modern spectra or SPT pollen.
+We see a broad peak at 1090-1100 cm^-1^ in the SPT spectra of all three cores.
+This peak is not present in either modern pollen or acetolysed pollen, additionally Tiefer See shows a similar peak at 1630.
 
@@
 
-```{r meanFRE, echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="100%"}
+```{r meanspectra, echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="100%"}
+knitr::include_graphics(here("figures", "mean_plots.png"))
+
 mean_plotsSM = readRDS(here("data","output","mean_plots.rds"))
-
-mean_plotsSM[[7]]
+mean_plots = readRDS(here("data","output","mean_plots_whole.rds"))
 ```
 
-```{r meanDAL, echo=FALSE, fig.cap="DAL whole core mean plots", message=FALSE, warning=FALSE, out.width="100%"}
-mean_plots = readRDS(here("data","output","mean_plots_whole.rds"))
-
-mean_plots[[1]]
-```
-
-```{r meanMFM, echo=FALSE, fig.cap="MFM whole core mean plots", message=FALSE, warning=FALSE, out.width="100%"}
-mean_plots[[2]]
-```
-
-```{r meanTSK, echo=FALSE, fig.cap="TSK whole core mean plots", message=FALSE, warning=FALSE, out.width="100%"}
-mean_plots[[3]]
+```{r ordinations, echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="100%"}
+knitr::include_graphics(here("figures", "ord_plots.png"))
 ```
@@
 
+\newpage
+\newpage
 # Mean spectra figures SM

### 2021-06-02 R/Regression.R

@@ -28,12 +28,32 @@ plot_scor = function(obj, meta){
   return(g)
 }
 
+create_layout = function(h,w,x,y,split = 2){
+  c = 1
+  h = h-1
+  w = w-1
+  layout = list("t" = rep(0,x*y), "l" =rep(0,x*y), "b" =rep(0,x*y), "r" =rep(0,x*y))
+  for (x in 1:x) {
+    for (y in 1:y) {
+      layout[["t"]][c] = y+(y-1)*(h)
+      layout[["l"]][c] =x+(x-1)*(w)
+      layout[["b"]][c] =y+h*y
+      layout[["r"]][c] =x+w*x
+      c = c+1
+    }
+  }
+  attr(layout,  "class") = "patch_area"
+  return(layout)
+}
+
 library(pls)
 
 FRE$treat = "fresh"
 DAL$treat = "fossil"
 MFM$treat = "fossil"
 TSK$treat = "fossil"
+
+ord_plots = list()
 
 # PLS fresh vs fossil
 
@@ -42,8 +62,10 @@ pls_data$y = model.matrix(~treat -1, data = pls_data)
 colnames(pls_data$sg2) = colnames(DAL$sg2)
 
 results = plsr(y ~sg2, 6, data = pls_data, validation = "LOO")
-plot_coef(results$coefficients)
-plot_scor(results$scores, pls_data)
+
+ord_plots[["allcoresord"]] = plot_scor(results$scores, pls_data)
+ord_plots[["allcorescoepc1"]] = plot_coef(results$coefficients)
+ord_plots[["allcorescoepc2"]] = plot_coef(results$coefficients, 2)
 
 # PLS SPT vs acet
 ## all cores
@@ -54,11 +76,13 @@ pls_data = filter(
     acet_sel(TSK)
     ), 
   species == "Pinus")
-pls_data$y = model.matrix(~treatment -1, data = pls_data)
+pls_data$y = model.matrix(~Label -1, data = pls_data)
 colnames(pls_data$sg2) = colnames(DAL$sg2)
 
-results = plsr(y ~sg2, 6, data = pls_data, validation = "LOO")
-plot_coef(results$coefficients)
+results = plsr(y ~ sg2, 6, data = pls_data, validation = "LOO")
+ord_plots[["allcoressptacetord"]] = plot_scor(results$scores, pls_data)
+ord_plots[["allcoressptacetpc1"]] = plot_coef(results$coefficients)
[…]

### 2021-06-02 manuscript.Rmd
adjusted ordination figure

@@
 
-```{r ordinations, echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="100%"}
+```{r ordinations, echo=FALSE, fig.cap="a.
+plsr ordination scores all samples diff fresh vs fossil b.
+regression coef; PC1 c.
+regression coef PC2; d plsr ordination scores only samples from cores diff spt vs acet; e.
+regression coef PC1; f.
+regression coef PC2", message=FALSE, warning=FALSE, out.width="100%"}
 knitr::include_graphics(here("figures", "ord_plots.png"))

### 2021-06-02 R/Regression.R

@@ -1,12 +1,23 @@
 # Stepwise Regression
 
-plot_coef = function(obj, comp = 1){
-  tmp = as.data.frame(obj[,,comp])
-  n = ncol(tmp)
-  tmp$wavenumbers = as.numeric(rownames(obj[,,comp]))
-  for (i in 1:n) {
-    tmp[,i] = tmp[,i] + i * max(abs(obj[,,comp]))
+plot_coef = function(obj, comp = 1, coef = T){
+  if (coef) {
+    tmp = as.data.frame(obj[,,comp])
+    n = ncol(tmp)
+    tmp$wavenumbers = as.numeric(rownames(obj[,,comp]))
+    for (i in 1:n) {
+      tmp[,i] = tmp[,i] + i * max(abs(obj[,,comp]))
+    }
+
+  } else {
+    tmp = as.data.frame(obj[,comp])
+    n = ncol(tmp)
+    tmp$wavenumbers = as.numeric(rownames(obj[,comp]))
+    for (i in 1:n) {
+      tmp[,i] = tmp[,i] + i * max(abs(obj[,comp]))
+    }
   }
+
   tmp2 = melt(tmp, id.vars = c("wavenumbers"))
   colnames(tmp2) = c("wavenumbers", "treatment", "coef")
   g = ggplot(tmp2, aes(x = wavenumbers, y = coef, color = treatment)) + 
@@ -175,27 +186,25 @@ ggsave(here("figures","mean_plots.png"), plot = g, width = 17, height = 25, unit
 plsr_plot = readRDS(here("data","output","ordplot.rds"))
 
 design = "
-11112222
-11112222
-33334444
-33334444
-55556666
-55556666
-77778888
-77778888
+1111112222
+1111112222
+1111113333
+1111113333
+4444445555
+4444445555
+4444446666
+4444446666
 "
 
 #create_layout(3,2,2,4)
 
 o = plsr_plot[[1]] +
   plsr_plot[[2]] +
-  plsr_plot[[4]] +
-  plsr_plot[[5]] +
-  plsr_plot[[7]] +
-  plsr_plot[[8]] +
-  plsr_plot[[10]] +
-  plsr_plot[[11]] +
+  plsr_plot[[3]] +
+  plsr_plot[[13]] +
+  plsr_plot[[14]] +
+  plsr_plot[[15]] +
   plot_layout(design = design , byrow = T, guides = "collect") + 
   plot_annotation(tag_levels = c("a"))
 
[…]

### 2021-06-02 manuscript.Rmd
removed pca fig and lda tables

@@
 
-```{r pcaacetsel, echo=FALSE, message=FALSE, warning=FALSE, fig.cap = "PCA all cores, only samples present in both acet and SPT"}
-pca_acet_select[1]
-```
-
-```{r ldatreatment, echo=FALSE, results='asis'}
-
-kable(lda_tables[["treatment"]], booktabs = T, format = "latex",
-      caption = "Accuracy of bootstrapped PCA LDA to distinguish treatment (SPT vs acet)")
-```
-
-```{r ldadepth, echo=FALSE, results='asis'}
-
-kable(lda_tables[["depth"]], booktabs = T, format = "latex",
-      caption = "Accuracy of bootstrapped PCA LDA to distinguish depths for each core. treatments are separate.")
-```
-
 ## Spectra mean plots

### 2021-06-03 manuscript.Rmd
large rewrite after figure adjustment

@@
 ---
-title: 'Chemistry of fossil Pinus pollen and the effect of acetolysis'
+title: 'Chemistry of fossil Pinus pollen in comparison to modern pollen'
 date: '2020/02/04'
@@
 With these newer machines a larger area is irradiated and spectra are recorded at higher resolution over the entire irradiated area.
-This allows capture of an image and matrix of spectra over the same sample area, which enables e.g.
-the extraction of spectra from several single grains from the same area on the sample without the need to relocate the detector of the microscope or move the slide.
+One such improvement is the use of a focal plane array detector (FPA).
+A FPA detector allows capture of an image and matrix of spectra over the same sample area instead of a single spectra at a specific point/area.
+This enables e.g. the extraction of spectra from several single grains from the same area on the sample without the need to relocate the detector of the microscope or move the slide.
 
-Palaeoecologists use a combination of several chemical extractions to isolate and concentrate pollen from the sediment matrix.
+[FM: needs transition or reorg of paragraphs]
+
+There are several extraction methods used to isolate pollen and other microfossils from sediment cores. 
+Traditionally, palaeoecologists use a combination of several chemical extractions to isolate and concentrate pollen from the sediment matrix.
 Most common is acetolysis in addition to HF treatment and HCl [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
-These strong acid treatments aim to remove other organic particulates, calcium carbonates and silicates, while preserving morphological integrity of pollen grains [REF].
+These strong acid treatments aim to remove other organic particles, calcium carbonates and silicates, while preserving the morphological integrity of pollen grains [REF].
 A limited number of studies have used acetolysis and other chemical treatments in combination with FTIR on pollen to simulate fossilization [@julierChemotaxonomyToolInterpreting2016; @jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or study the effects of acid extractions [@dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000;@schulzeosthoffPhenolsIntegratedCompounds1987;@jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
@@
 Density separation may be a an alternative to acid extractions for pollen studies concerned with chemical alteration of sporomorph chemistry.
-SPT is used to adjust the density of sample suspension and facilitace separation of less dense particulates (e.g. pollen and other microfossils) over mineral particles.
+SPT is used to adjust the density of sample suspension and facilitace separation of less dense particles (e.g. pollen and other microfossils) over mineral particles.
 SPT is removed from the sample with several washes after extraction and any contamination of the sample is easily identifiable by characteristic wavebands in the spectra.
 
-In this study we explore the use of SPT density separation as an alternative to acetolysis by investigating the chemical alteration of SPT and acetolyis on modern pollen and apply both extraction methods to three fossil cores.
+In this study we explore the use of SPT density separation as an alternative to acetolysis by investigating the chemical alteration of both extraction methods to three fossil cores.
 With this, we aim to investigate changes in fossil pollen chemistry, due to: i.
@@
 effect of diagenesis (age of fossil pollen).
+For this purpose we use an FTIR microscope with FPA detector to capture spectra of modern and fossil *Pinus sylvestris* pollen.
+<!--
 The effect of treatment with SPT on pollen chemistry is tested on modern material to establish if any chemical changes occur.
+-->
 The pollen was collected from three different cores in three locations (Germany and Norway), which allows us to examine potential differences in diagenesis on pollen chemistry at different locations.
@@
 
-i. Differences between cores
-ii.i Difference between acetolysed and fossil
-ii.ii Difference between fresh and fossil grains.
-iii effect of depth/age on chemistry
+```{r meanspectra, echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="100%"}
+knitr::include_graphics(here("figures", "mean_plots.png"))
 
-There are differences between modern and fossil grains (Fig. \@ref(fig:meanDAL), \@ref(fig:meanMFM) and \@ref(fig:meanTSK).
+mean_plotsSM = readRDS(here("data","output","mean_plots.rds"))
+mean_plots = readRDS(here("data","output","mean_plots_whole.rds"))
+```
+
+There are differences between modern and fossil grains (Fig. \@ref(fig:meanspectra)).
 The prominent and characteristic sporopollenin peaks (1600, 1510 and 1170 cm^-1^) are visible in the fossil pollen, while other parts of the spectra show substantial differences.
@@
 After visual inspection of the spectra, the chemical differences between modern and fossil pollen were explored with a PLSR model.
-The ordination showed that the fossil pollen are clearly separated from the modern pollen (Fig \@ref(fig:plsrord)).
-The regression coefficients reflect our observations from the visual inspection and show high importance of the 1745 lipid peak, the 1500 - 1800 region and the 1170 peak (Fig \@ref(fig:plsrcoef)).
+The ordination showed that the fossil pollen are clearly separated from the modern pollen (Fig \@ref(fig:ordinations) a).
+The regression coefficients reflect our observations from the visual inspection and show high importance of the 1745 lipid peak, the 1500 - 1800 region and the 1170 peak (Fig \@ref(fig:ordinations) b and c).
 
@@
 
+A peak at 1030 and ~1200 cm^-1^ is visible in the acetolysed pollen but not in the modern spectra or SPT pollen.
 -->
 
-A peak at 1030 and ~1200 cm^-1^ is visible in the acetolysed pollen but not in the modern spectra or SPT pollen.
-We see a broad peak at 1090-1100 cm^-1^ in the SPT spectra of all three cores.
-This peak is not present in either modern pollen or acetolysed pollen, additionally Tiefer See shows a similar peak at 1630.
+We see a broad peak at 1090-1100 cm^-1^ in the non acetolysed spectra of all three cores (Fig. \@ref(fig:meanspectra) b, c and d).
+This peak is not present in either modern pollen or acetolysed pollen, additionally Tiefer See shows a similar peak at 1630 cm^-1^.
 
-the 1090 cm^-1^ peak dissappears after acetolysis from all spectra as does the 1630 cm^-1^ peak in TSK. 
-a peak at 1040 cm^-1^ is now visible in acetolysed pollen that is not visible in the SPT pollen.
-The region 1000 - 1200 cm^-1^ is now dominated by two peaks at 1040 and 1170 and its shoulders.
-In the PCA ordination the acetolysed pollen is separated from the SPT pollen, stronger separation for TSK than for DAL and MFM.
-A LDA on each core separately had `r signif(mean(lda_tables[[1]]$acc),1)` % accuracy to distinguish between sample treatment (Table \@ref(tab:ldatreatment)).
+The 1090 cm^-1^ peak dissappears after acetolysis from all spectra as does the 1630 cm^-1^ peak in Tiefer See. 
+The region 1000 - 1200 cm^-1^ is now dominated in acetolysed fossil pollen by two peaks and its shoulders at 1040 and 1170 cm^-1^.
+This 1040 cm^-1^ has high importance for the PLSR model differentiating acetolysed and non acetolysed pollen (Fig. \@ref(fig:ordinations) e and f).
+Using this PLSR model the acetolysed pollen is clearly separated from the non acetolysed pollen, with a much higher distance between samples for acetolysed pollen compared to non acetolysed pollen (Fig.
+\@ref(fig:ordinations) d).
+In addition, there is some differentiation between the cores, samples from Tiefer See are more separate from samples of the other two cores, which show more similar scores in the ordination.
 
+<!--
 LDA to classify the pollen into depths showed higher accuracy for acetolysed pollen than SPT pollen (Table \@ref(tab:ldadepth)).
@@
 sporopollenin peaks are similar between depths, changes in the other parts of the spectra are driving separation.
-
-## Spectra mean plots
-
-```{r meanspectra, echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="100%"}
-knitr::include_graphics(here("figures", "mean_plots.png"))
-
-mean_plotsSM = readRDS(here("data","output","mean_plots.rds"))
-mean_plots = readRDS(here("data","output","mean_plots_whole.rds"))
-```
+-->
 
@@
 
+\newpage
+
 # Discussion
@@
 .Overall intro to discussion
+Our aim with this study was to use a new type of detector (FPA-MCT) in combination with a FTIR microscope to explore the chemistry of fossil pollen.
+We analysed modern and fossil material and explored the effects of sample processing methods on fossil pollen.
 
-.difference to fresh pollen
-the fossil pollen spectra show changes from fresh pollen.
-lipid and protein peaks have dissappeared and there is a lot of differences in the carbohydrate region.
-this part of the spectra has changed substantially and most fresh components are most likely removed.
-Nevertheless some peaks in the carb region indicate some amount of cellulose or grain material still present.
+.difference to fresh pollen (FM: have to mention Jardine 21 here)
+The fossil pollen spectra show changes from fresh pollen.
+Lipid and protein peaks have dissappeared and there are substantial changes in the carbohydrate region in fossil pollen.
+From the regression coefficients, we see that the largest changes are in the 1500 to 1750 cm^-1^ region, which have many lipid and protein peaks (eg. 1745 and 1545 cm^-1^ etc).
+Acetolysed pollen showed two dominant peaks in the carbohydrate region which are not visible in the modern spectra at 1040 and 1170 - 1190 cm^-1^ or the non acetolysed pollen.
+This change in the 1170 - 1190 cm^-1^ region especially is unique to acetolysized pollen and not visible in non acetolysed pollen and had higher regression coefficients that the surrounding wavenumbers (Fig.
+\@ref(fig:ordinations)).
+The presence of this new peak in the shoulder of the 1170 cm^-1^ peak, which can indicate sporopollenin and carbohydrates, suggest changes in the pollen chemistry.
+These changes may be related to removal and or alteration of grain wall carbohydrates (cellulose etc) as well as changes to sporopollenin due to acetolysis.
+[FM: this would be answered by acetolysis of the fresh grains!!]
+In non acetolysed pollen the 1190 peak is not visible in any of the cores, which indicates that it is not due to aging/diageneis, but acetolysis.
+Unfortunately, the non acetolysed pollen showed a wide peak covering part of the carbohydrate region at 1040 cm^-1^.
+This peak in the non acetolysed samples is most likely biogenic silica, when compared to this biogenic silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016] and hydrated silica spectra [@SilicaATRIRSpectrum].
+During the measurements with the microscope, we applied our extracted pollen as droplets onto ZnSe slides and waited for the water to evaporate.
+After the droplets had dried we observed a crystalline layer over the same area that covered all particles and pollen (ref Foto in SM).
+From the chemical spectra, we believe that it is dissolved silica that crystallised. 
+the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.
+
 the characteristic sporopollenin peaks are all present, but since so much of the rest of the spectra changes, we cant clearly associate any relative changes in these peaks to changes in composition of the sporopollenin or simply changes that only affect other components of the pollen (e.g.
@@
 
-.on bsi peak in spt samples
-wide peak indicates inorganic compound.
-This peak in the SPT samples is most likely biogenic silica, when compared to this biogenic silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016].
-we observed a crystalline layer over our pollen grains, when observed under the microscope. 
-this peak is removed with acetolysis (check references of this happening anywhere else)
-
-.acetolysis effects
-acetolysation had an effect on pollen chemistry.
-especially in the 1000-1200 region, which had the BSi peak but also the region from 1150 to 1250 changes, which is outside of the main effects of the BSi peak.
-These regions are mainly caused by carbohydrates, which could be cellulose from the grain wall.
-
-GCMS on these pollen can give more answers
-
-.core effects
-some effect of the core on chemistry, because TSK is more dissimilar than the other two core, could be connected to the fact that TSK is laminated sediment.
-BSi peak at 1630 was only present here.
-
-We saw large overlap between MFM and DAL in the PCA ordination and lower accuracies in LDA differentiating between cores, both in SPT and acetolysed pollen.
+We saw large overlap between MFM and DAL in the PLSR ordination , both in non acetolysed and acetolysed pollen.
 This seems to indicate that pine pollen from different cores are chemically quite similar, even cores that are far apart.
 TSK pollen on the other hand could be clearly separated from the other two cores, in both SPT and acetolysed pollen. 
-In fact acetolysed TSK pollen is even further apart from the rest of the acetolysed pollen than SPT pollen is.
+In fact, acetolysed TSK pollen is even further apart from the rest of the acetolysed pollen than SPT pollen is.
 There seem to be chemical differences between TSK and the other two cores that are present in SPT and acetolysed pollen.
-One reason for the dissimilarity of SPT TSK pollen is the BSi peak at 1630 which is only present in TSK, but even acetolysed pollen from TSK is quite different from the other cores according to PCA and LDA analysis.
+One reason for the dissimilarity of SPT TSK pollen is the BSi peak at 1630 which is only present in TSK. []
+Still, even acetolysed pollen from TSK is quite different from the other cores according to our PLSR model.
 The main difference of TSK pollen to the other cores is that TSK pollen is much younger than the other two cores.
@@
 
+<!--
 .depth patterns
@@
 for acetolysed pollen samples with the youngest and oldest pollen were used, while SPT had all samples.
+-->
 
@@
 Paragraph on practicality
-This is an improvement on previous single grain measurements with hyperion, where each grain had to be targeted individually.
-instead a large area could be measured and spectra selected post-hoc.
-This introduces also the possibilities of automization, such as machinelearning, which have shown to be very adapt at extracting object from larger images.
-the spectra can be seen as images, suitable for image segmentation/object location deep neural networks such as RCNN/YOLO etc.
-Nevertheless this introduces another challenge: concentration of pollen.
-Some of our samples had very low concentration of pollen and would need a concentration step in sample processing.
-Our results do show that acetolysis does not remove any of the sporopollenin peaks, but some of the other parts of the spectra, and could therefore be seen as a standardizing procedure.
+
+Our study is a demonstration of microscope FTIR for pollen chemical methods using fossil pollen.
+We focused on the chemistry of *Pinus* pollen, because *Pinus* is one of a few taxa, that is quite common in sediment cores and largely dominated by 1 species, *Pinus Sylvestris*.
+The pollen is also relatively large, which prevents any scattering problems found with smaller grains.
+We used *Pinus Sylvestris* to explore chemical differences between modern and fossil material, because this has not been fully explored yet.
+
+Here we used a detector in combination with a FTIR microscope that allows capture of several spectra of single grains at the same time.
+This is an improvement on previous single grain measurements, where each grain had to be targeted individually.
+The FPA-MCT detector on our FTIR microscope allows simultaneous capture of a larger area of the sample (XXX x XXX µm), compared to other microscopes (XXX x XXX µm).
+An image with pixel spectra is created, from which spectra of single grains were extracted.
+For this study the spectra were extracted by hand, but this process is possible to automate.
+Approaches using machinelearning and deep learing have shown to be capable of extracting objects from larger images, using deep neural networks specialised on image segmentation/object location [@renFasterRCNNRealTime2015].
+Conceptually the problem of locating pollen on a FTIR image is similar to locating an object on a optical image and these deep learning approaches can be applied.
+@kangSinglecellClassificationFoodborne2020 used deep neural network to classify bacteria based on images and spectral information, while @grosserueschkampSpatialMolecularResolution2017 used random forests for segmentation and identification of malignant tumor in histological thin sections using FTIR microscope equipped with a FPA detector.
+
+We encountered a number of challenges in this study that need to be addressed for the application of IR microscopes in palaeoecology.
+First, we found that the concentration of pollen in our samples was very variable and posed a major bottleneck for our analysis with the FPA detector.
+Capturing spectra from a larger area is only advantageous over single point measurements if several grains are located in the FTIR image.
+This was not always possible, we found, because the pollen was very scarce in some samples.
+
+Acetolysis is the conventional method to extract pollen in high concentration, but as discussed before, we believe there is value in extracting non acetolysed pollen from sediments, because of the substantial differences in pollen chemistry we found between acetolysed and non acetolysed pollen.
+We believe that density separation is a viable alternative [REF], but may have been one possible cause for our low concentration in some samples.
+The reported density of pollen is variable [REFS] and our extraction was done on the lower end of these reported densities (1.44 g cm ^-3^).
+We opted for this density because test extractions using a higher density (1.98 g cm ^-3^), also introduced even more organic material to clutter our samples.
+A refinement of the right extraction medium density in combination with sieving of the samples to remove non pollen organic material may be one approach.
+XXX have reported similar counts of pollen grains to conventional extraction methods using density separation.
+
+Second, the quality of the optical images captured by our microscope was much worse than convential LM, which made it difficult to identify smaller pollen grains on the slide from other organic debris.
+An objective with higher magnification (20x) may improve this aspect of using a FTIR microscope.
+The FPA detector in combination with automated localisation of potential pollen grains based on their chemical signature may help in solving this part of the problem.
+
+Third, as reported in other studies, FTIR measurements of small particles introduces scattering artifacts into the spectra.
+A number of pollen species are affected by this scattering because of their size.
+While sample processing and computational methods exist to minimize or avoid scattering artifacts, there may be substantial species specific information stored in scattering artifacts that could be useful for identification and classification.
+
 Further studies with GCMS methods on fossil pollen and several species have to be performed.

### 2021-06-08 manuscript_v2.Rmd
removed Valeria from author list

--- 2021-06-03 manuscript.Rmd
@@
 ---
-title: 'Chemistry of fossil Pinus pollen in comparison to modern pollen'
+title: 'Assessing variations in the chemistry of fossil- and modern-*Pinus* pollen'
 date: '2020/02/04'
@@
 
-The study of plant micro and macrofossil (spores, seed and pollen) is an integral part of palaeoecology and palaeobotany.
-Pollen records are an important tool to reconstruct past floristic, vegetational and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @noguesbravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013] and are important tools to address ecological processes with timescales that operate on centennial/decadal timescales.
+The study of plant microfossil (pollen and spores) and macrofossil (plant remains such as seeds) is an integral part of palaeoecology and palaeobotany.
+Pollen records are an important tool to reconstruct past floristic, vegetational and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @noguesbravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013] and provide insights into ecological processes which operate on decadal, centennial and millennial timescales (e.g.
+the post-glacial migration of tree taxa into new habitats following the Quaternary glacial maxima [ref]).
 
-The high preservation of pollen grains in the geological record is because of the resistant biopolymer sporopollenin, which is the main building block of the outer pollen grain wall (exine).
-The structure of sporopollenin has been described as similar to lignins, e.g.
-importance of aromatic compounds, cross-linked aliphatic components [@liMolecularStructurePlant2019], but the exact structure remains unknown.
+The high preservation potential of pollen grains in the geological record results from the presence of the resistant biopolymer sporopollenin, which is the main building block of the outer pollen-grain wall (exine).
+Although the exact structure of sporopollenin remains unknown [ref], sporopollenin has been described as being similar to lignins, incorporating both aromatic compounds and cross-linked aliphatic components [@liMolecularStructurePlant2019].
+Variations in the chemical composition of sporopollenin can reflect phylogenetic differences between plant groups [ref- suggest Neirop et al. 2019]. 
+There is also evidence of environmental signals that are recorded in the chemistry of pollen grains which may be used as proxies of past environmental conditions [@bagciogluMultiscaleVibrationalSpectroscopic2015; @kendelChemicalAnalysisPollen2020; @diehnCombiningChemicalInformation2020] (e.g.
+exposure to UV-B radiation [ref]).
+Thus, in recent years, the study of pollen-grain chemistry has gained increasing focus in palaeoecological research.
 
-In recent years, the study of pollen-grain chemistry has shown promising results for biological and palaeoecological research.
-Fourier Transform Infrared (FTIR) spectroscopy has become a popular method  for characterising the chemical variation within pollen and spores.
-A number of FTIR studies have identified pollen chemical differences on genus level, which may extent to species level differences [@zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019; @muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020].
-This is a promising finding for the application of pollen identification with chemical methods in palaeoecological research, because it has the potential to increase taxomonic resolution @dellannaPollenDiscriminationClassification2009; @zimmermannAnalysisAllergenicPollen2016.
+<!-- This is a promising finding for the application of pollen identification with chemical methods in palaeoecological research, because it has the potential to increase taxomonic resolution [@dellannaPollenDiscriminationClassification2009; @zimmermannAnalysisAllergenicPollen2016].
+-->
+<!-- Similarly, FTIR has become a popular method in the quantification of UV-B absorbing compounds to reconstruct changes in UV-B radiation from sediment sequences [ref]. -->
 <!--MOVE TO DISCUSSION: Further, chemical methods are suitable for automation and offer a more objective method than traditional morphological methods [ @mondolApplicationHighThroughputScreening2019; @zimmermannAnalysisAllergenicPollen2016; @guedesPollenRamanSpectra2014].
@@
 
-So far, most FTIR studies focused on identification of pollen have used modern material and bulk measurements, while some have used FTIR microscopes.
-These FTIR microscopes required manual targeting of small areas of the sample slides to record spectra of single grains or groups of pollen grains.
-FTIR as a method has advanced on a technical level over the past 5 to 10 years.
-The development and capabilities of microscopes with connected FTIR spectrometers now allow easier and faster capture of spectra of pollen grains.
-With these newer machines a larger area is irradiated and spectra are recorded at higher resolution over the entire irradiated area.
-One such improvement is the use of a focal plane array detector (FPA).
-A FPA detector allows capture of an image and matrix of spectra over the same sample area instead of a single spectra at a specific point/area.
-This enables e.g. the extraction of spectra from several single grains from the same area on the sample without the need to relocate the detector of the microscope or move the slide.
+Successful application of pollen chemistry techniques within palaeoecological research requires both a good understanding of the factors influencing chemical variations in the modern environment, and in the accurate estimation of the chemical spectra from pollen grains.
+Fourier Transform Infrared (FTIR) spectroscopy has become one of the most popular methods for characterising the chemical variations present within pollen and spores, and one common approach has been to use low-resolution FTIR instrumentation to obtain spectra from large quantities of modern-pollen samples (e.g.
+1-2 mg) [e.g.
+ref].
+These bulk FTIR techniques have been used to identify pollen chemical differences in modern pollen samples at the genus level, and in some cases, chemical variations between pollen types may even extend to species-level differences [@zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019; @muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020].
+<!-- are these references correct?-->
+Bulk FTIR approaches have the advantage of reducing scattering effects from measurements due to the analysis of large sample quantities, but these approaches may be less relevant for fossil sequences since the number of pollen grains analysed vastly exceeds the quantity which is practically available from a sediment sample.
 
-[FM: needs transition or reorg of paragraphs]
+An alternative approach involves the use FTIR microscopes, which require manual targeting of small areas of the sample slides to record spectra of single grains or groups of pollen grains.
+This approach has been successfully applied in various studies to show [...
+some sentence here summarising that it generally works], and may be better suited to sediment samples as a result of the capacity to generate FTIR spectra from individual grains.
+Indeed, microscopic FTIR techniques have now been applied to both modern and fossil-pollen grains preserved in sediments, with the ultimate goal to use observed chemical differences reconstruct changes in UV-B absorbing compounds in different periods across Earth's history [ref].
+However, such approaches may still result in interferences from scattering of FTIR spectra (is correct?), and the targeting of individual pollen grains on the slide can still be time consuming.
 
-There are several extraction methods used to isolate pollen and other microfossils from sediment cores. 
-Traditionally, palaeoecologists use a combination of several chemical extractions to isolate and concentrate pollen from the sediment matrix.
-Most common is acetolysis in addition to HF treatment and HCl [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
-These strong acid treatments aim to remove other organic particles, calcium carbonates and silicates, while preserving the morphological integrity of pollen grains [REF].
-A limited number of studies have used acetolysis and other chemical treatments in combination with FTIR on pollen to simulate fossilization [@julierChemotaxonomyToolInterpreting2016; @jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or study the effects of acid extractions [@dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000;@schulzeosthoffPhenolsIntegratedCompounds1987;@jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
-These results suggest that acetolysis may alter the chemical composition of sporopollenin.
+More recent developments in technologies related to FTIR may for even easier and faster capture of spectra of pollen grains. 
+For example, a focal plane array detector (FPA) allows capture of an image and matrix of spectra over the same sample area instead of a single spectra at a specific point/area [ref].
+This enables e.g.
+the extraction of spectra from several single grains from the same area on the sample without the need to relocate the detector of the microscope or move the slide, and is of potentially high relevance for palaeoecological research.
+When combined with imaging processing technology, it potentially allows for more rapid identification and generation of FTIR spectra of pollen grains from sediment samples.
+However, whilst FTIR spectroscopy using FPA detectors have been applied to pollen samples **in aerobiology [or others, check references]**, there has been limited investigation into their application on fossil-pollen residues taken from sediment cores.
 
-An alternative to chemical extraction is density separation using e.g.
-sodium polytungstate (SPT), which is used to isolate pollen from minerogenic sediments [REFS] or as preparations for pollen isotope studies [REFS].
-Density separation may be a an alternative to acid extractions for pollen studies concerned with chemical alteration of sporomorph chemistry.
-SPT is used to adjust the density of sample suspension and facilitace separation of less dense particles (e.g. pollen and other microfossils) over mineral particles.
-SPT is removed from the sample with several washes after extraction and any contamination of the sample is easily identifiable by characteristic wavebands in the spectra.
+An additional consideration for palaeoecologists aiming to apply FTIR techniques to their sediment records is that any insights gained from studies on modern grains are transferable under fossil settings [ref Jardine et al.
+2015, Jardine et al.
+2021]
+Palaeoecologists use a combination of chemical extraction procedures depending on sediment type in order to isolate and concentrate pollen from the sediment matrix [new pollen preparation method paper].
+One of the most common methods for sporopollenin extraction involves an acetolysis treatment, which in a combination with HF and HCl treatments enables removal of organic materials, calcium carbonates and silicates, while preserving the morphological integrity of pollen grains[@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
+In pollen-chemistry research, acetolysis has often been used in combination with FTIR on pollen as way to standardise between modern and fossil samples, [@julierChemotaxonomyToolInterpreting2016; @jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or to study the effects of acid extractions [@dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000;@schulzeosthoffPhenolsIntegratedCompounds1987;@jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
+Indeed, although acetolysis treatments at low temperatures did not create major chemical differences in Lycopodium spores [Jardine2015], it is uncommon for low temperature treatments to be applied to fossil sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
+Furthermore, a recent paper by Jardine et al. (2021) revealed potential chemically altering effects on fossil pollen composition.
+Whilst there is a developing understanding of the effects of a variety of chemical treatments on fossil- and modern-pollen grains, for many taxa these issues remain unresolved and more studies are needed which compare the chemical spectra of different pollen types.
 
-In this study we explore the use of SPT density separation as an alternative to acetolysis by investigating the chemical alteration of both extraction methods to three fossil cores.
-With this, we aim to investigate changes in fossil pollen chemistry, due to: i.
+In this study we aimed to explore the use an FPA detector to extract FTIR spectra from fossil-pollen *Pinus* grains.
+**Something about why Pinus is importnat here. You can use my JEcol paper for some text inspiration** 
+We aim to develop a workflow which enabled comparison of fossil-pollen grains taken from three locations sediment-core locations (Germany and Norway).
+Each core was sub sampled at different depths to allows us to investigate chemical changes in the pollen with age.
+the pollen was extracted from the sediment cores using two different approaches (density separation, and acetolysis) and the fossil pollen samples measured using the combined FTIR microscope with FPA detector were compared with modern-pollen samples.
+This enables us to investigate changes in fossil-pollen chemistry, due to: i.
 differences between modern and fossil material, ii.
@@
 effect of diagenesis (age of fossil pollen).
-For this purpose we use an FTIR microscope with FPA detector to capture spectra of modern and fossil *Pinus sylvestris* pollen.
-<!--
-The effect of treatment with SPT on pollen chemistry is tested on modern material to establish if any chemical changes occur.
+
+[**here we need Boris to add a few sentences here to give the study more context**]
+
+Chemical treatment...
+<!-- SPT density separation as an alternative to acetolysis by investigating the chemical alteration of both extraction methods to three fossil cores. -->
+<!-- With this, we aim to investigate changes in fossil pollen chemistry, due to: i.
+differences between modern and fossil material, ii.
+extraction method and iii.
+effect of diagenesis (age of fossil pollen).
 -->
-The pollen was collected from three different cores in three locations (Germany and Norway), which allows us to examine potential differences in diagenesis on pollen chemistry at different locations.
-Each core was sub sampled at different depths to allows us to investigate chemical changes in the pollen with age.
-The pollen was extracted from the sediment cores using density separation, with one replicate being acetolysed afterwards.
+<!-- For this purpose we use an FTIR microscope with FPA detector to capture spectra of modern and fossil *Pinus sylvestris* pollen. -->
+<!-- <!-- -->
+<!-- The effect of treatment with SPT on pollen chemistry is tested on modern material to establish if any chemical changes occur. -->
+<!-- --> -->
 
-RQs
-i. Differences between cores
-ii.i Difference between acetolysed and fossil
-ii.ii Difference between fresh and fossil grains.
-iii effect of depth/age on chemistry
+<!-- RQs -->
+<!-- i. Differences between cores -->
+<!-- ii.i Difference between acetolysed and fossil -->
+<!-- ii.ii Difference between fresh and fossil grains. -->
+<!-- iii effect of depth/age on chemistry -->
 
-iv effect of orientation on chemistry
+<!-- iv effect of orientation on chemistry -->
+
+#Alistair rewritten paragraphs- maybe move to methods
+
+<!-- An alternative to chemical extraction is density separation using e.g.
+sodium polytungstate (SPT), which is used to isolate pollen from minerogenic sediments [REFS] or as preparations for pollen isotope studies [REFS].
+-->
+<!-- Density separation may be a an alternative to acid extractions for pollen studies concerned with chemical alteration of sporomorph chemistry. -->
+<!-- SPT is used to adjust the density of sample suspension and facilitace separation of less dense particles (e.g. pollen and other microfossils) over mineral particles. -->
+<!-- SPT is removed from the sample with several washes after extraction and any contamination of the sample is easily identifiable by characteristic wavebands in the spectra.  -->
 
 # Methods and Materials
+The pollen was extracted from three sediment cores: Dalmuttlado in Norway, Tiefer See [@dragerVarveMicrofaciesVarve2017] and Meerfelder Maar [@brauerHighResolutionSediment1999] in Germany.
+The three sediment cores were selected because they contain *Pinus* pollen types, spanning three different periods in the late Quaternary **[it would be useful to include a summary table here, instead of the descriptions you had below]**
 
-The pollen was extracted from three cores: Dalmuttlado in Norway, Tiefer See [@dragerVarveMicrofaciesVarve2017] and Meerfelder Maar [@brauerHighResolutionSediment1999] in Germany.
-Tiefer See formed during the last glacial period, while the Meerfelder Maar record is from the largest maar of the Westeifel Volcaniv field and formed ~35000 BP.
-Tiefer See and Meerfelder Maar are varved cores and were the samples used in this study span two different time-periods.
-The sub samples from Tiefer see are from the last 80 yrs of the core (1933 - 2011), while the samples from Meerfelder Maar cover the timeperiod from the late Allerød to the Younger Dryas/Holocene transition (~12900 - ~11000 BP).
-The samples were extracted using density separation using dissolved Sodium Polytungstate (SPT) [@kampsAgentSeparationDissolved1985] at 1.43 g cm^-3^ as separation medium.
-During density separation a known volume of SPT (density 1.43 g cm^-3^) solution is added to the sample and homogenized/vortexed, followed by centrifugation.
-After centrifugation, particulates/material heavier than 1.43 g cm^-3^ has deposited at the bottom of the vial and the lighter fraction can be decanted and collected.
-In our case the lighter fraction contained most of the pollen.
-There is no agreement on the density of the pollen exine and reported values vary between  <1.4 g cm^-3^ [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996], 1.88 cm^-3^ [@nakagawaDensemediaSeparationMore1998] and 2.2 cm^-3^  [@juvigneDensiteExinesQuelques1973].
+Prior to chemical characterisation using FTIR, all sediment samples were processed in two ways.
+First, the samples were extracted using density separation using dissolved Sodium Polytungstate (SPT) [@kampsAgentSeparationDissolved1985] at 1.43 g cm^-3^ as separation medium.
+During density separation a known volume of SPT (density 1.43 g cm^-3^) solution was added to the sample and homogenized/vortexed, followed by centrifugation.
+After centrifugation, particulates/material heavier than 1.43 g cm^-3^ accumulate of the bottom of the sample vial, whilst the lighter fraction was decanted and collected.
+Note that there is no agreement on the density of the pollen exine, with reported values varyiing between  <1.4 g cm^-3^ [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996], 1.88 cm^-3^ [@nakagawaDensemediaSeparationMore1998] and 2.2 cm^-3^  [@juvigneDensiteExinesQuelques1973].
+In our case the lighter fraction contained majority of the pollen grains.
+Sample residues were then washed three times in using distilled H2O to remove any excess SPT, resulting in sample residues with a higher pollen concentration compared to other organic and inorganic materials on the slide.
+
+We then applied an acetolysis treatment [e.g. @erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989] to one subset of the samples following washing and extracting using SPT.
+Although acetolysis treatments at low temperatures did not create major chemical differences in Lycopodium spores [Jardine2015], it is uncommon for low temperature treatments to be applied to fossil sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
+Under these conditions it is possible that acetolysis may alter the chemical composition of sporopollenin, and so in many cases it may be preferable to use alternative approaches to extract fossil pollen grains before analysis using FTIR.
+Thus, incorporating the acetolysis step on this analysis allows us to compare the effects of pollen grains spectra from the both treated and untreated pollen grains.
+
+## Need to have something on the modern samples
+
+<!-- Put this into a table -->
+<!-- Tiefer See formed during the last glacial period, while the Meerfelder Maar record is from the largest maar of the Westeifel Volcaniv field and formed ~35000 BP. -->
+<!-- Tiefer See and Meerfelder Maar are varved cores and were the samples used in this study span two different time-periods. -->
+<!-- The sub samples from Tiefer see are from the last 80 yrs of the core (1933 - 2011), while the samples from Meerfelder Maar cover the timeperiod from the late Allerød to the Younger Dryas/Holocene transition (~12900 - ~11000 BP).
+-->
 
 ## FTIR measurements
-
 The spectra were measured with a Vertex 70 FTIR spectrometer connected to a Hyperion 3000 IR microscope (Bruker Optik, Ettlingen, Germany) equipped with a focal plane array (FPA) nitrogen cooled mercury cadmium telluride (MCT) detector and a 15x magnification objective.
@@
 The accompanied visible images were obtained by a charge coupled device (CCD) camera coupled to the microscope.
-Single grain pollen grain spectra were extracted from the FTIR images, by averaging all pixel spectra in the image area of the pollen grain. 
+Single-grain pollen grain spectra were extracted from the FTIR images, by averaging all pixel spectra in the image area of the pollen grain. 
 Data acquisition and instrument control was performed using the OPUS 7.5 software (Bruker Optik GmbH, Germany).
@@
 The prominent and characteristic sporopollenin peaks (1600, 1510 and 1170 cm^-1^) are visible in the fossil pollen, while other parts of the spectra show substantial differences.
-Most notibly the lipid peak at 1745 cm^-1^, caused by triglycerides, is not visible in the fossil pollen grains, neither in the SPT treated or acetolysed pollen.
+Most notably the lipid peak at 1745 cm^-1^, caused by triglycerides, is not visible in the fossil pollen grains, neither in the SPT treated or acetolysed pollen.
 There are also several differences in the "fingerprint" region (1500 to 1800 cm^-1^) and carbohydrate region from 1000 to 1250 cm^-1^ between modern and fossil pollen grains.
@@
 .Overall intro to discussion
-Our aim with this study was to use a new type of detector (FPA-MCT) in combination with a FTIR microscope to explore the chemistry of fossil pollen.
-We analysed modern and fossil material and explored the effects of sample processing methods on fossil pollen.
+Our aim with this study was to explore the use of a new type of detector (FPA-MCT) in combination with a FTIR microscope to explore the chemical variations of fossil pollen.
+Our goal was to also to explore the effects of two differemt sample-processing methods on fossil pollen and compare the spectra we obtained with modern samples.
 
 .difference to fresh pollen (FM: have to mention Jardine 21 here)
-The fossil pollen spectra show changes from fresh pollen.
-Lipid and protein peaks have dissappeared and there are substantial changes in the carbohydrate region in fossil pollen.
+One of the clearest differences in the results in our study was that fossil pollen spectra (either treated or untreated) show major differences from the fresh- (i.e. modern pollen samples).
+As expected, in the fossil-pollen samples (both treated and untreated), lipid and protein peaks disappeared and there are substantial changes in the carbohydrate region in fossil pollen.
 From the regression coefficients, we see that the largest changes are in the 1500 to 1750 cm^-1^ region, which have many lipid and protein peaks (eg. 1745 and 1545 cm^-1^ etc).
+**Here you need to add in more comparisons of the other published papers-the two Jardine papers and others which compare them**
+
+Within the fossil pollen samples, there were also differences in spectra identified between acetolysised and non-acetolysised pollen grains. 
 Acetolysed pollen showed two dominant peaks in the carbohydrate region which are not visible in the modern spectra at 1040 and 1170 - 1190 cm^-1^ or the non acetolysed pollen.
-This change in the 1170 - 1190 cm^-1^ region especially is unique to acetolysized pollen and not visible in non acetolysed pollen and had higher regression coefficients that the surrounding wavenumbers (Fig.
+This change in the 1170 - 1190 cm^-1^ region is unique to acetolysized pollen and not visible in non acetolysed pollen and had higher regression coefficients that the surrounding wavenumbers (Fig.
 \@ref(fig:ordinations)).
@@
 In non acetolysed pollen the 1190 peak is not visible in any of the cores, which indicates that it is not due to aging/diageneis, but acetolysis.
-Unfortunately, the non acetolysed pollen showed a wide peak covering part of the carbohydrate region at 1040 cm^-1^.
+
+.spectral interference
+One additional finding we did not expect was that the non acetolysed pollen showed a wide peak covering part of the carbohydrate region at 1040 cm^-1^.
 This peak in the non acetolysed samples is most likely biogenic silica, when compared to this biogenic silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016] and hydrated silica spectra [@SilicaATRIRSpectrum].
@@
 After the droplets had dried we observed a crystalline layer over the same area that covered all particles and pollen (ref Foto in SM).
-From the chemical spectra, we believe that it is dissolved silica that crystallised. 
-the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.
+From the chemical spectra, it is most likely that that it is dissolved silica that crystallised, although the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.
+**I think you need to explain this peak in the context of palaeoecological research a bit- could you talk about the correction factors that might be needed if you were to apply this method everywhere?
+And that different sediment context might create different baseline spectra so understanding sedimentary context is essential?**
 

### 2021-06-08 R/authors_and_affiliations.R

@@ -1,17 +1,17 @@
 dat <- data.frame(FirstName = c("Florian", 
-                                "Valeria", 
+                                #"Valeria", 
                                 "Boris", 
                                 "Alistair W R"),
                   LastName = c("Muthreich", 
-                               "Tafintseva", 
+                               #"Tafintseva", 
                                "Zimmermann", 
                                "Seddon"),
                   Affiliation1 = c("University of Bergen, Department of Biological Sciences, Norway", 
-                                   "University of Life Sciences, Ås, Norway",
+                                   #"University of Life Sciences, Ås, Norway",
                                    "University of Life Sciences, Ås, Norway",
                                    "University of Bergen, Department of Biological Sciences, Norway"),
                   Affiliation2 = c(NA, 
-                                   NA, 
+                                   #NA, 
                                    NA, 
                                    "Bjerknes Centre for Climate Research, Bergen"))
 dat <- dat %>% 

### 2021-06-08 manuscript.Rmd
added Als changes to the manuscript

@@
 <!-- PRE-PROCESS DATA -->
-```{r}
+```{r message=FALSE, warning=FALSE, include=FALSE}
 if (!hasName(DAL, "emsc")) {
@@
 <!-- PCA -->
-```{r}
+```{r message=FALSE, warning=FALSE, include=FALSE}
 

### 2021-06-08 manuscript.Rmd
work on introduciton; removed pollen counts from manuscript, added to SM; addressed AS comments in intro

@@
 
-```{r DALcounts, echo=FALSE, results='asis'}
-count_grains = function(core){
-  counts = core %>%
-    subset(type == "pollen" & species == "Pinus") %>% 
-    group_by(type, depth, treatment) %>%
-    summarise(count=n())
-  return(counts)
-}
-count_grains = function(core){
-  counts = core %>%
-    subset(type == "pollen") %>% 
-    group_by(treatment, depth, species) %>%
-    summarise(count=n())
-  return(counts)
-}
-kable(count_grains(DAL), booktabs = T, format = "latex",
-      caption = "Pinus pollen grain count in Dalmutladdo core.")
-#      digits = 1, format.args = list(scientific = F),
-#      col.names = c("")) %>% 
-#  add_header_above()
-```
-
-```{r TSKcounts, echo=FALSE, results='asis'}
-kable(count_grains(TSK), booktabs = T, format = "latex",
-      caption = "Pinus pollen grain count in Tiefer See core.")
-```
-
-```{r MFMcounts, echo=FALSE, results='asis'}
-kable(count_grains(MFM), booktabs = T, format = "latex",
-      caption = "Pinus pollen grain count in Meerfelder Maar core.")
-```
-
 \newpage
@@
 Successful application of pollen chemistry techniques within palaeoecological research requires both a good understanding of the factors influencing chemical variations in the modern environment, and in the accurate estimation of the chemical spectra from pollen grains.
-Fourier Transform Infrared (FTIR) spectroscopy has become one of the most popular methods for characterising the chemical variations present within pollen and spores, and one common approach has been to use low-resolution FTIR instrumentation to obtain spectra from large quantities of modern-pollen samples (e.g.
+Fourier Transform Infrared (FTIR) spectroscopy has become one of the most popular methods for characterising the chemical variations present within pollen and spores, and one common approach has been to obtain spectra from large quantities of modern-pollen samples (e.g.
 1-2 mg) [e.g.
 ref].
-These bulk FTIR techniques have been used to identify pollen chemical differences in modern pollen samples at the genus level, and in some cases, chemical variations between pollen types may even extend to species-level differences [@zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019; @muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020].
+These bulk FTIR techniques have been used to identify pollen chemical differences in modern pollen samples at the genus level, and in some cases, chemical variations between pollen types may even extend to species-level differences [@muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020].
 <!-- are these references correct?-->
@@
 
-An alternative approach involves the use FTIR microscopes, which require manual targeting of small areas of the sample slides to record spectra of single grains or groups of pollen grains.
-This approach has been successfully applied in various studies to show [...
-some sentence here summarising that it generally works], and may be better suited to sediment samples as a result of the capacity to generate FTIR spectra from individual grains.
-Indeed, microscopic FTIR techniques have now been applied to both modern and fossil-pollen grains preserved in sediments, with the ultimate goal to use observed chemical differences reconstruct changes in UV-B absorbing compounds in different periods across Earth's history [ref].
+An alternative approach involves the use of FTIR microscopes, which require manual targeting of small areas of the sample slides to record spectra of single grains or groups of pollen grains.
+Some microscopes are equipped with automated stages, where multiple locations on the slide can be targeted in sequence to record spectra.
+
+This approach has been successfully applied in various studies to show that chemical differences are found between pollen on genus and species level [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018;  @zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019], and may be better suited to sediment samples as a result of the capacity to generate FTIR spectra from individual grains.
+Indeed, microscopic FTIR techniques have now been applied to both modern and fossil-pollen grains preserved in sediments, with the ultimate goal to use observed chemical differences to reconstruct changes in UV-B absorbing compounds in different periods across Earth's history [@fraserUVBAbsorbingPigments2011; @jardinePollenSporesBiological2016; @jardineProxyReconstructionUltravioletB2019; @jardineSheddingLightSporopollenin2017; @willisQuantificationUVFlux2011].
 However, such approaches may still result in interferences from scattering of FTIR spectra (is correct?), and the targeting of individual pollen grains on the slide can still be time consuming.
 
-More recent developments in technologies related to FTIR may for even easier and faster capture of spectra of pollen grains. 
+More recent developments in technologies related to FTIR may allow for even easier and faster capture of spectra of pollen grains. 
 For example, a focal plane array detector (FPA) allows capture of an image and matrix of spectra over the same sample area instead of a single spectra at a specific point/area [ref].
@@
 
-An additional consideration for palaeoecologists aiming to apply FTIR techniques to their sediment records is that any insights gained from studies on modern grains are transferable under fossil settings [ref Jardine et al.
-2015, Jardine et al.
-2021]
-Palaeoecologists use a combination of chemical extraction procedures depending on sediment type in order to isolate and concentrate pollen from the sediment matrix [new pollen preparation method paper].
+An additional consideration for palaeoecologists aiming to apply FTIR techniques to their sediment records is that any insights gained from studies on modern grains are transferable under fossil settings [@jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
+Palaeoecologists use a combination of chemical extraction procedures depending on sediment type in order to isolate and concentrate pollen from the sediment matrix [@ridingGuidePreparationProtocols2021].
 One of the most common methods for sporopollenin extraction involves an acetolysis treatment, which in a combination with HF and HCl treatments enables removal of organic materials, calcium carbonates and silicates, while preserving the morphological integrity of pollen grains[@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
-In pollen-chemistry research, acetolysis has often been used in combination with FTIR on pollen as way to standardise between modern and fossil samples, [@julierChemotaxonomyToolInterpreting2016; @jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or to study the effects of acid extractions [@dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000;@schulzeosthoffPhenolsIntegratedCompounds1987;@jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
-Indeed, although acetolysis treatments at low temperatures did not create major chemical differences in Lycopodium spores [Jardine2015], it is uncommon for low temperature treatments to be applied to fossil sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
-Furthermore, a recent paper by Jardine et al. (2021) revealed potential chemically altering effects on fossil pollen composition.
+In pollen-chemistry research, acetolysis has often been used in combination with FTIR on pollen as a way to standardise between modern and fossil samples, [@jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or to study the effects of acid extractions [@dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000; @schulzeosthoffPhenolsIntegratedCompounds1987; @jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
+Indeed, although acetolysis treatments at low temperatures did not create major chemical differences in Lycopodium spores [@jardineImpactOxidationSpore2015], it is uncommon for low temperature treatments to be applied to fossil sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
+Furthermore, a recent paper by @jardineSporopolleninChemistryIts2021 revealed potential chemically altering effects on fossil pollen composition.
 Whilst there is a developing understanding of the effects of a variety of chemical treatments on fossil- and modern-pollen grains, for many taxa these issues remain unresolved and more studies are needed which compare the chemical spectra of different pollen types.
 
-In this study we aimed to explore the use an FPA detector to extract FTIR spectra from fossil-pollen *Pinus* grains.
-**Something about why Pinus is importnat here. You can use my JEcol paper for some text inspiration** 
+In this study we aimed to explore the use of an FPA detector to extract FTIR spectra from fossil-pollen *Pinus* grains.
+The development of a UV-proxy based on pollen chemistry is focused on *Pinus* pollen, which makes *Pinus* a interesting genus for our purposes. 
+Identifying chemical differences in fossil *Pinus* pollen may also allow for identification of cryptic *Pinus* species in the fossil record, because *Pinus* is currently mostly identified to type or even genus [@beugLeitfadenPollenbestimmungFur2004; @gieseckePostglacialChangeFloristic2019].
+
 We aim to develop a workflow which enabled comparison of fossil-pollen grains taken from three locations sediment-core locations (Germany and Norway).

### 2021-06-08 readme.Rmd
work on introduciton; removed pollen counts from manuscript, added to SM; addressed AS comments in intro

@@
 # Detailed Methods and Materials
+
+## Pollen Counts
+```{r DALcounts, echo=FALSE, results='asis'}
+count_grains = function(core){
+  counts = core %>%
+    subset(type == "pollen" & species == "Pinus") %>% 
+    group_by(type, depth, treatment) %>%
+    summarise(count=n())
+  return(counts)
+}
+count_grains = function(core){
+  counts = core %>%
+    subset(type == "pollen") %>% 
+    group_by(treatment, depth, species) %>%
+    summarise(count=n())
+  return(counts)
+}
+kable(count_grains(DAL), booktabs = T, format = "latex",
+      caption = "Pinus pollen grain count in Dalmutladdo core.")
+#      digits = 1, format.args = list(scientific = F),
+#      col.names = c("")) %>% 
+#  add_header_above()
+```
+
+```{r TSKcounts, echo=FALSE, results='asis'}
+kable(count_grains(TSK), booktabs = T, format = "latex",
+      caption = "Pinus pollen grain count in Tiefer See core.")
+```
+
+```{r MFMcounts, echo=FALSE, results='asis'}
+kable(count_grains(MFM), booktabs = T, format = "latex",
+      caption = "Pinus pollen grain count in Meerfelder Maar core.")
+```
 

### 2021-06-09 manuscript.Rmd
moved spectra plots to SM; wrote description table

@@
 # Methods and Materials
-The pollen was extracted from three sediment cores: Dalmuttlado in Norway, Tiefer See [@dragerVarveMicrofaciesVarve2017] and Meerfelder Maar [@brauerHighResolutionSediment1999] in Germany.
+The pollen was extracted from three sediment cores: Dalmuttlado in Norway [@bjuneHoloceneVegetationClimate2004], Tiefer See [@theuerkaufEffectsChangesLand2015]  and Meerfelder Maar [@brauerHighResolutionSediment1999] in Germany.
 The three sediment cores were selected because they contain *Pinus* pollen types, spanning three different periods in the late Quaternary **[it would be useful to include a summary table here, instead of the descriptions you had below]**
@@
 ## Need to have something on the modern samples
+
+```{r sites, echo=false, results='asis'}
+sites = data.frame(name = c("Tiefer See",
+                            "Dalmutladdo",
+                            "Meerfelder Maar"),
+                   abbreviation = c("TSK",
+                                    "DAL",
+                                    "MFM"),
+                   age_range = c("1932 - 2015 AD",
+                                 "9836 - 595 BP",
+                                 "13120 - 11400 BP"),
+                   description = c("varved sediment, formed during the last last glacial period, Mecklenburg-Vorpommern, Germany",
+                                   "formed after the last deglaciation, near Tromsø, Norway ",
+                                   "varved sediment, formed in a volcanic crater in the Westeifel, Germany"),
+                   publication = c("[@theuerkaufEffectsChangesLand2015]",
+                                   "[@bjuneHoloceneVegetationClimate2004]",
+                                   "[@brauerHighResolutionSediment1999]"))
+
+kable(sites, booktabs = T, format = "latex",
+      caption = "")
+```
 
@@
 knitr::include_graphics(here("figures", "mean_plots.png"))
-
-mean_plotsSM = readRDS(here("data","output","mean_plots.rds"))
-mean_plots = readRDS(here("data","output","mean_plots_whole.rds"))
 ```
@@
 
-\newpage
-\newpage
-# Mean spectra figures SM
-
-```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
-mean_plotsSM[[1]]
-```
-
-```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL acet"}
-mean_plotsSM[[2]]
-```
-
-```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK SPT"}
-mean_plotsSM[[3]]
-```
-
-```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK acet"}
-mean_plotsSM[[4]]
-```
-
-```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM SPT"}
-mean_plotsSM[[5]]
-```
-
-```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM acet"}
-mean_plotsSM[[6]]
-```
-
-\newpage
-
-```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
-blank_mean_plots = readRDS(here("data","output","blank_mean_plots.rds"))
-blank_mean_plots[[1]]
-```
-
-```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL acet"}
-blank_mean_plots[[2]]
-```
-
-```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK SPT"}
-blank_mean_plots[[3]]
-```
-
-```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK acet"}
-blank_mean_plots[[4]]
-```
-
-```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM SPT"}
-blank_mean_plots[[5]]
-```
-
-```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM acet"}
-blank_mean_plots[[6]]
-```
-
 # References

### 2021-06-09 readme.Rmd
moved spectra plots to SM; wrote description table

@@
       caption = "Pinus pollen grain count in Meerfelder Maar core.")
+```
+
+## Mean spectra
+
+```{r}
+mean_plotsSM = readRDS(here("data","output","mean_plots.rds"))
+mean_plots = readRDS(here("data","output","mean_plots_whole.rds"))
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
+mean_plotsSM[[1]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL acet"}
+mean_plotsSM[[2]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK SPT"}
+mean_plotsSM[[3]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK acet"}
+mean_plotsSM[[4]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM SPT"}
+mean_plotsSM[[5]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM acet"}
+mean_plotsSM[[6]]
+```
+
+\newpage
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL SPT"}
+blank_mean_plots = readRDS(here("data","output","blank_mean_plots.rds"))
+blank_mean_plots[[1]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="DAL acet"}
+blank_mean_plots[[2]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK SPT"}
+blank_mean_plots[[3]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="TSK acet"}
+blank_mean_plots[[4]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM SPT"}
+blank_mean_plots[[5]]
+```
+
+```{r echo = FALSE, message=FALSE, warning=FALSE, out.width = "100%", fig.cap="MFM acet"}
+blank_mean_plots[[6]]
 ```

### 2021-06-09 manuscript.Rmd

@@
 
-```{r sites, echo=false, results='asis'}
+```{r sites, echo=F, results='asis'}
 sites = data.frame(name = c("Tiefer See",
@@
                                    "varved sediment, formed in a volcanic crater in the Westeifel, Germany"),
-                   publication = c("[@theuerkaufEffectsChangesLand2015]",
-                                   "[@bjuneHoloceneVegetationClimate2004]",
-                                   "[@brauerHighResolutionSediment1999]"))
+                   publication = c("(ref:tsk)",
+                                   "(ref:dal)",
+                                   "(ref:mfm)"))
 
-kable(sites, booktabs = T, format = "latex",
-      caption = "")
+knitr::kable(sites, booktabs = T, format = "latex", 
+             caption = "",
+             col.names = c("Name", "Abbr.", "Age", "Description", "Publ.")) %>%
+  column_spec(4, width = "5cm")
 ```
+
+(ref:tsk) @theuerkaufEffectsChangesLand2015
+(ref:dal) @bjuneHoloceneVegetationClimate2004
+(ref:mfm) @brauerHighResolutionSediment1999
 

### 2021-06-09 manuscript.Rmd
merging results and discussion; before disc parts integration; removed old pca/lda text

@@
 
-<!-- Put this into a table -->
-<!-- Tiefer See formed during the last glacial period, while the Meerfelder Maar record is from the largest maar of the Westeifel Volcaniv field and formed ~35000 BP. -->
-<!-- Tiefer See and Meerfelder Maar are varved cores and were the samples used in this study span two different time-periods. -->
-<!-- The sub samples from Tiefer see are from the last 80 yrs of the core (1933 - 2011), while the samples from Meerfelder Maar cover the timeperiod from the late Allerød to the Younger Dryas/Holocene transition (~12900 - ~11000 BP).
--->
-
 ## FTIR measurements
@@
 
-```{r message=FALSE, warning=FALSE, include=FALSE}
-pca_acet_select = readRDS(here("data","output","pca_plots.rds"))
-lda_tables = readRDS(here("data","output","lda_tables.rds"))
-```
+# Results and Discussion
 
-# Results
-
-```{r meanspectra, echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="100%"}
+```{r meanspectra, echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="90%"}
 knitr::include_graphics(here("figures", "mean_plots.png"))
@@
 
-There are differences between modern and fossil grains (Fig. \@ref(fig:meanspectra)).
-The prominent and characteristic sporopollenin peaks (1600, 1510 and 1170 cm^-1^) are visible in the fossil pollen, while other parts of the spectra show substantial differences.
-Most notably the lipid peak at 1745 cm^-1^, caused by triglycerides, is not visible in the fossil pollen grains, neither in the SPT treated or acetolysed pollen.
-There are also several differences in the "fingerprint" region (1500 to 1800 cm^-1^) and carbohydrate region from 1000 to 1250 cm^-1^ between modern and fossil pollen grains.
-
-After visual inspection of the spectra, the chemical differences between modern and fossil pollen were explored with a PLSR model.
-The ordination showed that the fossil pollen are clearly separated from the modern pollen (Fig \@ref(fig:ordinations) a).
-The regression coefficients reflect our observations from the visual inspection and show high importance of the 1745 lipid peak, the 1500 - 1800 region and the 1170 peak (Fig \@ref(fig:ordinations) b and c).
-
-<!-- 
-Tiefer See pollen is separated from the other two cores, which cluster closer together.
-This is the same for acetolysed and SPT pollen. 
-A PCA-LDA model to differentiate fossil pollen into the three cores achieved ~60% accuracy to classify MFM and DAL pollen and 85% accuracy to identify TSK pollen (REF TO SM).
-
-A peak at 1030 and ~1200 cm^-1^ is visible in the acetolysed pollen but not in the modern spectra or SPT pollen.
--->
-
-We see a broad peak at 1090-1100 cm^-1^ in the non acetolysed spectra of all three cores (Fig. \@ref(fig:meanspectra) b, c and d).
-This peak is not present in either modern pollen or acetolysed pollen, additionally Tiefer See shows a similar peak at 1630 cm^-1^.
-
-The 1090 cm^-1^ peak dissappears after acetolysis from all spectra as does the 1630 cm^-1^ peak in Tiefer See. 
-The region 1000 - 1200 cm^-1^ is now dominated in acetolysed fossil pollen by two peaks and its shoulders at 1040 and 1170 cm^-1^.
-This 1040 cm^-1^ has high importance for the PLSR model differentiating acetolysed and non acetolysed pollen (Fig. \@ref(fig:ordinations) e and f).
-Using this PLSR model the acetolysed pollen is clearly separated from the non acetolysed pollen, with a much higher distance between samples for acetolysed pollen compared to non acetolysed pollen (Fig.
-\@ref(fig:ordinations) d).
-In addition, there is some differentiation between the cores, samples from Tiefer See are more separate from samples of the other two cores, which show more similar scores in the ordination.
-
-<!--
-LDA to classify the pollen into depths showed higher accuracy for acetolysed pollen than SPT pollen (Table \@ref(tab:ldadepth)).
-For the SPT pollen the accuracy increased when the depths were restricted to the same depths as the acetolysed samples.
-Generally lower number of depths to classify increased accuracy.
-sporopollenin peaks are similar between depths, changes in the other parts of the spectra are driving separation.
--->
+\newpage
 
@@
 \newpage
+
+## Fresh vs fossil pollen
+
+There are differences between modern and fossil grains (Fig. \@ref(fig:meanspectra)).
+The prominent and characteristic sporopollenin peaks (1600, 1510 and 1170 cm^-1^) are visible in the fossil pollen, while other parts of the spectra show substantial differences.
+Most notably the lipid peak at 1745 cm^-1^, caused by triglycerides, is not visible in the fossil pollen grains, neither in the SPT treated or acetolysed pollen.
+There are also several differences in the "fingerprint" region (1500 to 1800 cm^-1^) and carbohydrate region from 1000 to 1250 cm^-1^ between modern and fossil pollen grains.
+
+After visual inspection of the spectra, the chemical differences between modern and fossil pollen were explored with a PLSR model.
+The ordination showed that the fossil pollen are clearly separated from the modern pollen (Fig \@ref(fig:ordinations) a).
+The regression coefficients reflect our observations from the visual inspection and show high importance of the 1745 lipid peak, the 1500 - 1800 region and the 1170 peak (Fig \@ref(fig:ordinations) b and c).
+
+## Differences between fossil pollen
+Using this PLSR model the acetolysed pollen is clearly separated from the non acetolysed pollen, with a much higher distance between samples for acetolysed pollen compared to non acetolysed pollen (Fig.
+\@ref(fig:ordinations) d).
+In addition, there is some differentiation between the cores, samples from Tiefer See are more separate from samples of the other two cores, which show more similar scores in the ordination.
+
+## Anomalies in fossil pollen spectra
+
+We see a broad peak at 1090-1100 cm^-1^ in the non acetolysed spectra of all three cores (Fig. \@ref(fig:meanspectra) b, c and d).
+This peak is not present in either modern pollen or acetolysed pollen, additionally Tiefer See shows a similar peak at 1630 cm^-1^.
+
+The 1090 cm^-1^ peak dissappears after acetolysis from all spectra as does the 1630 cm^-1^ peak in Tiefer See. 
+The region 1000 - 1200 cm^-1^ is now dominated in acetolysed fossil pollen by two peaks and its shoulders at 1040 and 1170 cm^-1^.
+This 1040 cm^-1^ has high importance for the PLSR model differentiating acetolysed and non acetolysed pollen (Fig. \@ref(fig:ordinations) e and f).
+
+## Application for Palaeoecology
+
+# Conclusions
 

### 2021-06-09 manuscript.Rmd
rearranged R/D; added Jardine 2021 comparison

@@
 
+Our aim with this study was to explore the use of a new type of detector (FPA-MCT) in combination with a FTIR microscope to explore the chemical variations of fossil pollen.
+Our goal was also to explore the effects of two differemt sample-processing methods on fossil pollen and compare the spectra we obtained with modern samples.
+
 ```{r meanspectra, echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="90%"}
@@
 The ordination showed that the fossil pollen are clearly separated from the modern pollen (Fig \@ref(fig:ordinations) a).
-The regression coefficients reflect our observations from the visual inspection and show high importance of the 1745 lipid peak, the 1500 - 1800 region and the 1170 peak (Fig \@ref(fig:ordinations) b and c).
+The regression coefficients reflect our observations from the visual inspection and show that lipid  and protein peaks (1745 and 1500 - 1800 cm^-1^) have disappeared and there are substantial changes in the carbohydrate region (1000 to 1250 cm^-1^) in fossil pollen (Fig \@ref(fig:ordinations) b and c).
+
+ADD COMPARISON TO JARDINE PAPERS and other (e.g. Woutersen)
+Comparisons to other studies that looked at acetolysed and fossil samples show similarities between our acetolysed *Pinus* pollen and other taxa.
+E.g. the carb region 1000 - 1200 in our Pine is most similar to acetolysed extant poaceae set of @jardineSporopolleninChemistryIts2021, two large peaks at 1040 and 1190/1170.
+
+The fossil poaceae data set (Amazon Fan and Western Amazon) from @jardineSporopolleninChemistryIts2021 on the other looks quite different from our fossil pollen and their extant acetolysed set, with a large peak at 1600 and 1190 peak not visible.
+These fossil poaceae set was treated differently than ours and jardines extant poaceae set (combination of heavy liquid separation, Na~4~P~2~O~7~, lignite).
+
+@jardineImpactOxidationSpore2015 showed limited effect on pollen chemistry of cold acetolysis if it was shorter than 10 minutes.
+Hot acetolysis (90), which is standard procedure in palynology, had an effect after the first minute. 
+The simililarities of the acetolysed extant set to our acetolysed fossil set and dissimilarities to pollen treated with a different combination of extractions (Amazon Western and Fan), suggest that chemical treatment of the pollen has a large effect on the chemistry of pollen, both modern and fossil origin.
+This is important to consider going forward for the application of IR methods in palaeoecology, where extraction methods are often dependent on sediment type.
+The effects of these extraction methods need to be explored on modern and fossil material side by side to see if there are links to bridge the transition from modern to fossil pollen.
 
 ## Differences between fossil pollen
-Using this PLSR model the acetolysed pollen is clearly separated from the non acetolysed pollen, with a much higher distance between samples for acetolysed pollen compared to non acetolysed pollen (Fig.
-\@ref(fig:ordinations) d).
-In addition, there is some differentiation between the cores, samples from Tiefer See are more separate from samples of the other two cores, which show more similar scores in the ordination.
 
-## Anomalies in fossil pollen spectra
+### Anomalies in non acetolysed pollen spectra
 
@@
 This 1040 cm^-1^ has high importance for the PLSR model differentiating acetolysed and non acetolysed pollen (Fig. \@ref(fig:ordinations) e and f).
+This peak in the non acetolysed samples is most likely biogenic silica, when compared to this biogenic silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016] and hydrated silica spectra [@SilicaATRIRSpectrum].
+During the measurements with the microscope, we applied our extracted pollen as droplets onto ZnSe slides and waited for the water to evaporate.
+After the droplets had dried we observed a crystalline layer over the same area that covered all particles and pollen (ref Foto in SM).
+From the chemical spectra, it is most likely that that it is dissolved silica that crystallised, although the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.
 
-## Application for Palaeoecology
+[FM: would probably have to replicate this with BSi in different concentrations and maybe see if thickness of this BSi crystall has an effect on the spectra]
+**I think you need to explain this peak in the context of palaeoecological research a bit- could you talk about the correction factors that might be needed if you were to apply this method everywhere?
+And that different sediment context might create different baseline spectra so understanding sedimentary context is essential?**
 
-# Conclusions
+### pollen chemical differences
 
-# Discussion
-
-.Overall intro to discussion
-Our aim with this study was to explore the use of a new type of detector (FPA-MCT) in combination with a FTIR microscope to explore the chemical variations of fossil pollen.
-Our goal was to also to explore the effects of two differemt sample-processing methods on fossil pollen and compare the spectra we obtained with modern samples.
-
-.difference to fresh pollen (FM: have to mention Jardine 21 here)
-One of the clearest differences in the results in our study was that fossil pollen spectra (either treated or untreated) show major differences from the fresh- (i.e. modern pollen samples).
-As expected, in the fossil-pollen samples (both treated and untreated), lipid and protein peaks disappeared and there are substantial changes in the carbohydrate region in fossil pollen.
-From the regression coefficients, we see that the largest changes are in the 1500 to 1750 cm^-1^ region, which have many lipid and protein peaks (eg. 1745 and 1545 cm^-1^ etc).
-**Here you need to add in more comparisons of the other published papers-the two Jardine papers and others which compare them**
-
-Within the fossil pollen samples, there were also differences in spectra identified between acetolysised and non-acetolysised pollen grains. 
 Acetolysed pollen showed two dominant peaks in the carbohydrate region which are not visible in the modern spectra at 1040 and 1170 - 1190 cm^-1^ or the non acetolysed pollen.
@@
 \@ref(fig:ordinations)).
+
 The presence of this new peak in the shoulder of the 1170 cm^-1^ peak, which can indicate sporopollenin and carbohydrates, suggest changes in the pollen chemistry.
@@
 
-.spectral interference
-One additional finding we did not expect was that the non acetolysed pollen showed a wide peak covering part of the carbohydrate region at 1040 cm^-1^.
-This peak in the non acetolysed samples is most likely biogenic silica, when compared to this biogenic silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016] and hydrated silica spectra [@SilicaATRIRSpectrum].
-During the measurements with the microscope, we applied our extracted pollen as droplets onto ZnSe slides and waited for the water to evaporate.
-After the droplets had dried we observed a crystalline layer over the same area that covered all particles and pollen (ref Foto in SM).
-From the chemical spectra, it is most likely that that it is dissolved silica that crystallised, although the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.
-**I think you need to explain this peak in the context of palaeoecological research a bit- could you talk about the correction factors that might be needed if you were to apply this method everywhere?
-And that different sediment context might create different baseline spectra so understanding sedimentary context is essential?**
+Using the second PLSR model the acetolysed pollen is clearly separated from the non acetolysed pollen, with a much higher distance between samples for acetolysed pollen compared to non acetolysed pollen (Fig.
+\@ref(fig:ordinations) d).
+In addition, there is some differentiation between the cores, samples from Tiefer See are more separate from samples of the other two cores, which show more similar scores in the ordination.
 
-the characteristic sporopollenin peaks are all present, but since so much of the rest of the spectra changes, we cant clearly associate any relative changes in these peaks to changes in composition of the sporopollenin or simply changes that only affect other components of the pollen (e.g.
-carbohydrates).
+This seems to indicate that *Pinus* pollen from DAL and MFM are chemically quite similar, while TSK *Pinus* pollen is different.
+In fact, acetolysed TSK pollen is even further apart from the rest of the acetolysed pollen than non acetolysed pollen is.
+There seem to be chemical differences between TSK and the other two cores that are present in non acetolysed and acetolysed pollen.
+The main difference of TSK pollen to the other cores is that TSK pollen is much younger than the other two cores.
+Our TSK samples are from the past 80 years, while MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene, with the youngest sample several 100 years older than TSK pollen.
+We cannot see any evidence that sediment type has an effect on the chemistry of the pollen, both TSK and MFM cores are varved cores, while DAL is not.
+This would suggest more difference between varved and non varved pollen if different types of diagenesis had an effect. 
+Any difference may be overshadowed by the large difference in age between the samples from these cores. 
+Our findings suggest a chemical difference in fossil pollen grains that is persistent even after acetolysis, which may be explained by the age of the pollen.
+Further studies on pollen from different sediment types at fixed age/depths may provide a clearer answer on the effects of sediment type and age on pollen chemistry.
 
-We saw large overlap between MFM and DAL in the PLSR ordination , both in non acetolysed and acetolysed pollen.
-This seems to indicate that pine pollen from different cores are chemically quite similar, even cores that are far apart.
-TSK pollen on the other hand could be clearly separated from the other two cores, in both SPT and acetolysed pollen. 
-In fact, acetolysed TSK pollen is even further apart from the rest of the acetolysed pollen than SPT pollen is.
-There seem to be chemical differences between TSK and the other two cores that are present in SPT and acetolysed pollen.
-One reason for the dissimilarity of SPT TSK pollen is the BSi peak at 1630 which is only present in TSK. []
-Still, even acetolysed pollen from TSK is quite different from the other cores according to our PLSR model.
-The main difference of TSK pollen to the other cores is that TSK pollen is much younger than the other two cores.
-Our TSK samples are from the past 80 years, while MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene.
-The other difference is that TSK is a varved core, while MFM and DAL are not.
-Our findings suggest a chemical difference in fossil pollen grains, either caused by diagenesis (varved vs non-varved), age or combination of the two that is persistent even after acetolysis.
-
-<!--
-.depth patterns
-no clear patterns with depths chemically. the more depths are added, the more inaccurate the LDA became.
-for acetolysed pollen samples with the youngest and oldest pollen were used, while SPT had all samples.
--->
-
-.application for palaeoecology
-Paragraph on practicality
+## Application for Palaeoecology
 
@@
 
-Further studies with GCMS methods on fossil pollen and several species have to be performed.
-
-.conclusions
+# Conclusions
 

### 2021-06-10 R/plotting.R
mean plots in one graph and correct order

@@ -10,26 +10,44 @@ plot_mean_spectra = function(data, subset_sel, spec_id, id.vars = c("ids", "trea
   # wavenumbers to numeric  
   pdata$variable = as.numeric(as.character(pdata$variable))
   
+  # reorder
+  pdata %>%
+    mutate(group = factor(group, levels = c("DAL_SPT","DAL_acet",
+                                            "MFM_SPT","MFM_acet",
+                                            "TSK_SPT","TSK_acet",
+                                            "FRE_Bergen","FRE_Innsbruck"))) %>%
+    arrange(group, variable)
+  
   #offset
-  pdata = offset_spectra(pdata, group, "value", seq(0, by=max(pdata$value)/3, length.out = length(unique(pdata[,group]))))
+  pdata = offset_spectra(pdata, group, "value", seq(0, by=1, length.out = length(levels(pdata[,group])))) #max(pdata$value)/2
   
   # mean data over groups
   pdata_mean = pdata %>%
     group_by_at(c(group, "variable")) %>%
     summarise(mean = mean(value))
   
+  # reorder
+  pdata_mean %>%
+    mutate(group = factor(group, levels = c(
+                                            "DAL_SPT","DAL_acet",
+                                            "MFM_SPT","MFM_acet",
+                                            "TSK_SPT","TSK_acet",
+                                            "FRE_Bergen","FRE_Innsbruck"))) %>%
+    arrange(group)
+  
   g = ggplot(pdata, aes_string("variable", "value", color = group))+
     geom_line(aes(group = ids), alpha = .15) + #mean spectra
-    geom_line(data = pdata_mean, aes_string("variable", "mean", group = group), inherit.aes = F, size = 1) + #all spectra
+    #geom_line(data = pdata_mean, aes_string("variable", "mean", group = group), inherit.aes = F, size = 1) + #all spectra
     xlim(1800,900) +
-    theme_bw() + theme(legend.position = "bottom")  
+    theme_bw() + theme(legend.position = "bottom") +
+    guides(colour = guide_legend(override.aes = list(alpha = 1)))
   
   return(g)
 }  
 
 offset_spectra = function(data, variable, value, offset) {
-  for (i in 1:length(unique(data[, variable]))) {
[…]

### 2021-06-10 R/plotting.R
removed ordering from mean plot function

@@ -10,14 +10,6 @@ plot_mean_spectra = function(data, subset_sel, spec_id, id.vars = c("ids", "trea
   # wavenumbers to numeric  
   pdata$variable = as.numeric(as.character(pdata$variable))
   
-  # reorder
-  pdata %>%
-    mutate(group = factor(group, levels = c("DAL_SPT","DAL_acet",
-                                            "MFM_SPT","MFM_acet",
-                                            "TSK_SPT","TSK_acet",
-                                            "FRE_Bergen","FRE_Innsbruck"))) %>%
-    arrange(group, variable)
-  
   #offset
   pdata = offset_spectra(pdata, group, "value", seq(0, by=1, length.out = length(levels(pdata[,group])))) #max(pdata$value)/2
   
@@ -26,14 +18,6 @@ plot_mean_spectra = function(data, subset_sel, spec_id, id.vars = c("ids", "trea
     group_by_at(c(group, "variable")) %>%
     summarise(mean = mean(value))
   
-  # reorder
-  pdata_mean %>%
-    mutate(group = factor(group, levels = c(
-                                            "DAL_SPT","DAL_acet",
-                                            "MFM_SPT","MFM_acet",
-                                            "TSK_SPT","TSK_acet",
-                                            "FRE_Bergen","FRE_Innsbruck"))) %>%
-    arrange(group)
   
   g = ggplot(pdata, aes_string("variable", "value", color = group))+
     geom_line(aes(group = ids), alpha = .15) + #mean spectra

### 2021-06-10 R/plotting.R
readded mean lines

@@ -11,7 +11,7 @@ plot_mean_spectra = function(data, subset_sel, spec_id, id.vars = c("ids", "trea
   pdata$variable = as.numeric(as.character(pdata$variable))
   
   #offset
-  pdata = offset_spectra(pdata, group, "value", seq(0, by=1, length.out = length(levels(pdata[,group])))) #max(pdata$value)/2
+  pdata = offset_spectra(pdata, group, "value", seq(0, by=max(pdata$value)/5, length.out = length(levels(pdata[,group])))) #
   
   # mean data over groups
   pdata_mean = pdata %>%
@@ -21,7 +21,7 @@ plot_mean_spectra = function(data, subset_sel, spec_id, id.vars = c("ids", "trea
   
   g = ggplot(pdata, aes_string("variable", "value", color = group))+
     geom_line(aes(group = ids), alpha = .15) + #mean spectra
-    #geom_line(data = pdata_mean, aes_string("variable", "mean", group = group), inherit.aes = F, size = 1) + #all spectra
+    geom_line(data = pdata_mean, aes_string("variable", "mean", group = group), inherit.aes = F, size = 1) + #all spectra
     xlim(1800,900) +
     theme_bw() + theme(legend.position = "bottom") +
     guides(colour = guide_legend(override.aes = list(alpha = 1)))

### 2021-06-10 R/plotting.R
fixed axis titles and breaks

@@ -22,8 +22,12 @@ plot_mean_spectra = function(data, subset_sel, spec_id, id.vars = c("ids", "trea
   g = ggplot(pdata, aes_string("variable", "value", color = group))+
     geom_line(aes(group = ids), alpha = .15) + #mean spectra
     geom_line(data = pdata_mean, aes_string("variable", "mean", group = group), inherit.aes = F, size = 1) + #all spectra
-    xlim(1800,900) +
-    theme_bw() + theme(legend.position = "bottom") +
+    scale_x_reverse(name = "wavenumbers",limits = c(1800,900), breaks = seq(1800,900,-100)) +
+    scale_y_continuous(name = "Absorbance (arbitrary units)",breaks = NULL) +
+    theme_bw() + 
+    theme(legend.position = "bottom",
+          axis.ticks.y = element_blank(),
+          axis.text.y = element_blank()) +
     guides(colour = guide_legend(override.aes = list(alpha = 1)))
   
   return(g)

### 2021-06-10 R/plotting.R
first version with colors

@@ -24,6 +24,15 @@ plot_mean_spectra = function(data, subset_sel, spec_id, id.vars = c("ids", "trea
     geom_line(data = pdata_mean, aes_string("variable", "mean", group = group), inherit.aes = F, size = 1) + #all spectra
     scale_x_reverse(name = "wavenumbers",limits = c(1800,900), breaks = seq(1800,900,-100)) +
     scale_y_continuous(name = "Absorbance (arbitrary units)",breaks = NULL) +
+    scale_color_manual(name = NULL,
+                       values = c("#e66101","#ff6c01",
+                                  "#ac7d43","#fdb863",
+                                  "#b2abd2","#d8d0ff",
+                                  "#5e3c99","#9d64ff"),
+                       labels = c("DAL non acet.","DAl acet.",
+                                  "MFM non acet.","MFM acet.",
+                                  "TSK non acet.","TSK acet.",
+                                  "Bergen","Innsbruck")) +
     theme_bw() + 
     theme(legend.position = "bottom",
           axis.ticks.y = element_blank(),

### 2021-06-10 R/plotting.R
added save command

@@ -68,6 +68,7 @@ data = rbind(
   MFM,
   TSK
 )
+
 data= data%>%
   mutate(off = factor(off, levels = c(
     "DAL_SPT","DAL_acet",
@@ -77,14 +78,9 @@ data= data%>%
   arrange(off)
 colnames(data$emsc) = colnames(FRE$emsc)
 
-plot_mean_spectra(data, expression(species == "Pinus"), spec_id = "emsc", group = "off", id.vars = c("ids", "treatment", "Label", "depth", "age", "type", "off"))
+l = plot_mean_spectra(data, expression(species == "Pinus"), spec_id = "emsc", group = "off", id.vars = c("ids", "treatment", "Label", "depth", "age", "type", "off"))
 
-l[["DAL"]] = plot_mean_spectra(DAL, expression(species == "Pinus"), spec_id = "emsc", group = "treatment")
-
-l[["MFM"]] = plot_mean_spectra(MFM, expression(species == "Pinus"), spec_id = "emsc", group = "treatment")
-
-l[["TSK"]] = plot_mean_spectra(TSK, expression(species == "Pinus"), spec_id = "emsc", group = "treatment")
-
+ggsave(here("figures","mean_plots.png"), l, width = 17, height = 25,units = "cm")
 saveRDS(l, here("data","output","mean_plots_whole.rds"))
 
 ### creating Plots

### 2021-06-10 manuscript.Rmd
wrote stat analysis section

@@
 The pollen was extracted from three sediment cores: Dalmuttlado in Norway [@bjuneHoloceneVegetationClimate2004], Tiefer See [@theuerkaufEffectsChangesLand2015]  and Meerfelder Maar [@brauerHighResolutionSediment1999] in Germany.
-The three sediment cores were selected because they contain *Pinus* pollen types, spanning three different periods in the late Quaternary **[it would be useful to include a summary table here, instead of the descriptions you had below]**
+The three sediment cores were selected because they contain *Pinus* pollen types, spanning three different periods in the late Quaternary (Table \@ref(tab:sites)).
 
@@
 We then applied an acetolysis treatment [e.g. @erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989] to one subset of the samples following washing and extracting using SPT.
-Although acetolysis treatments at low temperatures did not create major chemical differences in Lycopodium spores [Jardine2015], it is uncommon for low temperature treatments to be applied to fossil sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
+Although acetolysis treatments at low temperatures did not create major chemical differences in Lycopodium spores [@jardineImpactOxidationSpore2015], it is uncommon for low temperature treatments to be applied to fossil sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
 Under these conditions it is possible that acetolysis may alter the chemical composition of sporopollenin, and so in many cases it may be preferable to use alternative approaches to extract fossil pollen grains before analysis using FTIR.
-Thus, incorporating the acetolysis step on this analysis allows us to compare the effects of pollen grains spectra from the both treated and untreated pollen grains.
+Thus, incorporating the acetolysis step on this analysis allows us to compare the effects of pollen grains spectra from both treated and untreated pollen grains.
 
@@
 
-Preprocessing:
-- SG 2nd degree, 2 poly, windowsize 5 or 6
-- EMSC linear (probably needs to involve background spectra to filter them out)
+The spectra were preprocessed before analysis consisted of smoothing and calculation of second derivative and normalization.
+We used the Savitzky Golay (SG) algorithm [@savitzkySmoothingDifferentiationData1964] as implemented by the extended multiplicative signal correction (EMSC) package in R [@lilandPackageEMSC2017] and used second degree polynomial and windowsize five as parameters.
+The second derivative spectra was then normalized using a EMSC model with a linear component [@lilandPackageEMSC2017].
+Preprocessing with SG and EMSC has been shown to improve performance of multivariate analysis, such as principal component analysis (PCA) or models in the partial least squares regression (PLSR) family [@woutersenNovelApproachStudy2018; @zimmermannOptimizingSavitzkyGolayParameters2013].
 
-PLSR second deriv on treatment (fresh vs fossil)
-
-PLS regression of spectra on depth
-
-repeat on different grain averages (1, 3, 5, 10)
-
-average all spectra and stack plots of different depth
+We used two PLSR models to explore the chemical differences between modern and fossil grains (model 1), and acetolyzed and non acetolyzed fossil grains (model 2).
+For these models the explanatory variables were the preprocessed spectra and the response variables were hot-one encoded response variables (Model 1: modern/fossil; Model 2: acet./ non acet.).
+By using PLSR we can maximize the distance between responses and identify chemical differences between the responses by looking at the regression coefficients.
 
@@
 @jardineImpactOxidationSpore2015 showed limited effect on pollen chemistry of cold acetolysis if it was shorter than 10 minutes.
-Hot acetolysis (90), which is standard procedure in palynology, had an effect after the first minute. 
+Hot acetolysis (90&deg;C), which is standard procedure in palynology, had an effect from the first minute. 
 The simililarities of the acetolysed extant set to our acetolysed fossil set and dissimilarities to pollen treated with a different combination of extractions (Amazon Western and Fan), suggest that chemical treatment of the pollen has a large effect on the chemistry of pollen, both modern and fossil origin.

### 2021-06-10 manuscript.Rmd
after readthrough and minor changes; added refs; fixed typos etc

@@
 Pollen records are an important tool to reconstruct past floristic, vegetational and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @noguesbravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013] and provide insights into ecological processes which operate on decadal, centennial and millennial timescales (e.g.
-the post-glacial migration of tree taxa into new habitats following the Quaternary glacial maxima [ref]).
+the post-glacial migration of tree taxa into new habitats following the Quaternary glacial maxima @gieseckeMovingFrontPopulation2005, @huntleyQuaternaryPalaeoecologyEcology1996, @tinnerCentralEuropeanVegetation2001).
 
@@
 Although the exact structure of sporopollenin remains unknown [ref], sporopollenin has been described as being similar to lignins, incorporating both aromatic compounds and cross-linked aliphatic components [@liMolecularStructurePlant2019].
-Variations in the chemical composition of sporopollenin can reflect phylogenetic differences between plant groups [ref- suggest Neirop et al. 2019]. 
-There is also evidence of environmental signals that are recorded in the chemistry of pollen grains which may be used as proxies of past environmental conditions [@bagciogluMultiscaleVibrationalSpectroscopic2015; @kendelChemicalAnalysisPollen2020; @diehnCombiningChemicalInformation2020] (e.g.
-exposure to UV-B radiation [ref]).
+Variations in the chemical composition of sporopollenin can reflect phylogenetic differences between plant groups [@nieropQuantitativeAnalysisDiverse2019]. 
+There is also evidence of environmental signals that are recorded in the chemistry of pollen grains which may be used as proxies of past environmental conditions, e.g.
+temperature [@bagciogluMultiscaleVibrationalSpectroscopic2015], nutrients [@kendelChemicalAnalysisPollen2020; @diehnCombiningChemicalInformation2020], UV-B radiation [@rozemaUVBAbsorbanceUVB2001; @blokkerDevelopmentProxySurface2005; @fraserUVBAbsorbingPigments2011; @jardineSheddingLightSporopollenin2017].
 Thus, in recent years, the study of pollen-grain chemistry has gained increasing focus in palaeoecological research.
@@
 Fourier Transform Infrared (FTIR) spectroscopy has become one of the most popular methods for characterising the chemical variations present within pollen and spores, and one common approach has been to obtain spectra from large quantities of modern-pollen samples (e.g.
-1-2 mg) [e.g.
-ref].
+1-2 mg) (e.g.
+@zimmermannCharacterizationPollenVibrational2010, @lahlaliATRFTIRSpectroscopyReveals2014).
 These bulk FTIR techniques have been used to identify pollen chemical differences in modern pollen samples at the genus level, and in some cases, chemical variations between pollen types may even extend to species-level differences [@muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020].
-<!-- are these references correct?-->
 Bulk FTIR approaches have the advantage of reducing scattering effects from measurements due to the analysis of large sample quantities, but these approaches may be less relevant for fossil sequences since the number of pollen grains analysed vastly exceeds the quantity which is practically available from a sediment sample.
@@
 The development of a UV-proxy based on pollen chemistry is focused on *Pinus* pollen, which makes *Pinus* a interesting genus for our purposes. 
-Identifying chemical differences in fossil *Pinus* pollen may also allow for identification of cryptic *Pinus* species in the fossil record, because *Pinus* is currently mostly identified to type or even genus [@beugLeitfadenPollenbestimmungFur2004; @gieseckePostglacialChangeFloristic2019].
+Identifying chemical differences in fossil *Pinus* pollen may also allow for identification of cryptic *Pinus* species in the fossil record, because *Pinus* is currently mostly identified to type or even only genus [@beugLeitfadenPollenbestimmungFur2004; @gieseckePostglacialChangeFloristic2019].
 
-We aim to develop a workflow which enabled comparison of fossil-pollen grains taken from three locations sediment-core locations (Germany and Norway).
-Each core was sub sampled at different depths to allows us to investigate chemical changes in the pollen with age.
+We aim to develop a workflow which enabled comparison of fossil-pollen grains taken from three sediment-core locations (Germany and Norway).
 the pollen was extracted from the sediment cores using two different approaches (density separation, and acetolysis) and the fossil pollen samples measured using the combined FTIR microscope with FPA detector were compared with modern-pollen samples.
-This enables us to investigate changes in fossil-pollen chemistry, due to: i.
-differences between modern and fossil material, ii.
-extraction method and iii.
-effect of diagenesis (age of fossil pollen).
+This enables us to investigate changes in fossil-pollen chemistry, due to: i. differences between modern and fossil material and ii. extraction method.
 
@@
 <!-- The effect of treatment with SPT on pollen chemistry is tested on modern material to establish if any chemical changes occur. -->
-<!-- --> -->
+<!-- -->
 
@@
 
+\newpage
+
 # Methods and Materials
-The pollen was extracted from three sediment cores: Dalmuttlado in Norway [@bjuneHoloceneVegetationClimate2004], Tiefer See [@theuerkaufEffectsChangesLand2015]  and Meerfelder Maar [@brauerHighResolutionSediment1999] in Germany.
-The three sediment cores were selected because they contain *Pinus* pollen types, spanning three different periods in the late Quaternary (Table \@ref(tab:sites)).
-
-Prior to chemical characterisation using FTIR, all sediment samples were processed in two ways.
-First, the samples were extracted using density separation using dissolved Sodium Polytungstate (SPT) [@kampsAgentSeparationDissolved1985] at 1.43 g cm^-3^ as separation medium.
-During density separation a known volume of SPT (density 1.43 g cm^-3^) solution was added to the sample and homogenized/vortexed, followed by centrifugation.
-After centrifugation, particulates/material heavier than 1.43 g cm^-3^ accumulate of the bottom of the sample vial, whilst the lighter fraction was decanted and collected.
-Note that there is no agreement on the density of the pollen exine, with reported values varyiing between  <1.4 g cm^-3^ [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996], 1.88 cm^-3^ [@nakagawaDensemediaSeparationMore1998] and 2.2 cm^-3^  [@juvigneDensiteExinesQuelques1973].
-In our case the lighter fraction contained majority of the pollen grains.
-Sample residues were then washed three times in using distilled H2O to remove any excess SPT, resulting in sample residues with a higher pollen concentration compared to other organic and inorganic materials on the slide.
-
-We then applied an acetolysis treatment [e.g. @erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989] to one subset of the samples following washing and extracting using SPT.
-Although acetolysis treatments at low temperatures did not create major chemical differences in Lycopodium spores [@jardineImpactOxidationSpore2015], it is uncommon for low temperature treatments to be applied to fossil sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
-Under these conditions it is possible that acetolysis may alter the chemical composition of sporopollenin, and so in many cases it may be preferable to use alternative approaches to extract fossil pollen grains before analysis using FTIR.
-Thus, incorporating the acetolysis step on this analysis allows us to compare the effects of pollen grains spectra from both treated and untreated pollen grains.
-
-## Need to have something on the modern samples
 
@@
 
+The pollen was extracted from three sediment cores: Dalmuttlado in Norway [@bjuneHoloceneVegetationClimate2004], Tiefer See [@theuerkaufEffectsChangesLand2015]  and Meerfelder Maar [@brauerHighResolutionSediment1999] in Germany.
+The three sediment cores were selected because they contain *Pinus sylvestris* pollen types, spanning three different periods in the late Quaternary (Table \@ref(tab:sites)).
+*Pinus* pollen in these cores has been identified as *Pinus sylvestris* in the original publication.
+
+Prior to chemical characterisation using FTIR, all sediment samples were processed in two ways.
+First, the samples were extracted using density separation using dissolved Sodium Polytungstate (SPT) [@kampsAgentSeparationDissolved1985] at 1.43 g cm^-3^ as separation medium.
+During density separation a known volume of SPT (density 1.43 g cm^-3^) solution was added to the sample and homogenized/vortexed, followed by centrifugation.
+After centrifugation, particulates/material heavier than 1.43 g cm^-3^ accumulate of the bottom of the sample vial, whilst the lighter fraction was decanted and collected.
+Note that there is no agreement on the density of the pollen exine, with reported values varyiing between <1.4 g cm^-3^ [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996], 1.88 cm^-3^ [@nakagawaDensemediaSeparationMore1998] and 2.2 cm^-3^  [@juvigneDensiteExinesQuelques1973].
+In our case the lighter fraction contained majority of the pollen grains.
+Sample residues were then washed three times in using distilled H~2~O to remove any excess SPT, resulting in sample residues with a higher pollen concentration compared to other organic and inorganic materials on the slide.
+
+We then applied an acetolysis treatment [e.g. @erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989] to one subset of the samples following washing and extracting using SPT.
+Although acetolysis treatments at low temperatures did not create major chemical differences in Lycopodium spores [@jardineImpactOxidationSpore2015], it is uncommon for low temperature treatments to be applied to fossil sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
+Under these conditions it is possible that acetolysis may alter the chemical composition of sporopollenin, and so in many cases it may be preferable to use alternative approaches to extract fossil pollen grains before analysis using FTIR.
+Thus, incorporating the acetolysis step on this analysis allows us to compare the effects of pollen grains spectra from both treated and untreated pollen grains.
+
+## Need to have something on the modern samples
+
+The modern *Pinus sylvestris* pollen samples were collected in Bergen and Innsbruck along a elevational gradient [@seddonPollenchemistryVariationsElevation].
+Pollen was sampled from four to six pollen cones and air-dried at room temperature for 24h.
+For this study, pollen from two trees at each location were selected and analysed
+
 ## FTIR measurements
 The spectra were measured with a Vertex 70 FTIR spectrometer connected to a Hyperion 3000 IR microscope (Bruker Optik, Ettlingen, Germany) equipped with a focal plane array (FPA) nitrogen cooled mercury cadmium telluride (MCT) detector and a 15x magnification objective.
-The FPA-MCT detector allows the capture of FTIR images, where each pixel is a full spectra, instead point measurements.
+The FPA-MCT detector allows the capture of FTIR images, where each pixel is a full spectra, instead of point measurements.
 For this study images of 128x128 pixels were taken and each spectra was recorded with 64 scans.
@@
 plsr ordination scores all samples diff fresh vs fossil b.
-regression coef; PC1 c.
-regression coef PC2; d plsr ordination scores only samples from cores diff spt vs acet; e.
-regression coef PC1; f.
-regression coef PC2", message=FALSE, warning=FALSE, out.width="100%"}
+regression coef; component 1 c.
+regression coef component 2; d plsr ordination scores only samples from cores diff spt vs acet; e.
+regression coef component 1; f.
+regression coef component 2", message=FALSE, warning=FALSE, out.width="100%"}
 knitr::include_graphics(here("figures", "ord_plots.png"))
@@
 Most notably the lipid peak at 1745 cm^-1^, caused by triglycerides, is not visible in the fossil pollen grains, neither in the SPT treated or acetolysed pollen.
-There are also several differences in the "fingerprint" region (1500 to 1800 cm^-1^) and carbohydrate region from 1000 to 1250 cm^-1^ between modern and fossil pollen grains.
+There are also several differences in the "fingerprint" region (1500 to 1800 cm^-1^) and carbohydrate region (1000 to 1250 cm^-1^) between modern and fossil pollen grains.
 
@@
 The regression coefficients reflect our observations from the visual inspection and show that lipid  and protein peaks (1745 and 1500 - 1800 cm^-1^) have disappeared and there are substantial changes in the carbohydrate region (1000 to 1250 cm^-1^) in fossil pollen (Fig \@ref(fig:ordinations) b and c).
+The region 1000 - 1200 cm^-1^ is now dominated in acetolysed fossil pollen by two peaks and its shoulders at 1040 and 1170 cm^-1^.
 
@@
 Comparisons to other studies that looked at acetolysed and fossil samples show similarities between our acetolysed *Pinus* pollen and other taxa.
-E.g. the carb region 1000 - 1200 in our Pine is most similar to acetolysed extant poaceae set of @jardineSporopolleninChemistryIts2021, two large peaks at 1040 and 1190/1170.
+E.g.
+the carb region 1000 - 1200 in our Pine is most similar to the acetolysed extant poaceae set of @jardineSporopolleninChemistryIts2021, where two large peaks at 1040 and 1190/1170 are visible similar to our samples.
 
-The fossil poaceae data set (Amazon Fan and Western Amazon) from @jardineSporopolleninChemistryIts2021 on the other looks quite different from our fossil pollen and their extant acetolysed set, with a large peak at 1600 and 1190 peak not visible.
-These fossil poaceae set was treated differently than ours and jardines extant poaceae set (combination of heavy liquid separation, Na~4~P~2~O~7~, lignite).
+The fossil poaceae data set (Amazon Fan and Western Amazon) from @jardineSporopolleninChemistryIts2021 on the other looks quite different from our fossil pollen and their extant acetolysed set, with a large peak at 1600 and 1190 peak not clearly visible.
+These fossil poaceae set were treated differently than our pollen and the extant poaceae set, a combination of heavy liquid separation, Na~4~P~2~O~7~ and lignite as opposed to mainly acetolysis.
 
-@jardineImpactOxidationSpore2015 showed limited effect on pollen chemistry of cold acetolysis if it was shorter than 10 minutes.
-Hot acetolysis (90&deg;C), which is standard procedure in palynology, had an effect from the first minute. 
+@jardineImpactOxidationSpore2015 showed limited effects on pollen chemistry of cold acetolysis if it was shorter than 10 minutes.
+Hot acetolysis (90&deg;C), which is standard procedure in palynology [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989], had an effect from the first minute. 
 The simililarities of the acetolysed extant set to our acetolysed fossil set and dissimilarities to pollen treated with a different combination of extractions (Amazon Western and Fan), suggest that chemical treatment of the pollen has a large effect on the chemistry of pollen, both modern and fossil origin.
@@
 The effects of these extraction methods need to be explored on modern and fossil material side by side to see if there are links to bridge the transition from modern to fossil pollen.
+It is crucial to understand the changes pollen undergoes during diagenesis, e.g. for the purposes of artificially fossilizing grains to curate reference collections of fossil grains of known taxa.
 
@@
 This peak is not present in either modern pollen or acetolysed pollen, additionally Tiefer See shows a similar peak at 1630 cm^-1^.
-
 The 1090 cm^-1^ peak dissappears after acetolysis from all spectra as does the 1630 cm^-1^ peak in Tiefer See.
-The region 1000 - 1200 cm^-1^ is now dominated in acetolysed fossil pollen by two peaks and its shoulders at 1040 and 1170 cm^-1^.
-This 1040 cm^-1^ has high importance for the PLSR model differentiating acetolysed and non acetolysed pollen (Fig. \@ref(fig:ordinations) e and f).
-This peak in the non acetolysed samples is most likely biogenic silica, when compared to this biogenic silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016] and hydrated silica spectra [@SilicaATRIRSpectrum].
+This peak (1090 cm^-1^) in the non acetolysed samples is most likely biogenic silica, when compared to this biogenic silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016] and hydrated silica spectra [@SilicaATRIRSpectrum].
 During the measurements with the microscope, we applied our extracted pollen as droplets onto ZnSe slides and waited for the water to evaporate.
 After the droplets had dried we observed a crystalline layer over the same area that covered all particles and pollen (ref Foto in SM).
-From the chemical spectra, it is most likely that that it is dissolved silica that crystallised, although the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.
+From the chemical spectra, it is most likely that it is dissolved silica that crystallised, although the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.
 
@@
 
-Acetolysed pollen showed two dominant peaks in the carbohydrate region which are not visible in the modern spectra at 1040 and 1170 - 1190 cm^-1^ or the non acetolysed pollen.
+Acetolysed pollen showed two dominant peaks in the carbohydrate region at 1040 and 1170 - 1190 cm^-1^, which are not visible in the modern spectra or the non acetolysed pollen.
 This change in the 1170 - 1190 cm^-1^ region is unique to acetolysized pollen and not visible in non acetolysed pollen and had higher regression coefficients that the surrounding wavenumbers (Fig.
 \@ref(fig:ordinations)).
+The 1040 cm^-1^ peak had high importance for the PLSR model differentiating acetolysed and non acetolysed pollen (Fig. \@ref(fig:ordinations) e and f).
 
 The presence of this new peak in the shoulder of the 1170 cm^-1^ peak, which can indicate sporopollenin and carbohydrates, suggest changes in the pollen chemistry.
-These changes may be related to removal and or alteration of grain wall carbohydrates (cellulose etc) as well as changes to sporopollenin due to acetolysis.
+Presence of these two dominant peaks may be related to removal and or alteration of grain wall carbohydrates (cellulose etc) as well as changes to sporopollenin due to acetolysis.
 [FM: this would be answered by acetolysis of the fresh grains!!]
 In non acetolysed pollen the 1190 peak is not visible in any of the cores, which indicates that it is not due to aging/diageneis, but acetolysis.
+Unfortunately, the BSi peak is covering the area of the 1040 cm^-1^ peak in our DAL and TSK samples, while it is visible in the MFM spectra.
 
@@
 Our study is a demonstration of microscope FTIR for pollen chemical methods using fossil pollen.
-We focused on the chemistry of *Pinus* pollen, because *Pinus* is one of a few taxa, that is quite common in sediment cores and largely dominated by 1 species, *Pinus Sylvestris*.
+We focused on the chemistry of *Pinus* pollen, because *Pinus* is one of a few taxa, that is quite common in sediment cores and largely dominated by 1 species, *Pinus sylvestris*.
 The pollen is also relatively large, which prevents any scattering problems found with smaller grains.
-We used *Pinus Sylvestris* to explore chemical differences between modern and fossil material, because this has not been fully explored yet.
+We used *Pinus sylvestris* to explore chemical differences between modern and fossil material, because this has not been fully explored yet.
 
@@
 This is an improvement on previous single grain measurements, where each grain had to be targeted individually.
-The FPA-MCT detector on our FTIR microscope allows simultaneous capture of a larger area of the sample (XXX x XXX µm), compared to other microscopes (XXX x XXX µm).
-An image with pixel spectra is created, from which spectra of single grains were extracted.
+The FPA-MCT detector on our FTIR microscope allows simultaneous capture of a larger area of the sample (XXX x XXX µm), compared to other detectors (XXX x XXX µm).
+An image with a spectra for each pixel is created, from which spectra of single grains were extracted.
 For this study the spectra were extracted by hand, but this process is possible to automate.
-Approaches using machinelearning and deep learing have shown to be capable of extracting objects from larger images, using deep neural networks specialised on image segmentation/object location [@renFasterRCNNRealTime2015].
+Approaches using deep learing have shown to be capable of extracting objects from larger images, using deep neural networks specialised on image segmentation/object location [@renFasterRCNNRealTime2015].
 Conceptually the problem of locating pollen on a FTIR image is similar to locating an object on a optical image and these deep learning approaches can be applied.
 @kangSinglecellClassificationFoodborne2020 used deep neural network to classify bacteria based on images and spectral information, while @grosserueschkampSpatialMolecularResolution2017 used random forests for segmentation and identification of malignant tumor in histological thin sections using FTIR microscope equipped with a FPA detector.
+These applications from other disciplines with a similar task show the potential of FPA detectors for chemical palynology using FTIR microscopy.
 
@@
 An objective with higher magnification (20x) may improve this aspect of using a FTIR microscope.
-The FPA detector in combination with automated localisation of potential pollen grains based on their chemical signature may help in solving this part of the problem.
+The FPA detector in combination with automated localisation of potential pollen grains based on their chemical signature may also help in solving this part of the problem.
 

### 2021-06-10 manuscript.Rmd
added spt refs to disco

@@
 Acetolysis is the conventional method to extract pollen in high concentration, but as discussed before, we believe there is value in extracting non acetolysed pollen from sediments, because of the substantial differences in pollen chemistry we found between acetolysed and non acetolysed pollen.
-We believe that density separation is a viable alternative [REF], but may have been one possible cause for our low concentration in some samples.
-The reported density of pollen is variable [REFS] and our extraction was done on the lower end of these reported densities (1.44 g cm ^-3^).
+We believe that density separation is a viable alternative [@caffreyUseLithiumHeteropolytungstate2013; @leipeTestingPerformanceSodium2019; @regnellPreparativeCentrifugationNew1996], but may have been one possible cause for our low concentration in some samples.
+The reported density of pollen is variable [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996; @nakagawaDensemediaSeparationMore1998; @juvigneDensiteExinesQuelques1973] and our extraction was done on the lower end of these reported densities (1.44 g cm ^-3^).
 We opted for this density because test extractions using a higher density (1.98 g cm ^-3^), also introduced even more organic material to clutter our samples.
 A refinement of the right extraction medium density in combination with sieving of the samples to remove non pollen organic material may be one approach.
-XXX have reported similar counts of pollen grains to conventional extraction methods using density separation.
+@caffreyUseLithiumHeteropolytungstate2013 and @leipeTestingPerformanceSodium2019 have reported similar counts of pollen grains to conventional extraction methods using density separation.
 

### 2021-06-11 manuscript.Rmd
adjusted coef/loadings plot function

@@
 This is an improvement on previous single grain measurements, where each grain had to be targeted individually.
-The FPA-MCT detector on our FTIR microscope allows simultaneous capture of a larger area of the sample (XXX x XXX µm), compared to other detectors (XXX x XXX µm).
+The FPA-MCT detector on our FTIR microscope allows simultaneous capture of a larger area of the sample (XXX x XXX µm ), compared to other detectors (e.g.
+100 x 100 µm @jardineChemotaxonomyDomesticatedGrasses2019, 50 x 50 ?µm?
+@zimmermannChemicalCharacterizationIdentification2018).
 An image with a spectra for each pixel is created, from which spectra of single grains were extracted.
@@
 Conceptually the problem of locating pollen on a FTIR image is similar to locating an object on a optical image and these deep learning approaches can be applied.
-@kangSinglecellClassificationFoodborne2020 used deep neural network to classify bacteria based on images and spectral information, while @grosserueschkampSpatialMolecularResolution2017 used random forests for segmentation and identification of malignant tumor in histological thin sections using FTIR microscope equipped with a FPA detector.
+@kangSinglecellClassificationFoodborne2020 used deep neural network to classify bacteria based on images and spectral information, while @grosserueschkampSpatialMolecularResolution2017 used random forests for segmentation and identification of malignant tumor in histological thin sections using FTIR microscope equipped with a FPA detector (715 x 715 µm).
 These applications from other disciplines with a similar task show the potential of FPA detectors for chemical palynology using FTIR microscopy.

### 2021-06-11 R/Regression.R

@@ -12,7 +12,7 @@ plot_coef = function(obj, comp = 1, coef = T){
   } else {
     tmp = as.data.frame(obj[,comp])
     n = ncol(tmp)
-    tmp$wavenumbers = as.numeric(rownames(obj[,comp]))
+    tmp$wavenumbers = as.numeric(rownames(tmp))
     for (i in 1:n) {
       tmp[,i] = tmp[,i] + i * max(abs(obj[,comp]))
     }
@@ -21,20 +21,23 @@ plot_coef = function(obj, comp = 1, coef = T){
   tmp2 = melt(tmp, id.vars = c("wavenumbers"))
   colnames(tmp2) = c("wavenumbers", "treatment", "coef")
   g = ggplot(tmp2, aes(x = wavenumbers, y = coef, color = treatment)) + 
-    geom_line() + xlim(1900,900) + theme_bw() +
+    geom_line() +
+    scale_x_reverse(name = "wavenumbers",limits = c(1800,900), breaks = seq(1800,900,-100)) +
+    scale_y_continuous(name = "Arbitrary units",breaks = NULL)
+    theme_bw() +
     theme(axis.ticks.y = element_blank(),
           axis.text.y = element_blank())
     
   return(g)
 }
 
-plot_scor = function(obj, meta){
+plot_scor = function(obj, meta, comps = c("PC1","PC2")){
   meta = meta[,c("Label","treatment")]
   tmp = as.data.frame(obj[])
   n = ncol(tmp)
   colnames(tmp) = c("PC1","PC2","PC3","PC4","PC5","PC6")
   tmp = cbind(tmp, meta)
-  g = ggplot(tmp, aes_string(x = "PC1", y =  "PC2", color = "Label")) +
+  g = ggplot(tmp, aes_string(x = comps[1], y = comps[2], color = "Label")) +
     geom_point(aes_string(shape = "treatment"), size = 1) + scale_shape_manual(values = c(17, 19, 15)) + theme_bw()
   return(g)
 }
@@ -198,10 +201,10 @@ design = "
 
 #create_layout(3,2,2,4)
 
-o = plsr_plot[[1]] +
+o = plsr_plot[[1]] + xlab("Comp. 1") + ylab("Comp. 2") +
   plsr_plot[[2]] +
   plsr_plot[[3]] +
-  plsr_plot[[13]] +
+  plsr_plot[[13]] + xlab("Comp. 1") + ylab("Comp. 2") +
[…]

### 2021-06-11 manuscript.Rmd
typo fixes

@@
 After centrifugation, particulates/material heavier than 1.43 g cm^-3^ accumulate of the bottom of the sample vial, whilst the lighter fraction was decanted and collected.
-Note that there is no agreement on the density of the pollen exine, with reported values varyiing between <1.4 g cm^-3^ [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996], 1.88 cm^-3^ [@nakagawaDensemediaSeparationMore1998] and 2.2 cm^-3^  [@juvigneDensiteExinesQuelques1973].
+Note, that there is no agreement on the density of the pollen exine, with reported values varying between <1.4 g cm^-3^ [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996], 1.88 cm^-3^ [@nakagawaDensemediaSeparationMore1998] and 2.2 cm^-3^  [@juvigneDensiteExinesQuelques1973].
 In our case the lighter fraction contained majority of the pollen grains.

### 2021-06-11 R/plotting.R
added peak vlines to meanplot

@@ -78,7 +78,15 @@ data= data%>%
   arrange(off)
 colnames(data$emsc) = colnames(FRE$emsc)
 
+vlines = data.frame(wn = c(1745, 1710, 1605, 1515, 1170, 1040),
+                    label = as.character(c(1745, 1710, 1605, 1515, 1170, 1040)),
+                    y = rep(0.01, 6))
+
 l = plot_mean_spectra(data, expression(species == "Pinus"), spec_id = "emsc", group = "off", id.vars = c("ids", "treatment", "Label", "depth", "age", "type", "off"))
+
+l = l + ylim(.05, 2.3) +
+  geom_vline(data = vlines, aes(xintercept = wn), linetype = "dashed", alpha = .3) +
+  geom_text(data = vlines, aes(wn, y, label = label), inherit.aes = F)
 
 ggsave(here("figures","mean_plots.png"), l, width = 17, height = 25,units = "cm")
 saveRDS(l, here("data","output","mean_plots_whole.rds"))

### 2021-06-11 manuscript.Rmd
added woutersen to jardine paragraph

@@
 
-ADD COMPARISON TO JARDINE PAPERS and other (e.g. Woutersen)
 Comparisons to other studies that looked at acetolysed and fossil samples show similarities between our acetolysed *Pinus* pollen and other taxa.
 E.g.
-the carb region 1000 - 1200 in our Pine is most similar to the acetolysed extant poaceae set of @jardineSporopolleninChemistryIts2021, where two large peaks at 1040 and 1190/1170 are visible similar to our samples.
+the carb region 1000 - 1200 in our Pine is most similar to the acetolysed extant poaceae set of @jardineSporopolleninChemistryIts2021 and acetolyzed herbarium *Nitrariaceae* pollen [@woutersenNovelApproachStudy2018], where two peaks at 1040 and 1190/1170 are visible and similar to our samples.
 
@@
 
-### pollen chemical differences
+### Pollen chemical differences
 

### 2021-06-15 manuscript.Rmd
Als comments

--- 2021-06-11 manuscript.Rmd
@@
 The study of plant microfossil (pollen and spores) and macrofossil (plant remains such as seeds) is an integral part of palaeoecology and palaeobotany.
-Pollen records are an important tool to reconstruct past floristic, vegetational and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @noguesbravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013] and provide insights into ecological processes which operate on decadal, centennial and millennial timescales (e.g.
+Pollen records are an important tool to reconstruct past floristic, vegetation and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @noguesbravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013] and provide insights into ecological processes which operate on decadal, centennial and millennial timescales (e.g.
 the post-glacial migration of tree taxa into new habitats following the Quaternary glacial maxima @gieseckeMovingFrontPopulation2005, @huntleyQuaternaryPalaeoecologyEcology1996, @tinnerCentralEuropeanVegetation2001).
 
-The high preservation potential of pollen grains in the geological record results from the presence of the resistant biopolymer sporopollenin, which is the main building block of the outer pollen-grain wall (exine).
-Although the exact structure of sporopollenin remains unknown [ref], sporopollenin has been described as being similar to lignins, incorporating both aromatic compounds and cross-linked aliphatic components [@liMolecularStructurePlant2019].
+The high preservation-potential of pollen grains in the geological record results from the presence of the resistant biopolymer sporopollenin, which is the main building block of the outer pollen-grain wall (exine).
+Although the exact structure of sporopollenin remains unknown [ref], sporopollenin has been described as being similar to lignin, incorporating both aromatic compounds and cross-linked aliphatic components [@liMolecularStructurePlant2019].
 Variations in the chemical composition of sporopollenin can reflect phylogenetic differences between plant groups [@nieropQuantitativeAnalysisDiverse2019]. 
-There is also evidence of environmental signals that are recorded in the chemistry of pollen grains which may be used as proxies of past environmental conditions, e.g.
-temperature [@bagciogluMultiscaleVibrationalSpectroscopic2015], nutrients [@kendelChemicalAnalysisPollen2020; @diehnCombiningChemicalInformation2020], UV-B radiation [@rozemaUVBAbsorbanceUVB2001; @blokkerDevelopmentProxySurface2005; @fraserUVBAbsorbingPigments2011; @jardineSheddingLightSporopollenin2017].
-Thus, in recent years, the study of pollen-grain chemistry has gained increasing focus in palaeoecological research.
+There is also evidence of environmental signals that are recorded in the chemistry of pollen grains, including temperature [@bagciogluMultiscaleVibrationalSpectroscopic2015], nutrients [@kendelChemicalAnalysisPollen2020; @diehnCombiningChemicalInformation2020],  and UV-B radiation [@rozemaUVBAbsorbanceUVB2001; @blokkerDevelopmentProxySurface2005; @fraserUVBAbsorbingPigments2011; @jardineSheddingLightSporopollenin2017].
+Thus, in recent years, the study of pollen-grain chemistry has gained increasing focus in palaeoecological research since it is possible that chemical variations observed in sporopollenin may be used as proxies for past environmental conditions, with a particular focus on using sporopollenin chemistry to reconstruct past changes in UV-B radiation **[ref: here my 2019 paper would be a good one to cite]** and for improving the taxonomic precision available for sediment records [@dellannaPollenDiscriminationClassification2009; @zimmermannAnalysisAllergenicPollen2016].
 
-<!-- This is a promising finding for the application of pollen identification with chemical methods in palaeoecological research, because it has the potential to increase taxomonic resolution [@dellannaPollenDiscriminationClassification2009; @zimmermannAnalysisAllergenicPollen2016].
--->
+<!-- This is a promising finding for the application of pollen identification with chemical methods in palaeoecological research, because it has the potential to increase taxomonic resolution  -->
 <!-- Similarly, FTIR has become a popular method in the quantification of UV-B absorbing compounds to reconstruct changes in UV-B radiation from sediment sequences [ref]. -->
@@
 
-Successful application of pollen chemistry techniques within palaeoecological research requires both a good understanding of the factors influencing chemical variations in the modern environment, and in the accurate estimation of the chemical spectra from pollen grains.
-Fourier Transform Infrared (FTIR) spectroscopy has become one of the most popular methods for characterising the chemical variations present within pollen and spores, and one common approach has been to obtain spectra from large quantities of modern-pollen samples (e.g.
+Successful application of pollen-chemistry techniques within palaeoecological research requires a thorough understanding of the factors influencing chemical variations in the modern environment and in the accurate estimation of the chemical spectra from pollen grains.
+Any insights gained from studies on modern grains must also be transferable under fossil settings [@jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
+Fourier Transform Infrared (FTIR) spectroscopy has become one of the most popular methods for characterising the chemical variations present within pollen and spores in the modern environment.
+Bulk FTIR techniques, which involve the analysis of large quantities of modern-pollen samples (e.g.
 1-2 mg) (e.g.
-@zimmermannCharacterizationPollenVibrational2010, @lahlaliATRFTIRSpectroscopyReveals2014).
-These bulk FTIR techniques have been used to identify pollen chemical differences in modern pollen samples at the genus level, and in some cases, chemical variations between pollen types may even extend to species-level differences [@muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020].
-Bulk FTIR approaches have the advantage of reducing scattering effects from measurements due to the analysis of large sample quantities, but these approaches may be less relevant for fossil sequences since the number of pollen grains analysed vastly exceeds the quantity which is practically available from a sediment sample.
+@zimmermannCharacterizationPollenVibrational2010, @lahlaliATRFTIRSpectroscopyReveals2014) **[is it possible to have one very short description here about what these techniques entail- e.g.
+the resolution of the laser?]**, have the advantage of reducing scattering effects from measurements due to the analysis of large sample quantities.
+As a result, FTIR spectra obtained from bulk pollen samples have been used identify chemical differences at the genus, and in some cases, species level [@muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020].
+However, these approaches may be less relevant for fossil sequences since the number of pollen grains analysed vastly exceeds the quantity which is practically available from a sediment-core sample.
 
 An alternative approach involves the use of FTIR microscopes, which require manual targeting of small areas of the sample slides to record spectra of single grains or groups of pollen grains.
-Some microscopes are equipped with automated stages, where multiple locations on the slide can be targeted in sequence to record spectra.
+Because of their higher spatial resolution, FTIR-microscopes are better suited to sediment samples as a result of the capacity to generate FTIR spectra from individual grains, although spectral-processing procedures may be required to reduce the scattering effects when obtaining spectra from individual pollen grains **[ref]**.
+Nevertheless, microscopic FTIR techniques have now been applied to both modern and fossil-pollen grains preserved in sediments.
+The approach has been successfully used to show that chemical differences are found between pollen on genus and species level [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018;  @zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019], and microscopic-FTIR techniques have also been used to observe chemical differences to reconstruct changes in UV-B absorbing compounds in different periods across Earth's history [@fraserUVBAbsorbingPigments2011; @jardinePollenSporesBiological2016; @jardineProxyReconstructionUltravioletB2019; @jardineSheddingLightSporopollenin2017].
+**[Does this sentence belong in the next paragraph]**. Some microscopes are equipped with automated stages, where multiple locations on the slide can be targeted in sequence to record spectra.
 
-This approach has been successfully applied in various studies to show that chemical differences are found between pollen on genus and species level [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018;  @zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019], and may be better suited to sediment samples as a result of the capacity to generate FTIR spectra from individual grains.
-Indeed, microscopic FTIR techniques have now been applied to both modern and fossil-pollen grains preserved in sediments, with the ultimate goal to use observed chemical differences to reconstruct changes in UV-B absorbing compounds in different periods across Earth's history [@fraserUVBAbsorbingPigments2011; @jardinePollenSporesBiological2016; @jardineProxyReconstructionUltravioletB2019; @jardineSheddingLightSporopollenin2017; @willisQuantificationUVFlux2011].
-However, such approaches may still result in interferences from scattering of FTIR spectra (is correct?), and the targeting of individual pollen grains on the slide can still be time consuming.
+One of key challenges faced by chemical palynologists is the time taken to extract and analyse fossil-pollen samples.
+However, recent developments in technologies related to FTIR may allow for easier and faster capture of spectra of pollen grains.
+A focal plane array detector (FPA) allows capture of an image and matrix of spectra over the same sample area instead of a single spectra at a specific point/area [ref].
+This enables the extraction of spectra from several single grains from the same area on the sample without the need to relocate the detector of the microscope or move the slide, and is therefore of potentially high relevance for palaeoecological research.
+When combined with imaging processing technology, it potentially allows for more rapid identification and generation of FTIR spectra of pollen grains from sediment samples, but whilst FTIR spectroscopy using FPA detectors have been applied to pollen samples **in aerobiology [or others, check references]**, there has been limited investigation into their application on fossil-pollen residues taken from sediment cores.
 
-More recent developments in technologies related to FTIR may allow for even easier and faster capture of spectra of pollen grains. 
-For example, a focal plane array detector (FPA) allows capture of an image and matrix of spectra over the same sample area instead of a single spectra at a specific point/area [ref].
-This enables e.g.
-the extraction of spectra from several single grains from the same area on the sample without the need to relocate the detector of the microscope or move the slide, and is of potentially high relevance for palaeoecological research.
-When combined with imaging processing technology, it potentially allows for more rapid identification and generation of FTIR spectra of pollen grains from sediment samples.
-However, whilst FTIR spectroscopy using FPA detectors have been applied to pollen samples **in aerobiology [or others, check references]**, there has been limited investigation into their application on fossil-pollen residues taken from sediment cores.
-
-An additional consideration for palaeoecologists aiming to apply FTIR techniques to their sediment records is that any insights gained from studies on modern grains are transferable under fossil settings [@jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
-Palaeoecologists use a combination of chemical extraction procedures depending on sediment type in order to isolate and concentrate pollen from the sediment matrix [@ridingGuidePreparationProtocols2021].
+Palaeoecologists may also use a variety of chemical processing techniques in order to speed up analyses through the isolation and concentration of pollen from the sediment matrix [@ridingGuidePreparationProtocols2021].
 One of the most common methods for sporopollenin extraction involves an acetolysis treatment, which in a combination with HF and HCl treatments enables removal of organic materials, calcium carbonates and silicates, while preserving the morphological integrity of pollen grains[@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
-In pollen-chemistry research, acetolysis has often been used in combination with FTIR on pollen as a way to standardise between modern and fossil samples, [@jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or to study the effects of acid extractions [@dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000; @schulzeosthoffPhenolsIntegratedCompounds1987; @jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
-Indeed, although acetolysis treatments at low temperatures did not create major chemical differences in Lycopodium spores [@jardineImpactOxidationSpore2015], it is uncommon for low temperature treatments to be applied to fossil sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
-Furthermore, a recent paper by @jardineSporopolleninChemistryIts2021 revealed potential chemically altering effects on fossil pollen composition.
-Whilst there is a developing understanding of the effects of a variety of chemical treatments on fossil- and modern-pollen grains, for many taxa these issues remain unresolved and more studies are needed which compare the chemical spectra of different pollen types.
+Acetolysis has also often been used in combination with FTIR on pollen as a way to standardise between modern and fossil samples, [@jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or to study the effects of acid extractions [@dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000; @schulzeosthoffPhenolsIntegratedCompounds1987; @jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
+Indeed, although acetolysis treatments at low temperatures did not create major chemical differences in Lycopodium spores [@jardineImpactOxidationSpore2015], it is uncommon for low-temperature treatments to be applied to fossil-sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
+Such treatments may then result in chemically altering effects on fossil-pollen composition [@jardineSporopolleninChemistryIts2021], whilst variations in sediment types mean that different combinations of sample preparation techniques may be required for different sediment types [@ridingGuidePreparationProtocols2021].
+Given the large variations in FTIR spectra observed across taxa, and the variety of sediment types even from Holocene sequences in Europe, more studies are needed investigate the potential effects of the chemical preparation procedures used to isolate pollen-grains from different pollen types.
 
 In this study we aimed to explore the use of an FPA detector to extract FTIR spectra from fossil-pollen *Pinus* grains.
-The development of a UV-proxy based on pollen chemistry is focused on *Pinus* pollen, which makes *Pinus* a interesting genus for our purposes. 
-Identifying chemical differences in fossil *Pinus* pollen may also allow for identification of cryptic *Pinus* species in the fossil record, because *Pinus* is currently mostly identified to type or even only genus [@beugLeitfadenPollenbestimmungFur2004; @gieseckePostglacialChangeFloristic2019].
+We aim to develop a workflow which enables comparison of fossil-pollen grains taken from three sediment-core locations **[TABLE 1]**, and compare the FTIR spectra generated using the FPA detector from both modern and fossil-pollen samples.
+Pollen was extracted from the sediment cores using two different approaches (density separation using sodium polytungstate, and acetolysis) and the fossil pollen samples measured using the combined FTIR microscope with FPA detector were compared with modern-pollen samples.
+Thus, in addition to testing the feasibility of the FPA detector on different sediment types, we were were also able to investigate how the inferred FTIR spectra in *Pinus* pollen can vary as a result of two different extraction procedures.
 
-We aim to develop a workflow which enabled comparison of fossil-pollen grains taken from three sediment-core locations (Germany and Norway).
-the pollen was extracted from the sediment cores using two different approaches (density separation, and acetolysis) and the fossil pollen samples measured using the combined FTIR microscope with FPA detector were compared with modern-pollen samples.
-This enables us to investigate changes in fossil-pollen chemistry, due to: i. differences between modern and fossil material and ii. extraction method.
-
-[**here we need Boris to add a few sentences here to give the study more context**]
-
-Chemical treatment...
+[**here we need Boris to add a few sentences here to give the study more context**]- not sure this is needed now? unless he is not happy with FPA detector section?
+<!-- Chemical treatment... -->
 <!-- SPT density separation as an alternative to acetolysis by investigating the chemical alteration of both extraction methods to three fossil cores. -->
@@
 
-#Alistair rewritten paragraphs- maybe move to methods
+<!-- #Alistair rewritten paragraphs- maybe move to methods -->
 
@@
 *Pinus* pollen in these cores has been identified as *Pinus sylvestris* in the original publication.
+In this study we focus on *Pinus* pollen for two reasons.
+First, *Pinus* spp.
+is a shade-intolerant, high pollen producer, and so a number of studies have proposed that *Pinus* spp.
+pollen may be a useful taxon to use as a biological indicator for reconstructing past changes of UV-B radiation [ref- willis 2011, seddon 2017, seddon 2021].
+Second, although prominent in Quaternary sediment records from Europe, and despite it's high species diversity in Europe (**Check number of species**), *Pinus* is currently mostly only split into two different pollen types using fossil pollen, reflecting the two phylogenitic lineages [@beugLeitfadenPollenbestimmungFur2004; @gieseckePostglacialChangeFloristic2019].
+Thus, identifying chemical differences in fossil *Pinus* pollen may also allow for identification of crytpic *Pinus* species from the pollen types observed in the fossil record.
 
@@
 Sample residues were then washed three times in using distilled H~2~O to remove any excess SPT, resulting in sample residues with a higher pollen concentration compared to other organic and inorganic materials on the slide.
+Since SPT is a chemically inert salt, it was expected that, after washingm this treatment should have minimal chemical effect on the chemical spectra inferred from the fossil-pollen grains.
 
@@
 Under these conditions it is possible that acetolysis may alter the chemical composition of sporopollenin, and so in many cases it may be preferable to use alternative approaches to extract fossil pollen grains before analysis using FTIR.
-Thus, incorporating the acetolysis step on this analysis allows us to compare the effects of pollen grains spectra from both treated and untreated pollen grains.
+Thus, incorporating the acetolysis step on this analysis allows us to compare the effects of pollen-grain spectra from both treated and untreated pollen grains.
 
-## Need to have something on the modern samples
-
-The modern *Pinus sylvestris* pollen samples were collected in Bergen and Innsbruck along a elevational gradient [@seddonPollenchemistryVariationsElevation].
+The modern *Pinus sylvestris* pollen samples were collected from locations close to Bergen, (Norway) in spring 2019 and Innsbruck, (Austria) in spring 2020.
 Pollen was sampled from four to six pollen cones and air-dried at room temperature for 24h.
-For this study, pollen from two trees at each location were selected and analysed
+For this study, pollen from two trees at each location were selected and analysed.
 
 ## FTIR measurements
-The spectra were measured with a Vertex 70 FTIR spectrometer connected to a Hyperion 3000 IR microscope (Bruker Optik, Ettlingen, Germany) equipped with a focal plane array (FPA) nitrogen cooled mercury cadmium telluride (MCT) detector and a 15x magnification objective.
+The spectra were measured with a Vertex 70 FTIR spectrometer connected to a Hyperion 3000 IR microscope (Bruker Optik, Ettlingen, Germany) equipped with a focal-plane-array (FPA) nitrogen-cooled mercury-cadmium-telluride (MCT) detector and a 15x magnification objective.
 The FPA-MCT detector allows the capture of FTIR images, where each pixel is a full spectra, instead of point measurements.
@@
 The second derivative spectra was then normalized using a EMSC model with a linear component [@lilandPackageEMSC2017].
-Preprocessing with SG and EMSC has been shown to improve performance of multivariate analysis, such as principal component analysis (PCA) or models in the partial least squares regression (PLSR) family [@woutersenNovelApproachStudy2018; @zimmermannOptimizingSavitzkyGolayParameters2013].
+Preprocessing with SG and EMSC has been shown to improve the performance of multivariate analyses, such as principal component analysis (PCA) or models in the partial least squares regression (PLSR) family [@woutersenNovelApproachStudy2018; @zimmermannOptimizingSavitzkyGolayParameters2013].
 
 We used two PLSR models to explore the chemical differences between modern and fossil grains (model 1), and acetolyzed and non acetolyzed fossil grains (model 2).
-For these models the explanatory variables were the preprocessed spectra and the response variables were hot-one encoded response variables (Model 1: modern/fossil; Model 2: acet./ non acet.).
+For these models the explanatory variables were the pre-processed spectra and the response variables were hot-one encoded response variables (Model 1: modern/fossil; Model 2: acet./ non acet.).
 By using PLSR we can maximize the distance between responses and identify chemical differences between the responses by looking at the regression coefficients.
@@
 
-Our aim with this study was to explore the use of a new type of detector (FPA-MCT) in combination with a FTIR microscope to explore the chemical variations of fossil pollen.
-Our goal was also to explore the effects of two differemt sample-processing methods on fossil pollen and compare the spectra we obtained with modern samples.
+<!-- Our aim with this study was to explore the use of a new type of detector (FPA-MCT) in combination with a FTIR microscope to explore the chemical variations of fossil pollen. -->
+<!-- Our goal was also to explore the effects of two different sample-processing methods on fossil pollen and compare the spectra we obtained with modern samples. -->
 
@@
 
-There are differences between modern and fossil grains (Fig. \@ref(fig:meanspectra)).
+Our analyses revealed clear differences between modern- and fossil-pollen grains (Fig. \@ref(fig:meanspectra)).
 The prominent and characteristic sporopollenin peaks (1600, 1510 and 1170 cm^-1^) are visible in the fossil pollen, while other parts of the spectra show substantial differences.
-Most notably the lipid peak at 1745 cm^-1^, caused by triglycerides, is not visible in the fossil pollen grains, neither in the SPT treated or acetolysed pollen.
+Most notably, the lipid peak at 1745 cm^-1^, caused by triglycerides, is not visible in the fossil pollen grains, neither in the non-acetolysed and the acetolysed fossil pollen.
 There are also several differences in the "fingerprint" region (1500 to 1800 cm^-1^) and carbohydrate region (1000 to 1250 cm^-1^) between modern and fossil pollen grains.
 
-After visual inspection of the spectra, the chemical differences between modern and fossil pollen were explored with a PLSR model.
-The ordination showed that the fossil pollen are clearly separated from the modern pollen (Fig \@ref(fig:ordinations) a).
+These chemical differences are also clearly demonstrated in the results of the PLSR model (Fig \@ref(fig:ordinations) a).
+Fossil pollen are clearly separated from the modern pollen along the first component of the PLSR model.
 The regression coefficients reflect our observations from the visual inspection and show that lipid  and protein peaks (1745 and 1500 - 1800 cm^-1^) have disappeared and there are substantial changes in the carbohydrate region (1000 to 1250 cm^-1^) in fossil pollen (Fig \@ref(fig:ordinations) b and c).
@@
 
-ADD COMPARISON TO JARDINE PAPERS and other (e.g. Woutersen)
-Comparisons to other studies that looked at acetolysed and fossil samples show similarities between our acetolysed *Pinus* pollen and other taxa.
-E.g.
-the carb region 1000 - 1200 in our Pine is most similar to the acetolysed extant poaceae set of @jardineSporopolleninChemistryIts2021, where two large peaks at 1040 and 1190/1170 are visible similar to our samples.
+**The main important discussion point to add here is to note the main differences that you see in the fossil and modern Pine pollen spectra- taken using bulk analyses, and using micro-FTIR.
+Then you address the issue of comparing the FPA detector.
+I think the story si that the fresh spectra you have here are quite similar to spectra in other studies, correct?**
 
-The fossil poaceae data set (Amazon Fan and Western Amazon) from @jardineSporopolleninChemistryIts2021 on the other looks quite different from our fossil pollen and their extant acetolysed set, with a large peak at 1600 and 1190 peak not clearly visible.
-These fossil poaceae set were treated differently than our pollen and the extant poaceae set, a combination of heavy liquid separation, Na~4~P~2~O~7~ and lignite as opposed to mainly acetolysis.
+## Effects of acetolysis on the fossil-pollen samples
+In addition to the differences between fresh and fossil pollen samples, our analysis revealed large differences in chemical variations as a result of the acetolysis treatment on the fossil *Pinus* spp.
+pollen grains.
+For example, acetolysed pollen showed two dominant peaks in the carbohydrate region at 1040 and 1170 - 1190 cm^-1^, which are not visible in the modern spectra or the non acetolysed pollen.
+This change in the 1170 - 1190 cm^-1^ region is unique to acetolysized pollenand had higher regression coefficients that the surrounding wavenumbers (Fig. \@ref(fig:ordinations)).
+The 1040 cm^-1^ peak also had high importance for the PLSR model differentiating acetolysed and non acetolysed pollen (Fig. \@ref(fig:ordinations) e and f).
 
-@jardineImpactOxidationSpore2015 showed limited effects on pollen chemistry of cold acetolysis if it was shorter than 10 minutes.
-Hot acetolysis (90&deg;C), which is standard procedure in palynology [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989], had an effect from the first minute. 
+The presence of this new peak in the shoulder of the 1170 cm^-1^ wavenumbers in the acetolysised pollen grains can be indicative of sporopollenin and carbohydrates, and suggests an overall change in the pollen chemistry of the sample.
+The presence of these two dominant peaks may be related to the removal and or alteration of grain-wall carbohydrates (cellulose etc), as well as changes to chemical composition of sporopollenin due to the acetolysis treatment.
+[FM: this would be answered by acetolysis of the fresh grains!!]
+Since the 1190  cm^-1^ peak is not visible in any of the samples from the  non-acetolysed pollen, this likely indicates that the acetolysis treatment is creating this peak.
+<!-- Unfortunately, the BSi peak is covering the area of the 1040 cm^-1^ peak in our DAL and TSK samples, while it is visible in the MFM spectra.  -->
+
+Using the second PLSR model, the acetolysed pollen is clearly separated from the non-acetolysed pollen, with a much greater distance in the ordination space between samples for acetolysed pollen compared to non-acetolysed pollen (Fig.
+\@ref(fig:ordinations) d).
+In addition, there is some differentiation between the cores.
+Samples from Tiefer See are more separate from samples of the other two cores, which show more similar scores in the ordination.
+One interpretation of this is that *Pinus* pollen from DAL and MFM are chemically quite similar, while TSK *Pinus* pollen is different in chemical structure.
+Since it is likely that the main contributions to the pollen some in the TSK core is *Pinus sylvestris*, it is likely that the same species is represented by all three cores, and the main difference between of the TSK pollen is age.
+The TSK samples are from the past 80 years, while MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene, with the youngest sample several 100 years older than TSK pollen.
+<!-- We cannot see any evidence that sediment type has an effect on the chemistry of the pollen, both TSK and MFM cores are varved cores, while DAL is not. -->
+<!-- This would suggest more difference between varved and non varved pollen if different types of diagenesis had an effect.  -->
+<!-- Any difference may be overshadowed by the large difference in age between the samples from these cores.  -->
+This finding could suggest a chemical difference in fossil pollen grains that is persistent even after acetolysis, which may be explained by the age of the pollen.
+Further studies on pollen from different sediment types at fixed age/depths may provide a clearer answer on the effects of sediment type and age on pollen chemistry.
+
+**[one thing to think about- could it be that the SPT residues on TSK were actually influencing the acetolysis effect in some way?]**
+
+In general there are similarities between the FTIR spectra from the acetolysed *Pinus* pollen from our study and in the acetolysed pollen grains of other taxa
+For example, the carbohydrate region at 1000 - 1200 cm^-1^ our acetolysised *Pinus* pollen samples is most similar to the acetolysed extant Poaceae set of @jardineSporopolleninChemistryIts2021 and acetolyzed herbarium *Nitrariaceae* pollen [@woutersenNovelApproachStudy2018], where two peaks at 1040 and 1190/1170 are visible and similar to our samples.
+However, spectra from a modern Poaceae data set (sampled from the the Amazon Fan and Western Amazon) [@jardineSporopolleninChemistryIts2021] is quite different in composition to both our fossil pollen, and any extant acetolysed pollen  samples, in which the large peak at 1600 and 1190 peak are not clearly visible.
+These fossil Poaceae were treated differently than our pollen and the extant Poaceae set, a combination of heavy liquid separation, Na~4~P~2~O~7~ and lignite as opposed to mainly acetolysis.
 The simililarities of the acetolysed extant set to our acetolysed fossil set and dissimilarities to pollen treated with a different combination of extractions (Amazon Western and Fan), suggest that chemical treatment of the pollen has a large effect on the chemistry of pollen, both modern and fossil origin.
@@
 
-## Differences between fossil pollen
+<!-- @jardineImpactOxidationSpore2015 showed limited effects on pollen chemistry of cold acetolysis if it was shorter than 10 minutes. -->
+<!-- Hot acetolysis (90&deg;C), which is standard procedure in palynology [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989], had an effect from the first minute.  -->
 
-### Anomalies in non acetolysed pollen spectra
+## Other effects observed in the fossil pollen spectra
+Spectra from the fossil-pollen samples showed differences between both sediment-core location and fossil-pollen treatment.
+These differences are likely due to a combination of sample processing, chemical variations, and also sediment-core processing effects.
+For example, we observed a broad peak at 1090-1100 cm^-1^ in the non acetolysed spectra of all three sediment cores (Fig. \@ref(fig:meanspectra) b, c and d).
+In addition, samples from Tiefer See also shows peak at 1630 cm^-1^, but these peaks are not present in either modern- or in the acetolysed-pollen samples.
 
-We see a broad peak at 1090-1100 cm^-1^ in the non acetolysed spectra of all three cores (Fig. \@ref(fig:meanspectra) b, c and d).
-This peak is not present in either modern pollen or acetolysed pollen, additionally Tiefer See shows a similar peak at 1630 cm^-1^.
-The 1090 cm^-1^ peak dissappears after acetolysis from all spectra as does the 1630 cm^-1^ peak in Tiefer See.
-This peak (1090 cm^-1^) in the non acetolysed samples is most likely biogenic silica, when compared to this biogenic silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016] and hydrated silica spectra [@SilicaATRIRSpectrum].
-During the measurements with the microscope, we applied our extracted pollen as droplets onto ZnSe slides and waited for the water to evaporate.
-After the droplets had dried we observed a crystalline layer over the same area that covered all particles and pollen (ref Foto in SM).
-From the chemical spectra, it is most likely that it is dissolved silica that crystallised, although the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.
+Comparisons of FTIR spectra from biogenic-silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016] and hydrated silica spectra [@SilicaATRIRSpectrum] reveal that the peak 1090 cm^-1^ in the non acetolysed samples is most likely biogenic silica, which may have precipitated onto the onto ZnSe slides as the sediment residues being prepared for analysis.
+During the measurements with the microscope, we applied our extracted pollen as droplets onto ZnSe slides and waited for the water to evaporate and we observed a crystalline layer over the same area that covered all particles and pollen **(ref Foto in SM)**.
+Interestinglyl, it appeared that the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.
+**[Could we say something here like- for pollen chemistry research this may present a challenge if we are aiming to directly compare spectra from unprocessed pollen samples since it may create differences in the spectra beween modern and fossil.
+I also dont know if it could interfere with the other sporopollenin peaks.
+A solution would be to apply some baseline correction procedures on this peak first, whihc may be important since even if we found it in these core, the quantity of biogenic silica may vary between different sediment cores from different lakes and regions.
+This is something to be explored]**
 
 [FM: would probably have to replicate this with BSi in different concentrations and maybe see if thickness of this BSi crystall has an effect on the spectra]
-**I think you need to explain this peak in the context of palaeoecological research a bit- could you talk about the correction factors that might be needed if you were to apply this method everywhere?
-And that different sediment context might create different baseline spectra so understanding sedimentary context is essential?**
+**[what about the 1630 peak- unclear if this is a different peak or something else?]**
 
-### pollen chemical differences
+A similar contamination effect was observed in the Tiefer See samples (Figure XX).
+This is SPT.
+Need to add in discussion about importance of cleaning- maybe the sediment type meant more washing was required for this specific sediment type?
 
-Acetolysed pollen showed two dominant peaks in the carbohydrate region at 1040 and 1170 - 1190 cm^-1^, which are not visible in the modern spectra or the non acetolysed pollen.
-This change in the 1170 - 1190 cm^-1^ region is unique to acetolysized pollen and not visible in non acetolysed pollen and had higher regression coefficients that the surrounding wavenumbers (Fig.
-\@ref(fig:ordinations)).
-The 1040 cm^-1^ peak had high importance for the PLSR model differentiating acetolysed and non acetolysed pollen (Fig. \@ref(fig:ordinations) e and f).
+In addition to these potential contamination effects, some of the largest differences in fossil pollen samples were the results chemical differences between acetolysised and non-acetolysised samples.
 
-The presence of this new peak in the shoulder of the 1170 cm^-1^ peak, which can indicate sporopollenin and carbohydrates, suggest changes in the pollen chemistry.
-Presence of these two dominant peaks may be related to removal and or alteration of grain wall carbohydrates (cellulose etc) as well as changes to sporopollenin due to acetolysis.
-[FM: this would be answered by acetolysis of the fresh grains!!]
-In non acetolysed pollen the 1190 peak is not visible in any of the cores, which indicates that it is not due to aging/diageneis, but acetolysis.
-Unfortunately, the BSi peak is covering the area of the 1040 cm^-1^ peak in our DAL and TSK samples, while it is visible in the MFM spectra.
+**If this section is discussing differences between fossil vs fresh pollen then the next three paragraphs don;t fit.
+They are more discussing acetolysised and non acetolysed differences.
+As result the discussion feels like it is wandering, Therefore, I moved the differences between fossil pollen treatments up**
 
-Using the second PLSR model the acetolysed pollen is clearly separated from the non acetolysed pollen, with a much higher distance between samples for acetolysed pollen compared to non acetolysed pollen (Fig.
-\@ref(fig:ordinations) d).
-In addition, there is some differentiation between the cores, samples from Tiefer See are more separate from samples of the other two cores, which show more similar scores in the ordination.
+## Conclusions/ Applications for palaeoecological research
+<!-- Our study is a demonstration of microscope FTIR for pollen chemical methods using fossil pollen. -->
+<!-- We focused on the chemistry of *Pinus* pollen, because *Pinus* is one of a few taxa, that is quite common in sediment cores and largely dominated by 1 species, *Pinus sylvestris*. -->
+<!-- The pollen is also relatively large, which prevents any scattering problems found with smaller grains. -->
+<!-- We used *Pinus sylvestris* to explore chemical differences between modern and fossil material, because this has not been fully explored yet. -->
 
-This seems to indicate that *Pinus* pollen from DAL and MFM are chemically quite similar, while TSK *Pinus* pollen is different.
-In fact, acetolysed TSK pollen is even further apart from the rest of the acetolysed pollen than non acetolysed pollen is.
-There seem to be chemical differences between TSK and the other two cores that are present in non acetolysed and acetolysed pollen.
-The main difference of TSK pollen to the other cores is that TSK pollen is much younger than the other two cores.
-Our TSK samples are from the past 80 years, while MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene, with the youngest sample several 100 years older than TSK pollen.
-We cannot see any evidence that sediment type has an effect on the chemistry of the pollen, both TSK and MFM cores are varved cores, while DAL is not.
-This would suggest more difference between varved and non varved pollen if different types of diagenesis had an effect. 
-Any difference may be overshadowed by the large difference in age between the samples from these cores. 
-Our findings suggest a chemical difference in fossil pollen grains that is persistent even after acetolysis, which may be explained by the age of the pollen.
-Further studies on pollen from different sediment types at fixed age/depths may provide a clearer answer on the effects of sediment type and age on pollen chemistry.
-
-## Application for Palaeoecology
-
-Our study is a demonstration of microscope FTIR for pollen chemical methods using fossil pollen.
-We focused on the chemistry of *Pinus* pollen, because *Pinus* is one of a few taxa, that is quite common in sediment cores and largely dominated by 1 species, *Pinus sylvestris*.
-The pollen is also relatively large, which prevents any scattering problems found with smaller grains.
-We used *Pinus sylvestris* to explore chemical differences between modern and fossil material, because this has not been fully explored yet.
-
-Here we used a detector in combination with a FTIR microscope that allows capture of several spectra of single grains at the same time.
-This is an improvement on previous single grain measurements, where each grain had to be targeted individually.
+In this study we investigated the use of a detector in combination with a FTIR microscope that allows capture of several spectra in order to characterise the chemistry of pollen grains from a sediment sample.
+The approach has the potential to be an improvement on previous single-grain measurements using FTIR, where each grain is required to be targeted individually, or even manually extracted from the sediment matrix using a micro-pippette.
 The FPA-MCT detector on our FTIR microscope allows simultaneous capture of a larger area of the sample (XXX x XXX µm ), compared to other detectors (e.g.
@@
 An image with a spectra for each pixel is created, from which spectra of single grains were extracted.
+
+Our analysis revealed that
+
 For this study the spectra were extracted by hand, but this process is possible to automate.
@@
 
-# Conclusions
-
 # References

### 2021-06-15 manuscript.Rmd
went over ALs comments; expanded SPT/BSI paragraph

@@
 There is also evidence of environmental signals that are recorded in the chemistry of pollen grains, including temperature [@bagciogluMultiscaleVibrationalSpectroscopic2015], nutrients [@kendelChemicalAnalysisPollen2020; @diehnCombiningChemicalInformation2020],  and UV-B radiation [@rozemaUVBAbsorbanceUVB2001; @blokkerDevelopmentProxySurface2005; @fraserUVBAbsorbingPigments2011; @jardineSheddingLightSporopollenin2017].
-Thus, in recent years, the study of pollen-grain chemistry has gained increasing focus in palaeoecological research since it is possible that chemical variations observed in sporopollenin may be used as proxies for past environmental conditions, with a particular focus on using sporopollenin chemistry to reconstruct past changes in UV-B radiation **[ref: here my 2019 paper would be a good one to cite]** and for improving the taxonomic precision available for sediment records [@dellannaPollenDiscriminationClassification2009; @zimmermannAnalysisAllergenicPollen2016].
+Thus, in recent years, the study of pollen-grain chemistry has gained increasing focus in palaeoecological research since it is possible that chemical variations observed in sporopollenin may be used as proxies for past environmental conditions, with a particular focus on using sporopollenin chemistry to reconstruct past changes in UV-B radiation [@seddonFossilPollenSpores2019] and for improving the taxonomic precision available for sediment records [@dellannaPollenDiscriminationClassification2009; @zimmermannAnalysisAllergenicPollen2016].
 
@@
 @zimmermannCharacterizationPollenVibrational2010, @lahlaliATRFTIRSpectroscopyReveals2014) **[is it possible to have one very short description here about what these techniques entail- e.g.
-the resolution of the laser?]**, have the advantage of reducing scattering effects from measurements due to the analysis of large sample quantities.
+the resolution of the laser?
+FM: spectra resolution is the same as micro afaik, but signal to noise is better]**, have the advantage of reducing scattering effects from measurements due to the analysis of large sample quantities, improving signal to noise compared to microscopes.
 As a result, FTIR spectra obtained from bulk pollen samples have been used identify chemical differences at the genus, and in some cases, species level [@muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020].
@@
 An alternative approach involves the use of FTIR microscopes, which require manual targeting of small areas of the sample slides to record spectra of single grains or groups of pollen grains.
-Because of their higher spatial resolution, FTIR-microscopes are better suited to sediment samples as a result of the capacity to generate FTIR spectra from individual grains, although spectral-processing procedures may be required to reduce the scattering effects when obtaining spectra from individual pollen grains **[ref]**.
+Because of their higher spatial resolution, FTIR-microscopes are better suited to sediment samples as a result of the capacity to generate FTIR spectra from individual grains, although spectral-processing procedures may be required to reduce the scattering effects when obtaining spectra from individual pollen grains [@zimmermannVibrationalMicrospectroscopyEnables2015; @zimmermannAnalysisAllergenicPollen2016].
 Nevertheless, microscopic FTIR techniques have now been applied to both modern and fossil-pollen grains preserved in sediments.
 The approach has been successfully used to show that chemical differences are found between pollen on genus and species level [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018;  @zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019], and microscopic-FTIR techniques have also been used to observe chemical differences to reconstruct changes in UV-B absorbing compounds in different periods across Earth's history [@fraserUVBAbsorbingPigments2011; @jardinePollenSporesBiological2016; @jardineProxyReconstructionUltravioletB2019; @jardineSheddingLightSporopollenin2017].
-**[Does this sentence belong in the next paragraph]**. Some microscopes are equipped with automated stages, where multiple locations on the slide can be targeted in sequence to record spectra.
 
 One of key challenges faced by chemical palynologists is the time taken to extract and analyse fossil-pollen samples.
+Some microscopes are equipped with automated stages, where multiple point locations on the slide can be targeted manually to record spectra automatically.
 However, recent developments in technologies related to FTIR may allow for easier and faster capture of spectra of pollen grains.
-A focal plane array detector (FPA) allows capture of an image and matrix of spectra over the same sample area instead of a single spectra at a specific point/area [ref].
+A focal plane array detector (FPA) allows capture of an image and matrix of spectra over the same sample area instead of a single spectra at a specific point/area [@zimmermannAnalysisAllergenicPollen2016;@zimmermannChemicalCharacterizationIdentification2018; @zimmermannVibrationalMicrospectroscopyEnables2015].
 This enables the extraction of spectra from several single grains from the same area on the sample without the need to relocate the detector of the microscope or move the slide, and is therefore of potentially high relevance for palaeoecological research.
@@
 In this study we aimed to explore the use of an FPA detector to extract FTIR spectra from fossil-pollen *Pinus* grains.
-We aim to develop a workflow which enables comparison of fossil-pollen grains taken from three sediment-core locations **[TABLE 1]**, and compare the FTIR spectra generated using the FPA detector from both modern and fossil-pollen samples.
+We aim to develop a workflow which enables comparison of fossil-pollen grains taken from three sediment-core locations (Table \@ref(tab:sites)), and compare the FTIR spectra generated using the FPA detector from both modern and fossil-pollen samples.
 Pollen was extracted from the sediment cores using two different approaches (density separation using sodium polytungstate, and acetolysis) and the fossil pollen samples measured using the combined FTIR microscope with FPA detector were compared with modern-pollen samples.
@@
 
+Our modern spectra are very similar to *Pinus sylvestris* spectra measured using bulk sample [@bagciogluMultiscaleVibrationalSpectroscopic2015; @zimmermannCharacterizationPollenVibrational2010] and FTIR microscopy using single point measurements [@zimmermannChemicalCharacterizationIdentification2018].
+The same characteristic peaks described above are visible in all modern spectra captured with bulk and FTIR microscopy, e.g. 1745, 1515, 1170 and 1000 to 1250 cm^-1^.
 **The main important discussion point to add here is to note the main differences that you see in the fossil and modern Pine pollen spectra- taken using bulk analyses, and using micro-FTIR.
+FM: I dont know of any bulk fossil spectra of Pinus published.
 Then you address the issue of comparing the FPA detector.
@@
 
-**[one thing to think about- could it be that the SPT residues on TSK were actually influencing the acetolysis effect in some way?]**
+**[one thing to think about- could it be that the SPT residues on TSK were actually influencing the acetolysis effect in some way?
+FM: I think it is unlikely.
+The first step of acetolysis is acetic acid wash, which removes most of the water.
+I would assume that it removes most if not all of the spt there before the heavy acid step.]**
 
@@
 These differences are likely due to a combination of sample processing, chemical variations, and also sediment-core processing effects.
-For example, we observed a broad peak at 1090-1100 cm^-1^ in the non acetolysed spectra of all three sediment cores (Fig. \@ref(fig:meanspectra) b, c and d).
+For example, we observed a broad peak at 1090-1100 cm^-1^ in the non acetolysed spectra of all three sediment cores (Fig. \@ref(fig:meanspectra)).
 In addition, samples from Tiefer See also shows peak at 1630 cm^-1^, but these peaks are not present in either modern- or in the acetolysed-pollen samples.
@@
 During the measurements with the microscope, we applied our extracted pollen as droplets onto ZnSe slides and waited for the water to evaporate and we observed a crystalline layer over the same area that covered all particles and pollen **(ref Foto in SM)**.
-Interestinglyl, it appeared that the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.
-**[Could we say something here like- for pollen chemistry research this may present a challenge if we are aiming to directly compare spectra from unprocessed pollen samples since it may create differences in the spectra beween modern and fossil.
-I also dont know if it could interfere with the other sporopollenin peaks.
-A solution would be to apply some baseline correction procedures on this peak first, whihc may be important since even if we found it in these core, the quantity of biogenic silica may vary between different sediment cores from different lakes and regions.
-This is something to be explored]**
+Interestingly, it appeared that the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.
+For pollen chemistry research this may present a challenge if we are aiming to directly compare spectra from unprocessed pollen samples since it may create differences in the spectra beween modern and fossil.
+Fortunately, the BSi peak does not cover any of the sporopollenin peaks, but obscures the region 1070 to 1110 cm^-1^.
+The quantity of biogenic silica may vary between different sediment cores from different lakes and regions, which may influence the severety of the interference caused by BSi.
+A solution would be to apply correction procedures during preprocessing, or explore extendend washing iterations after extraction of the pollen to dilute BSi.
 
-[FM: would probably have to replicate this with BSi in different concentrations and maybe see if thickness of this BSi crystall has an effect on the spectra]
-**[what about the 1630 peak- unclear if this is a different peak or something else?]**
+A similar contamination effect was observed in the Tiefer See samples (Fig. \@ref(fig:meanspectra)) with the peak at peak at 1630 cm^-1^. 
+This peak belongs very likely to SPT, when compared to this sodium metatungstate spectra [@SodiumMetatungstateHydrate].
+It is possible that the three washes with water after decanting of the pollen fraction were not enough in this case, but absence of this peak in the other spectra seem to suggest that further washes may remove successfully remove SPT fully.
+This 1630 cm^-1^ peak was responsible for part of the separation between Tiefer See non acetolysed samples from the other cores, but does not change the larger differentiation of acetolysed Tiefer See samples (without the 1630 cm^-1^ peak) from the other cores.
 
-A similar contamination effect was observed in the Tiefer See samples (Figure XX).
-This is SPT.
-Need to add in discussion about importance of cleaning- maybe the sediment type meant more washing was required for this specific sediment type?
-
-In addition to these potential contamination effects, some of the largest differences in fossil pollen samples were the results chemical differences between acetolysised and non-acetolysised samples.
-
-**If this section is discussing differences between fossil vs fresh pollen then the next three paragraphs don;t fit.
-They are more discussing acetolysised and non acetolysed differences.
-As result the discussion feels like it is wandering, Therefore, I moved the differences between fossil pollen treatments up**
+In addition to these potential contamination effects, some of the largest differences in fossil pollen samples were the results of chemical differences between acetolysised and non-acetolysised samples.
 
@@
 
-Third, as reported in other studies, FTIR measurements of small particles introduces scattering artifacts into the spectra.
+Third, as reported in other studies, FTIR measurements of small particles (5-25 µm) introduces scattering artifacts into the spectra [@lukacsRecoveryAbsorbanceSpectra2015; @zimmermannVibrationalMicrospectroscopyEnables2015].
 A number of pollen species are affected by this scattering because of their size.

### 2021-06-15 manuscript.Rmd
added SM figure for BSi

@@
 ## Other effects observed in the fossil pollen spectra
+
+```{r SM, echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="90%"}
+knitr::include_graphics(here("figures", "SM_BSI.png"))
+```
+
 Spectra from the fossil-pollen samples showed differences between both sediment-core location and fossil-pollen treatment.
@@
 Comparisons of FTIR spectra from biogenic-silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016] and hydrated silica spectra [@SilicaATRIRSpectrum] reveal that the peak 1090 cm^-1^ in the non acetolysed samples is most likely biogenic silica, which may have precipitated onto the onto ZnSe slides as the sediment residues being prepared for analysis.
-During the measurements with the microscope, we applied our extracted pollen as droplets onto ZnSe slides and waited for the water to evaporate and we observed a crystalline layer over the same area that covered all particles and pollen **(ref Foto in SM)**.
+During the measurements with the microscope, we applied our extracted pollen as droplets onto ZnSe slides and waited for the water to evaporate and we observed a crystalline layer over the same area that covered all particles and pollen (Fig.
+SM \@ref(fig:SM)).
 Interestingly, it appeared that the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.

### 2021-06-16 manuscript.Rmd
wrote conclusion paragraph

@@
 the resolution of the laser?
-FM: spectra resolution is the same as micro afaik, but signal to noise is better]**, have the advantage of reducing scattering effects from measurements due to the analysis of large sample quantities, improving signal to noise compared to microscopes.
+FM: spectra resolution is the same as micro afaik, but signal to noise is better]**, have the advantage of reducing scattering effects from measurements due to the analysis of large sample quantities, which improves signal to noise compared to microscopes.
 As a result, FTIR spectra obtained from bulk pollen samples have been used identify chemical differences at the genus, and in some cases, species level [@muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020].
@@
 
-```{r SM, echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="90%"}
+```{r SM, echo=FALSE, fig.cap="Image captured from FTIR microscope of sample slide Dalmutladdo 60 cm depth.
+Pinus pollen is visible as well as debris and dried residue, presumably BSi.", message=FALSE, warning=FALSE, out.width="90%"}
 knitr::include_graphics(here("figures", "SM_BSI.png"))
@@
 
-Our analysis revealed that
-
 For this study the spectra were extracted by hand, but this process is possible to automate.
@@
 A refinement of the right extraction medium density in combination with sieving of the samples to remove non pollen organic material may be one approach.
-@caffreyUseLithiumHeteropolytungstate2013 and @leipeTestingPerformanceSodium2019 have reported similar counts of pollen grains to conventional extraction methods using density separation.
+@caffreyUseLithiumHeteropolytungstate2013 and @leipeTestingPerformanceSodium2019 have reported similar counts of pollen grains to conventional extraction methods using density separation, but also mention that ideal extraction density varies from core to core.
 
-Second, the quality of the optical images captured by our microscope was much worse than convential LM, which made it difficult to identify smaller pollen grains on the slide from other organic debris.
+Second, the quality of the optical images captured by our microscope was much worse than convential LM, which made it difficult to identify smaller pollen grains on the slide from other organic debris, e.g.
+*Betula*/*Corylus*, *Alnus* or *Quercus*.
 An objective with higher magnification (20x) may improve this aspect of using a FTIR microscope.
-The FPA detector in combination with automated localisation of potential pollen grains based on their chemical signature may also help in solving this part of the problem.
+The FPA detector in combination with automated localisation of potential pollen grains based on their chemical signature may also help in solving this part of the problem, because identification by eye won't be necessary.
+Such
 
@@
 
+Fossil pollen from three sediment cores and modern material was analysed by FTIR microspectroscopy using a new detector that allows simultaneous capture of multiple single grain spectra.
+We explore the potential of a workflow for chemical palynology by analysing modern and fossil *Pinus* pollen.
+Our methods reproduces modern *Pinus* spectra well, compared to bulk analysis and other microspectroscopy approaches.
+Acetolysed fossil pollen was compared to non-acetolysed fossil pollen, which revealed differences in their chemical composition (e.g.
+changes in the 1000-1200 cm^-2^ region) that are most similar to reported acetolysed spectra from other taxa, suggesting that acetolysis alters fossil pollen chemistry.
+We also report contaminations observed in non-acetolysized samples and the applications of high density liquid extraction for FTIR microspectroscopy and Palynology.
+
 # References

### 2021-06-21 R/plotting.R
facetwrap for mean figure

@@ -11,19 +11,19 @@ plot_mean_spectra = function(data, subset_sel, spec_id, id.vars = c("ids", "trea
   pdata$variable = as.numeric(as.character(pdata$variable))
   
   #offset
-  pdata = offset_spectra(pdata, group, "value", seq(0, by=max(pdata$value)/5, length.out = length(levels(pdata[,group])))) #
+  pdata = offset_spectra(pdata, group, "value", rep(c(0, .2), length(levels(pdata[,group])))) #
   
   # mean data over groups
   pdata_mean = pdata %>%
     group_by_at(c(group, "variable")) %>%
-    summarise(mean = mean(value))
+    summarise(mean = mean(value), Label = first(Label))
   
   
   g = ggplot(pdata, aes_string("variable", "value", color = group))+
-    geom_line(aes(group = ids), alpha = .15) + #mean spectra
-    geom_line(data = pdata_mean, aes_string("variable", "mean", group = group), inherit.aes = F, size = 1) + #all spectra
+    geom_line(aes(group = ids), alpha = .15) + facet_wrap(vars(Label), ncol = 2) +#mean spectra
+    geom_line(data = pdata_mean, aes_string("variable", "mean", group = group), inherit.aes = F, size = 1) + facet_wrap(vars(Label), ncol = 2) +#all spectra
     scale_x_reverse(name = "wavenumbers",limits = c(1800,900), breaks = seq(1800,900,-100)) +
-    scale_y_continuous(name = "Absorbance (arbitrary units)",breaks = NULL) +
+    scale_y_continuous(name = "Absorbance (arbitrary units)",limits = c(0,0.5), breaks = NULL) +
     scale_color_manual(name = NULL,
                        values = c("#e66101","#ff6c01",
                                   "#ac7d43","#fdb863",
@@ -36,15 +36,17 @@ plot_mean_spectra = function(data, subset_sel, spec_id, id.vars = c("ids", "trea
     theme_bw() + 
     theme(legend.position = "bottom",
           axis.ticks.y = element_blank(),
-          axis.text.y = element_blank()) +
+          axis.text.y = element_blank()
+          ) +
     guides(colour = guide_legend(override.aes = list(alpha = 1)))
   
   return(g)
 }  
 
 offset_spectra = function(data, variable, value, offset) {
+  off2 = c(0.11,0.11,0.11,0.11,0.22,0.22,.55,.55)
   for (i in 1:length(levels(data[, variable]))) {
[…]

### 2021-06-21 manuscript.Rmd
facetwrap for mean figure
addressed Al comments; more adjustments to mean spectra plot

@@
 
-The study of plant microfossil (pollen and spores) and macrofossil (plant remains such as seeds) is an integral part of palaeoecology and palaeobotany.
+The study of plant microfossils (pollen and spores) is an integral part of palaeoecology and palaeobotany.
 Pollen records are an important tool to reconstruct past floristic, vegetation and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @noguesbravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013] and provide insights into ecological processes which operate on decadal, centennial and millennial timescales (e.g.
@@
 Variations in the chemical composition of sporopollenin can reflect phylogenetic differences between plant groups [@nieropQuantitativeAnalysisDiverse2019]. 
-There is also evidence of environmental signals that are recorded in the chemistry of pollen grains, including temperature [@bagciogluMultiscaleVibrationalSpectroscopic2015], nutrients [@kendelChemicalAnalysisPollen2020; @diehnCombiningChemicalInformation2020],  and UV-B radiation [@rozemaUVBAbsorbanceUVB2001; @blokkerDevelopmentProxySurface2005; @fraserUVBAbsorbingPigments2011; @jardineSheddingLightSporopollenin2017].
-Thus, in recent years, the study of pollen-grain chemistry has gained increasing focus in palaeoecological research since it is possible that chemical variations observed in sporopollenin may be used as proxies for past environmental conditions, with a particular focus on using sporopollenin chemistry to reconstruct past changes in UV-B radiation [@seddonFossilPollenSpores2019] and for improving the taxonomic precision available for sediment records [@dellannaPollenDiscriminationClassification2009; @zimmermannAnalysisAllergenicPollen2016].
+There is also evidence that environmental signals are recorded in the chemistry of pollen grains, including temperature [@bagciogluMultiscaleVibrationalSpectroscopic2015], nutrients [@kendelChemicalAnalysisPollen2020; @diehnCombiningChemicalInformation2020],  and UV-B radiation [@rozemaUVBAbsorbanceUVB2001; @blokkerDevelopmentProxySurface2005; @fraserUVBAbsorbingPigments2011; @jardineSheddingLightSporopollenin2017].
+Thus, in recent years, the study of pollen-grain chemistry has gained increasing focus in palaeoecological research since it is possible that chemical variations observed in sporopollenin may be used as proxies for past environmental conditions.
+There has been a particular focus on using sporopollenin chemistry to reconstruct past changes in UV-B radiation [@seddonFossilPollenSpores2019] and for improving the taxonomic precision available for sediment records [@dellannaPollenDiscriminationClassification2009; @zimmermannAnalysisAllergenicPollen2016].
 
@@
 1-2 mg) (e.g.
-@zimmermannCharacterizationPollenVibrational2010, @lahlaliATRFTIRSpectroscopyReveals2014) **[is it possible to have one very short description here about what these techniques entail- e.g.
-the resolution of the laser?
-FM: spectra resolution is the same as micro afaik, but signal to noise is better]**, have the advantage of reducing scattering effects from measurements due to the analysis of large sample quantities, which improves signal to noise compared to microscopes.
-As a result, FTIR spectra obtained from bulk pollen samples have been used identify chemical differences at the genus, and in some cases, species level [@muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020].
-However, these approaches may be less relevant for fossil sequences since the number of pollen grains analysed vastly exceeds the quantity which is practically available from a sediment-core sample.
+@zimmermannCharacterizationPollenVibrational2010, @lahlaliATRFTIRSpectroscopyReveals2014), have the advantage of reducing scattering effects from measurements due to the analysis of large sample quantities, which improves the signal-to-noise ratio in terms of spectra acquisition..
+As a result, FTIR spectra obtained from bulk pollen samples have been used to identify chemical differences at the genus, and in some cases, species level [@muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020].
+However, these approaches may be less relevant for pollen samples extracted from fossil sequences since the number of pollen grains analysed vastly exceeds the quantity which is practically available from a sediment-core sample.
 
@@
 Because of their higher spatial resolution, FTIR-microscopes are better suited to sediment samples as a result of the capacity to generate FTIR spectra from individual grains, although spectral-processing procedures may be required to reduce the scattering effects when obtaining spectra from individual pollen grains [@zimmermannVibrationalMicrospectroscopyEnables2015; @zimmermannAnalysisAllergenicPollen2016].
-Nevertheless, microscopic FTIR techniques have now been applied to both modern and fossil-pollen grains preserved in sediments.
+Nevertheless, microscopic FTIR techniques have now been applied to both modern- and fossil-pollen grains preserved in sediments.
 The approach has been successfully used to show that chemical differences are found between pollen on genus and species level [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018;  @zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019], and microscopic-FTIR techniques have also been used to observe chemical differences to reconstruct changes in UV-B absorbing compounds in different periods across Earth's history [@fraserUVBAbsorbingPigments2011; @jardinePollenSporesBiological2016; @jardineProxyReconstructionUltravioletB2019; @jardineSheddingLightSporopollenin2017].
 
-One of key challenges faced by chemical palynologists is the time taken to extract and analyse fossil-pollen samples.
+One of key challenges faced by chemical palynologists using micro-FTIR and other approaches is the time taken to extract and analyse fossil-pollen samples.
 Some microscopes are equipped with automated stages, where multiple point locations on the slide can be targeted manually to record spectra automatically.
@@
 This enables the extraction of spectra from several single grains from the same area on the sample without the need to relocate the detector of the microscope or move the slide, and is therefore of potentially high relevance for palaeoecological research.
-When combined with imaging processing technology, it potentially allows for more rapid identification and generation of FTIR spectra of pollen grains from sediment samples, but whilst FTIR spectroscopy using FPA detectors have been applied to pollen samples **in aerobiology [or others, check references]**, there has been limited investigation into their application on fossil-pollen residues taken from sediment cores.
+When combined with imaging processing technology, it potentially allows for more rapid identification and generation of FTIR spectra of pollen grains from sediment samples, but whilst FTIR spectroscopy using FPA detectors have been widely applied with microplastics [@loderEnzymaticPurificationMicroplastics2017; @machadoMicroplasticsEmergingThreat2018; @primpkeAutomatedApproachMicroplastics2017] and some studies on pollen [@chiappePollenGrainsFunctionalized2017], there has been limited investigation into their application on fossil-pollen residues taken from sediment cores.
 
-Palaeoecologists may also use a variety of chemical processing techniques in order to speed up analyses through the isolation and concentration of pollen from the sediment matrix [@ridingGuidePreparationProtocols2021].
-One of the most common methods for sporopollenin extraction involves an acetolysis treatment, which in a combination with HF and HCl treatments enables removal of organic materials, calcium carbonates and silicates, while preserving the morphological integrity of pollen grains[@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
+Palaeoecologists may also use a variety of chemical-processing techniques in order to speed-up analyses through the isolation and concentration of pollen from the sediment matrix [@ridingGuidePreparationProtocols2021].
+One of the most common methods for sporopollenin extraction involves an acetolysis treatment, with which a combination with HF and HCl treatments enables removal of organic materials, calcium carbonates and silicates, while preserving the morphological integrity of pollen grains [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
 Acetolysis has also often been used in combination with FTIR on pollen as a way to standardise between modern and fossil samples, [@jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or to study the effects of acid extractions [@dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000; @schulzeosthoffPhenolsIntegratedCompounds1987; @jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
-Indeed, although acetolysis treatments at low temperatures did not create major chemical differences in Lycopodium spores [@jardineImpactOxidationSpore2015], it is uncommon for low-temperature treatments to be applied to fossil-sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
-Such treatments may then result in chemically altering effects on fossil-pollen composition [@jardineSporopolleninChemistryIts2021], whilst variations in sediment types mean that different combinations of sample preparation techniques may be required for different sediment types [@ridingGuidePreparationProtocols2021].
-Given the large variations in FTIR spectra observed across taxa, and the variety of sediment types even from Holocene sequences in Europe, more studies are needed investigate the potential effects of the chemical preparation procedures used to isolate pollen-grains from different pollen types.
+Indeed, although acetolysis treatments at low temperatures did not create major chemical differences in *Lycopodium* spores [@jardineImpactOxidationSpore2015], it is uncommon for low-temperature treatments to be applied to fossil-sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
+Such treatments may then result in chemically altering effects on fossil-pollen composition [@jardineSporopolleninChemistryIts2021], whilst variations in sediment types mean that different combinations of sample-preparation techniques may be required for different sediment types [@ridingGuidePreparationProtocols2021].
+Given the large variations in FTIR spectra observed across taxa [@diehnCombiningChemicalInformation2020; @muthreichChemicalVariationsQuercus2020], and the variety of sediment types even from Holocene sequences in Europe, more studies are needed investigate the potential effects of the chemical preparation procedures used to isolate pollen-grains from different pollen types.
 
 In this study we aimed to explore the use of an FPA detector to extract FTIR spectra from fossil-pollen *Pinus* grains.
-We aim to develop a workflow which enables comparison of fossil-pollen grains taken from three sediment-core locations (Table \@ref(tab:sites)), and compare the FTIR spectra generated using the FPA detector from both modern and fossil-pollen samples.
-Pollen was extracted from the sediment cores using two different approaches (density separation using sodium polytungstate, and acetolysis) and the fossil pollen samples measured using the combined FTIR microscope with FPA detector were compared with modern-pollen samples.
+We develop a workflow which enables comparison of fossil-pollen grains taken from three sediment-core locations (Table \@ref(tab:sites)), and compare the FTIR spectra generated using the FPA detector from both modern- and fossil-pollen samples.
+Pollen was extracted from the sediment cores using two different approaches (density separation using sodium polytungstate and acetolysis) and the fossil-pollen samples measured using the combined FTIR microscope with FPA detector were compared with modern-pollen samples.
 Thus, in addition to testing the feasibility of the FPA detector on different sediment types, we were were also able to investigate how the inferred FTIR spectra in *Pinus* pollen can vary as a result of two different extraction procedures.
 
-[**here we need Boris to add a few sentences here to give the study more context**]- not sure this is needed now? unless he is not happy with FPA detector section?
 <!-- Chemical treatment... -->
@@
 The pollen was extracted from three sediment cores: Dalmuttlado in Norway [@bjuneHoloceneVegetationClimate2004], Tiefer See [@theuerkaufEffectsChangesLand2015]  and Meerfelder Maar [@brauerHighResolutionSediment1999] in Germany.
-The three sediment cores were selected because they contain *Pinus sylvestris* pollen types, spanning three different periods in the late Quaternary (Table \@ref(tab:sites)).
-*Pinus* pollen in these cores has been identified as *Pinus sylvestris* in the original publication.
+The three sediment cores were selected because they contain *Pinus sylvestris*-pollen types, spanning three different periods in the late Quaternary (Table \@ref(tab:sites)).
+*Pinus* pollen in these cores has been described as most likely representing *Pinus sylvestris* in Dalmutladdo and Tiefer See, while it is described as *Pinus* in Meerfelder Maar record.
 In this study we focus on *Pinus* pollen for two reasons.
@@
 is a shade-intolerant, high pollen producer, and so a number of studies have proposed that *Pinus* spp.
-pollen may be a useful taxon to use as a biological indicator for reconstructing past changes of UV-B radiation [ref- willis 2011, seddon 2017, seddon 2021].
+pollen may be a useful taxon to use as a biological indicator for reconstructing past changes of UV-B radiation [@jardineProxyReconstructionUltravioletB2020; @willisQuantificationUVFlux2011; @seddonPollenchemistryVariationsElevation2021].
 Second, although prominent in Quaternary sediment records from Europe, and despite it's high species diversity in Europe (**Check number of species**), *Pinus* is currently mostly only split into two different pollen types using fossil pollen, reflecting the two phylogenitic lineages [@beugLeitfadenPollenbestimmungFur2004; @gieseckePostglacialChangeFloristic2019].
@@
 Note, that there is no agreement on the density of the pollen exine, with reported values varying between <1.4 g cm^-3^ [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996], 1.88 cm^-3^ [@nakagawaDensemediaSeparationMore1998] and 2.2 cm^-3^  [@juvigneDensiteExinesQuelques1973].
-In our case the lighter fraction contained majority of the pollen grains.
+In our case the lighter fraction (<1.4 g cm^-3^) contained majority of the pollen grains, after checking both fractions.
 Sample residues were then washed three times in using distilled H~2~O to remove any excess SPT, resulting in sample residues with a higher pollen concentration compared to other organic and inorganic materials on the slide.
-Since SPT is a chemically inert salt, it was expected that, after washingm this treatment should have minimal chemical effect on the chemical spectra inferred from the fossil-pollen grains.
+Since SPT is a chemically inert salt, it was expected that, after washing this treatment should have minimal chemical effect on the chemical spectra inferred from the fossil-pollen grains.
 
 We then applied an acetolysis treatment [e.g. @erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989] to one subset of the samples following washing and extracting using SPT.
-Although acetolysis treatments at low temperatures did not create major chemical differences in Lycopodium spores [@jardineImpactOxidationSpore2015], it is uncommon for low temperature treatments to be applied to fossil sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
-Under these conditions it is possible that acetolysis may alter the chemical composition of sporopollenin, and so in many cases it may be preferable to use alternative approaches to extract fossil pollen grains before analysis using FTIR.
-Thus, incorporating the acetolysis step on this analysis allows us to compare the effects of pollen-grain spectra from both treated and untreated pollen grains.
+The acetolysis was performed at 105&deg;C for 5 minutes.
+Incorporating the acetolysis step on this analysis allows us to compare the effects of pollen-grain spectra from both treated and untreated pollen grains.
 
@@
 The accompanied visible images were obtained by a charge coupled device (CCD) camera coupled to the microscope.
-Single-grain pollen grain spectra were extracted from the FTIR images, by averaging all pixel spectra in the image area of the pollen grain. 
+Single-grain FTIR spectra were extracted from the FTIR images, by averaging all pixel spectra in the image area of the pollen grain. 
 Data acquisition and instrument control was performed using the OPUS 7.5 software (Bruker Optik GmbH, Germany).
@@
 The extracted pollen was applied to ZnSe slides and a background spectra was measured before each sample on an empty area of the slide.
-The pollen treated with SPT and acetolysis were applied onto the slide suspended in water and dried out before measurement under a fume hood.
+The pollen treated with SPT and acetolysis were suspended in water, applied to the slide and then dried out before measurement under a fume hood.
 The extracted sediment samples contained other small material than pollen (e.g. diatom fragments, organic material fragments, etc).
@@
 The spectra were preprocessed before analysis consisted of smoothing and calculation of second derivative and normalization.
-We used the Savitzky Golay (SG) algorithm [@savitzkySmoothingDifferentiationData1964] as implemented by the extended multiplicative signal correction (EMSC) package in R [@lilandPackageEMSC2017] and used second degree polynomial and windowsize five as parameters.
+We used the Savitzky Golay (SG) algorithm [@savitzkySmoothingDifferentiationData1964] as implemented by the extended multiplicative signal correction (EMSC) package in R [@lilandPackageEMSC2017] and used second degree polynomial and window size five as parameters.
 The second derivative spectra was then normalized using a EMSC model with a linear component [@lilandPackageEMSC2017].
@@
 
-We used two PLSR models to explore the chemical differences between modern and fossil grains (model 1), and acetolyzed and non acetolyzed fossil grains (model 2).
+We used two PLSR models to explore the chemical differences between modern and fossil grains (model 1), and acetolysised and non-acetolysised fossil grains (model 2).
 For these models the explanatory variables were the pre-processed spectra and the response variables were hot-one encoded response variables (Model 1: modern/fossil; Model 2: acet./ non acet.).
-By using PLSR we can maximize the distance between responses and identify chemical differences between the responses by looking at the regression coefficients.
+By using PLSR we aimed to maximize the distance between responses and identify chemical differences between the responses by looking at the regression coefficients.
 
@@
 
-```{r meanspectra, echo=FALSE, fig.cap="FRE", message=FALSE, warning=FALSE, out.width="90%"}
+```{r meanspectra, echo=FALSE, fig.cap="Mean spectra summarised by core and chemical treatment", message=FALSE, warning=FALSE, out.width="90%"}
 knitr::include_graphics(here("figures", "mean_plots.png"))
@@
 Our analyses revealed clear differences between modern- and fossil-pollen grains (Fig. \@ref(fig:meanspectra)).
-The prominent and characteristic sporopollenin peaks (1600, 1510 and 1170 cm^-1^) are visible in the fossil pollen, while other parts of the spectra show substantial differences.
-Most notably, the lipid peak at 1745 cm^-1^, caused by triglycerides, is not visible in the fossil pollen grains, neither in the non-acetolysed and the acetolysed fossil pollen.
-There are also several differences in the "fingerprint" region (1500 to 1800 cm^-1^) and carbohydrate region (1000 to 1250 cm^-1^) between modern and fossil pollen grains.
+Although the prominent and characteristic sporopollenin peaks (1600, 1510 and 1170 cm^-1^) are visible in both the modern and fossil pollen, other parts of the spectra show substantial differences.
+Most notably, the lipid peak at 1745 cm^-1^, caused by triglycerides, is not visible in the fossil-pollen grains, neither in the non-acetolysed and the acetolysed fossil pollen.
+There are also several differences in the "fingerprint" region (1800 - 1500 cm^-1^) and carbohydrate region (1250 - 1000 cm^-1^) between modern and fossil pollen grains.
 
-These chemical differences are also clearly demonstrated in the results of the PLSR model (Fig \@ref(fig:ordinations) a).
+These chemical differences are also clearly demonstrated in the results of the PLSR (Fig \@ref(fig:ordinations) a).
 Fossil pollen are clearly separated from the modern pollen along the first component of the PLSR model.
-The regression coefficients reflect our observations from the visual inspection and show that lipid  and protein peaks (1745 and 1500 - 1800 cm^-1^) have disappeared and there are substantial changes in the carbohydrate region (1000 to 1250 cm^-1^) in fossil pollen (Fig \@ref(fig:ordinations) b and c).
-The region 1000 - 1200 cm^-1^ is now dominated in acetolysed fossil pollen by two peaks and its shoulders at 1040 and 1170 cm^-1^.
+The regression coefficients reflect our observations from the visual inspection and show that lipid  and protein peaks (1745 and 1800 - 1500 cm^-1^) are no longer present and there are also substantial changes in the carbohydrate region (1250 - 1000 cm^-1^) in the fossil pollen (Fig \@ref(fig:ordinations) b and c).
+The region 1200 - 1000 cm^-1^ is now dominated in acetolysed fossil pollen by two peaks and its shoulders at 1170 and 1040 cm^-1^.
 
-Our modern spectra are very similar to *Pinus sylvestris* spectra measured using bulk sample [@bagciogluMultiscaleVibrationalSpectroscopic2015; @zimmermannCharacterizationPollenVibrational2010] and FTIR microscopy using single point measurements [@zimmermannChemicalCharacterizationIdentification2018].
-The same characteristic peaks described above are visible in all modern spectra captured with bulk and FTIR microscopy, e.g. 1745, 1515, 1170 and 1000 to 1250 cm^-1^.
-**The main important discussion point to add here is to note the main differences that you see in the fossil and modern Pine pollen spectra- taken using bulk analyses, and using micro-FTIR.
-FM: I dont know of any bulk fossil spectra of Pinus published.
-Then you address the issue of comparing the FPA detector.
-I think the story si that the fresh spectra you have here are quite similar to spectra in other studies, correct?**
+Our modern spectra are very similar to *Pinus sylvestris* spectra measured using FTIR-bulk sample methodologies [@bagciogluMultiscaleVibrationalSpectroscopic2015; @zimmermannCharacterizationPollenVibrational2010] and FTIR microscopy using single point measurements [@zimmermannChemicalCharacterizationIdentification2018].
+The same characteristic peaks described above are visible in all modern spectra captured with bulk and FTIR microscopy, e.g. 1745, 1515, 1170 and 1250 - 1000 cm^-1^.
 
@@
 pollen grains.
-For example, acetolysed pollen showed two dominant peaks in the carbohydrate region at 1040 and 1170 - 1190 cm^-1^, which are not visible in the modern spectra or the non acetolysed pollen.
-This change in the 1170 - 1190 cm^-1^ region is unique to acetolysized pollenand had higher regression coefficients that the surrounding wavenumbers (Fig. \@ref(fig:ordinations)).
+For example, acetolysed pollen showed two dominant peaks in the carbohydrate region at 1190 - 1170 and 1040 cm^-1^, which are not visible in the modern spectra or the non acetolysed pollen.
+This change in the 1190 - 1170 cm^-1^ region is unique to acetolysized pollenand had higher regression coefficients that the surrounding wavenumbers (Fig. \@ref(fig:ordinations)).
 The 1040 cm^-1^ peak also had high importance for the PLSR model differentiating acetolysed and non acetolysed pollen (Fig. \@ref(fig:ordinations) e and f).
@@
 The presence of these two dominant peaks may be related to the removal and or alteration of grain-wall carbohydrates (cellulose etc), as well as changes to chemical composition of sporopollenin due to the acetolysis treatment.
-[FM: this would be answered by acetolysis of the fresh grains!!]
-Since the 1190  cm^-1^ peak is not visible in any of the samples from the  non-acetolysed pollen, this likely indicates that the acetolysis treatment is creating this peak.
+Since the 1190 cm^-1^ peak is not visible in any of the samples from the  non-acetolysed pollen, this likely indicates that the acetolysis treatment is creating this peak in the FTIR spectra.
 <!-- Unfortunately, the BSi peak is covering the area of the 1040 cm^-1^ peak in our DAL and TSK samples, while it is visible in the MFM spectra.  -->
@@
 In addition, there is some differentiation between the cores.
-Samples from Tiefer See are more separate from samples of the other two cores, which show more similar scores in the ordination.
+Samples from Tiefer See are more distinct from samples of the other two cores, which show more similar scores in the ordination.
 One interpretation of this is that *Pinus* pollen from DAL and MFM are chemically quite similar, while TSK *Pinus* pollen is different in chemical structure.
 Since it is likely that the main contributions to the pollen some in the TSK core is *Pinus sylvestris*, it is likely that the same species is represented by all three cores, and the main difference between of the TSK pollen is age.
-The TSK samples are from the past 80 years, while MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene, with the youngest sample several 100 years older than TSK pollen.
+The TSK samples are from the past 80 years, while MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene, with the youngest sample more than 100 years older than TSK pollen.
 <!-- We cannot see any evidence that sediment type has an effect on the chemistry of the pollen, both TSK and MFM cores are varved cores, while DAL is not. -->
@@
 <!-- Any difference may be overshadowed by the large difference in age between the samples from these cores.  -->
-This finding could suggest a chemical difference in fossil pollen grains that is persistent even after acetolysis, which may be explained by the age of the pollen.
+This finding could suggest a chemical difference in fossil-pollen grains that is persistent even after acetolysis, which may be explained by the age of the pollen.
+@jardineSporopolleninChemistryIts2021 have shown chemical differences between pollen on much larger timescales, where modern pollen had more in common with quarternary pollen than deep time samples.
+Our results also indicate chemical difference with age, but on a much shorter timescale.
 Further studies on pollen from different sediment types at fixed age/depths may provide a clearer answer on the effects of sediment type and age on pollen chemistry.
 
-**[one thing to think about- could it be that the SPT residues on TSK were actually influencing the acetolysis effect in some way?
-FM: I think it is unlikely.
-The first step of acetolysis is acetic acid wash, which removes most of the water.
-I would assume that it removes most if not all of the spt there before the heavy acid step.]**
+If we compare our *Pinus* spectra to other acetolysised fossil *Pinus* pollen, we observed some differences.
+@jardineProxyReconstructionUltravioletB2020 extracted *Pinus* pollen from a varved sediment core using acetolysis and HF to study the UV-B response in *Pinus* pollen chemistry.
+Their spectra are similar to ours in the 1800 - 1500 cm^-1^ region, a prominent peak at 1515 cm^-1^ and peaks at 1700 and 1600 cm^-1^.
+The region between 1500 - 1300 cm^-1^ is fairly flat in our spectra, while a strong peak is visible at 1450 in @jardineProxyReconstructionUltravioletB2020 spectra.
+Lastly, the region between 1300 - 900 cm^-1^ in our spectra mainly consists of two large peaks at 1200 and 1040 cm^-1^, while the *Pinus* pollen from @jardineProxyReconstructionUltravioletB2020 shows large peaks at 1250, 1150, 1100 and 1010.
+Direct comparisons between spectra from different spectrometers are difficult, because peak heights are not always directly comparable.
+Nevertheless, we do see substantial differences in the 1500 - 1000 cm^-1^ region, while there are similarities in the 1800 - 1500 cm^-1^ region
 
-In general there are similarities between the FTIR spectra from the acetolysed *Pinus* pollen from our study and in the acetolysed pollen grains of other taxa
-For example, the carbohydrate region at 1000 - 1200 cm^-1^ our acetolysised *Pinus* pollen samples is most similar to the acetolysed extant Poaceae set of @jardineSporopolleninChemistryIts2021 and acetolyzed herbarium *Nitrariaceae* pollen [@woutersenNovelApproachStudy2018], where two peaks at 1040 and 1190/1170 are visible and similar to our samples.
+Additionally, there are similarities between the FTIR spectra from the acetolysed *Pinus* pollen from our study to the acetolysed pollen grains of other taxa.
+For example, the carbohydrate region at 1200 - 1000 cm^-1^ in our acetolysised *Pinus* pollen samples is most similar to the acetolysed extant Poaceae set of @jardineSporopolleninChemistryIts2021 and acetolyzed herbarium *Nitrariaceae* pollen [@woutersenNovelApproachStudy2018], where two peaks at 1040 and 1190/1170 are visible and similar to our samples.
 However, spectra from a modern Poaceae data set (sampled from the the Amazon Fan and Western Amazon) [@jardineSporopolleninChemistryIts2021] is quite different in composition to both our fossil pollen, and any extant acetolysed pollen  samples, in which the large peak at 1600 and 1190 peak are not clearly visible.
-These fossil Poaceae were treated differently than our pollen and the extant Poaceae set, a combination of heavy liquid separation, Na~4~P~2~O~7~ and lignite as opposed to mainly acetolysis.
-The simililarities of the acetolysed extant set to our acetolysed fossil set and dissimilarities to pollen treated with a different combination of extractions (Amazon Western and Fan), suggest that chemical treatment of the pollen has a large effect on the chemistry of pollen, both modern and fossil origin.
+These fossil Poaceae were treated differently than our pollen and the extant Poaceae set, a combination of density separation, Na~4~P~2~O~7~ and lignite as opposed to mainly acetolysis.
+
+Taken together, the simililarities of the acetolysed extant set to our acetolysed fossil set and dissimilarities to pollen treated with a different combination of extractions (Amazon Western and Fan), suggest that chemical extraction methods have a large effect on the chemistry of pollen, both modern and fossil origin.
 This is important to consider going forward for the application of IR methods in palaeoecology, where extraction methods are often dependent on sediment type.
-The effects of these extraction methods need to be explored on modern and fossil material side by side to see if there are links to bridge the transition from modern to fossil pollen.
-It is crucial to understand the changes pollen undergoes during diagenesis, e.g. for the purposes of artificially fossilizing grains to curate reference collections of fossil grains of known taxa.
+Further studies, in which the effects of these extraction methods need to be explored on modern and fossil material side by side to see if there are methods that can bridge the transition from modern to fossil pollen.
+It seems from the comparison of our spectra to other published fossil spectra, that chemical treatments have a large effect on pollen chemistry and may affect fossil and modern grains differently.
+Chemical treatments that can mimic diagenesis will be crucial to understanding the chemical changes of pollen under diagenesis and help to create species specific reference collections of artificial fossilized modern pollen grains.
+This would enable us to compare fossil pollen spectra with fossilized pollen spectra of known species for identification.
 
@@
 
-## Other effects observed in the fossil pollen spectra
+## Other effects observed in the fossil-pollen spectra
 
@@
 These differences are likely due to a combination of sample processing, chemical variations, and also sediment-core processing effects.
-For example, we observed a broad peak at 1090-1100 cm^-1^ in the non acetolysed spectra of all three sediment cores (Fig. \@ref(fig:meanspectra)).
-In addition, samples from Tiefer See also shows peak at 1630 cm^-1^, but these peaks are not present in either modern- or in the acetolysed-pollen samples.
+For example, we observed a broad peak at 1100 - 1090 cm^-1^ in the non acetolysed spectra of all three sediment cores (Fig. \@ref(fig:meanspectra)).
+In addition, samples from Tiefer See also shows a peak at 1630 cm^-1^, but these peaks are not present in either modern- or in the acetolysed-pollen samples.
 
@@
 Interestingly, it appeared that the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.
-For pollen chemistry research this may present a challenge if we are aiming to directly compare spectra from unprocessed pollen samples since it may create differences in the spectra beween modern and fossil.
-Fortunately, the BSi peak does not cover any of the sporopollenin peaks, but obscures the region 1070 to 1110 cm^-1^.
+For pollen-chemistry research this may present a challenge if we are aiming to directly compare spectra from unprocessed pollen samples since it may create differences in the spectra beween modern and fossil grains.
+Fortunately, the BSi peak does not cover any of the sporopollenin peaks, but obscures the region 1110 - 1070 cm^-1^.
 The quantity of biogenic silica may vary between different sediment cores from different lakes and regions, which may influence the severety of the interference caused by BSi.
@@
 
-A similar contamination effect was observed in the Tiefer See samples (Fig. \@ref(fig:meanspectra)) with the peak at peak at 1630 cm^-1^. 
+An additional contamination effect was observed in the Tiefer See samples (Fig. \@ref(fig:meanspectra)) with the peak at peak at 1630 cm^-1^. 
 This peak belongs very likely to SPT, when compared to this sodium metatungstate spectra [@SodiumMetatungstateHydrate].
 It is possible that the three washes with water after decanting of the pollen fraction were not enough in this case, but absence of this peak in the other spectra seem to suggest that further washes may remove successfully remove SPT fully.
-This 1630 cm^-1^ peak was responsible for part of the separation between Tiefer See non acetolysed samples from the other cores, but does not change the larger differentiation of acetolysed Tiefer See samples (without the 1630 cm^-1^ peak) from the other cores.
+This 1630 cm^-1^ peak was responsible for part of the separation between Tiefer See non-acetolysised samples from the other cores, but does not change the larger differentiation of acetolysed Tiefer See samples (without the 1630 cm^-1^ peak) from the other cores.
 
-In addition to these potential contamination effects, some of the largest differences in fossil pollen samples were the results of chemical differences between acetolysised and non-acetolysised samples.
-
-## Conclusions/ Applications for palaeoecological research
+## Applications for palaeoecological research
 <!-- Our study is a demonstration of microscope FTIR for pollen chemical methods using fossil pollen. -->
@@
 An image with a spectra for each pixel is created, from which spectra of single grains were extracted.
+In general the FTIR spectra obtained using this approach created similar spectra to spectra of *Pinus* pollen from bulk approaches, and the method was also sensitive to differences in chemical treatments on the fossil-pollen grains.
 
-For this study the spectra were extracted by hand, but this process is possible to automate.
-Approaches using deep learing have shown to be capable of extracting objects from larger images, using deep neural networks specialised on image segmentation/object location [@renFasterRCNNRealTime2015].
-Conceptually the problem of locating pollen on a FTIR image is similar to locating an object on a optical image and these deep learning approaches can be applied.
+Thus, moving forward we see potential in the application of the FDA detector in chemical palaeoecology. In particular, one potential future area of research lies in the possibility for automation. 
+Although in this study the spectra were extracted by hand, approaches using deep learning have shown to be capable of extracting objects from larger images, using deep neural networks specialised on image segmentation/object location [@renFasterRCNNRealTime2015].
+Conceptually the problem of locating pollen on a FTIR image is similar to locating an object on a optical image.
+@primpkeAutomatedApproachMicroplastics2017 devloped an automated approach to identify microplastic with a FPA detector.
 @kangSinglecellClassificationFoodborne2020 used deep neural network to classify bacteria based on images and spectral information, while @grosserueschkampSpatialMolecularResolution2017 used random forests for segmentation and identification of malignant tumor in histological thin sections using FTIR microscope equipped with a FPA detector (715 x 715 µm).
@@
 
-We encountered a number of challenges in this study that need to be addressed for the application of IR microscopes in palaeoecology.
+Nevertheless, we encountered a number of challenges that will need to be addressed if the experimental set-up we used here can be more comprensively applied in palaeoecology.
 First, we found that the concentration of pollen in our samples was very variable and posed a major bottleneck for our analysis with the FPA detector.
-Capturing spectra from a larger area is only advantageous over single point measurements if several grains are located in the FTIR image.
+Capturing spectra from a larger area is only advantageous over single-point measurements if several grains are located in the FTIR image.
 This was not always possible, we found, because the pollen was very scarce in some samples.
 
-Acetolysis is the conventional method to extract pollen in high concentration, but as discussed before, we believe there is value in extracting non acetolysed pollen from sediments, because of the substantial differences in pollen chemistry we found between acetolysed and non acetolysed pollen.
-We believe that density separation is a viable alternative [@caffreyUseLithiumHeteropolytungstate2013; @leipeTestingPerformanceSodium2019; @regnellPreparativeCentrifugationNew1996], but may have been one possible cause for our low concentration in some samples.
-The reported density of pollen is variable [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996; @nakagawaDensemediaSeparationMore1998; @juvigneDensiteExinesQuelques1973] and our extraction was done on the lower end of these reported densities (1.44 g cm ^-3^).
-We opted for this density because test extractions using a higher density (1.98 g cm ^-3^), also introduced even more organic material to clutter our samples.
+Acetolysis is the conventional method to extract pollen in high concentration in Quarternary pollen samples, but the results presented here demonstrate the value in extracting non-acetolysised pollen from sediments, because of the substantial differences in pollen chemistry we found between acetolysised and non-acetolysised pollen.
+We believe that density separation is a viable alternative to acetolysis and other acid or oxidizing extractions [@caffreyUseLithiumHeteropolytungstate2013; @leipeTestingPerformanceSodium2019; @regnellPreparativeCentrifugationNew1996].
+The reported density of pollen is variable and may be dependent on age of the sample [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996; @nakagawaDensemediaSeparationMore1998; @juvigneDensiteExinesQuelques1973].
+After testing multiple densities, 1.44 g cm^-3^ yielded the most pollen from our samples.
+This density may be optimal for *Pinus* in our samples, but other pollen may have been discarded with the heavier fraction.
+Our results show that there is a need to refine extraction methods, which minimize chemical alteration to the pollen grain.
 A refinement of the right extraction medium density in combination with sieving of the samples to remove non pollen organic material may be one approach.
@@
 The FPA detector in combination with automated localisation of potential pollen grains based on their chemical signature may also help in solving this part of the problem, because identification by eye won't be necessary.
-Such
 
@@
 
-Fossil pollen from three sediment cores and modern material was analysed by FTIR microspectroscopy using a new detector that allows simultaneous capture of multiple single grain spectra.
-We explore the potential of a workflow for chemical palynology by analysing modern and fossil *Pinus* pollen.
+In summary, fossil pollen from three sediment cores and modern material was analysed by FTIR microspectroscopy using a new detector that allows simultaneous capture of multiple single grain spectra.
 Our methods reproduces modern *Pinus* spectra well, compared to bulk analysis and other microspectroscopy approaches.
 Acetolysed fossil pollen was compared to non-acetolysed fossil pollen, which revealed differences in their chemical composition (e.g.
-changes in the 1000-1200 cm^-2^ region) that are most similar to reported acetolysed spectra from other taxa, suggesting that acetolysis alters fossil pollen chemistry.
-We also report contaminations observed in non-acetolysized samples and the applications of high density liquid extraction for FTIR microspectroscopy and Palynology.
+changes in the 1200 - 1000 cm^-2^ region) that are most similar to reported acetolysed spectra from other taxa, suggesting that acetolysis alters fossil pollen chemistry.
+We also report contaminations observed in non-acetolysized samples and the applications of high density liquid extraction for FTIR microspectroscopy and palynology.
 

### 2021-06-21 R/plotting.R

@@ -86,11 +86,11 @@ vlines = data.frame(wn = c(1745, 1710, 1605, 1515, 1170, 1040),
 
 l = plot_mean_spectra(data, expression(species == "Pinus"), spec_id = "emsc", group = "off", id.vars = c("ids", "treatment", "Label", "depth", "age", "type", "off"))
 
-l = l + ylim(.05, 2.3) +
+l = l + #ylim(.05, 2.3) +
   geom_vline(data = vlines, aes(xintercept = wn), linetype = "dashed", alpha = .3) +
-  geom_text(data = vlines, aes(wn, y, label = label), inherit.aes = F)
+  geom_text(data = vlines, aes(wn, y, label = label), angle = 90, inherit.aes = F)
 
-ggsave(here("figures","mean_plots.png"), l, width = 17, height = 25,units = "cm")
+ggsave(here("figures","mean_plots.png"), l, width = 17, height = 17,units = "cm")
 saveRDS(l, here("data","output","mean_plots_whole.rds"))
 
 ### creating Plots

### 2021-06-25 manuscript.Rmd
Als comments

@@
 Fourier Transform Infrared (FTIR) spectroscopy has become one of the most popular methods for characterising the chemical variations present within pollen and spores in the modern environment.
-Bulk FTIR techniques, which involve the analysis of large quantities of modern-pollen samples (e.g.
-1-2 mg) (e.g.
+Bulk FTIR techniques, which involve the analysis of large quantities of modern-pollen samples (approximately 1-2 mg) (e.g.
 @zimmermannCharacterizationPollenVibrational2010, @lahlaliATRFTIRSpectroscopyReveals2014), have the advantage of reducing scattering effects from measurements due to the analysis of large sample quantities, which improves the signal-to-noise ratio in terms of spectra acquisition..
@@
 Nevertheless, microscopic FTIR techniques have now been applied to both modern- and fossil-pollen grains preserved in sediments.
-The approach has been successfully used to show that chemical differences are found between pollen on genus and species level [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018;  @zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019], and microscopic-FTIR techniques have also been used to observe chemical differences to reconstruct changes in UV-B absorbing compounds in different periods across Earth's history [@fraserUVBAbsorbingPigments2011; @jardinePollenSporesBiological2016; @jardineProxyReconstructionUltravioletB2019; @jardineSheddingLightSporopollenin2017].
+The approach has been successfully used to show that chemical differences are found between pollen at the genus and even species level [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018;  @zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019], and microscopic-FTIR techniques have also been used to observe chemical differences to reconstruct changes in UV-B absorbing compounds in different periods across Earth's history [@fraserUVBAbsorbingPigments2011; @jardinePollenSporesBiological2016; @jardineProxyReconstructionUltravioletB2019; @jardineSheddingLightSporopollenin2017].
 
@@
 This enables the extraction of spectra from several single grains from the same area on the sample without the need to relocate the detector of the microscope or move the slide, and is therefore of potentially high relevance for palaeoecological research.
-When combined with imaging processing technology, it potentially allows for more rapid identification and generation of FTIR spectra of pollen grains from sediment samples, but whilst FTIR spectroscopy using FPA detectors have been widely applied with microplastics [@loderEnzymaticPurificationMicroplastics2017; @machadoMicroplasticsEmergingThreat2018; @primpkeAutomatedApproachMicroplastics2017] and some studies on pollen [@chiappePollenGrainsFunctionalized2017], there has been limited investigation into their application on fossil-pollen residues taken from sediment cores.
+When combined with imaging processing technology, it may allow for more rapid identification and generation of FTIR spectra of pollen grains from sediment samples, but whilst FTIR spectroscopy using FPA detectors have been widely applied with microplastics [@loderEnzymaticPurificationMicroplastics2017; @machadoMicroplasticsEmergingThreat2018; @primpkeAutomatedApproachMicroplastics2017] and some studies on pollen [@chiappePollenGrainsFunctionalized2017], there has been limited investigation into their application on fossil-pollen residues taken from sediment cores.
 
 Palaeoecologists may also use a variety of chemical-processing techniques in order to speed-up analyses through the isolation and concentration of pollen from the sediment matrix [@ridingGuidePreparationProtocols2021].
-One of the most common methods for sporopollenin extraction involves an acetolysis treatment, with which a combination with HF and HCl treatments enables removal of organic materials, calcium carbonates and silicates, while preserving the morphological integrity of pollen grains [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
+One of the most common methods for sporopollenin extraction involves an acetolysis treatment, which, alongside a combination of HF and HCl treatments, enables removal of organic materials, calcium carbonates and silicates, while preserving the morphological integrity of pollen grains [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
 Acetolysis has also often been used in combination with FTIR on pollen as a way to standardise between modern and fossil samples, [@jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or to study the effects of acid extractions [@dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000; @schulzeosthoffPhenolsIntegratedCompounds1987; @jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
@@
 Such treatments may then result in chemically altering effects on fossil-pollen composition [@jardineSporopolleninChemistryIts2021], whilst variations in sediment types mean that different combinations of sample-preparation techniques may be required for different sediment types [@ridingGuidePreparationProtocols2021].
-Given the large variations in FTIR spectra observed across taxa [@diehnCombiningChemicalInformation2020; @muthreichChemicalVariationsQuercus2020], and the variety of sediment types even from Holocene sequences in Europe, more studies are needed investigate the potential effects of the chemical preparation procedures used to isolate pollen-grains from different pollen types.
+Given the large variations in FTIR spectra observed across taxa [@diehnCombiningChemicalInformation2020; @muthreichChemicalVariationsQuercus2020], and the variety of sediment types even from Holocene sequences in Europe, more studies are needed investigate the potential effects of the chemical preparation procedures used to isolate pollen grains from different pollen types.
 
@@
 We develop a workflow which enables comparison of fossil-pollen grains taken from three sediment-core locations (Table \@ref(tab:sites)), and compare the FTIR spectra generated using the FPA detector from both modern- and fossil-pollen samples.
-Pollen was extracted from the sediment cores using two different approaches (density separation using sodium polytungstate and acetolysis) and the fossil-pollen samples measured using the combined FTIR microscope with FPA detector were compared with modern-pollen samples.
+Pollen was extracted from the sediment cores using two different approaches (density separation using sodium polytungstate and acetolysis) and the fossil-pollen samples were measured using a combined FTIR microscope with FPA detector.
+We also compared the FTIR spectra obtained from fossil-pollen grains with  modern *Pinus* pollen samples.
 Thus, in addition to testing the feasibility of the FPA detector on different sediment types, we were were also able to investigate how the inferred FTIR spectra in *Pinus* pollen can vary as a result of two different extraction procedures.
@@
 pollen may be a useful taxon to use as a biological indicator for reconstructing past changes of UV-B radiation [@jardineProxyReconstructionUltravioletB2020; @willisQuantificationUVFlux2011; @seddonPollenchemistryVariationsElevation2021].
-Second, although prominent in Quaternary sediment records from Europe, and despite it's high species diversity in Europe (**Check number of species**), *Pinus* is currently mostly only split into two different pollen types using fossil pollen, reflecting the two phylogenitic lineages [@beugLeitfadenPollenbestimmungFur2004; @gieseckePostglacialChangeFloristic2019].
+Second, although prominent in Quaternary sediment records from Europe, and despite it's high species diversity in Europe (**Check number of species**), *Pinus* is currently mostly only split into two different pollen types using fossil pollen, reflecting the two phylogenetic lineages [@beugLeitfadenPollenbestimmungFur2004; @gieseckePostglacialChangeFloristic2019].
 Thus, identifying chemical differences in fossil *Pinus* pollen may also allow for identification of crytpic *Pinus* species from the pollen types observed in the fossil record.
@@
 Prior to chemical characterisation using FTIR, all sediment samples were processed in two ways.
-First, the samples were extracted using density separation using dissolved Sodium Polytungstate (SPT) [@kampsAgentSeparationDissolved1985] at 1.43 g cm^-3^ as separation medium.
+First, the samples were extracted using density separation using dissolved sodium polytungstate (SPT) [@kampsAgentSeparationDissolved1985] at 1.43 g cm^-3^ as a separation medium.
 During density separation a known volume of SPT (density 1.43 g cm^-3^) solution was added to the sample and homogenized/vortexed, followed by centrifugation.
-After centrifugation, particulates/material heavier than 1.43 g cm^-3^ accumulate of the bottom of the sample vial, whilst the lighter fraction was decanted and collected.
+After centrifugation, particulates heavier than 1.43 g cm^-3^ accumulate of the bottom of the sample vial, whilst the lighter fraction was decanted and collected.
 Note, that there is no agreement on the density of the pollen exine, with reported values varying between <1.4 g cm^-3^ [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996], 1.88 cm^-3^ [@nakagawaDensemediaSeparationMore1998] and 2.2 cm^-3^  [@juvigneDensiteExinesQuelques1973].
@@
 Sample residues were then washed three times in using distilled H~2~O to remove any excess SPT, resulting in sample residues with a higher pollen concentration compared to other organic and inorganic materials on the slide.
-Since SPT is a chemically inert salt, it was expected that, after washing this treatment should have minimal chemical effect on the chemical spectra inferred from the fossil-pollen grains.
+Since SPT is a chemically inert salt, it was expected that, after washing, this treatment should have minimal chemical effect on the chemical spectra inferred from the fossil-pollen grains.
 
@@
 Most notably, the lipid peak at 1745 cm^-1^, caused by triglycerides, is not visible in the fossil-pollen grains, neither in the non-acetolysed and the acetolysed fossil pollen.
-There are also several differences in the "fingerprint" region (1800 - 1500 cm^-1^) and carbohydrate region (1250 - 1000 cm^-1^) between modern and fossil pollen grains.
+There are also several differences in the "fingerprint" region (1800 - 1500 cm^-1^) and carbohydrate region (1250 - 1000 cm^-1^) between modern- and fossil-pollen grains.
 
@@
 Fossil pollen are clearly separated from the modern pollen along the first component of the PLSR model.
-The regression coefficients reflect our observations from the visual inspection and show that lipid  and protein peaks (1745 and 1800 - 1500 cm^-1^) are no longer present and there are also substantial changes in the carbohydrate region (1250 - 1000 cm^-1^) in the fossil pollen (Fig \@ref(fig:ordinations) b and c).
-The region 1200 - 1000 cm^-1^ is now dominated in acetolysed fossil pollen by two peaks and its shoulders at 1170 and 1040 cm^-1^.
-
-Our modern spectra are very similar to *Pinus sylvestris* spectra measured using FTIR-bulk sample methodologies [@bagciogluMultiscaleVibrationalSpectroscopic2015; @zimmermannCharacterizationPollenVibrational2010] and FTIR microscopy using single point measurements [@zimmermannChemicalCharacterizationIdentification2018].
-The same characteristic peaks described above are visible in all modern spectra captured with bulk and FTIR microscopy, e.g. 1745, 1515, 1170 and 1250 - 1000 cm^-1^.
+The regression coefficients reflect our observations from the visual inspection and show that lipid  and protein peaks (1745 and 1800 - 1500 cm^-1^) are no longer present.
+There are also substantial changes in the carbohydrate region (1250 - 1000 cm^-1^) in the fossil pollen compared to the modern (Fig \@ref(fig:ordinations) b and c).
+The region 1250 - 1000 cm^-1^ is now dominated in acetolysed fossil pollen by two peaks and its shoulders at 1170 and 1040 cm^-1^.
 
 ## Effects of acetolysis on the fossil-pollen samples
+
 In addition to the differences between fresh and fossil pollen samples, our analysis revealed large differences in chemical variations as a result of the acetolysis treatment on the fossil *Pinus* spp.
@@
 For example, acetolysed pollen showed two dominant peaks in the carbohydrate region at 1190 - 1170 and 1040 cm^-1^, which are not visible in the modern spectra or the non acetolysed pollen.
-This change in the 1190 - 1170 cm^-1^ region is unique to acetolysized pollenand had higher regression coefficients that the surrounding wavenumbers (Fig. \@ref(fig:ordinations)).
-The 1040 cm^-1^ peak also had high importance for the PLSR model differentiating acetolysed and non acetolysed pollen (Fig. \@ref(fig:ordinations) e and f).
+This change in the 1190 - 1170 cm^-1^ region is unique to acetolysized pollen and had higher regression coefficients that the surrounding wavenumbers (Fig. \@ref(fig:ordinations)).
+The 1040 cm^-1^ peak also had high importance for the PLSR model differentiating acetolysed and non-acetolysed fossil pollen (Fig. \@ref(fig:ordinations) e and f).
 
-The presence of this new peak in the shoulder of the 1170 cm^-1^ wavenumbers in the acetolysised pollen grains can be indicative of sporopollenin and carbohydrates, and suggests an overall change in the pollen chemistry of the sample.
+The presence of this new peak in the 1170 cm^-1^ shoulder in the spectra from the acetolysised pollen grains suggests an overall change in the pollen chemistry of the sample.
 The presence of these two dominant peaks may be related to the removal and or alteration of grain-wall carbohydrates (cellulose etc), as well as changes to chemical composition of sporopollenin due to the acetolysis treatment.
-Since the 1190 cm^-1^ peak is not visible in any of the samples from the  non-acetolysed pollen, this likely indicates that the acetolysis treatment is creating this peak in the FTIR spectra.
+Since the 1190 cm^-1^ peak is not visible in any of the samples from the  non-acetolysed pollen, this likely indicates that it is processes related to the acetolysis treatment that are creating this new peaks in the FTIR spectra.
 <!-- Unfortunately, the BSi peak is covering the area of the 1040 cm^-1^ peak in our DAL and TSK samples, while it is visible in the MFM spectra.  -->
@@
 One interpretation of this is that *Pinus* pollen from DAL and MFM are chemically quite similar, while TSK *Pinus* pollen is different in chemical structure.
-Since it is likely that the main contributions to the pollen some in the TSK core is *Pinus sylvestris*, it is likely that the same species is represented by all three cores, and the main difference between of the TSK pollen is age.
-The TSK samples are from the past 80 years, while MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene, with the youngest sample more than 100 years older than TSK pollen.
+Since it is likely that the main contributions to the pollen some in the TSK core is *Pinus sylvestris*, it is likely that the same species is represented by all three cores, and one of the main differences between the TSK pollen and the other two locations is sample age.
+Whilst the TSK samples are from the past 80 years, MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene, with the youngest sample likely more than 100 years older than TSK pollen.
 <!-- We cannot see any evidence that sediment type has an effect on the chemistry of the pollen, both TSK and MFM cores are varved cores, while DAL is not. -->
@@
 This finding could suggest a chemical difference in fossil-pollen grains that is persistent even after acetolysis, which may be explained by the age of the pollen.
-@jardineSporopolleninChemistryIts2021 have shown chemical differences between pollen on much larger timescales, where modern pollen had more in common with quarternary pollen than deep time samples.
-Our results also indicate chemical difference with age, but on a much shorter timescale.
+Interestingly, @jardineSporopolleninChemistryIts2021 also found differences in FTIR spectra with different sample ages, although they found that that modern-pollen spectra had more in common with spectra from Quaternary pollen, compared to pollen from deeper time samples.
 Further studies on pollen from different sediment types at fixed age/depths may provide a clearer answer on the effects of sediment type and age on pollen chemistry.
-
-If we compare our *Pinus* spectra to other acetolysised fossil *Pinus* pollen, we observed some differences.
-@jardineProxyReconstructionUltravioletB2020 extracted *Pinus* pollen from a varved sediment core using acetolysis and HF to study the UV-B response in *Pinus* pollen chemistry.
-Their spectra are similar to ours in the 1800 - 1500 cm^-1^ region, a prominent peak at 1515 cm^-1^ and peaks at 1700 and 1600 cm^-1^.
-The region between 1500 - 1300 cm^-1^ is fairly flat in our spectra, while a strong peak is visible at 1450 in @jardineProxyReconstructionUltravioletB2020 spectra.
-Lastly, the region between 1300 - 900 cm^-1^ in our spectra mainly consists of two large peaks at 1200 and 1040 cm^-1^, while the *Pinus* pollen from @jardineProxyReconstructionUltravioletB2020 shows large peaks at 1250, 1150, 1100 and 1010.
-Direct comparisons between spectra from different spectrometers are difficult, because peak heights are not always directly comparable.
-Nevertheless, we do see substantial differences in the 1500 - 1000 cm^-1^ region, while there are similarities in the 1800 - 1500 cm^-1^ region
-
-Additionally, there are similarities between the FTIR spectra from the acetolysed *Pinus* pollen from our study to the acetolysed pollen grains of other taxa.
-For example, the carbohydrate region at 1200 - 1000 cm^-1^ in our acetolysised *Pinus* pollen samples is most similar to the acetolysed extant Poaceae set of @jardineSporopolleninChemistryIts2021 and acetolyzed herbarium *Nitrariaceae* pollen [@woutersenNovelApproachStudy2018], where two peaks at 1040 and 1190/1170 are visible and similar to our samples.
-However, spectra from a modern Poaceae data set (sampled from the the Amazon Fan and Western Amazon) [@jardineSporopolleninChemistryIts2021] is quite different in composition to both our fossil pollen, and any extant acetolysed pollen  samples, in which the large peak at 1600 and 1190 peak are not clearly visible.
-These fossil Poaceae were treated differently than our pollen and the extant Poaceae set, a combination of density separation, Na~4~P~2~O~7~ and lignite as opposed to mainly acetolysis.
-
-Taken together, the simililarities of the acetolysed extant set to our acetolysed fossil set and dissimilarities to pollen treated with a different combination of extractions (Amazon Western and Fan), suggest that chemical extraction methods have a large effect on the chemistry of pollen, both modern and fossil origin.
-This is important to consider going forward for the application of IR methods in palaeoecology, where extraction methods are often dependent on sediment type.
-Further studies, in which the effects of these extraction methods need to be explored on modern and fossil material side by side to see if there are methods that can bridge the transition from modern to fossil pollen.
-It seems from the comparison of our spectra to other published fossil spectra, that chemical treatments have a large effect on pollen chemistry and may affect fossil and modern grains differently.
-Chemical treatments that can mimic diagenesis will be crucial to understanding the chemical changes of pollen under diagenesis and help to create species specific reference collections of artificial fossilized modern pollen grains.
-This would enable us to compare fossil pollen spectra with fossilized pollen spectra of known species for identification.
-
-<!-- @jardineImpactOxidationSpore2015 showed limited effects on pollen chemistry of cold acetolysis if it was shorter than 10 minutes. -->
-<!-- Hot acetolysis (90&deg;C), which is standard procedure in palynology [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989], had an effect from the first minute.  -->
 
@@
 These differences are likely due to a combination of sample processing, chemical variations, and also sediment-core processing effects.
-For example, we observed a broad peak at 1100 - 1090 cm^-1^ in the non acetolysed spectra of all three sediment cores (Fig. \@ref(fig:meanspectra)).
-In addition, samples from Tiefer See also shows a peak at 1630 cm^-1^, but these peaks are not present in either modern- or in the acetolysed-pollen samples.
-
-Comparisons of FTIR spectra from biogenic-silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016] and hydrated silica spectra [@SilicaATRIRSpectrum] reveal that the peak 1090 cm^-1^ in the non acetolysed samples is most likely biogenic silica, which may have precipitated onto the onto ZnSe slides as the sediment residues being prepared for analysis.
+For example, we observed a broad peak at 1100 - 1090 cm^-1^ in the non-acetolysed spectra of all three sediment cores (Fig. \@ref(fig:meanspectra)).
+Comparisons of FTIR spectra from biogenic-silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016] and hydrated silica spectra [@SilicaATRIRSpectrum] reveal that the peak at 1090 cm^-1^ in the non acetolysed samples is most likely biogenic silica, which may have precipitated onto the onto ZnSe slides as the sediment residues being prepared for analysis.
 During the measurements with the microscope, we applied our extracted pollen as droplets onto ZnSe slides and waited for the water to evaporate and we observed a crystalline layer over the same area that covered all particles and pollen (Fig.
@@
 Fortunately, the BSi peak does not cover any of the sporopollenin peaks, but obscures the region 1110 - 1070 cm^-1^.
-The quantity of biogenic silica may vary between different sediment cores from different lakes and regions, which may influence the severety of the interference caused by BSi.
-A solution would be to apply correction procedures during preprocessing, or explore extendend washing iterations after extraction of the pollen to dilute BSi.
+The quantity of biogenic silica may vary between different sediment cores from different lakes and regions, which may influence the severity of the interference caused by BSi.
+A solution would be to apply correction procedures during pre-processing, or explore extended-washing iterations after extraction of the pollen to dilute BSi.
 
-An additional contamination effect was observed in the Tiefer See samples (Fig. \@ref(fig:meanspectra)) with the peak at peak at 1630 cm^-1^. 
-This peak belongs very likely to SPT, when compared to this sodium metatungstate spectra [@SodiumMetatungstateHydrate].
+An additional contamination effect was observed in the Tiefer See samples (Fig. \@ref(fig:meanspectra)) with a broad peak at 1630 cm^-1^. 
+This peak belongs very likely to SPT  [see spectra from @SodiumMetatungstateHydrate].
+This 1630 cm^-1^ peak was responsible for part of the separation between Tiefer See non-acetolysised samples from the other cores, but does not change the larger differentiation of acetolysed Tiefer See samples (without the 1630 cm^-1^ peak) from the other cores.
 It is possible that the three washes with water after decanting of the pollen fraction were not enough in this case, but absence of this peak in the other spectra seem to suggest that further washes may remove successfully remove SPT fully.
-This 1630 cm^-1^ peak was responsible for part of the separation between Tiefer See non-acetolysised samples from the other cores, but does not change the larger differentiation of acetolysed Tiefer See samples (without the 1630 cm^-1^ peak) from the other cores.
+
+## Comparison of *Pinus* spectra with FTIR spectra from other studies
+
+The FTIR spectra obtained using the experimental set-up used in this study can be compared to FTIR spectra from *Pinus* pollen from other sources. 
+For example, the spectra obtained from our modern samples are very similar to *Pinus sylvestris* spectra measured using FTIR-bulk sample methodologies [@bagciogluMultiscaleVibrationalSpectroscopic2015; @zimmermannCharacterizationPollenVibrational2010] and FTIR microscopy using single-point measurements [@zimmermannChemicalCharacterizationIdentification2018].
+The same characteristic peaks described above are visible in all modern spectra captured with bulk and FTIR microscopy, e.g.
+1745, 1515, 1170 and 1250 - 1000 cm^-1^, and these results build on previous studies which suggest that overall potential for the use of the FPA detector in pollen research **[add references]**.
+
+Interestingly, when comparing our acetolysed *Pinus* spectra to other acetolysised fossil *Pinus* pollen, we observed some differences.
+For example, @jardineProxyReconstructionUltravioletB2020 extracted *Pinus* pollen from a varved sediment core using acetolysis and HF to study the UV-B response in *Pinus* pollen chemistry.
+But while their spectra are similar to ours in the 1800 - 1500 cm^-1^ region, with similar prominent peaks at 1515 cm^-1^, 1700 and 1600 cm^-1^, there are major differences in the region between 1500 - 1300 cm^-1^.
+The latter region is fairly flat in our spectra, while a strong peak is visible at 1450 in @jardineProxyReconstructionUltravioletB2020 spectra.
+Lastly, the region between 1300 - 900 cm^-1^ in our spectra mainly consists of two large peaks at 1200 and 1040 cm^-1^, while the *Pinus* pollen from @jardineProxyReconstructionUltravioletB2020 shows large peaks at 1250, 1150, 1100 and 1010.
+Direct comparisons between spectra from different spectrometers are difficult, because peak heights are not always directly comparable, but further research is required to understand the reasons between these differences between the acetolysed *Pinus* pollen from across the sample sets.
+
+Indeed, we actually found greater similarities between the FTIR spectra from the acetolysed *Pinus* pollen from our study to  acetolysed pollen grains of other pollen types.
+For example, the carbohydrate region at 1200 - 1000 cm^-1^ in our acetolysised *Pinus* pollen samples is most similar to an acetolysed *extant* Poaceae set of @jardineSporopolleninChemistryIts2021, and in addition to acetolyzed herbarium *Nitrariaceae* pollen [@woutersenNovelApproachStudy2018], where two peaks at 1040 and 1190/1170 are visible and similar to our samples.
+However, spectra from a fossil Poaceae data set (sampled from the the Amazon Fan and Western Amazon; AF: Na~4~P~2~O~7~, density sep., lignite, WA: HCl, Na~4~P~2~O~7~, density sep., acetolysis) [@jardineSporopolleninChemistryIts2021] is quite different in composition to both our fossil pollen, and any extant acetolysed-pollen samples, since the large peak at 1600 and 1190 peak are not clearly visible.
+
+Taken together, these differences in the effects of acetolysis acorss taxa suggest that the chemical extraction methods have a large effect on the chemistry of pollen, both modern and fossil origin.
+This is important to consider going forward for the application of IR methods in palaeoecology, where extraction methods are often dependent on sediment type.
+It would be useful to explore the effects of these extraction methods on modern and fossil material side-by-side, to investigate if there are methods that can bridge the transition from modern to fossil pollen.
+Chemical treatments that can mimic diagenesis will be crucial to understanding the chemical changes of pollen under diagenesis and help to create species specific reference collections of artificial fossilized modern pollen grains.
+This would enable us to compare fossil pollen spectra with fossilized pollen spectra of known species for identification.
+
+<!-- @jardineImpactOxidationSpore2015 showed limited effects on pollen chemistry of cold acetolysis if it was shorter than 10 minutes. -->
+<!-- Hot acetolysis (90&deg;C), which is standard procedure in palynology [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989], had an effect from the first minute.  -->
 
@@
 In this study we investigated the use of a detector in combination with a FTIR microscope that allows capture of several spectra in order to characterise the chemistry of pollen grains from a sediment sample.
-The approach has the potential to be an improvement on previous single-grain measurements using FTIR, where each grain is required to be targeted individually, or even manually extracted from the sediment matrix using a micro-pippette.
+The approach has the potential to be an improvement on previous single-grain measurements using FTIR, where each grain is required to be targeted individually, or even manually extracted from the sediment matrix using a micro-pipette.
 The FPA-MCT detector on our FTIR microscope allows simultaneous capture of a larger area of the sample (XXX x XXX µm ), compared to other detectors (e.g.
@@
 This was not always possible, we found, because the pollen was very scarce in some samples.
-
-Acetolysis is the conventional method to extract pollen in high concentration in Quarternary pollen samples, but the results presented here demonstrate the value in extracting non-acetolysised pollen from sediments, because of the substantial differences in pollen chemistry we found between acetolysised and non-acetolysised pollen.
-We believe that density separation is a viable alternative to acetolysis and other acid or oxidizing extractions [@caffreyUseLithiumHeteropolytungstate2013; @leipeTestingPerformanceSodium2019; @regnellPreparativeCentrifugationNew1996].
-The reported density of pollen is variable and may be dependent on age of the sample [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996; @nakagawaDensemediaSeparationMore1998; @juvigneDensiteExinesQuelques1973].
-After testing multiple densities, 1.44 g cm^-3^ yielded the most pollen from our samples.
-This density may be optimal for *Pinus* in our samples, but other pollen may have been discarded with the heavier fraction.
-Our results show that there is a need to refine extraction methods, which minimize chemical alteration to the pollen grain.
-A refinement of the right extraction medium density in combination with sieving of the samples to remove non pollen organic material may be one approach.
-@caffreyUseLithiumHeteropolytungstate2013 and @leipeTestingPerformanceSodium2019 have reported similar counts of pollen grains to conventional extraction methods using density separation, but also mention that ideal extraction density varies from core to core.
+Acetolysis is the conventional method to extract pollen in high concentration in Quaternary pollen samples, but the results presented here demonstrate the value in extracting non-acetolysised pollen from sediments, because of the substantial differences in pollen chemistry we found between acetolysised and non-acetolysised pollen (see above).
+Given the limited chemical influence on the fossil pollen grains, we believe that density separation is a viable alternative to acetolysis and other acid or oxidizing extractions [@caffreyUseLithiumHeteropolytungstate2013; @leipeTestingPerformanceSodium2019; @regnellPreparativeCentrifugationNew1996], alhtough the results in this paper clearly demonstrate the need to identify efficient dilution processes.
+Similarly, since the reported density of pollen is variable and may be dependent on age of the sample [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996; @nakagawaDensemediaSeparationMore1998; @juvigneDensiteExinesQuelques1973], it is possible that specific extraction densities and procedures may be required from core to core.
+<!-- After testing multiple densities, 1.44 g cm^-3^ yielded the most pollen from our samples. -->
+<!-- This density may be optimal for *Pinus* in our samples, but other pollen may have been discarded with the heavier fraction. -->
+<!-- Our results show that there is a need to refine extraction methods, which minimize chemical alteration to the pollen grain. -->
+<!-- A refinement of the right extraction medium density in combination with sieving of the samples to remove non pollen organic material may be one approach. -->
+<!-- @caffreyUseLithiumHeteropolytungstate2013 and @leipeTestingPerformanceSodium2019 have reported similar counts of pollen grains to conventional extraction methods using density separation, but also mention that ideal extraction density varies from core to core.
+-->
 
@@
 
-In summary, fossil pollen from three sediment cores and modern material was analysed by FTIR microspectroscopy using a new detector that allows simultaneous capture of multiple single grain spectra.
+## Summary
+
+Fossil pollen from three sediment cores and modern material was analysed by FTIR microspectroscopy using a new detector that allows simultaneous capture of multiple single grain spectra.
 Our methods reproduces modern *Pinus* spectra well, compared to bulk analysis and other microspectroscopy approaches.

### 2021-06-25 manuscript.Rmd
Al splitting Results and Discussion

@@
 
-# Results and Discussion
-
+# Results
 <!-- Our aim with this study was to explore the use of a new type of detector (FPA-MCT) in combination with a FTIR microscope to explore the chemical variations of fossil pollen. -->
@@
 \newpage
-
-## Fresh vs fossil pollen
-
+## Chemical differences between pollen samples
 Our analyses revealed clear differences between modern- and fossil-pollen grains (Fig. \@ref(fig:meanspectra)).
@@
 
-## Effects of acetolysis on the fossil-pollen samples
-
 In addition to the differences between fresh and fossil pollen samples, our analysis revealed large differences in chemical variations as a result of the acetolysis treatment on the fossil *Pinus* spp.
 pollen grains.
-For example, acetolysed pollen showed two dominant peaks in the carbohydrate region at 1190 - 1170 and 1040 cm^-1^, which are not visible in the modern spectra or the non acetolysed pollen.
+For example, acetolysed pollen showed two dominant peaks in the carbohydrate region at 1190 - 1170 and 1040 cm^-1^, which are not visible in the modern spectra or the non acetolysed pollen, and when using the second PLSR model, the acetolysed pollen is clearly separated from the non-acetolysed pollen, with a much greater distance in the ordination space between samples (Fig.
+\@ref(fig:ordinations) d).
 This change in the 1190 - 1170 cm^-1^ region is unique to acetolysized pollen and had higher regression coefficients that the surrounding wavenumbers (Fig. \@ref(fig:ordinations)).
@@
 
-The presence of this new peak in the 1170 cm^-1^ shoulder in the spectra from the acetolysised pollen grains suggests an overall change in the pollen chemistry of the sample.
-The presence of these two dominant peaks may be related to the removal and or alteration of grain-wall carbohydrates (cellulose etc), as well as changes to chemical composition of sporopollenin due to the acetolysis treatment.
-Since the 1190 cm^-1^ peak is not visible in any of the samples from the  non-acetolysed pollen, this likely indicates that it is processes related to the acetolysis treatment that are creating this new peaks in the FTIR spectra.
-<!-- Unfortunately, the BSi peak is covering the area of the 1040 cm^-1^ peak in our DAL and TSK samples, while it is visible in the MFM spectra.  -->
-
-Using the second PLSR model, the acetolysed pollen is clearly separated from the non-acetolysed pollen, with a much greater distance in the ordination space between samples for acetolysed pollen compared to non-acetolysed pollen (Fig.
-\@ref(fig:ordinations) d).
-In addition, there is some differentiation between the cores.
-Samples from Tiefer See are more distinct from samples of the other two cores, which show more similar scores in the ordination.
-One interpretation of this is that *Pinus* pollen from DAL and MFM are chemically quite similar, while TSK *Pinus* pollen is different in chemical structure.
-Since it is likely that the main contributions to the pollen some in the TSK core is *Pinus sylvestris*, it is likely that the same species is represented by all three cores, and one of the main differences between the TSK pollen and the other two locations is sample age.
-Whilst the TSK samples are from the past 80 years, MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene, with the youngest sample likely more than 100 years older than TSK pollen.
-<!-- We cannot see any evidence that sediment type has an effect on the chemistry of the pollen, both TSK and MFM cores are varved cores, while DAL is not. -->
-<!-- This would suggest more difference between varved and non varved pollen if different types of diagenesis had an effect.  -->
-<!-- Any difference may be overshadowed by the large difference in age between the samples from these cores.  -->
-This finding could suggest a chemical difference in fossil-pollen grains that is persistent even after acetolysis, which may be explained by the age of the pollen.
-Interestingly, @jardineSporopolleninChemistryIts2021 also found differences in FTIR spectra with different sample ages, although they found that that modern-pollen spectra had more in common with spectra from Quaternary pollen, compared to pollen from deeper time samples.
-Further studies on pollen from different sediment types at fixed age/depths may provide a clearer answer on the effects of sediment type and age on pollen chemistry.
-
-## Other effects observed in the fossil-pollen spectra
+## Contamination effects
 
@@
 
-Spectra from the fossil-pollen samples showed differences between both sediment-core location and fossil-pollen treatment.
-These differences are likely due to a combination of sample processing, chemical variations, and also sediment-core processing effects.
-For example, we observed a broad peak at 1100 - 1090 cm^-1^ in the non-acetolysed spectra of all three sediment cores (Fig. \@ref(fig:meanspectra)).
+In addition to the clear differences between samples, we also observed two potential contamination effects from the sediment core processing.
+First, in the Tiefer See non-acetolysed fossil pollen samples (Fig. \@ref(fig:meanspectra)) there was a broad peak at 1630 cm^-1^. 
+This 1630 cm^-1^ peak was responsible for part of the separation between Tiefer See non-acetolysised samples from the other cores, but does not change the larger differentiation of acetolysed Tiefer See samples (without the 1630 cm^-1^ peak) from the other cores, and the peak very likely indicates the presence of SPT [see spectra from @SodiumMetatungstateHydrate].
+Thus, it is possible that the three washes with water after decanting of the pollen fraction were not enough in this case, but absence of this peak in the other spectra seem to suggest that further washes may remove successfully remove SPT fully.
+
+Second, we also observed a broad peak at 1100 - 1090 cm^-1^ in the non-acetolysed spectra of all three sediment cores (Fig. \@ref(fig:meanspectra)).
 Comparisons of FTIR spectra from biogenic-silica spectra from diatoms [@alipourInfraredMicrospectroscopyBionanomaterials2016] and hydrated silica spectra [@SilicaATRIRSpectrum] reveal that the peak at 1090 cm^-1^ in the non acetolysed samples is most likely biogenic silica, which may have precipitated onto the onto ZnSe slides as the sediment residues being prepared for analysis.
@@
 Interestingly, it appeared that the acetolysis treatment removed this peak, which may be because of the extended washing with several chemicals and water during standard acetolysis.
-For pollen-chemistry research this may present a challenge if we are aiming to directly compare spectra from unprocessed pollen samples since it may create differences in the spectra beween modern and fossil grains.
-Fortunately, the BSi peak does not cover any of the sporopollenin peaks, but obscures the region 1110 - 1070 cm^-1^.
-The quantity of biogenic silica may vary between different sediment cores from different lakes and regions, which may influence the severity of the interference caused by BSi.
-A solution would be to apply correction procedures during pre-processing, or explore extended-washing iterations after extraction of the pollen to dilute BSi.
 
-An additional contamination effect was observed in the Tiefer See samples (Fig. \@ref(fig:meanspectra)) with a broad peak at 1630 cm^-1^. 
-This peak belongs very likely to SPT  [see spectra from @SodiumMetatungstateHydrate].
-This 1630 cm^-1^ peak was responsible for part of the separation between Tiefer See non-acetolysised samples from the other cores, but does not change the larger differentiation of acetolysed Tiefer See samples (without the 1630 cm^-1^ peak) from the other cores.
-It is possible that the three washes with water after decanting of the pollen fraction were not enough in this case, but absence of this peak in the other spectra seem to suggest that further washes may remove successfully remove SPT fully.
+# Discussion
 
-## Comparison of *Pinus* spectra with FTIR spectra from other studies
+## Extraction of *Pinus* FTIR spectra using the FDA detector
+In this study we investigated the use of an FDA detector in combination with a FTIR microscope that allows capture of several spectra in order to characterise the chemistry of pollen grains from a sediment sample.
+The approach has the potential to be an improvement on previous single-grain measurements using FTIR, where each grain is required to be targeted individually, or even manually extracted from the sediment matrix using a micro-pipette.
+The FPA-MCT detector on our FTIR microscope allows simultaneous capture of a larger area of the sample (XXX x XXX µm ), compared to other detectors (e.g.
+100 x 100 µm @jardineChemotaxonomyDomesticatedGrasses2019, 50 x 50 ?µm?
+@zimmermannChemicalCharacterizationIdentification2018).
+An image with a spectra for each pixel is created, from which spectra of single grains were extracted.
 
-The FTIR spectra obtained using the experimental set-up used in this study can be compared to FTIR spectra from *Pinus* pollen from other sources. 
+In general the FTIR spectra obtained using this approach created similar spectra to spectra of *Pinus* pollen from bulk approaches, and the method was also sensitive to differences in chemical treatments on the fossil-pollen grains.
 For example, the spectra obtained from our modern samples are very similar to *Pinus sylvestris* spectra measured using FTIR-bulk sample methodologies [@bagciogluMultiscaleVibrationalSpectroscopic2015; @zimmermannCharacterizationPollenVibrational2010] and FTIR microscopy using single-point measurements [@zimmermannChemicalCharacterizationIdentification2018].
@@
 1745, 1515, 1170 and 1250 - 1000 cm^-1^, and these results build on previous studies which suggest that overall potential for the use of the FPA detector in pollen research **[add references]**.
+Thus, moving forward we see potential in the application of the FDA detector in chemical palaeoecology.
+
+## Variable effects of acetolysis
+Another important result in our study is the large differences in FTIR spectra between acetolysised and non-acetolysised grains.
+The presence of the peak in the 1170 cm^-1^ shoulder in the spectra from the acetolysised pollen grains suggests an overall change in the pollen chemistry of the sample.
+The presence of these two dominant peaks may be related to the removal and or alteration of grain-wall carbohydrates (cellulose etc), as well as changes to chemical composition of sporopollenin due to the acetolysis treatment.
+Since the 1190 cm^-1^ peak is not visible in any of the samples from the non-acetolysed pollen, this likely indicates that it is processes related to the acetolysis treatment that are creating this new peaks in the FTIR spectra.
 
@@
 Lastly, the region between 1300 - 900 cm^-1^ in our spectra mainly consists of two large peaks at 1200 and 1040 cm^-1^, while the *Pinus* pollen from @jardineProxyReconstructionUltravioletB2020 shows large peaks at 1250, 1150, 1100 and 1010.
-Direct comparisons between spectra from different spectrometers are difficult, because peak heights are not always directly comparable, but further research is required to understand the reasons between these differences between the acetolysed *Pinus* pollen from across the sample sets.
 
-Indeed, we actually found greater similarities between the FTIR spectra from the acetolysed *Pinus* pollen from our study to  acetolysed pollen grains of other pollen types.
+Conversely, we found greater similarities between the FTIR spectra from the acetolysed *Pinus* pollen from our study to  acetolysed pollen grains of other pollen types.
 For example, the carbohydrate region at 1200 - 1000 cm^-1^ in our acetolysised *Pinus* pollen samples is most similar to an acetolysed *extant* Poaceae set of @jardineSporopolleninChemistryIts2021, and in addition to acetolyzed herbarium *Nitrariaceae* pollen [@woutersenNovelApproachStudy2018], where two peaks at 1040 and 1190/1170 are visible and similar to our samples.
@@
 
-Taken together, these differences in the effects of acetolysis acorss taxa suggest that the chemical extraction methods have a large effect on the chemistry of pollen, both modern and fossil origin.
+Taken together, the differences in the effects of acetolysis across different taxa and sample types suggest that the chemical extraction methods have a large effect on the chemistry of pollen, both modern and fossil origin.
+Direct comparisons between spectra from different spectrometers are difficult, because peak heights are not always directly comparable, but these results indicate that further research is required to understand the reasons between these differences between the acetolysed *Pinus* pollen from across the sample sets.
 This is important to consider going forward for the application of IR methods in palaeoecology, where extraction methods are often dependent on sediment type.
-It would be useful to explore the effects of these extraction methods on modern and fossil material side-by-side, to investigate if there are methods that can bridge the transition from modern to fossil pollen.
+For example, it would be useful to explore the effects of these extraction methods on modern and fossil material side-by-side, to investigate if there are methods that can bridge the transition from modern to fossil pollen.
 Chemical treatments that can mimic diagenesis will be crucial to understanding the chemical changes of pollen under diagenesis and help to create species specific reference collections of artificial fossilized modern pollen grains.
@@
 
-## Applications for palaeoecological research
-<!-- Our study is a demonstration of microscope FTIR for pollen chemical methods using fossil pollen. -->
-<!-- We focused on the chemistry of *Pinus* pollen, because *Pinus* is one of a few taxa, that is quite common in sediment cores and largely dominated by 1 species, *Pinus sylvestris*. -->
-<!-- The pollen is also relatively large, which prevents any scattering problems found with smaller grains. -->
-<!-- We used *Pinus sylvestris* to explore chemical differences between modern and fossil material, because this has not been fully explored yet. -->
+The chemical effects of acetolysis that identified across studies suggests that it may be preferable to use an alternative approach to extract fossil pollen grains from sediment samples.
+Given the limited chemical influence on the fossil pollen grains, we believe that density separation is a viable alternative to acetolysis and other acid or oxidizing extractions [@caffreyUseLithiumHeteropolytungstate2013; @leipeTestingPerformanceSodium2019; @regnellPreparativeCentrifugationNew1996], although given the fact that an SPT peak was observed in one of the cores, the results in this paper clearly demonstrate the need to identify efficient dilution processes that may need to be tailored to specific sediment types.
+Similarly, since the reported density of pollen is variable and may be dependent on age of the sample [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996; @nakagawaDensemediaSeparationMore1998; @juvigneDensiteExinesQuelques1973], it is possible that specific extraction densities and procedures may be required from core to core.
 
-In this study we investigated the use of a detector in combination with a FTIR microscope that allows capture of several spectra in order to characterise the chemistry of pollen grains from a sediment sample.
-The approach has the potential to be an improvement on previous single-grain measurements using FTIR, where each grain is required to be targeted individually, or even manually extracted from the sediment matrix using a micro-pipette.
-The FPA-MCT detector on our FTIR microscope allows simultaneous capture of a larger area of the sample (XXX x XXX µm ), compared to other detectors (e.g.
-100 x 100 µm @jardineChemotaxonomyDomesticatedGrasses2019, 50 x 50 ?µm?
-@zimmermannChemicalCharacterizationIdentification2018).
-An image with a spectra for each pixel is created, from which spectra of single grains were extracted.
-In general the FTIR spectra obtained using this approach created similar spectra to spectra of *Pinus* pollen from bulk approaches, and the method was also sensitive to differences in chemical treatments on the fossil-pollen grains.
+Since our non-acetolysised samples are characterised by a BSi peak in the FTIR spectra, the density-separation approach also appears to introduce an additional important effect.
+Although the BSi peak does not obscure any of the sporopollenin peaks in the FTIR spectra, it does obscure the region 1110 - 1070 cm^-1^.
+The quantity of biogenic silica may vary between different sediment cores from different lakes and regions, which may influence the severity of the interference caused by BSi. 
+For pollen-chemistry research this may present a challenge if we are aiming to directly compare spectra from unprocessed pollen samples since it may create differences in the spectra beween modern and fossil grains.
+A solution would be to apply correction procedures during pre-processing, or explore extended-washing iterations after extraction of the pollen to dilute BSi.
 
-Thus, moving forward we see potential in the application of the FDA detector in chemical palaeoecology. In particular, one potential future area of research lies in the possibility for automation. 
+## Future developments
+In this study we focused on the chemistry of *Pinus* pollen, because of it's importance in studies related the UV-B-related research and due to its high ecological importance in Quaternary sequences, particularly in the northern hemisphere.
+*Pinus* pollen is also relatively large, which prevents any scattering problems found with smaller grains and made for easy identification of pollen granis in from the images created in our experimental set up.
+However, the quality of the optical images captured by our microscope was much worse than convential LM, which made it difficult to identify smaller pollen grains on the slide from other organic debris, e.g.
+*Betula*/*Corylus*, *Alnus* or *Quercus*.
+An objective with higher magnification (20x) may improve this aspect of using a FTIR microscope.
+Since FTIR measurements of small particles (5-25 µm) introduces scattering artifacts into the spectra [@lukacsRecoveryAbsorbanceSpectra2015; @zimmermannVibrationalMicrospectroscopyEnables2015], sample processing and computational methods exist to minimize or avoid scattering artifacts for these smaller taxa using FTIR.
+
+Finally, a potential future area of research would then be to investigate the possibility for automation to streamline the spectral-collection process. 
 Although in this study the spectra were extracted by hand, approaches using deep learning have shown to be capable of extracting objects from larger images, using deep neural networks specialised on image segmentation/object location [@renFasterRCNNRealTime2015].
@@
 
-Nevertheless, we encountered a number of challenges that will need to be addressed if the experimental set-up we used here can be more comprensively applied in palaeoecology.
-First, we found that the concentration of pollen in our samples was very variable and posed a major bottleneck for our analysis with the FPA detector.
-Capturing spectra from a larger area is only advantageous over single-point measurements if several grains are located in the FTIR image.
-This was not always possible, we found, because the pollen was very scarce in some samples.
-Acetolysis is the conventional method to extract pollen in high concentration in Quaternary pollen samples, but the results presented here demonstrate the value in extracting non-acetolysised pollen from sediments, because of the substantial differences in pollen chemistry we found between acetolysised and non-acetolysised pollen (see above).
-Given the limited chemical influence on the fossil pollen grains, we believe that density separation is a viable alternative to acetolysis and other acid or oxidizing extractions [@caffreyUseLithiumHeteropolytungstate2013; @leipeTestingPerformanceSodium2019; @regnellPreparativeCentrifugationNew1996], alhtough the results in this paper clearly demonstrate the need to identify efficient dilution processes.
-Similarly, since the reported density of pollen is variable and may be dependent on age of the sample [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996; @nakagawaDensemediaSeparationMore1998; @juvigneDensiteExinesQuelques1973], it is possible that specific extraction densities and procedures may be required from core to core.
 <!-- After testing multiple densities, 1.44 g cm^-3^ yielded the most pollen from our samples. -->
@@
 
-Second, the quality of the optical images captured by our microscope was much worse than convential LM, which made it difficult to identify smaller pollen grains on the slide from other organic debris, e.g.
-*Betula*/*Corylus*, *Alnus* or *Quercus*.
-An objective with higher magnification (20x) may improve this aspect of using a FTIR microscope.
-The FPA detector in combination with automated localisation of potential pollen grains based on their chemical signature may also help in solving this part of the problem, because identification by eye won't be necessary.
-
-Third, as reported in other studies, FTIR measurements of small particles (5-25 µm) introduces scattering artifacts into the spectra [@lukacsRecoveryAbsorbanceSpectra2015; @zimmermannVibrationalMicrospectroscopyEnables2015].
-A number of pollen species are affected by this scattering because of their size.
-While sample processing and computational methods exist to minimize or avoid scattering artifacts, there may be substantial species specific information stored in scattering artifacts that could be useful for identification and classification.
-
 ## Summary
-
 Fossil pollen from three sediment cores and modern material was analysed by FTIR microspectroscopy using a new detector that allows simultaneous capture of multiple single grain spectra.
@@
 
+## I haven't found a way to integrate these ideas in there. Perhaps we don't need them? I think perhaps we could leave it out
+In addition, there is some differentiation between the cores.
+Samples from Tiefer See are more distinct from samples of the other two cores, which show more similar scores in the ordination.
+One interpretation of this is that *Pinus* pollen from DAL and MFM are chemically quite similar, while TSK *Pinus* pollen is different in chemical structure.
+Since it is likely that the main contributions to the pollen some in the TSK core is *Pinus sylvestris*, it is likely that the same species is represented by all three cores, and one of the main differences between the TSK pollen and the other two locations is sample age.
+Whilst the TSK samples are from the past 80 years, MFM covers the YD/Holocene transition and DAL samples are from throughout the Holocene, with the youngest sample likely more than 100 years older than TSK pollen.
+
+This finding could suggest a chemical difference in fossil-pollen grains that is persistent even after acetolysis, which may be explained by the age of the pollen.
+Interestingly, @jardineSporopolleninChemistryIts2021 also found differences in FTIR spectra with different sample ages, although they found that that modern-pollen spectra had more in common with spectra from Quaternary pollen, compared to pollen from deeper time samples.
+Further studies on pollen from different sediment types at fixed age/depths may provide a clearer answer on the effects of sediment type and age on pollen chemistry.
+
 # References

### 2021-06-25 manuscript.Rmd
added abstract and Pinus no of species ref

@@
 # Abstract
-200 words no abreviations/refs
+
+The study of pollen-grain chemistry has gained increasing focus in palaeoecological research since it is possible that chemical variations observed in sporopollenin may be used for identification of pollen grains.
+Fourier Transform Infrared (FTIR) spectroscopy has become one of the most popular methods for characterising the chemical variations present within pollen and spores.
+Fossil pollen from three sediment cores and modern material was analysed by FTIR microspectroscopy using a new detector that allows simultaneous capture of multiple single grain spectra.
+We also investigate the effect of two pollen extraction methods (acetolysis and density separation) on fossil pollen chemistry.
+Our methods reproduce modern *Pinus* spectra well, compared to bulk analysis and other microspectroscopy approaches.
+We report chemical differences between fossil and modern pollen, mainly lack of lipid and protein wavebands (1800 - 1500 cm^-1^), in addition to changes in the carbohydrate region (1200 - 1000 cm^-2^).
+Acetolysed fossil pollen was compared to non-acetolysed fossil pollen, which revealed differences in their chemical composition (e.g.
+changes in the 1200 - 1000 cm^-2^ region) that are most similar to reported acetolysed spectra from other taxa, suggesting that acetolysis alters fossil pollen chemistry.
+We also report contaminations observed in non-acetolysed samples and the applications of high density separation extraction for FTIR microspectroscopy and palynology.
 
@@
 pollen may be a useful taxon to use as a biological indicator for reconstructing past changes of UV-B radiation [@jardineProxyReconstructionUltravioletB2020; @willisQuantificationUVFlux2011; @seddonPollenchemistryVariationsElevation2021].
-Second, although prominent in Quaternary sediment records from Europe, and despite it's high species diversity in Europe (**Check number of species**), *Pinus* is currently mostly only split into two different pollen types using fossil pollen, reflecting the two phylogenetic lineages [@beugLeitfadenPollenbestimmungFur2004; @gieseckePostglacialChangeFloristic2019].
+Second, although prominent in Quaternary sediment records from Europe, and despite the highest number of species amoing conifers [110 species, @gernandtPhylogenyClassificationPinus2005], *Pinus* is currently mostly only split into two different pollen types using fossil pollen, reflecting the two phylogenetic lineages [@beugLeitfadenPollenbestimmungFur2004; @gieseckePostglacialChangeFloristic2019].
 Thus, identifying chemical differences in fossil *Pinus* pollen may also allow for identification of crytpic *Pinus* species from the pollen types observed in the fossil record.
@@
 We used two PLSR models to explore the chemical differences between modern and fossil grains (model 1), and acetolysised and non-acetolysised fossil grains (model 2).
-For these models the explanatory variables were the pre-processed spectra and the response variables were hot-one encoded response variables (Model 1: modern/fossil; Model 2: acet./ non acet.).
+For these models the explanatory variables were the pre-processed spectra and the response variables were hot-one encoded response variables (Model 1: modern/fossil; Model 2: acet./non acet.).
 By using PLSR we aimed to maximize the distance between responses and identify chemical differences between the responses by looking at the regression coefficients.

### 2021-06-25 manuscript.Rmd
revert acetolysised to acetolysed

@@
 
-We used two PLSR models to explore the chemical differences between modern and fossil grains (model 1), and acetolysised and non-acetolysised fossil grains (model 2).
+We used two PLSR models to explore the chemical differences between modern and fossil grains (model 1), and acetolysed and non-acetolysed fossil grains (model 2).
 For these models the explanatory variables were the pre-processed spectra and the response variables were hot-one encoded response variables (Model 1: modern/fossil; Model 2: acet./non acet.).
@@
 First, in the Tiefer See non-acetolysed fossil pollen samples (Fig. \@ref(fig:meanspectra)) there was a broad peak at 1630 cm^-1^. 
-This 1630 cm^-1^ peak was responsible for part of the separation between Tiefer See non-acetolysised samples from the other cores, but does not change the larger differentiation of acetolysed Tiefer See samples (without the 1630 cm^-1^ peak) from the other cores, and the peak very likely indicates the presence of SPT [see spectra from @SodiumMetatungstateHydrate].
+This 1630 cm^-1^ peak was responsible for part of the separation between Tiefer See non-acetolysed samples from the other cores, but does not change the larger differentiation of acetolysed Tiefer See samples (without the 1630 cm^-1^ peak) from the other cores, and the peak very likely indicates the presence of SPT [see spectra from @SodiumMetatungstateHydrate].
 Thus, it is possible that the three washes with water after decanting of the pollen fraction were not enough in this case, but absence of this peak in the other spectra seem to suggest that further washes may remove successfully remove SPT fully.
@@
 ## Variable effects of acetolysis
-Another important result in our study is the large differences in FTIR spectra between acetolysised and non-acetolysised grains.
-The presence of the peak in the 1170 cm^-1^ shoulder in the spectra from the acetolysised pollen grains suggests an overall change in the pollen chemistry of the sample.
+Another important result in our study is the large differences in FTIR spectra between acetolysed and non-acetolysed grains.
+The presence of the peak in the 1170 cm^-1^ shoulder in the spectra from the acetolysed pollen grains suggests an overall change in the pollen chemistry of the sample.
 The presence of these two dominant peaks may be related to the removal and or alteration of grain-wall carbohydrates (cellulose etc), as well as changes to chemical composition of sporopollenin due to the acetolysis treatment.
@@
 
-Interestingly, when comparing our acetolysed *Pinus* spectra to other acetolysised fossil *Pinus* pollen, we observed some differences.
+Interestingly, when comparing our acetolysed *Pinus* spectra to other acetolysed fossil *Pinus* pollen, we observed some differences.
 For example, @jardineProxyReconstructionUltravioletB2020 extracted *Pinus* pollen from a varved sediment core using acetolysis and HF to study the UV-B response in *Pinus* pollen chemistry.
@@
 Conversely, we found greater similarities between the FTIR spectra from the acetolysed *Pinus* pollen from our study to  acetolysed pollen grains of other pollen types.
-For example, the carbohydrate region at 1200 - 1000 cm^-1^ in our acetolysised *Pinus* pollen samples is most similar to an acetolysed *extant* Poaceae set of @jardineSporopolleninChemistryIts2021, and in addition to acetolyzed herbarium *Nitrariaceae* pollen [@woutersenNovelApproachStudy2018], where two peaks at 1040 and 1190/1170 are visible and similar to our samples.
+For example, the carbohydrate region at 1200 - 1000 cm^-1^ in our acetolysed *Pinus* pollen samples is most similar to an acetolysed *extant* Poaceae set of @jardineSporopolleninChemistryIts2021, and in addition to acetolyzed herbarium *Nitrariaceae* pollen [@woutersenNovelApproachStudy2018], where two peaks at 1040 and 1190/1170 are visible and similar to our samples.
 However, spectra from a fossil Poaceae data set (sampled from the the Amazon Fan and Western Amazon; AF: Na~4~P~2~O~7~, density sep., lignite, WA: HCl, Na~4~P~2~O~7~, density sep., acetolysis) [@jardineSporopolleninChemistryIts2021] is quite different in composition to both our fossil pollen, and any extant acetolysed-pollen samples, since the large peak at 1600 and 1190 peak are not clearly visible.
@@
 
-Since our non-acetolysised samples are characterised by a BSi peak in the FTIR spectra, the density-separation approach also appears to introduce an additional important effect.
+Since our non-acetolysed samples are characterised by a BSi peak in the FTIR spectra, the density-separation approach also appears to introduce an additional important effect.
 Although the BSi peak does not obscure any of the sporopollenin peaks in the FTIR spectra, it does obscure the region 1110 - 1070 cm^-1^.
@@
 changes in the 1200 - 1000 cm^-2^ region) that are most similar to reported acetolysed spectra from other taxa, suggesting that acetolysis alters fossil pollen chemistry.
-We also report contaminations observed in non-acetolysized samples and the applications of high density liquid extraction for FTIR microspectroscopy and palynology.
+We also report contaminations observed in non-acetolysed samples and the applications of high density liquid extraction for FTIR microspectroscopy and palynology.
 

### 2021-06-25 manuscript.Rmd
changed letters in gernandt ref that caused crashes

@@
 
-## Extraction of *Pinus* FTIR spectra using the FDA detector
-In this study we investigated the use of an FDA detector in combination with a FTIR microscope that allows capture of several spectra in order to characterise the chemistry of pollen grains from a sediment sample.
+## Extraction of *Pinus* FTIR spectra using the FPA detector
+In this study we investigated the use of an FPA detector in combination with a FTIR microscope that allows capture of several spectra in order to characterise the chemistry of pollen grains from a sediment sample.
 The approach has the potential to be an improvement on previous single-grain measurements using FTIR, where each grain is required to be targeted individually, or even manually extracted from the sediment matrix using a micro-pipette.
@@
 1745, 1515, 1170 and 1250 - 1000 cm^-1^, and these results build on previous studies which suggest that overall potential for the use of the FPA detector in pollen research **[add references]**.
-Thus, moving forward we see potential in the application of the FDA detector in chemical palaeoecology.
+Thus, moving forward we see potential in the application of the FPA detector in chemical palaeoecology.
 

### 2021-06-25 manuscript.Rmd
added some comments from Boris

@@
 Bulk FTIR techniques, which involve the analysis of large quantities of modern-pollen samples (approximately 1-2 mg) (e.g.
-@zimmermannCharacterizationPollenVibrational2010, @lahlaliATRFTIRSpectroscopyReveals2014), have the advantage of reducing scattering effects from measurements due to the analysis of large sample quantities, which improves the signal-to-noise ratio in terms of spectra acquisition..
-As a result, FTIR spectra obtained from bulk pollen samples have been used to identify chemical differences at the genus, and in some cases, species level [@muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020].
+@zimmermannCharacterizationPollenVibrational2010, @lahlaliATRFTIRSpectroscopyReveals2014), have the advantage of reducing scattering effects from measurements due to contact measurement of large sample quantities, which improves the signal-to-noise ratio in terms of spectra acquisition.
+As a result, FTIR spectra obtained from bulk pollen samples have been used to identify chemical differences at the genus, and in some cases, species and population level [@muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020; @bagciogluMonitoringPlantEnvironment2017; @zimmermannHighthroughputFTIRSpectroscopy2017].
 However, these approaches may be less relevant for pollen samples extracted from fossil sequences since the number of pollen grains analysed vastly exceeds the quantity which is practically available from a sediment-core sample.
 
-An alternative approach involves the use of FTIR microscopes, which require manual targeting of small areas of the sample slides to record spectra of single grains or groups of pollen grains.
-Because of their higher spatial resolution, FTIR-microscopes are better suited to sediment samples as a result of the capacity to generate FTIR spectra from individual grains, although spectral-processing procedures may be required to reduce the scattering effects when obtaining spectra from individual pollen grains [@zimmermannVibrationalMicrospectroscopyEnables2015; @zimmermannAnalysisAllergenicPollen2016].
+An alternative approach involves the use of FTIR microscopes, where a FTIR spectrometer is coupled to a microscope. 
+This enables acquisition of infrared spectra from a small area, with aperture covering several pollen grains, a single pollen grain, or even pollen grain substructure [@zimmermannVibrationalMicrospectroscopyEnables2015].
+FTIR-microscopes are better suited to sediment samples as a result of the capacity to generate FTIR spectra from individual grains.
+However, FTIR microspectroscopy is hindered by strong Mie scattering of IR radiation on pollen grains, resulting with anomalous features in their FTIR spectra [@zimmermannVibrationalMicrospectroscopyEnables2015; @blumelObservationMieRipples2018a; @lukacsRecoveryAbsorbanceSpectra2015].
+This issues can be tackled by a number of analytical and numerical solutions [@diehnDiscriminationGrassPollen2020; @zimmermannChemicalCharacterizationIdentification2018; @lukacsRecoveryAbsorbanceSpectra2015; @zimmermannAnalysisAllergenicPollen2016], but nevertheless FTIR measurement of individual pollen grains remains challenging.
 Nevertheless, microscopic FTIR techniques have now been applied to both modern- and fossil-pollen grains preserved in sediments.

### 2021-06-25 manuscript.Rmd
excluding docx from tracking, changed a sentence in FPA section

@@
 output:
+  bookdown::word_document2:
   bookdown::pdf_document2:
@@
 This enables the extraction of spectra from several single grains from the same area on the sample without the need to relocate the detector of the microscope or move the slide, and is therefore of potentially high relevance for palaeoecological research.
-When combined with imaging processing technology, it may allow for more rapid identification and generation of FTIR spectra of pollen grains from sediment samples, but whilst FTIR spectroscopy using FPA detectors have been widely applied with microplastics [@loderEnzymaticPurificationMicroplastics2017; @machadoMicroplasticsEmergingThreat2018; @primpkeAutomatedApproachMicroplastics2017] and some studies on pollen [@chiappePollenGrainsFunctionalized2017], there has been limited investigation into their application on fossil-pollen residues taken from sediment cores.
+When combined with imaging processing technology, it may allow for more rapid identification and extraction of FTIR spectra of pollen grains from sediment samples.
+Whilst FTIR spectroscopy using FPA detectors have been widely applied with microplastics [@loderEnzymaticPurificationMicroplastics2017; @machadoMicroplasticsEmergingThreat2018; @primpkeAutomatedApproachMicroplastics2017] and some studies on pollen [@chiappePollenGrainsFunctionalized2017], there has been no investigation into application of FPA detectors on fossil-pollen residues taken from sediment cores.
 

### 2021-07-06 manuscript.Rmd
boris comm to intro

@@
 output:
-  bookdown::word_document2:
   bookdown::pdf_document2:
@@
     extra_dependencies: ["flafter"]
+  bookdown::word_document2:
 bibliography: references.bib
@@
 Any insights gained from studies on modern grains must also be transferable under fossil settings [@jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
-Fourier Transform Infrared (FTIR) spectroscopy has become one of the most popular methods for characterising the chemical variations present within pollen and spores in the modern environment.
+Fourier transform infrared (FTIR) spectroscopy has become one of the most popular methods for characterising the chemical variations present within pollen and spores in the modern environment.
 Bulk FTIR techniques, which involve the analysis of large quantities of modern-pollen samples (approximately 1-2 mg) (e.g.
-@zimmermannCharacterizationPollenVibrational2010, @lahlaliATRFTIRSpectroscopyReveals2014), have the advantage of reducing scattering effects from measurements due to contact measurement of large sample quantities, which improves the signal-to-noise ratio in terms of spectra acquisition.
+@zimmermannCharacterizationPollenVibrational2010, @lahlaliATRFTIRSpectroscopyReveals2014), predominantly result with high-quality spectra with detailed biochemical ‘fingerprint’ of pollen, comprising specific signals of sporopollenins, lipids, proteins, carbohydrates and other chemical constituents [@bagciogluMultiscaleVibrationalSpectroscopic2015].
 As a result, FTIR spectra obtained from bulk pollen samples have been used to identify chemical differences at the genus, and in some cases, species and population level [@muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020; @bagciogluMonitoringPlantEnvironment2017; @zimmermannHighthroughputFTIRSpectroscopy2017].
@@
 However, FTIR microspectroscopy is hindered by strong Mie scattering of IR radiation on pollen grains, resulting with anomalous features in their FTIR spectra [@zimmermannVibrationalMicrospectroscopyEnables2015; @blumelObservationMieRipples2018a; @lukacsRecoveryAbsorbanceSpectra2015].
-This issues can be tackled by a number of analytical and numerical solutions [@diehnDiscriminationGrassPollen2020; @zimmermannChemicalCharacterizationIdentification2018; @lukacsRecoveryAbsorbanceSpectra2015; @zimmermannAnalysisAllergenicPollen2016], but nevertheless FTIR measurement of individual pollen grains remains challenging.
+Although the scattering issues can be tackled by a number of analytical and numerical solutions [@diehnDiscriminationGrassPollen2020; @zimmermannChemicalCharacterizationIdentification2018; @lukacsRecoveryAbsorbanceSpectra2015; @zimmermannAnalysisAllergenicPollen2016], FTIR measurement of individual pollen grains remains challenging.
 Nevertheless, microscopic FTIR techniques have now been applied to both modern- and fossil-pollen grains preserved in sediments.
-The approach has been successfully used to show that chemical differences are found between pollen at the genus and even species level [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018;  @zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019], and microscopic-FTIR techniques have also been used to observe chemical differences to reconstruct changes in UV-B absorbing compounds in different periods across Earth's history [@fraserUVBAbsorbingPigments2011; @jardinePollenSporesBiological2016; @jardineProxyReconstructionUltravioletB2019; @jardineSheddingLightSporopollenin2017].
+The approach has been successfully used to show that chemical differences are found between pollen at the genus and even species level [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018;  @zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019].
+Moreover, microscopic-FTIR techniques have been used to observe chemical differences in pollen samples to reconstruct changes in UV-B absorbing compounds in different periods across Earth's history [@fraserUVBAbsorbingPigments2011; @jardinePollenSporesBiological2016; @jardineProxyReconstructionUltravioletB2019; @jardineSheddingLightSporopollenin2017].
 
 One of key challenges faced by chemical palynologists using micro-FTIR and other approaches is the time taken to extract and analyse fossil-pollen samples.
-Some microscopes are equipped with automated stages, where multiple point locations on the slide can be targeted manually to record spectra automatically.
-However, recent developments in technologies related to FTIR may allow for easier and faster capture of spectra of pollen grains.
-A focal plane array detector (FPA) allows capture of an image and matrix of spectra over the same sample area instead of a single spectra at a specific point/area [@zimmermannAnalysisAllergenicPollen2016;@zimmermannChemicalCharacterizationIdentification2018; @zimmermannVibrationalMicrospectroscopyEnables2015].
-This enables the extraction of spectra from several single grains from the same area on the sample without the need to relocate the detector of the microscope or move the slide, and is therefore of potentially high relevance for palaeoecological research.
+Some FTIR microscopes are equipped with automated sample stages, enabling automated measurement of predetermined measurement positions.
+Measurements are usually acquired by consecutive recording of spectra by a single element detector [@zimmermannAnalysisAllergenicPollen2016;@zimmermannChemicalCharacterizationIdentification2018; @zimmermannVibrationalMicrospectroscopyEnables2015].
+When spectra are recorded with equidistant spatial steps (i.e.
+scanning measurement), the individual spectra can be assembled after the recording to obtain a chemical image [@jardineSporopolleninChemistryIts2021; @zimmermannVibrationalMicrospectroscopyEnables2015; @julierChemotaxonomyToolInterpreting2016].
+However, recent developments in technologies related to FTIR allow for easier and faster capture of spectra of pollen grains. 
+A focal plane array detector (FPA) is made of a 2D array of IR detectors, and allows true chemical imaging.
+This enables simultaneous measurement of several single pollen grains from a relatively large surface area without the need to move the sample stage.
+Therefore, it is of potentially high relevance for palaeoecological research.
+
 When combined with imaging processing technology, it may allow for more rapid identification and extraction of FTIR spectra of pollen grains from sediment samples.
-Whilst FTIR spectroscopy using FPA detectors have been widely applied with microplastics [@loderEnzymaticPurificationMicroplastics2017; @machadoMicroplasticsEmergingThreat2018; @primpkeAutomatedApproachMicroplastics2017] and some studies on pollen [@chiappePollenGrainsFunctionalized2017], there has been no investigation into application of FPA detectors on fossil-pollen residues taken from sediment cores.
+Whilst FTIR microspectroscopy with FPA detectors has been widely applied with microplastics [@primpkeAutomatedApproachMicroplastics2017] application on pollen samples was limited just to chemical characterization of isolated sporopollenin capsules  [@chiappePollenGrainsFunctionalized2017].
+In particular, there has been no investigation into application of FPA detectors on measurement of fossil-pollen residues taken from sediment cores.
 
 Palaeoecologists may also use a variety of chemical-processing techniques in order to speed-up analyses through the isolation and concentration of pollen from the sediment matrix [@ridingGuidePreparationProtocols2021].
-One of the most common methods for sporopollenin extraction involves an acetolysis treatment, which, alongside a combination of HF and HCl treatments, enables removal of organic materials, calcium carbonates and silicates, while preserving the morphological integrity of pollen grains [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
+One of the most common methods for sporopollenin extraction involves an acetolysis treatment, which, alongside a combination of treatments with hydrofluoric acid (HF) and hydrochloric acid (HCl), enables removal of organic materials, calcium carbonates and silicates, while preserving the morphological integrity of pollen grains [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
 Acetolysis has also often been used in combination with FTIR on pollen as a way to standardise between modern and fossil samples, [@jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or to study the effects of acid extractions [@dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000; @schulzeosthoffPhenolsIntegratedCompounds1987; @jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
@@
 Such treatments may then result in chemically altering effects on fossil-pollen composition [@jardineSporopolleninChemistryIts2021], whilst variations in sediment types mean that different combinations of sample-preparation techniques may be required for different sediment types [@ridingGuidePreparationProtocols2021].
-Given the large variations in FTIR spectra observed across taxa [@diehnCombiningChemicalInformation2020; @muthreichChemicalVariationsQuercus2020], and the variety of sediment types even from Holocene sequences in Europe, more studies are needed investigate the potential effects of the chemical preparation procedures used to isolate pollen grains from different pollen types.
+Given the large variations in FTIR spectra observed across taxa [@diehnCombiningChemicalInformation2020; @muthreichChemicalVariationsQuercus2020], and the variety of sediment types even from Holocene sequences in Europe, more studies are needed to investigate the potential effects of the chemical preparation procedures used to isolate pollen grains from different pollen types.
 
-In this study we aimed to explore the use of an FPA detector to extract FTIR spectra from fossil-pollen *Pinus* grains.
-We develop a workflow which enables comparison of fossil-pollen grains taken from three sediment-core locations (Table \@ref(tab:sites)), and compare the FTIR spectra generated using the FPA detector from both modern- and fossil-pollen samples.
-Pollen was extracted from the sediment cores using two different approaches (density separation using sodium polytungstate and acetolysis) and the fossil-pollen samples were measured using a combined FTIR microscope with FPA detector.
-We also compared the FTIR spectra obtained from fossil-pollen grains with modern *Pinus* pollen samples.
-Thus, in addition to testing the feasibility of the FPA detector on different sediment types, we were were also able to investigate how the inferred FTIR spectra in *Pinus* pollen can vary as a result of two different extraction procedures.
+In this study we aimed to explore the use of a FTIR imaging to obtain IR spectra from fossil-pollen *Pinus* grains.
+We develop a workflow which enables comparison of fossil-pollen grains taken from three sediment-core locations (Table \@ref(tab:sites)), and compare the IR spectra from both modern- and fossil-pollen samples.
+Pollen was extracted from the sediment cores using two different approaches, namely density separation using sodium polytungstate and acetolysis, followed by FTIR microscopy imaging.
+Thus, in addition to testing the feasibility of the FTIR imaging approach on different sediment types, we were were also able to investigate how the IR spectra of the *Pinus* pollen vary as a result of two different extraction procedures.
 

### 2021-07-08 manuscript.Rmd
moving refs and removing whitespace

@@
 ---
-title: 'Assessing variations in the chemistry of fossil- and modern-*Pinus* pollen'
-date: '2020/02/04'
+title: 'Assessing variations in the chemistry of sub-fossil and modern *Pinus* pollen'
+#date: '2020/02/04'
 output:
@@
 `r Affiliations`
-
+<!-- 
 2500 words not including in tables title page references
@@
 
-<!-- 
 A  200
@@
 The study of plant microfossils (pollen and spores) is an integral part of palaeoecology and palaeobotany.
-Pollen records are an important tool to reconstruct past floristic, vegetation and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @noguesbravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013] and provide insights into ecological processes which operate on decadal, centennial and millennial timescales (e.g.
-the post-glacial migration of tree taxa into new habitats following the Quaternary glacial maxima @gieseckeMovingFrontPopulation2005, @huntleyQuaternaryPalaeoecologyEcology1996, @tinnerCentralEuropeanVegetation2001).
+Pollen records are an important tool to reconstruct past floristic, vegetation and environmental changes [@magriNewScenarioQuaternary2006; @tzedakisVegetationChangeGlacialInterglacial1994; @huntleyMigrationSpeciesResponse1989; @noguesbravoCrackingCodeBiodiversity2018; @tinnerEcologyAbiesAlba2013] and provide insights into ecological processes which operate on decadal, centennial and millennial timescales [e.g.
+the post-glacial migration of tree taxa into new habitats following the Quaternary glacial maxima @gieseckeMovingFrontPopulation2005, @huntleyQuaternaryPalaeoecologyEcology1996, @tinnerCentralEuropeanVegetation2001].
 
@@
 @zimmermannCharacterizationPollenVibrational2010, @lahlaliATRFTIRSpectroscopyReveals2014), predominantly result with high-quality spectra with detailed biochemical ‘fingerprint’ of pollen, comprising specific signals of sporopollenins, lipids, proteins, carbohydrates and other chemical constituents [@bagciogluMultiscaleVibrationalSpectroscopic2015].
-As a result, FTIR spectra obtained from bulk pollen samples have been used to identify chemical differences at the genus, and in some cases, species and population level [@muthreichChemicalVariationsQuercus2020; @zimmermannInfraredSpectroscopyPollen2014; @depciuchIdentificationBirchPollen2018; @diehnDiscriminationGrassPollen2020; @bagciogluMonitoringPlantEnvironment2017; @zimmermannHighthroughputFTIRSpectroscopy2017].
+As a result, FTIR spectra obtained from bulk pollen samples have been used to identify chemical differences at the genus, and in some cases, species and population level [@zimmermannInfraredSpectroscopyPollen2014; @bagciogluMonitoringPlantEnvironment2017; @zimmermannHighthroughputFTIRSpectroscopy2017; @depciuchIdentificationBirchPollen2018; @muthreichChemicalVariationsQuercus2020; @diehnDiscriminationGrassPollen2020].
 However, these approaches may be less relevant for pollen samples extracted from fossil sequences since the number of pollen grains analysed vastly exceeds the quantity which is practically available from a sediment-core sample.
@@
 FTIR-microscopes are better suited to sediment samples as a result of the capacity to generate FTIR spectra from individual grains.
-However, FTIR microspectroscopy is hindered by strong Mie scattering of IR radiation on pollen grains, resulting with anomalous features in their FTIR spectra [@zimmermannVibrationalMicrospectroscopyEnables2015; @blumelObservationMieRipples2018a; @lukacsRecoveryAbsorbanceSpectra2015].
-Although the scattering issues can be tackled by a number of analytical and numerical solutions [@diehnDiscriminationGrassPollen2020; @zimmermannChemicalCharacterizationIdentification2018; @lukacsRecoveryAbsorbanceSpectra2015; @zimmermannAnalysisAllergenicPollen2016], FTIR measurement of individual pollen grains remains challenging.
+However, FTIR microspectroscopy is hindered by strong Mie scattering of IR radiation on pollen grains, resulting with anomalous features in their FTIR spectra [@lukacsRecoveryAbsorbanceSpectra2015; @zimmermannVibrationalMicrospectroscopyEnables2015; @blumelObservationMieRipples2018a].
+Although the scattering issues can be tackled by a number of analytical and numerical solutions [@lukacsRecoveryAbsorbanceSpectra2015; @zimmermannAnalysisAllergenicPollen2016; @zimmermannChemicalCharacterizationIdentification2018; @diehnDiscriminationGrassPollen2020], FTIR measurement of individual pollen grains remains challenging.
 Nevertheless, microscopic FTIR techniques have now been applied to both modern- and fossil-pollen grains preserved in sediments.
 The approach has been successfully used to show that chemical differences are found between pollen at the genus and even species level [@julierChemotaxonomyToolInterpreting2016; @woutersenNovelApproachStudy2018; @zimmermannChemicalCharacterizationIdentification2018; @jardineChemotaxonomyDomesticatedGrasses2019].
-Moreover, microscopic-FTIR techniques have been used to observe chemical differences in pollen samples to reconstruct changes in UV-B absorbing compounds in different periods across Earth's history [@fraserUVBAbsorbingPigments2011; @jardinePollenSporesBiological2016; @jardineProxyReconstructionUltravioletB2019; @jardineSheddingLightSporopollenin2017].
+Moreover, microscopic-FTIR techniques have been used to observe chemical differences in pollen samples to reconstruct changes in UV-B absorbing compounds in different periods across Earth's history [@fraserUVBAbsorbingPigments2011; @jardinePollenSporesBiological2016;@jardineSheddingLightSporopollenin2017; @jardineProxyReconstructionUltravioletB2019].
 
@@
 Some FTIR microscopes are equipped with automated sample stages, enabling automated measurement of predetermined measurement positions.
-Measurements are usually acquired by consecutive recording of spectra by a single element detector [@zimmermannAnalysisAllergenicPollen2016;@zimmermannChemicalCharacterizationIdentification2018; @zimmermannVibrationalMicrospectroscopyEnables2015].
+Measurements are usually acquired by consecutive recording of spectra by a single element detector [@zimmermannVibrationalMicrospectroscopyEnables2015; @zimmermannAnalysisAllergenicPollen2016; @zimmermannChemicalCharacterizationIdentification2018].
 When spectra are recorded with equidistant spatial steps (i.e.
-scanning measurement), the individual spectra can be assembled after the recording to obtain a chemical image [@jardineSporopolleninChemistryIts2021; @zimmermannVibrationalMicrospectroscopyEnables2015; @julierChemotaxonomyToolInterpreting2016].
+scanning measurement), the individual spectra can be assembled after the recording to obtain a chemical image [@zimmermannVibrationalMicrospectroscopyEnables2015; @julierChemotaxonomyToolInterpreting2016; @jardineSporopolleninChemistryIts2021].
 However, recent developments in technologies related to FTIR allow for easier and faster capture of spectra of pollen grains. 
@@
 One of the most common methods for sporopollenin extraction involves an acetolysis treatment, which, alongside a combination of treatments with hydrofluoric acid (HF) and hydrochloric acid (HCl), enables removal of organic materials, calcium carbonates and silicates, while preserving the morphological integrity of pollen grains [@erdtmanSixNotesPollen1959; @faegriTextbookPollenAnalysis1989].
-Acetolysis has also often been used in combination with FTIR on pollen as a way to standardise between modern and fossil samples, [@jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or to study the effects of acid extractions [@dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000; @schulzeosthoffPhenolsIntegratedCompounds1987; @jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
+Acetolysis has also often been used in combination with FTIR on pollen as a way to standardise between modern and fossil samples, [@jardinePollenSporesBiological2016; @woutersenNovelApproachStudy2018] or to study the effects of acid extractions [@schulzeosthoffPhenolsIntegratedCompounds1987; @dominguezIsolationIntactPollen1998; @dominguezPollenSporopolleninDegradation1999; @loaderPreparationPollenStable2000;  @jardineImpactOxidationSpore2015; @jardineSporopolleninChemistryIts2021].
 Indeed, although acetolysis treatments at low temperatures did not create major chemical differences in *Lycopodium* spores [@jardineImpactOxidationSpore2015], it is uncommon for low-temperature treatments to be applied to fossil-sediment samples since the higher temperatures are required for oxidation of the non-pollen organic material.
@@
 In this study we aimed to explore the use of a FTIR imaging to obtain IR spectra from fossil-pollen *Pinus* grains.
-We develop a workflow which enables comparison of fossil-pollen grains taken from three sediment-core locations (Table \@ref(tab:sites)), and compare the IR spectra from both modern- and fossil-pollen samples.
+We develop a workflow which enables comparison of fossil-pollen grains taken from three sediment-core locations (Table \@ref(tab:sites)), and compare the IR spectra from both modern and fossil pollen samples.
 Pollen was extracted from the sediment cores using two different approaches, namely density separation using sodium polytungstate and acetolysis, followed by FTIR microscopy imaging.
@@
 knitr::kable(sites, booktabs = T, format = "latex", 
-             caption = "",
+             caption = "Sampled sediment cores for this study. ",
              col.names = c("Name", "Abbr.", "Age", "Description", "Publ.")) %>%
@@
 is a shade-intolerant, high pollen producer, and so a number of studies have proposed that *Pinus* spp.
-pollen may be a useful taxon to use as a biological indicator for reconstructing past changes of UV-B radiation [@jardineProxyReconstructionUltravioletB2020; @willisQuantificationUVFlux2011; @seddonPollenchemistryVariationsElevation2021].
+pollen may be a useful taxon to use as a biological indicator for reconstructing past changes of UV-B radiation [@willisQuantificationUVFlux2011; @jardineProxyReconstructionUltravioletB2020; @seddonPollenchemistryVariationsElevation2021].
 Second, although prominent in Quaternary sediment records from Europe, and despite the highest number of species amoing conifers [110 species, @gernandtPhylogenyClassificationPinus2005], *Pinus* is currently mostly only split into two different pollen types using fossil pollen, reflecting the two phylogenetic lineages [@beugLeitfadenPollenbestimmungFur2004; @gieseckePostglacialChangeFloristic2019].
@@
 After centrifugation, particulates heavier than 1.43 g cm^-3^ accumulate of the bottom of the sample vial, whilst the lighter fraction was decanted and collected.
-Note, that there is no agreement on the density of the pollen exine, with reported values varying between <1.4 g cm^-3^ [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996], 1.88 cm^-3^ [@nakagawaDensemediaSeparationMore1998] and 2.2 cm^-3^  [@juvigneDensiteExinesQuelques1973].
+Note, that there is no agreement on the density of the pollen exine, with reported values varying between <1.4 g cm^-3^ [@flenley1971measurements; @faegriTextbookPollenAnalysis1989; @regnellPreparativeCentrifugationNew1996], 1.88 cm^-3^ [@nakagawaDensemediaSeparationMore1998] and 2.2 cm^-3^ [@juvigneDensiteExinesQuelques1973].
 In our case the lighter fraction (<1.4 g cm^-3^) contained majority of the pollen grains, after checking both fractions.
@@
 The second derivative spectra was then normalized using a EMSC model with a linear component [@lilandPackageEMSC2017].
-Preprocessing with SG and EMSC has been shown to improve the performance of multivariate analyses, such as principal component analysis (PCA) or models in the partial least squares regression (PLSR) family [@woutersenNovelApproachStudy2018; @zimmermannOptimizingSavitzkyGolayParameters2013].
+Preprocessing with SG and EMSC has been shown to improve the performance of multivariate analyses, such as principal component analysis (PCA) or models in the partial least squares regression (PLSR) family [@zimmermannOptimizingSavitzkyGolayParameters2013; @woutersenNovelApproachStudy2018].
 
@@
 ```{r ordinations, echo=FALSE, fig.cap="a.
-plsr ordination scores all samples diff fresh vs fossil b.
-regression coef; component 1 c.
-regression coef component 2; d plsr ordination scores only samples from cores diff spt vs acet; e.
-regression coef component 1; f.
-regression coef component 2", message=FALSE, warning=FALSE, out.width="100%"}
+PLSR ordination scores of all samples showing the difference between fresh vs fossil pollen b.
+regression coefficient of component 1 c.
+regression coefficient of component 2; d PLSR ordination scores using only samples from sediment cores.
+Difference between non-acetolysed and acetolyses; e.
+regression coefficient component 1; f.
+regression coefficient component 2", message=FALSE, warning=FALSE, out.width="100%"}
 knitr::include_graphics(here("figures", "ord_plots.png"))
@@
 ```{r SM, echo=FALSE, fig.cap="Image captured from FTIR microscope of sample slide Dalmutladdo 60 cm depth.
-Pinus pollen is visible as well as debris and dried residue, presumably BSi.", message=FALSE, warning=FALSE, out.width="90%"}
+\\textit{Pinus} pollen is visible as well as debris and dried residue, presumably BSi.", message=FALSE, warning=FALSE, out.width="90%"}
 knitr::include_graphics(here("figures", "SM_BSI.png"))
@@
 The approach has the potential to be an improvement on previous single-grain measurements using FTIR, where each grain is required to be targeted individually, or even manually extracted from the sediment matrix using a micro-pipette.
-The FPA-MCT detector on our FTIR microscope allows simultaneous capture of a larger area of the sample (XXX x XXX µm ), compared to other detectors (e.g.
-100 x 100 µm @jardineChemotaxonomyDomesticatedGrasses2019, 50 x 50 ?µm?
-@zimmermannChemicalCharacterizationIdentification2018).
+The FPA-MCT detector on our FTIR microscope allows simultaneous capture of a larger area of the sample (500 x 500 µm ), compared to other detectors [e.g.
+100 x 100 µm @jardineChemotaxonomyDomesticatedGrasses2019, 50 x 50 µm @zimmermannChemicalCharacterizationIdentification2018].
 An image with a spectra for each pixel is created, from which spectra of single grains were extracted.
@@
 In general the FTIR spectra obtained using this approach created similar spectra to spectra of *Pinus* pollen from bulk approaches, and the method was also sensitive to differences in chemical treatments on the fossil-pollen grains.
-For example, the spectra obtained from our modern samples are very similar to *Pinus sylvestris* spectra measured using FTIR-bulk sample methodologies [@bagciogluMultiscaleVibrationalSpectroscopic2015; @zimmermannCharacterizationPollenVibrational2010] and FTIR microscopy using single-point measurements [@zimmermannChemicalCharacterizationIdentification2018].
+For example, the spectra obtained from our modern samples are very similar to *Pinus sylvestris* spectra measured using FTIR-bulk sample methodologies [@zimmermannCharacterizationPollenVibrational2010; @bagciogluMultiscaleVibrationalSpectroscopic2015] and FTIR microscopy using single-point measurements [@zimmermannChemicalCharacterizationIdentification2018].
 The same characteristic peaks described above are visible in all modern spectra captured with bulk and FTIR microscopy, e.g.
@@
 
-Taken together, the differences in the effects of acetolysis across different taxa and sample types suggest that the chemical extraction methods have a large effect on the chemistry of pollen, both modern and fossil origin.
+Taken together, the differences in the effects of acetolysis across different taxa and sample types suggest that chemical extraction methods have an effect on the chemistry of pollen, both modern and fossil origin.
 Direct comparisons between spectra from different spectrometers are difficult, because peak heights are not always directly comparable, but these results indicate that further research is required to understand the reasons between these differences between the acetolysed *Pinus* pollen from across the sample sets.
@@
 The chemical effects of acetolysis that identified across studies suggests that it may be preferable to use an alternative approach to extract fossil pollen grains from sediment samples.
-Given the limited chemical influence on the fossil pollen grains, we believe that density separation is a viable alternative to acetolysis and other acid or oxidizing extractions [@caffreyUseLithiumHeteropolytungstate2013; @leipeTestingPerformanceSodium2019; @regnellPreparativeCentrifugationNew1996], although given the fact that an SPT peak was observed in one of the cores, the results in this paper clearly demonstrate the need to identify efficient dilution processes that may need to be tailored to specific sediment types.
-Similarly, since the reported density of pollen is variable and may be dependent on age of the sample [@faegriTextbookPollenAnalysis1989; @flenley1971measurements; @regnellPreparativeCentrifugationNew1996; @nakagawaDensemediaSeparationMore1998; @juvigneDensiteExinesQuelques1973], it is possible that specific extraction densities and procedures may be required from core to core.
+Given the limited chemical influence on the fossil pollen grains, we believe that density separation is a viable alternative to acetolysis and other acid or oxidizing extractions [@regnellPreparativeCentrifugationNew1996; @caffreyUseLithiumHeteropolytungstate2013; @leipeTestingPerformanceSodium2019], although given the fact that an SPT peak was observed in one of the cores, the results in this paper clearly demonstrate the need to identify efficient dilution processes that may need to be tailored to specific sediment types.
+Similarly, since the reported density of pollen is variable and may be dependent on age of the sample [@flenley1971measurements; @juvigneDensiteExinesQuelques1973; @faegriTextbookPollenAnalysis1989; @regnellPreparativeCentrifugationNew1996; @nakagawaDensemediaSeparationMore1998], it is possible that specific extraction densities and procedures may be required from core to core.
 
@@
 
+<!--
 ## I haven't found a way to integrate these ideas in there. Perhaps we don't need them? I think perhaps we could leave it out
@@
 Further studies on pollen from different sediment types at fixed age/depths may provide a clearer answer on the effects of sediment type and age on pollen chemistry.
-
+-->
 # References
